### 2023-06-14 report/markdown/report.Rmd
Initial analysis version
Analysis scripts and analysis report backbone

---
title: "Artificial inteligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Analysis report"
author: "Piotr Tymoszuk for the Department of Radiology"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPDF", 
                      dpi = 600)

set_flextable_defaults(font.family = "Helvetica", 
                       font.size = 10)

```

<center>

![](logo_large.png){#id .class width="40mm"}

</center>

<hr style = "border-top: 5px solid DarkGray" />

# Analysis summary

<hr style = "border-top: 5px solid #417a8b" />

# Characteristic of the cohort and exploratory data analysis

## Characteristic of the CovILD cohort

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-cohort, tab.cap = 'Baseline characteristics and COVID-19 course in the CovILD cohort. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:5, width = 3.0, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  width(7, width = 2.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('BMI: body mass index; COPD: chronic obstructive pulmonary disease; CKD: chronic kidney disease'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:6, 
           value = as_paragraph('Comparison between COVID-19 severity groups. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple testing with the false discovery rate method'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(4, 1, 
           value = as_paragraph('young adult: 18 - 30 years; middle aged: 31 - 65 years; elderly: 66 years and more'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  footnote(8, 1, 
           value = as_paragraph('normal: BMI < 25 kg/m\u00B2; overweight: BMI 25 - 30 kg/m\u00B2; obesity: BMI > 30 kg/m\u00B2'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Chest computed tomography paramaters and findings

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-cohort-ct, tab.cap = 'Chest computed tomography variables at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_ct) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3:6, width = 3.8, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity; CTSS: CT severity score; AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Lung function testing paramaters and abnormalities

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-cohort-lft, tab.cap = 'Lung function testing variables at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(tables$cohort_lft) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3:6, width = 3.8, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second; DLCO: diffusion capacity for CO; FEV1:FVC: FEV1 to FVC ratio'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid #417a8b" />

# Detection of radiological lung findings with AI-determined opacity and high opacity

## Optimal cutoff, sensitivity and specificity of AI opacity and high opacity at detection of radiological findings

```{r tab-ai-cutoffs, tab.cap = 'Optimal cutoffs for AI-determined lung opacity and high opacity at detection of chest computed tomography abnormalities. The optimal cutoff was determined by Youden criterion.'}

flextable::flextable(tables$ai_cutoffs) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 4.8, unit = 'cm') %>% 
  width(4:8, width = 2.5, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 7, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 8, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 9, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Performance of AI opacity and high opacity in COVID-19 severity groups and at consecutive follow-ups

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-ai-cutoffs-severity, tab.cap = 'Performance of AI-determined lung opacity and high opacity at detection of chest computed tomography abnormalities in the COVID-19 severity subsets. The opacity and high opacity values were stratified by their optimal cutoffs determined by Youden criterion in the entire CovILD cohort. Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}

flextable::flextable(tables$ai_cutoffs_severity) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:7, width = 5.2, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-ai-cutoffs-fup, tab.cap = 'Performance of AI-determined lung opacity and high opacity at detection of chest computed tomography abnormalities at the consecutive follow-ups. The opacity and high opacity values were stratified by their optimal cutoffs determined by Youden criterion in the entire CovILD cohort. Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}

flextable::flextable(tables$ai_cutoffs_fup) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:7, width = 5.2, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid #417a8b" />

# Chest computed tomography severity score, AI-determined opacity and high opacity

## Chest computed tomography severity score and interstitial lung disease

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-ild-cutoff, tab.cap = 'Detecion of interstitial lung disease defined by lung opacity > 5% by chest computed tomography severity score (CTSS). The optimal cutoff for CTSS was determined by Youden criterion.'}

flextable::flextable(tables$ild_cutoff) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 4.8, unit = 'cm') %>% 
  width(3:8, width = 2.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 6, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 7, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 8, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-ild-cutoff-severity, tab.cap = 'Detecion of interstitial lung disease defined by lung opacity > 5% by chest computed tomography severity score (CTSS) in the COVID-19 severity subsets. The optimal cutoff for CTSS was determined by Youden criterion in the entire CovILD cohort. Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}

flextable::flextable(tables$ild_cutoff_severity) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2:5, width = 5, unit = 'cm') %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-ild-cutoff-fup, tab.cap = 'Detecion of interstitial lung disease defined by lung opacity > 5% by chest computed tomography severity score (CTSS) at the consecutive follow-ups. The optimal cutoff for CTSS was determined by Youden criterion in the entire CovILD cohort. Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}

flextable::flextable(tables$ild_cutoff_fup) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2:5, width = 5, unit = 'cm') %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Mathematical relationship of the severity score, and AI-determined opacity and high opacity

<hr style = "border-top: 5px solid #417a8b" />

# Prediction of lung function with chest computed tomography

## Development of multi-paramater Random Forest models of lung function variables and abnormalities

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-rf-tuning, tab.cap = 'Optimal values of the mtry, splitrule and min.node.size paramaters for Random Forest models of lung function abnormalities and paramaters found by cross-validation tuning.'}

flextable::flextable(tables$rf_tuning) %>% 
  width(1, width = 2.5, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:6, width = 2.7, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Performance of Random Forest models at prediction of lung function

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-bin-models, tab.cap = 'Performance of Random Forest classifiers at prediction of lung function testing (LFT) abnormalities in the training dataset and cross-validation (CV). CV results for any LFT abnormalities and diffusion capacity for CO (DLCO) < 80% reference are presented. The entire table is available in a supplementary Excel file.'}

tables$bin_models %>% 
  filter(`LFT variable` %in% c('LFT findings', 'DLCO < 80%'), 
         `Data type` == 'CV') %>% 
  select(-`Data type`) %>% 
  flextable %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3, width = 4, unit = 'cm') %>% 
  width(4, width = 5.5, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('C-index: concordance index; \u03BA: Cohen \u03BA inter-rater relaibility statistic; AUC, ROC: area under the curve, receiver - operating characteristic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-reg-models, tab.cap = 'Performance of Random Forest regression models at prediction of lung function testing (LFT) variables in the training dataset and cross-validation (CV). CV results for diffusion capacity for CO (DLCO), forced vital capacity (FVC) and forced expiratory volume in 1 second (FEV1) are presented. The entire table is available in a supplementary Excel file.'}

tables$reg_models %>% 
  filter(`Data type` == 'CV') %>% 
  select(-`Data type`) %>% 
  flextable %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3, width = 4, unit = 'cm') %>% 
  width(4, width = 5.5, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('MAE: mean absolute error; RMSE: root mean square error; Pearson r: Pearson correlation coefficient between fitter and observed values'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid DarkGray" />

## Severity score, and AI-determined opacity and high opacity are the key predictors of lung function

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-lft-binary-uni, tab.cap = 'Differences in chest computed tomography severity score (CTSS), and AI-determined lung opacity and high opacity in CovILD study participants with and without lung function testing (LFT) abnormalities. Numeric variariables are presented as medians with interquartile ranges (IQR) and ranges.'}

flextable::flextable(tables$lft_binary_uni) %>% 
  width(1, width = 2.8, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3:4, width = 3.8, unit = 'cm') %>% 
  width(5, width = 2.7, unit = 'cm') %>% 
  width(6, width = 2.5, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: chest computed tomography severity score; AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Mann-Whitney test with r effect size statistic. P values corrected for multiple testing with the false discovery method'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-lft-correlation-uni, tab.cap = 'Correlation of LFT variables with chest computed tomography severity score, and AI-determined opacity and high opactity.'}

flextable::flextable(tables$lft_correlation_uni) %>% 
  width(1:2, width = 4.5, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm') %>% 
  width(4, width = 5, unit = 'cm') %>% 
  width(5, width = 2.7, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCOP: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: chest computed tomography severity score; AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Spearman correlation coefficient and Spearman rank test. P values corrected for multiple testing with the false discovery method'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

<hr style = "border-top: 1px solid DarkGray" />

```{r tab-lft-roc, tab.cap = 'Detection of lung function testing (LFT) abnormalities by chest computed tomography variables in receiver-operator characteristic (ROC) analysis.'}

flextable::flextable(tables$lft_roc) %>% 
  width(1:2, width = 2.8, unit = 'cm') %>% 
  width(3, width = 5, unit = 'cm') %>% 
  width(4:9, width = 2.5, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT:: lung function testting; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: chest computed tomography severity score; AI: artificual intelligence'),
           part = 'header',
           ref_symbols = 'b') %>% 
   footnote(1, 3, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 7, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 8, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
   footnote(1, 9, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'f') %>% 
  theme_vanilla

```

<hr style = "border-top: 5px solid #417a8b" />

# Methods

<hr style = "border-top: 5px solid #417a8b" />

# Data and code availability

The entire R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/CovILD_AI).

<hr style = "border-top: 5px solid #417a8b" />

# References

### 2023-06-14 CTSS scripts/ctss_classes.R

@@ -1,0 +1,94 @@
+# Opacity and high opacity in the CTSS severity classes
+
+  insert_head()
+  
+# container ------
+  
+  ctss_classes <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  ctss_classes$variables <- c('opacity_percent', 'high_opacity_percent')
+  
+  ctss_classes$variables <- 
+    set_names(ctss_classes$variables, 
+              ctss_classes$variables)
+  
+  ## analysis table
+  
+  ctss_classes$analysis_tbl <- ctss_globals$analysis_tbl %>% 
+    filter(severity_class == 'cohort') %>% 
+    select(ID, CTSS_class, all_of(ctss_classes$variables)) %>%
+    filter(complete.cases(.))
+
+# descriptive stats -----
+  
+  insert_msg('Descriptive stats')
+  
+  ## n numbers 
+  
+  ctss_classes$n_numbers <- ctss_classes$analysis_tbl %>% 
+    count(CTSS_class)
+  
+  ## normality: clearly non-normal distribution
+  
+  ctss_classes$normality <- ctss_classes$analysis_tbl %>% 
+    explore(split_factor = 'CTSS_class', 
+            variables = ctss_classes$variables, 
+            what = 'normality', 
+            pub_styled = TRUE)
+  
+  ## descriptive stats
+  
+  ctss_classes$stats <- ctss_classes$analysis_tbl %>% 
+    explore(split_factor = 'CTSS_class', 
+            variables = ctss_classes$variables, 
+            what = 'table', 
+            pub_styled = TRUE)
+  
+# Testing for trend with Kruskal-Wallis test -----
+  
+  insert_msg('Testing')
+  
+  ctss_classes$test <- ctss_classes$analysis_tbl %>% 
+    compare_variables(variables = ctss_classes$variables, 
+                      split_factor = 'CTSS_class', 
+                      what = 'eff_size', 
+                      types = 'kruskal_etasq', 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      adj_method = 'BH') %>% 
+    mutate(plot_cap = paste(eff_size, significance, sep = ', '))
+  
+# Plots -------
+  
+  insert_msg('Plots')
+  
+  ctss_classes$plots <-
+    list(variable = ctss_classes$variables, 
+         plot_title = exchange(ctss_classes$variables, 
[…]

### 2023-06-14 CTSS scripts/ctss_relationship.R

@@ -1,0 +1,160 @@
+# Trial to establish a function to link opacity and high opacity
+# with CTSS
+
+  insert_head()
+  
+# container -------
+  
+  ctss_fun <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## explanatory variables
+  
+  ctss_fun$responses <- c('opacity_percent', 'high_opacity_percent')
+  
+  ## analysis table
+
+  ctss_fun$analysis_tbl <- ctss_globals$analysis_tbl %>% 
+    filter(severity_class == 'cohort') %>% 
+    select(ID, CTSS, all_of(ctss_fun$responses)) %>% 
+    filter(complete.cases(.))
+
+# Modeling AI opacity -------
+  
+  insert_msg('Modeling AI opacity')
+  
+  ctss_fun$models$opacity_lm_poly <- 
+    make_lm(data = ctss_fun$analysis_tbl, 
+            formula = sqrt(opacity_percent) ~ CTSS + I(CTSS^2) + I(CTSS^3) + I(CTSS^4), 
+            mod_fun = lm, 
+            family = NULL)
+  
+  ctss_fun$models$opacity_rlm_poly <- 
+    make_lm(data = ctss_fun$analysis_tbl, 
+            formula = sqrt(opacity_percent) ~ CTSS + I(CTSS^2) + I(CTSS^3) + I(CTSS^4), 
+            mod_fun = rlm, 
+            family = NULL)
+  
+  ctss_fun$models$opacity_gam_tp <- 
+    make_lm(data = ctss_fun$analysis_tbl, 
+            formula = sqrt(opacity_percent) ~ CTSS + s(CTSS, bs = 'tp', k = 8), 
+            mod_fun = gam, 
+            family = NULL)
+
+# Modeling AI high opacity -------
+  
+  insert_msg('Modeling AI high opacity')
+  
+  ctss_fun$models$high_opacity_lm_poly <- 
+    make_lm(data = ctss_fun$analysis_tbl, 
+            formula = sqrt(high_opacity_percent) ~ CTSS + I(CTSS^2) + I(CTSS^3) + I(CTSS^4), 
+            mod_fun = lm, 
+            family = NULL)
+  
+  ctss_fun$models$high_opacity_rlm_poly <- 
+    make_lm(data = ctss_fun$analysis_tbl, 
[…]

### 2023-06-14 CTSS scripts/ild_cutoff.R

@@ -1,0 +1,131 @@
+# Checking for the optimal CTSS cutoff to detect interstitial lung disease
+# defined as opacity in >5% of the lung.
+#
+# Done with Youden criterion for the entire cohort
+
+  insert_head()
+  
+# container ------
+  
+  ild_cutoff <- list()
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ild_cutoff$analysis_tbl <- ctss_globals$analysis_tbl %>% 
+    filter(severity_class == 'cohort') %>% 
+    select(ID, CTSS, opacity_class, opacity_percent) %>% 
+    filter(complete.cases(.))
+  
+# Number of cases ------
+  
+  insert_msg('Number of cases')
+  
+  ild_cutoff$n_numbers <- ild_cutoff$analysis_tbl %>% 
+    count(opacity_class)
+  
+  ild_cutoff$n_caps <- 
+    paste0('total: n = ', sum(ild_cutoff$n_numbers$n), 
+           ', findings: n = ', ild_cutoff$n_numbers$n[2])
+  
+# optimal cutpoint object ------
+  
+  insert_msg('Optrimal cutpoint object')
+  
+  ild_cutoff$cut_obj <- 
+    optimal.cutpoints('CTSS', 
+                      status = 'opacity_class', 
+                      tag.healthy = '0-5%', 
+                      methods = 'Youden', 
+                      data = as.data.frame(ild_cutoff$analysis_tbl))
+
+# Cutpoint stats ------
+  
+  insert_msg('Cutpoint stats')
+  
+  ild_cutoff$stats <- ild_cutoff$cut_obj %>% 
+    extract_cut
+  
+  ild_cutoff$stats$optimal_cutoff <- ild_cutoff$stats$optimal_cutoff %>% 
+    mutate(response = 'opacity_percent', 
+           auc_lab = paste0('AUC = ', signif(auc, 2), 
+                            ' [', signif(lower_ci, 3), 
+                            ' - ', signif(upper_ci, 2), ']'), 
+           plot_lab = paste0('Se = ', signif(Se, 2), 
+                             '\nSp = ', signif(Sp, 2), 
+                             '\n', auc_lab))
+  
+# Plotting cutpoint stats for all possible cutoffs -------
+  
+  insert_msg('Plotting cutoff stats for all possible cutoffs')
+  
+  ild_cutoff$all_plots <- 
+    plot_cut_search(data = ild_cutoff$stats$all_cutoffs, 
[…]

### 2023-06-14 CTSS scripts/ild_strata.R

@@ -1,0 +1,130 @@
+# Performance of the CTSS cutoff defining the ILD in the severity strata
+# and at the follow-ups
+
+  insert_head()
+  
+# container -----
+  
+  ild_strata <- list()
+  
+# parallel backend -----
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## optimal cutoff
+  
+  ild_strata$cutoff <- ild_cutoff$stats$optimal_cutoff$cutoff
+
+  ## analysis table
+  
+  ild_strata$analysis_tbl <- ctss_globals$analysis_tbl %>% 
+    filter(severity_class != 'cohort') %>% 
+    mutate(CTSS = cut(CTSS, 
+                      c(-Inf, ild_strata$cutoff, Inf), 
+                      c('no', 'yes'), 
+                      right = FALSE), 
+           opacity_class = car::recode(opacity_class, 
+                                       "'0-5%' = 'no'; '>5%' = 'yes'")) %>% 
+    select(ID, severity_class, follow_up, CTSS, opacity_class) %>% 
+    filter(complete.cases(.))
+    
+  ild_strata$analysis_tbl <- ild_strata$analysis_tbl %>% 
+    mutate(severity_class = 'cohort', 
+           follow_up = 'cohort') %>% 
+    rbind(ild_strata$analysis_tbl, .) %>% 
+    mutate(follow_up = factor(follow_up, 
+                              c('cohort', 
+                                levels(ild_strata$analysis_tbl$follow_up))))
+  
+# calculation of bootstrapped stats for the severity classes --------
+  
+  insert_msg('Inter-rater stats for the severity classes')
+  
+  ## working with safely: especially in mild cases, 
+  ## where ILD may not appear
+  
+  ild_strata$severity_stats <- ild_strata$analysis_tbl %>% 
+    blast(severity_class) %>% 
+    map(rater_stats, 
+        d_variable = 'opacity_class',
+        m_variable = 'CTSS', 
+        positive = 'yes', 
+        B = 2000) %>% 
+    compress(names_to = 'severity_class') %>%
+    mutate(severity_class = factor(severity_class, 
+                                   levels(ild_strata$analysis_tbl$severity_class)), 
+           response = 'opacity_class')
+  
+# calculation of bootstrapped stats for the follow-ups --------
+  
+  insert_msg('Inter-rater stats for the follow-ups')
+  
[…]

### 2023-06-14 LFT scripts/binary_importance.R

@@ -1,0 +1,151 @@
+# Permutation importance, binary classifiers
+
+  insert_head()
+  
+# container -----
+  
+  bin_importance <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## optimal tuning parameters needed to re-construct Random Forest models
+  
+  bin_importance$optima <- 
+    c(lft_rf$optima, 
+      dlco_red_rf$optima, 
+      fvc_red_rf$optima, 
+      fev1_red_rf$optima) %>% 
+    set_names(c('lft_matched', 
+                'lft_unmatched', 
+                'dlco_matched', 
+                'dlco_unmatched', 
+                'fvc_matched', 
+                'fvc_unmatched', 
+                'fev1_matched', 
+                'fev1_unmatched')) %>% 
+    compress(names_to = 'model')
+  
+  ## model formulas
+  
+  bin_importance$formulas <- 
+    list('lft_matched' = lft_globals$formulas$LFT_findings, 
+         'lft_unmatched' = lft_globals$formulas$LFT_findings, 
+         'dlco_matched' = lft_globals$formulas$DLCO_reduced, 
+         'dlco_unmatched' = lft_globals$formulas$DLCO_reduced, 
+         'fvc_matched' = lft_globals$formulas$FVC_reduced, 
+         'fvc_unmatched' = lft_globals$formulas$FVC_reduced, 
+         'fev1_matched' = lft_globals$formulas$FEV1_reduced, 
+         'fev1_unmatched' = lft_globals$formulas$FEV1_reduced)
+  
+  ## lexicon
+  
+  bin_importance$lexicon <- 
+    c('lft_matched' = 'LFT findings, matched', 
+      'lft_unmatched' = 'LFT findings, unmatched', 
+      'dlco_matched' = 'DLCO < 80%, matched', 
+      'dlco_unmatched' = 'DLCO < 80%, unmatched', 
+      'fvc_matched' = 'FVC < 80%, matched', 
+      'fvc_unmatched' = 'FVC < 80%, unmatched', 
+      'fev1_matched' = 'FEV1 < 80%, matched', 
+      'fev1_unmatched' = 'FEV1 < 80%, unmatched') %>% 
+    compress(names_to = 'variable', 
+             values_to = 'label') %>% 
[…]

### 2023-06-14 LFT scripts/binary_performance.R

@@ -1,0 +1,264 @@
+# Checking performance of Random Forest models for the binary responses: 
+# any CT findings, reduced DLCO, reduced FVC and reduced FEV1
+
+  insert_head()
+  
+# container ------
+  
+  bin_models <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## caret models
+  
+  bin_models$models <- 
+    c(lft_rf$models, 
+      dlco_red_rf$models, 
+      fvc_red_rf$models, 
+      fev1_red_rf$models) %>% 
+    set_names(c('lft_matched', 
+                'lft_unmatched', 
+                'dlco_matched', 
+                'dlco_unmatched', 
+                'fvc_matched', 
+                'fvc_unmatched', 
+                'fev1_matched', 
+                'fev1_unmatched'))
+  
+  ## model name lexicon
+  
+  bin_models$lexicon <- 
+    c('lft_matched' = 'LFT findings, matched', 
+      'lft_unmatched' = 'LFT findings, unmatched', 
+      'dlco_matched' = 'DLCO < 80%, matched', 
+      'dlco_unmatched' = 'DLCO < 80%, unmatched', 
+      'fvc_matched' = 'FVC < 80%, matched', 
+      'fvc_unmatched' = 'FVC < 80%, unmatched', 
+      'fev1_matched' = 'FEV1 < 80%, matched', 
+      'fev1_unmatched' = 'FEV1 < 80%, unmatched') %>% 
+    compress(names_to = 'variable', 
+             values_to = 'label') %>% 
+    mutate(label_short = stri_replace(label, regex = ',.*', replacement = ''))
+  
+# Predictions ------
+  
+  insert_msg('Predictions')
+  
+  bin_models$predictions <- bin_models$models %>% 
+    future_map(predict.caretx, 
+               .options = furrr_options(seed = TRUE)) %>% 
+    map(~.x[c('train', 'cv')]) %>% 
+    transpose
+  
+# Model residuals: plots -----
+  
+  insert_msg('Model residuals plots')
+  
+  bin_models$resid_plots <- bin_models$models %>% 
+    future_map(plot.caretx, 
+               type = 'diagnostic', 
+               cust_theme = globals$common_theme, 
+               .options = furrr_options(seed = TRUE)) %>% 
+    transpose
+  
[…]

### 2023-06-14 LFT scripts/regression_importance.R

@@ -1,0 +1,142 @@
+# Permutation importance of explanatory variables for regression 
+# Random Forest models
+
+  insert_head()
+  
+# container -------
+  
+  reg_importance <- list()
+  
+# parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  ## optimal tuning parameters needed to re-construct Random Forest models
+  
+  reg_importance$optima <- 
+    c(dlco_rf$optima, 
+      fvc_rf$optima, 
+      fev1_rf$optima) %>% 
+    set_names(c('dlco_matched', 
+                'dlco_unmatched', 
+                'fvc_matched', 
+                'fvc_unmatched', 
+                'fev1_matched', 
+                'fev1_unmatched')) %>% 
+    compress(names_to = 'model')
+  
+  ## model formulas
+  
+  reg_importance$formulas <- 
+    list('dlco_matched' = lft_globals$formulas$DLCO_percent, 
+         'dlco_unmatched' = lft_globals$formulas$DLCO_percent, 
+         'fvc_matched' = lft_globals$formulas$FVC_percent, 
+         'fvc_unmatched' = lft_globals$formulas$FVC_percent, 
+         'fev1_matched' = lft_globals$formulas$FEV1_percent, 
+         'fev1_unmatched' = lft_globals$formulas$FEV1_percent)
+  
+  ## lexicon
+  
+  reg_importance$lexicon <- 
+    c('dlco_matched' = 'DLCO, matched', 
+      'dlco_unmatched' = 'DLCO, unmatched', 
+      'fvc_matched' = 'FVC, matched', 
+      'fvc_unmatched' = 'FVC, unmatched', 
+      'fev1_matched' = 'FEV1, matched', 
+      'fev1_unmatched' = 'FEV1, unmatched') %>% 
+    compress(names_to = 'variable', 
+             values_to = 'label') %>% 
+    mutate(label_short = stri_replace(label, regex = ',.*', replacement = ''))
+  
+# Ranger models -----
+  
+  insert_msg('Ranger models')
+  
+  reg_importance$models <- 
+    list(formula = reg_importance$formulas, 
+         mtry = reg_importance$optima$mtry, 
+         splitrule = reg_importance$optima$splitrule, 
+         min.node.size = reg_importance$optima$min.node.size) %>% 
+    pmap(ranger, 
[…]

### 2023-06-14 LFT scripts/regression_performance.R

@@ -1,0 +1,270 @@
+# Checking performance of Random Forest models for the numeric responses: 
+# any DLCO, FVC and FEV1
+
+  insert_head()
+  
+# container -----
+  
+  reg_models <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## caret models
+  
+  reg_models$models <- 
+    c(dlco_rf$models, 
+      fvc_rf$models, 
+      fev1_rf$models) %>% 
+    set_names(c('dlco_matched', 
+                'dlco_unmatched', 
+                'fvc_matched', 
+                'fvc_unmatched', 
+                'fev1_matched', 
+                'fev1_unmatched'))
+  
+  ## model name lexicon
+  
+  reg_models$lexicon <- 
+    c('dlco_matched' = 'DLCO, matched', 
+      'dlco_unmatched' = 'DLCO, unmatched', 
+      'fvc_matched' = 'FVC, matched', 
+      'fvc_unmatched' = 'FVC, unmatched', 
+      'fev1_matched' = 'FEV1, matched', 
+      'fev1_unmatched' = 'FEV1, unmatched') %>% 
+    compress(names_to = 'variable', 
+             values_to = 'label') %>% 
+    mutate(label_short = stri_replace(label, regex = ',.*', replacement = ''))
+  
+# Predictions ------
+  
+  insert_msg('Predictions')
+  
+  reg_models$predictions <- reg_models$models %>% 
+    future_map(predict.caretx, 
+               .options = furrr_options(seed = TRUE)) %>% 
+    map(~.x[c('train', 'cv')]) %>% 
+    transpose
+  
+# Model residuals: plots -----
+  
+  insert_msg('Model residuals plots')
+  
+  ## working with the cache: this is a time-consuming step
+  
+  if(!file.exists('./cache/reg_resid_plots.RData')) {
+    
+    reg_models$resid_plots <- reg_models$models %>% 
+      future_map(plot.caretx, 
+                 type = 'diagnostic', 
+                 cust_theme = globals$common_theme, 
+                 .options = furrr_options(seed = TRUE)) %>% 
+      transpose
+    
+    ## caching
+    
+    reg_resid_plots <-  reg_models$resid_plots
+    
+    save(reg_resid_plots, file = './cache/reg_resid_plots.RData')
+    
+  } else {
[…]

### 2023-06-14 LFT scripts/rf_DLCO.R

@@ -1,0 +1,88 @@
+# Modeling of DLCO with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of DLCO
+# Unmatched CV folds: modeling simple presence/absence of DLCO
+
+  insert_head()
+
+# container ------
+
+  dlco_rf <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  dlco_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  dlco_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(DLCO_percent, all_of(dlco_rf$variables))
+  
+  dlco_rf$formula <- lft_globals$formulas$DLCO_percent
+  
+  dlco_rf$train_control <- lft_globals$train_control
+
+# tuning grid ------
+
+  insert_msg('Tuning grids')
+  
+  dlco_rf$tune_grid <- 
+    expand.grid(splitrule = c('variance', 'extratrees', 'maxstat'), 
+                mtry = seq(2, length(dlco_rf$variables), by = 3), 
+                min.node.size = c(3, 5, 7), 
+                stringsAsFactors = FALSE)
+
+# modeling ------
+
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  dlco_rf$models <- dlco_rf$train_control %>% 
+    map(~train(form = dlco_rf$formula, 
+               data = dlco_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'RMSE', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = dlco_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  dlco_rf$models <- dlco_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+
+  insert_msg('Extracting tuning results')
+  
+  dlco_rf$stats <- dlco_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  dlco_rf$optima <- dlco_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+
+# Plotting tuning results ------
+
+  insert_msg('Plotting tuning results')
+  
+  dlco_rf$plots <- 
+    list(data = dlco_rf$stats, 
[…]

### 2023-06-14 LFT scripts/rf_DLCO_findings.R

@@ -1,0 +1,88 @@
+# Modeling of the DLCO findings with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of the DLCO findings
+# Unmatched CV folds: modeling simple presence/absence of the DLCO findings
+
+  insert_head()
+
+# container ------
+
+  dlco_red_rf <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  dlco_red_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  dlco_red_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(DLCO_reduced, all_of(dlco_red_rf$variables))
+  
+  dlco_red_rf$formula <- lft_globals$formulas$DLCO_reduced
+  
+  dlco_red_rf$train_control <- lft_globals$train_control
+
+# tuning grid ------
+
+  insert_msg('Tuning grids')
+  
+  dlco_red_rf$tune_grid <- 
+    expand.grid(splitrule = c('gini', 'extratrees'), 
+                mtry = seq(2, length(dlco_red_rf$variables), by = 3), 
+                min.node.size = c(1, 3), 
+                stringsAsFactors = FALSE)
+
+# modeling ------
+
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  dlco_red_rf$models <- dlco_red_rf$train_control %>% 
+    map(~train(form = dlco_red_rf$formula, 
+               data = dlco_red_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'Kappa', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = dlco_red_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  dlco_red_rf$models <- dlco_red_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+
+  insert_msg('Extracting tuning results')
+  
+  dlco_red_rf$stats <- dlco_red_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  dlco_red_rf$optima <- dlco_red_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+
+# Plotting tuning results ------
+
+  insert_msg('Plotting tuning results')
[…]

### 2023-06-14 LFT scripts/rf_FEV1.R

@@ -1,0 +1,88 @@
+# Modeling of FEV1 with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of FEV1
+# Unmatched CV folds: modeling simple presence/absence of FEV1
+
+  insert_head()
+
+# container ------
+
+  fev1_rf <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  fev1_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fev1_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(FEV1_percent, all_of(fev1_rf$variables))
+  
+  fev1_rf$formula <- lft_globals$formulas$FEV1_percent
+  
+  fev1_rf$train_control <- lft_globals$train_control
+
+# tuning grid ------
+
+  insert_msg('Tuning grids')
+  
+  fev1_rf$tune_grid <- 
+    expand.grid(splitrule = c('variance', 'extratrees', 'maxstat'), 
+                mtry = seq(2, length(fev1_rf$variables), by = 3), 
+                min.node.size = c(3, 5, 7), 
+                stringsAsFactors = FALSE)
+
+# modeling ------
+
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  fev1_rf$models <- fev1_rf$train_control %>% 
+    map(~train(form = fev1_rf$formula, 
+               data = fev1_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'RMSE', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = fev1_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  fev1_rf$models <- fev1_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+
+  insert_msg('Extracting tuning results')
+  
+  fev1_rf$stats <- fev1_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  fev1_rf$optima <- fev1_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+
+# Plotting tuning results ------
+
+  insert_msg('Plotting tuning results')
+  
+  fev1_rf$plots <- 
+    list(data = fev1_rf$stats, 
[…]

### 2023-06-14 LFT scripts/rf_FEV1_findings.R

@@ -1,0 +1,88 @@
+# Modeling of the FEV1 findings with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of the FEV1 findings
+# Unmatched CV folds: modeling simple presence/absence of the LFT findings
+
+  insert_head()
+  
+# container ------
+  
+  fev1_red_rf <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  fev1_red_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fev1_red_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(FEV1_reduced, all_of(fev1_red_rf$variables))
+  
+  fev1_red_rf$formula <- lft_globals$formulas$FEV1_reduced
+  
+  fev1_red_rf$train_control <- lft_globals$train_control
+  
+# tuning grid ------
+  
+  insert_msg('Tuning grids')
+  
+  fev1_red_rf$tune_grid <- 
+    expand.grid(splitrule = c('gini', 'extratrees'), 
+                mtry = seq(2, length(fev1_red_rf$variables), by = 3), 
+                min.node.size = c(1, 3), 
+                stringsAsFactors = FALSE)
+  
+# modeling ------
+  
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  fev1_red_rf$models <- fev1_red_rf$train_control %>% 
+    map(~train(form = fev1_red_rf$formula, 
+               data = fev1_red_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'Kappa', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = fev1_red_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  fev1_red_rf$models <- fev1_red_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+  
+  insert_msg('Extracting tuning results')
+  
+  fev1_red_rf$stats <- fev1_red_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  fev1_red_rf$optima <- fev1_red_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+  
+# Plotting tuning results ------
+  
[…]

### 2023-06-14 LFT scripts/rf_FVC.R

@@ -1,0 +1,88 @@
+# Modeling of FVC with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of FVC
+# Unmatched CV folds: modeling simple presence/absence of FVC
+
+  insert_head()
+
+# container ------
+
+  fvc_rf <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  fvc_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fvc_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(FVC_percent, all_of(fvc_rf$variables))
+  
+  fvc_rf$formula <- lft_globals$formulas$FVC_percent
+  
+  fvc_rf$train_control <- lft_globals$train_control
+
+# tuning grid ------
+
+  insert_msg('Tuning grids')
+  
+  fvc_rf$tune_grid <- 
+    expand.grid(splitrule = c('variance', 'extratrees', 'maxstat'), 
+                mtry = seq(2, length(fvc_rf$variables), by = 3), 
+                min.node.size = c(3, 5, 7), 
+                stringsAsFactors = FALSE)
+
+# modeling ------
+
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  fvc_rf$models <- fvc_rf$train_control %>% 
+    map(~train(form = fvc_rf$formula, 
+               data = fvc_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'RMSE', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = fvc_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  fvc_rf$models <- fvc_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+
+  insert_msg('Extracting tuning results')
+  
+  fvc_rf$stats <- fvc_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  fvc_rf$optima <- fvc_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+
+# Plotting tuning results ------
+
+  insert_msg('Plotting tuning results')
+  
+  fvc_rf$plots <- 
+    list(data = fvc_rf$stats, 
[…]

### 2023-06-14 LFT scripts/rf_FVC_findings.R

@@ -1,0 +1,88 @@
+# Modeling of the FVC findings with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of the FVC findings
+# Unmatched CV folds: modeling simple presence/absence of the LFT findings
+
+  insert_head()
+  
+# container ------
+  
+  fvc_red_rf <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  fvc_red_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fvc_red_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(FVC_reduced, all_of(fvc_red_rf$variables))
+  
+  fvc_red_rf$formula <- lft_globals$formulas$FVC_reduced
+  
+  fvc_red_rf$train_control <- lft_globals$train_control
+  
+# tuning grid ------
+  
+  insert_msg('Tuning grids')
+  
+  fvc_red_rf$tune_grid <- 
+    expand.grid(splitrule = c('gini', 'extratrees'), 
+                mtry = seq(2, length(fvc_red_rf$variables), by = 3), 
+                min.node.size = c(1, 3), 
+                stringsAsFactors = FALSE)
+  
+# modeling ------
+  
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  fvc_red_rf$models <- fvc_red_rf$train_control %>% 
+    map(~train(form = fvc_red_rf$formula, 
+               data = fvc_red_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'Kappa', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = fvc_red_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  fvc_red_rf$models <- fvc_red_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+  
+  insert_msg('Extracting tuning results')
+  
+  fvc_red_rf$stats <- fvc_red_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  fvc_red_rf$optima <- fvc_red_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+  
+# Plotting tuning results ------
+  
[…]

### 2023-06-14 LFT scripts/rf_LFT_findings.R

@@ -1,0 +1,88 @@
+# Modeling of the LF findings with Random Forests (Ranger)
+# Matched CV folds: modeling trajectories of the LFT findings
+# Unmatched CV folds: modeling simple presence/absence of the LFT findings
+
+  insert_head()
+  
+# container ------
+  
+  lft_rf <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  lft_rf$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  lft_rf$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(LFT_findings, all_of(lft_rf$variables))
+  
+  lft_rf$formula <- lft_globals$formulas$LFT_findings
+  
+  lft_rf$train_control <- lft_globals$train_control
+  
+# tuning grid ------
+  
+  insert_msg('Tuning grids')
+  
+  lft_rf$tune_grid <- 
+    expand.grid(splitrule = c('gini', 'extratrees'), 
+                mtry = seq(2, length(lft_rf$variables), by = 3), 
+                min.node.size = c(1, 3), 
+                stringsAsFactors = FALSE)
+  
+# modeling ------
+  
+  insert_msg('Modeling')
+  
+  registerDoParallel(cores = 7)
+  
+  set.seed(12345)
+  
+  lft_rf$models <- lft_rf$train_control %>% 
+    map(~train(form = lft_rf$formula, 
+               data = lft_rf$analysis_tbl, 
+               method = 'ranger', 
+               metric = 'Kappa', 
+               trControl = .x, 
+               num.trees = 1000, 
+               tuneGrid = lft_rf$tune_grid))
+  
+  stopImplicitCluster()
+  
+  lft_rf$models <- lft_rf$models %>% 
+    map(as_caretx)
+
+# Tuning results (kappa) ------
+  
+  insert_msg('Extracting tuning results')
+  
+  lft_rf$stats <- lft_rf$models %>% 
+    map(~.x$results) %>% 
+    map(as_tibble)
+  
+  lft_rf$optima <- lft_rf$models %>% 
+    map(~.x$bestTune) %>% 
+    map(as_tibble)
+  
+# Plotting tuning results ------
+  
+  insert_msg('Plotting tuning results')
+  
+  lft_rf$plots <- 
+    list(data = lft_rf$stats, 
+         plot_title = c('ID-matched models', 
[…]

### 2023-06-14 LFT scripts/roc_ct.R

@@ -1,0 +1,231 @@
+# Optimal cutpoints for opacity, high opacity and CTSS
+# for detection of any CT abnormalities, reduced DLCO, 
+# reduced FVC and reduced FEV1
+# 
+# The optimal cutoff is determined by the Youden criterion
+
+  insert_head()
+  
+# container -------
+  
+  lft_roc <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## responses and marker variables
+  
+  lft_roc$variables <- 
+    c('opacity_percent', 'high_opacity_percent', 'CTSS')
+  
+  lft_roc$variables <- 
+    set_names(lft_roc$variables, 
+              lft_roc$variables)
+  
+  lft_roc$responses <- 
+    c('LFT_findings', 
+      'DLCO_reduced', 
+      'FVC_reduced', 
+      'FEV1_reduced')
+  
+  lft_roc$responses <- 
+    set_names(lft_roc$responses, 
+              lft_roc$responses)
+  
+  ## analysis table
+  
+  lft_roc$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(all_of(lft_roc$responses), 
+           all_of(lft_roc$variables)) 
+  
+  ## marker colors
+  
+  lft_roc$variable_colors <- 
+    c('opacity_percent' = 'orangered3', 
+      'high_opacity_percent' = 'gray30', 
+      'CTSS' = 'steelblue')
+  
+# N numbers ------
+  
+  insert_msg('N numbers')
+  
+  lft_roc$n_numbers <- lft_roc$responses %>% 
+    map(~count(lft_roc$analysis_tbl, .data[[.x]]))
+  
+  lft_roc$n_caps <- lft_roc$n_numbers %>% 
+    map(~paste0('total: n = ', sum(.x$n), 
+                ', findings: n = ', .x$n[2]))
+  
+# Optimal cutpoint objects ------
+  
+  insert_msg('Optimal cutpoint objects')
+  
+  lft_roc$cut_obj <- lft_roc$responses %>% 
+    map(function(resp) lft_roc$variables %>% 
+          map(~optimal.cutpoints(.x, 
+                                 status = resp, 
+                                 tag.healthy = 'no', 
+                                 methods = 'Youden', 
+                                 data = lft_roc$analysis_tbl)))
+  
+# Cutoff stats -----
+  
[…]

### 2023-06-14 LFT scripts/univariable_DLCO.R

@@ -1,0 +1,216 @@
+# Association of explanatory variables with DLCO expressed as percentage 
+# of reference value
+# Numeric variables: Spearman's correlation
+# Categorical variables: comparison of DLCO values between the categories 
+# by Mann-Whitney or Kruskal-Wallis test.
+
+  insert_head()
+
+# container -------
+
+  dlco_uni <- list()
+
+# analysis globals -------
+
+insert_msg('Analysis globals')
+  
+  ## variables and analysis table
+
+  dlco_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+
+  dlco_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'DLCO_percent', dlco_uni$variables)]
+  
+  ## variable lexicon, appended with the plot type
+  ## and type of the statistical test
+  
+  dlco_uni$lexicon <- lft_globals$lexicon %>% 
+    filter(variable %in% dlco_uni$variables)
+  
+  dlco_uni$lexicon$level_no <- 
+    dlco_uni$analysis_tbl[dlco_uni$lexicon$variable] %>% 
+    map(levels) %>% 
+    map_dbl(length)
+  
+  dlco_uni$lexicon <- dlco_uni$lexicon %>% 
+    mutate(plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'factor', 
+                          '% of reference', 
+                          table_label), 
+           analysis_type = ifelse(format == 'factor', 
+                                  'comparison', 'correlation'), 
+           analysis_type = factor(analysis_type, 
+                                  c('comparison', 'correlation')), 
+           test_type = ifelse(level_no == 0, 
+                              'spearman', 
+                              ifelse(level_no == 2, 
+                                     'wilcoxon_r', 'kruskal_etasq')))
+  
+  dlco_uni$lexicon$variable <- 
+    set_names(dlco_uni$lexicon$variable, 
+              dlco_uni$lexicon$variable)
+  
+  dlco_uni$lexicon <- dlco_uni$lexicon %>% 
+    blast(analysis_type)
+
+# Comparison -------
+  
+  insert_msg('Comparison')
+  
+  ## descriptive stats
+  
+  dlco_uni$comparison$stats <- dlco_uni$lexicon$comparison$variable %>% 
+    map(~explore(data = dlco_uni$analysis_tbl, 
[…]

### 2023-06-14 LFT scripts/univariable_DLCO_findings.R

@@ -1,0 +1,137 @@
+# Comparison of explanatory variables between observations with and without
+# DLCO abnormalities
+# Numeric variables: Mann-Whitney test with r effect size statistic
+# Categorical variables: Chi-squared test with Cramer's V effect size statistic
+
+  insert_head()
+
+# container -------
+
+  dlco_red_uni <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  dlco_red_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  dlco_red_uni$lexicon <- lft_globals$lexicon %>% 
+    mutate(test_type = ifelse(format == 'numeric', 
+                              'wilcoxon_r', 'cramer_v'), 
+           plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'numeric', 
+                          unit, '% of strata')) %>% 
+    filter(variable %in% dlco_red_uni$variables)
+  
+  dlco_red_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'DLCO_reduced', dlco_red_uni$variables)] %>% 
+    mutate(LFT_findings = car::recode(DLCO_reduced, 
+                                      "'yes' = 'DLCO low'; 'no' = 'DLCO high'"), 
+           LFT_findings = factor(DLCO_reduced, c('DLCO high', 'DLCO low')))
+
+# Descriptive stats ------
+
+  insert_msg('Descriptive stats')
+  
+  ## n numbers
+  
+  dlco_red_uni$n_numbers <- dlco_red_uni$analysis_tbl %>% 
+    count(DLCO_reduced)
+  
+  dlco_red_uni$n_caps <- 
+    paste0('total: n = ', sum(dlco_red_uni$n_numbers$n), 
+           ', events: n = ', dlco_red_uni$n_numbers$n[2])
+
+  ## stats
+  
+  dlco_red_uni$stats <- dlco_red_uni$analysis_tbl %>% 
+    explore(split_factor = 'DLCO_reduced', 
+            variables = dlco_red_uni$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    set_names(c('variable', 'DLCO \u2265 80%', 'DLCO < 80%'))
+
+# Testing --------
+
[…]

### 2023-06-14 LFT scripts/univariable_FEV1.R

@@ -1,0 +1,208 @@
+# Association of explanatory variables with FEV1 expressed as percentage 
+# of reference value
+# Numeric variables: Spearman's correlation
+# Categorical variables: comparison of FEV1 values between the categories 
+# by Mann-Whitney or Kruskal-Wallis test.
+
+  insert_head()
+
+# container -------
+
+  fev1_uni <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  ## variables and analysis table
+
+  fev1_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+
+  fev1_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'FEV1_percent', fev1_uni$variables)]
+  
+  ## variable lexicon, appended with the plot type
+  ## and type of the statistical test
+  
+  fev1_uni$lexicon <- lft_globals$lexicon %>% 
+    filter(variable %in% fev1_uni$variables)
+  
+  fev1_uni$lexicon$level_no <- 
+    fev1_uni$analysis_tbl[fev1_uni$lexicon$variable] %>% 
+    map(levels) %>% 
+    map_dbl(length)
+  
+  fev1_uni$lexicon <- fev1_uni$lexicon %>% 
+    mutate(plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'factor', 
+                          '% of reference', 
+                          table_label), 
+           analysis_type = ifelse(format == 'factor', 
+                                  'comparison', 'correlation'), 
+           analysis_type = factor(analysis_type, 
+                                  c('comparison', 'correlation')), 
+           test_type = ifelse(level_no == 0, 
+                              'spearman', 
+                              ifelse(level_no == 2, 
+                                     'wilcoxon_r', 'kruskal_etasq')))
+  
+  fev1_uni$lexicon$variable <- 
+    set_names(fev1_uni$lexicon$variable, 
+              fev1_uni$lexicon$variable)
+  
+  fev1_uni$lexicon <- fev1_uni$lexicon %>% 
+    blast(analysis_type)
+
+# Comparison -------
+  
+  insert_msg('Comparison')
+  
+  ## descriptive stats
+  
+  fev1_uni$comparison$stats <- fev1_uni$lexicon$comparison$variable %>% 
+    map(~explore(data = fev1_uni$analysis_tbl, 
[…]

### 2023-06-14 LFT scripts/univariable_FEV1_findings.R

@@ -1,0 +1,137 @@
+# Comparison of explanatory variables between observations with and without
+# FEV1 abnormalities
+# Numeric variables: Mann-Whitney test with r effect size statistic
+# Categorical variables: Chi-squared test with Cramer's V effect size statistic
+
+  insert_head()
+  
+# container -------
+  
+  fev1_red_uni <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  fev1_red_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fev1_red_uni$lexicon <- lft_globals$lexicon %>% 
+    mutate(test_type = ifelse(format == 'numeric', 
+                              'wilcoxon_r', 'cramer_v'), 
+           plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'numeric', 
+                          unit, '% of strata')) %>% 
+    filter(variable %in% fev1_red_uni$variables)
+  
+  fev1_red_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'FEV1_reduced', fev1_red_uni$variables)] %>% 
+    mutate(FEV1_reduced = car::recode(FEV1_reduced, 
+                                      "'yes' = 'FEV1 low'; 'no' = 'FEV1 high'"), 
+           FEV1_reduced = factor(FEV1_reduced, c('FEV1 high', 'FEV1 low')))
+  
+# Descriptive stats ------
+  
+  insert_msg('Descriptive stats')
+  
+  ## n numbers
+  
+  fev1_red_uni$n_numbers <- fev1_red_uni$analysis_tbl %>% 
+    count(FEV1_reduced)
+  
+  fev1_red_uni$n_caps <- 
+    paste0('total: n = ', sum(fev1_red_uni$n_numbers$n), 
+           ', events: n = ', fev1_red_uni$n_numbers$n[2])
+  
+  ## stats
+  
+  fev1_red_uni$stats <- fev1_red_uni$analysis_tbl %>% 
+    explore(split_factor = 'FEV1_reduced', 
+            variables = fev1_red_uni$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    set_names(c('variable', 'FEV1 \u2265 80%', 'FEV1 < 80%'))
+  
+# Testing --------
[…]

### 2023-06-14 LFT scripts/univariable_FVC.R

@@ -1,0 +1,216 @@
+# Association of explanatory variables with FVC expressed as percentage 
+# of reference value
+# Numeric variables: Spearman's correlation
+# Categorical variables: comparison of FVC values between the categories 
+# by Mann-Whitney or Kruskal-Wallis test.
+
+  insert_head()
+
+# container -------
+
+  fvc_uni <- list()
+
+# analysis globals -------
+
+insert_msg('Analysis globals')
+  
+  ## variables and analysis table
+
+  fvc_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+
+  fvc_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'FVC_percent', fvc_uni$variables)]
+  
+  ## variable lexicon, appended with the plot type
+  ## and type of the statistical test
+  
+  fvc_uni$lexicon <- lft_globals$lexicon %>% 
+    filter(variable %in% fvc_uni$variables)
+  
+  fvc_uni$lexicon$level_no <- 
+    fvc_uni$analysis_tbl[fvc_uni$lexicon$variable] %>% 
+    map(levels) %>% 
+    map_dbl(length)
+  
+  fvc_uni$lexicon <- fvc_uni$lexicon %>% 
+    mutate(plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'factor', 
+                          '% of reference', 
+                          table_label), 
+           analysis_type = ifelse(format == 'factor', 
+                                  'comparison', 'correlation'), 
+           analysis_type = factor(analysis_type, 
+                                  c('comparison', 'correlation')), 
+           test_type = ifelse(level_no == 0, 
+                              'spearman', 
+                              ifelse(level_no == 2, 
+                                     'wilcoxon_r', 'kruskal_etasq')))
+  
+  fvc_uni$lexicon$variable <- 
+    set_names(fvc_uni$lexicon$variable, 
+              fvc_uni$lexicon$variable)
+  
+  fvc_uni$lexicon <- fvc_uni$lexicon %>% 
+    blast(analysis_type)
+
+# Comparison -------
+  
+  insert_msg('Comparison')
+  
+  ## descriptive stats
+  
+  fvc_uni$comparison$stats <- fvc_uni$lexicon$comparison$variable %>% 
+    map(~explore(data = fvc_uni$analysis_tbl, 
[…]

### 2023-06-14 LFT scripts/univariable_FVC_findings.R

@@ -1,0 +1,137 @@
+# Comparison of explanatory variables between observations with and without
+# FVC abnormalities
+# Numeric variables: Mann-Whitney test with r effect size statistic
+# Categorical variables: Chi-squared test with Cramer's V effect size statistic
+
+  insert_head()
+  
+# container -------
+  
+  fvc_red_uni <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  fvc_red_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  fvc_red_uni$lexicon <- lft_globals$lexicon %>% 
+    mutate(test_type = ifelse(format == 'numeric', 
+                              'wilcoxon_r', 'cramer_v'), 
+           plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'numeric', 
+                          unit, '% of strata')) %>% 
+    filter(variable %in% fvc_red_uni$variables)
+  
+  fvc_red_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'FVC_reduced', fvc_red_uni$variables)] %>% 
+    mutate(FVC_reduced = car::recode(FVC_reduced, 
+                                      "'yes' = 'FVC low'; 'no' = 'FVC high'"), 
+           FVC_reduced = factor(FVC_reduced, c('FVC high', 'FVC low')))
+  
+# Descriptive stats ------
+  
+  insert_msg('Descriptive stats')
+  
+  ## n numbers
+  
+  fvc_red_uni$n_numbers <- fvc_red_uni$analysis_tbl %>% 
+    count(FVC_reduced)
+  
+  fvc_red_uni$n_caps <- 
+    paste0('total: n = ', sum(fvc_red_uni$n_numbers$n), 
+           ', events: n = ', fvc_red_uni$n_numbers$n[2])
+  
+  ## stats
+  
+  fvc_red_uni$stats <- fvc_red_uni$analysis_tbl %>% 
+    explore(split_factor = 'FVC_reduced', 
+            variables = fvc_red_uni$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    set_names(c('variable', 'FVC \u2265 80%', 'FVC < 80%'))
+  
+# Testing --------
+  
+  insert_msg('Testing')
[…]

### 2023-06-14 LFT scripts/univariable_LFT_findings.R

@@ -1,0 +1,137 @@
+# Comparison of explanatory variables between observations with and without
+# LFT abnormalities
+# Numeric variables: Mann-Whitney test with r effect size statistic
+# Categorical variables: Chi-squared test with Cramer's V effect size statistic
+
+  insert_head()
+  
+# container -------
+  
+  lft_uni <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  lft_uni$variables <- 
+    lft_globals[c("ct_variables", "baseline_variables")] %>% 
+    reduce(c)
+  
+  lft_uni$lexicon <- lft_globals$lexicon %>% 
+    mutate(test_type = ifelse(format == 'numeric', 
+                              'wilcoxon_r', 'cramer_v'), 
+           plot_type = ifelse(format == 'numeric', 
+                              'box', 'stack'), 
+           y_lab = ifelse(format == 'numeric', 
+                          unit, '% of strata')) %>% 
+    filter(variable %in% lft_uni$variables)
+  
+  lft_uni$analysis_tbl <- 
+    lft_globals$analysis_tbl[c('ID', 'LFT_findings', lft_uni$variables)] %>% 
+    mutate(LFT_findings = car::recode(LFT_findings, 
+                                      "'yes' = 'LFT+'; 'no' = 'LFT-'"), 
+           LFT_findings = factor(LFT_findings, c('LFT-', 'LFT+')))
+  
+# Descriptive stats ------
+  
+  insert_msg('Descriptive stats')
+  
+  ## n numbers
+  
+  lft_uni$n_numbers <- lft_uni$analysis_tbl %>% 
+    count(LFT_findings)
+  
+  lft_uni$n_caps <- 
+    paste0('total: n = ', sum(lft_uni$n_numbers$n), 
+           ', events: n = ', lft_uni$n_numbers$n[2])
+  
+  ## stats
+  
+  lft_uni$stats <- lft_uni$analysis_tbl %>% 
+    explore(split_factor = 'LFT_findings', 
+            variables = lft_uni$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    set_names(c('variable', 'LFT findings absent', 'LFT findings present'))
+  
+# Testing --------
+  
+  insert_msg('Testing')
+  
+  lft_uni$test <- lft_uni$analysis_tbl %>% 
+    compare_variables(variables = lft_uni$lexicon$variable, 
[…]

### 2023-06-14 ctss.R

@@ -1,0 +1,95 @@
+# Concordance of CTSS, AI-determined opacity and high opacity
+#
+# A general problem with the CTSS and opacity relationship: 
+# integration of these two variables.
+# An extreme case: let's assume, one lap of the lung is completely destroyed, 
+# this gives CTSS = 5, and opacity of approx. 20%. 
+# Another extreme case: single subtle GGOs in each of the laps, 
+# this gives CTSS = 5, and opacity of approx. 5%.
+# So those two parameters measure clearly different outcomes. 
+# Integration of CTSS by sum over all 5 lung laps may lead to problems, 
+# since it does not discriminate between locally severe alteratons and globally
+# mild abnormalities!
+
+# tools -------
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+  
+  library(caret)
+  library(exda)
+  library(rstatix)
+
+  library(plotROC)
+  library(OptimalCutpoints)
+  
+  library(soucer)
+  library(furrr)
+  library(MASS)
+  
+  library(lmqc)
+
+  insert_head()
+  
+  select <- dplyr::select
+  
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# Analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ctss_globals <- list()
+  
+  ## variables
+  
+  ctss_globals$variables <- 
+    c('CTSS', 'CTSS_class', 
+      'opacity_percent', 
+      'high_opacity_percent', 
+      'opacity_class')
+  
+  ctss_globals$lexicon <- covild$ct_lexicon %>% 
+    filter(variable %in% ctss_globals$variables)
+  
+  ## analysis tables
+  
+  ctss_globals$analysis_tbl <- 
+    left_join(covild$ct,
+              covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    select(ID, severity_class, follow_up, any_of(ctss_globals$variables))
+
+  ctss_globals$analysis_tbl <- ctss_globals$analysis_tbl %>% 
+    mutate(severity_class = 'cohort') %>%
+    rbind(ctss_globals$analysis_tbl) %>% 
+    mutate(severity_class = factor(severity_class, 
+                                   c('cohort', 
[…]

### 2023-06-14 cutpoint scripts/cutoff_high.R

@@ -1,0 +1,166 @@
+# Searching for the optimal cutoff of AI high opacity 
+# to detect any CT abnormalities, 
+# GGOs, reticulation, consolidation and bronchiectasis. Youden criterion 
+# implemented by the optimalCutpoint package
+
+  insert_head()
+  
+# container -------
+  
+  cut_high <- list()
+  
+# Analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## CT findings defined manually to serve as responses
+  
+  cut_high$responses <- cut_globals$responses
+
+  ## analysis table
+  
+  cut_high$analysis_tbl <- covild$ct %>% 
+    select(ID, follow_up, high_opacity_percent, all_of(cut_high$responses)) %>% 
+    filter(complete.cases(.), 
+           follow_up != 'acute COVID-19')
+  
+# N numbers ------
+  
+  insert_msg('N numbers')
+  
+  cut_high$n_numbers <- cut_high$responses %>% 
+    map(~count(cut_high$analysis_tbl, .data[[.x]]))
+  
+  cut_high$n_caps <- cut_high$n_numbers %>% 
+    map(~paste0('total: n = ', sum(.x$n), 
+                ', findings: n = ', .x$n[2]))
+  
+# Optimal cutpoint objects ------
+  
+  insert_msg('Optimal cutpoint objects')
+  
+  cut_high$cut_obj <- cut_high$responses %>% 
+    map(~optimal.cutpoints('high_opacity_percent', 
+                           status = .x, 
+                           tag.healthy = 'no', 
+                           methods = 'Youden', 
+                           data = as.data.frame(cut_high$analysis_tbl)))
+  
+# Extracting the optimal cutpoint values and its stats ------
+  
+  insert_msg('Extracting the optimal cutpoint values and stats')
+  
+  cut_high$stats <- cut_high$cut_obj %>% 
+    map(extract_cut) %>% 
+    map(~.x[c('all_cutoffs', 'optimal_cutoff')]) %>% 
+    transpose
+  
+  cut_high$stats$optimal_cutoff <- cut_high$stats$optimal_cutoff %>% 
+    compress(names_to = 'response') %>% 
+    mutate(response = factor(response, cut_high$responses), 
+           auc_lab = paste0('AUC = ', signif(auc, 2), 
+                            ' [', signif(lower_ci, 3), 
+                            ' - ', signif(upper_ci, 2), ']'), 
+           plot_lab = paste0('Se = ', signif(Se, 2), 
[…]

### 2023-06-14 cutpoint scripts/cutoff_opacity.R

@@ -1,0 +1,165 @@
+# Searching for the optimal cutoff of AI opacity to detect any CT abnormalities, 
+# GGOs, reticulation, consolidation and bronchiectasis. Youden criterion 
+# implemented by the optimalCutpoint package
+
+  insert_head()
+  
+# container -------
+  
+  cut_opacity <- list()
+  
+# Analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## CT findings defined manually to serve as responses
+  
+  cut_opacity$responses <- cut_globals$responses
+
+  ## analysis table
+  
+  cut_opacity$analysis_tbl <- covild$ct %>% 
+    select(ID, follow_up, opacity_percent, all_of(cut_opacity$responses)) %>% 
+    filter(complete.cases(.),
+           follow_up != 'acute COVID-19')
+  
+# N numbers ------
+  
+  insert_msg('N numbers')
+  
+  cut_opacity$n_numbers <- cut_opacity$responses %>% 
+    map(~count(cut_opacity$analysis_tbl, .data[[.x]]))
+  
+  cut_opacity$n_caps <- cut_opacity$n_numbers %>% 
+    map(~paste0('total: n = ', sum(.x$n), 
+                ', findings: n = ', .x$n[2]))
+  
+# Optimal cutpoint objects ------
+  
+  insert_msg('Optimal cutpoint objects')
+  
+  cut_opacity$cut_obj <- cut_opacity$responses %>% 
+    map(~optimal.cutpoints('opacity_percent', 
+                           status = .x, 
+                           tag.healthy = 'no', 
+                           methods = 'Youden', 
+                           data = as.data.frame(cut_opacity$analysis_tbl)))
+  
+# Extracting the optimal cutpoint values and its stats ------
+  
+  insert_msg('Extracting the optimal cutpoint values and stats')
+  
+  cut_opacity$stats <- cut_opacity$cut_obj %>% 
+    map(extract_cut) %>% 
+    map(~.x[c('all_cutoffs', 'optimal_cutoff')]) %>% 
+    transpose
+  
+  cut_opacity$stats$optimal_cutoff <- cut_opacity$stats$optimal_cutoff %>% 
+    compress(names_to = 'response') %>% 
+    mutate(response = factor(response, cut_opacity$responses), 
+           auc_lab = paste0('AUC = ', signif(auc, 2), 
+                            ' [', signif(lower_ci, 3), 
+                            ' - ', signif(upper_ci, 2), ']'), 
[…]

### 2023-06-14 cutpoint scripts/high_strata.R

@@ -1,0 +1,160 @@
+# Testing for performance of the opacity cutoff at detecting any CT findings,  
+# GGOs, consolidations etc. in the severity strata and timepoints after 
+# infection
+
+  insert_head()
+  
+# container -----
+  
+  high_strata <- list()
+  
+# Parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## optimal cutoffs identified by the Youden criterion
+  
+  high_strata$responses <- cut_globals$responses
+  
+  high_strata$cutoffs <- cut_high$stats$optimal_cutoff$cutoff %>% 
+    set_names(cut_high$stats$optimal_cutoff$response)
+  
+  high_strata$markers <- stri_replace(high_strata$responses, 
+                                    regex = '_find.*$', 
+                                    replacement = '_marker')
+  
+  ## analysis tables with the binarized AI-determined opacity
+  
+  high_strata$analysis_tbl <- covild$ct %>% 
+    select(ID, follow_up, high_opacity_percent, all_of(high_strata$responses)) %>% 
+    left_join(covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    filter(complete.cases(.), 
+           follow_up != 'acute COVID-19') %>% 
+    mutate(follow_up = droplevels(follow_up))
+  
+  ## binarization
+  
+  for(i in pmap(list(high_strata$marker, high_strata$cutoffs), list)) {
+    
+    high_strata$analysis_tbl <- high_strata$analysis_tbl %>% 
+      mutate(!!i[[1]] := cut(high_opacity_percent, 
+                             c(-Inf, i[[2]], Inf), 
+                             c('no', 'yes'), 
+                             right = FALSE))
+    
+  }
+  
+  ## merging with the data table for cohort stats
+  
+  high_strata$analysis_tbl <- high_strata$analysis_tbl %>% 
+    rbind(high_strata$analysis_tbl %>% 
+            mutate(follow_up = 'cohort', 
+                   severity_class = 'cohort'), .) %>% 
+    mutate(follow_up = factor(follow_up, 
+                              c('cohort', 
+                                levels(high_strata$analysis_tbl$follow_up))), 
+           severity_class = factor(severity_class, 
+                                   c('cohort', 
+                                     levels(high_strata$analysis_tbl$severity_class))))
+
[…]

### 2023-06-14 cutpoint scripts/opacity_strata.R

@@ -1,0 +1,160 @@
+# Testing for performance of the opacity cutoff at detecting any CT findings,  
+# GGOs, consolidations etc. in the severity strata and timepoints after 
+# infection
+
+  insert_head()
+  
+# container -----
+  
+  op_strata <- list()
+  
+# Parallel backend ------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## optimal cutoffs identified by the Youden criterion
+  
+  op_strata$responses <- cut_globals$responses
+  
+  op_strata$cutoffs <- cut_opacity$stats$optimal_cutoff$cutoff %>% 
+    set_names(cut_opacity$stats$optimal_cutoff$response)
+  
+  op_strata$markers <- stri_replace(op_strata$responses, 
+                                    regex = '_find.*$', 
+                                    replacement = '_marker')
+  
+  ## analysis tables with the binarized AI-determined opacity
+  
+  op_strata$analysis_tbl <- covild$ct %>% 
+    select(ID, follow_up, opacity_percent, all_of(op_strata$responses)) %>% 
+    left_join(covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    filter(complete.cases(.), 
+           follow_up != 'acute COVID-19') %>% 
+    mutate(follow_up = droplevels(follow_up))
+  
+  ## binarization
+  
+  for(i in pmap(list(op_strata$marker, op_strata$cutoffs), list)) {
+    
+    op_strata$analysis_tbl <- op_strata$analysis_tbl %>% 
+      mutate(!!i[[1]] := cut(opacity_percent, 
+                             c(-Inf, i[[2]], Inf), 
+                             c('no', 'yes'), 
+                             right = FALSE))
+    
+  }
+  
+  ## merging with the data table for cohort stats
+  
+  op_strata$analysis_tbl <- op_strata$analysis_tbl %>% 
+    rbind(op_strata$analysis_tbl %>% 
+            mutate(follow_up = 'cohort', 
+                   severity_class = 'cohort'), .) %>% 
+    mutate(follow_up = factor(follow_up, 
+                              c('cohort', 
+                                levels(op_strata$analysis_tbl$follow_up))), 
+           severity_class = factor(severity_class, 
+                                   c('cohort', 
+                                     levels(op_strata$analysis_tbl$severity_class))))
+
[…]

### 2023-06-14 cutpoint.R

@@ -1,0 +1,75 @@
+# Inter-rater reliability of the AI-detected opacity and high opacity 
+# as markers of manually-detected, clinically relevant CT abnormalities
+#
+# 1) Definition of the optimal cutoff of AI-opacity and AI-high opacity
+# by ROC and Youden criterion. The entire cohort and timepoints are used
+# for the cutoff search
+#
+# 2) Computation of the sensitivity, specificity, kappa and accuracy for the 
+# optimal cutoffs in the COVID-19 severity strata and for particular timepoints
+
+# tools -----
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+
+  library(OptimalCutpoints)
+  library(plotROC)
+  library(caret)
+
+  library(soucer)
+  library(furrr)
+
+  insert_head()
+  
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+
+  cut_globals <- list()
+  
+  ## responses
+  
+  cut_globals$responses <- 
+    c('CT_findings', 
+      'GGO_finding', 
+      'reticulation_finding', 
+      'consolidation_finding', 
+      'bronchiectasis_finding') 
+  
+  
+  cut_globals$responses <- 
+    set_names(cut_globals$responses, 
+              cut_globals$responses)
+  
+  ## their dictionary and colors
+  
+  cut_globals$lexicon <- covild$ct_lexicon %>% 
+    filter(variable %in% cut_globals$responses)
+  
+  cut_globals$response_colors <- 
+    c('CT_findings' = 'coral3', 
+      'GGO_finding' = 'firebrick3', 
+      'reticulation_finding' = 'coral4', 
+      'consolidation_finding' = 'gray40', 
+      'bronchiectasis_finding' = 'steelblue4')
+    
+# analysis scripts ------
+  
+  insert_msg('Analysis scripts')
+  
+  c('./cutpoint scripts/cutoff_opacity.R', 
+    './cutpoint scripts/cutoff_high.R', 
+    './cutpoint scripts/opacity_strata.R', 
+    './cutpoint scripts/high_strata.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# END -----
+  
+  insert_tail()

### 2023-06-14 exec.R

@@ -1,0 +1,14 @@
+# Launches the entire analysis pipeline
+
+  library(soucer)
+
+  print(source_all(c('import.R', 
+                     'exploration.R', 
+                     'kinetic.R', 
+                     'cutpoint.R', 
+                     'ctss.R', 
+                     'lft.R', 
+                     'report.R'), 
+                   message = TRUE, crash = TRUE))
+  
+  save.image()

### 2023-06-14 exploration scripts/baseline_cohort.R

@@ -1,0 +1,79 @@
+# Baseline characteristic of the study cohort
+# Comparison of the study readouts between the severity strata
+# with KruskalWallis or Chi-Squared test
+
+  insert_head()
+  
+# container -----
+  
+  cohort <- list()
+  
+# analysis globals ------
+  
+  insert_msg('Analysis globals')
+
+  cohort$lexicon <- covild$baseline_lexicon %>% 
+    mutate(test_type = ifelse(format == 'numeric', 
+                              'kruskal_eta', 'cramer_v'))
+  
+  ## analysis table: the entire cohort
+  ## and particular severity strata
+  
+  cohort$analysis_tbl <- 
+    c(list(cohort = covild$baseline), 
+      blast(covild$baseline, severity_class)) %>% 
+    compress(names_to = 'cohort_split') %>% 
+    mutate(cohort_split = factor(cohort_split, 
+                                 c('cohort', 
+                                   levels(covild$baseline$severity_class))))
+  
+# Descriptive statistics ------
+  
+  insert_msg('Descriptive statistic')
+  
+  ## n numbers
+  
+  cohort$n_numbers <- cohort$analysis_tbl %>% 
+    count(cohort_split)
+  
+  ## stats
+  
+  cohort$stats <- cohort$analysis_tbl %>% 
+    explore(split_factor = 'cohort_split', 
+            variables = cohort$lexicon$variable, 
+            what = 'table', 
+            pub_styled = TRUE) %>% 
+    reduce(left_join, by = 'variable') %>% 
+    set_names(c('variable', levels(cohort$analysis_tbl$cohort_split)))
+  
+# Testing for differences between the severity strata -----
+  
+  insert_msg('Testing for differences between the severity strata')
+  
+  cohort$test <- covild$baseline %>% 
+    compare_variables(variables = cohort$lexicon$variable, 
+                      split_factor = 'severity_class', 
+                      what = 'eff_size', 
+                      types = cohort$lexicon$test_type, 
+                      exact = FALSE, 
+                      ci = FALSE, 
+                      pub_styled = TRUE, 
+                      adj_method = 'BH')
+  
+# Result table ------
+  
+  insert_msg('Result table')
+  
+  cohort$result_tbl <- 
+    left_join(cohort$stats, 
+              cohort$test[c('variable', 'significance', 'eff_size')], 
+              by = 'variable') %>% 
[…]

### 2023-06-14 exploration scripts/ct.R

@@ -1,0 +1,149 @@
+# Descriptive stats for the CT parameters split be the timepoint 
+# and COVID-19 severity. No statistical testing is done - a 'proper' 
+# statistical analysis is done later on with participant-matched data.
+#
+# This analytic task is here to get an idea, how the variables change
+# in time and severity
+
+  insert_head()
+  
+# container ------
+  
+  cohort_ct <- list()
+  
+# Analysis globals ------
+  
+  insert_msg('Analysis globals')
+  
+  cohort_ct$lexicon <- covild$ct_lexicon
+  
+  ## numeric and categorical variables
+  
+  cohort_ct$variables <- cohort_ct$lexicon %>% 
+    blast(format) %>% 
+    map(~.x$variable)
+  
+  ## analysis table, introducing the severity variable
+  ## removing the acute timepoint - there was no study CT but just 
+  ## a diagnostic scan for some participants
+  
+  cohort_ct$analysis_tbl <- covild$ct %>% 
+    left_join(covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    mutate(cohort_split = severity_class)
+  
+  cohort_ct$analysis_tbl <- cohort_ct$analysis_tbl %>% 
+    full_rbind(covild$ct) %>% 
+    filter(follow_up != 'acute COVID-19') %>% 
+    mutate(cohort_split = ifelse(is.na(cohort_split), 
+                                 'cohort', as.character(cohort_split)), 
+           cohort_split = factor(cohort_split, 
+                                 c('cohort', 
+                                   levels(cohort_ct$analysis_tbl$severity_class))), 
+           follow_up = droplevels(follow_up), 
+           analysis_split = interaction(follow_up, cohort_split)) %>% 
+    relocate(cohort_split) %>% 
+    relocate(analysis_split) %>% 
+    relocate(ID)
+
+# Number of complete cases per follow-up ------
+  
+  insert_msg('Number of complete cases per follow-up')
+  
+  cohort_ct$n_numbers <- cohort_ct$analysis_tbl %>% 
+    filter(cohort_split != 'cohort') %>% 
+    blast(severity_class, follow_up) %>%
+    map(~filter(.x, complete.cases(.x))) %>% 
+    map_dbl(nrow) %>% 
+    compress(names_to = 'strata', 
+             value = 'n') %>% 
+    mutate(severity_class = stri_split_fixed(strata, 
[…]

### 2023-06-14 exploration scripts/distribution_baseline.R

@@ -1,0 +1,92 @@
+# Distribution of the numeric baseline variables
+
+  insert_head()
+  
+# container ------
+  
+  distr_base <- list()
+  
+# analysis globals -----
+  
+  insert_msg('Analysis globals')
+  
+  ## numeric variables
+  
+  distr_base$lexicon <- covild$baseline_lexicon %>% 
+    filter(format == 'numeric', 
+           !variable %in% c('ID', 'follow_up'))
+  
+  distr_base$analysis_tbl <- covild$baseline %>% 
+    select(all_of(distr_base$lexicon$variable)) %>% 
+    mutate(weight_change_kg = -min(weight_change_kg) + weight_change_kg)
+  
+# Shapiro-Wilk test, mean and normality -----
+  
+  insert_msg('Shairo-Wilk test')
+  
+  distr_base$test <- distr_base$lexicon$variable %>% 
+    map(test_long_distribution, 
+        data = distr_base$analysis_tbl, 
+        split_factor = NULL) %>% 
+    set_names(distr_base$lexicon$variable) %>% 
+    compress(names_to = 'variable')
+  
+# Plotting the Shapiro-Wilk testing results ------
+  
+  insert_msg('Shapiro-Wilk testing results')
+  
+  distr_base$shapiro_plot <- distr_base$test %>% 
+    filter(best_transf == 'yes') %>% 
+    mutate(var_lab = exchange(variable, dict = distr_base$lexicon), 
+           var_lab  = paste(var_lab, transformation, sep = ', ')) %>% 
+    ggplot(aes(x = statistic, 
+               y = reorder(var_lab, statistic))) + 
+    geom_bar(stat = 'identity', 
+             color = 'black', 
+             fill = 'steelblue') + 
+    geom_vline(xintercept = 0.95, 
+               linetype = 'dashed') + 
+    globals$common_theme + 
+    theme(axis.title.y = element_blank()) + 
+    labs(title = 'Normality testing with Shapiro-Wilk test', 
+         subtitle = 'entire baseline dataset', 
+         x = 'Test statistic, W')
+  
+# Plotting the means and variances -----
+  
+  insert_msg('Plotting the means and variances')
+  
+  distr_base$mean_var_plot <- distr_base$test %>% 
+    mutate(var_lab = exchange(variable, dict = distr_base$lexicon)) %>% 
+    blast(transformation) %>% 
+    map2(., 
+         paste('Mean and variance,', names(.)), 
+         ~ggplot(.x, 
[…]

### 2023-06-14 exploration scripts/distribution_longitudinal.R

@@ -1,0 +1,118 @@
+# Distribution tests: checking normality (splitting by severity and timepoint) 
+# of the longitudinally recorded 
+# CT and LFT variables
+
+  insert_head()
+  
+# container ------
+  
+  distr_long <- list()
+  
+# Parallel backend -----
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis globals ----
+  
+  insert_msg('Analysis globals')
+  
+  ## variable lexicon, selecting only the numeric ones
+  
+  distr_long$lexicon <- 
+    rbind(covild$ct_lexicon, 
+          covild$lft_lexicon) %>% 
+    filter(!variable %in% c('ID', 'follow_up', 'severity_class')) %>% 
+    filter(format == 'numeric') %>% 
+    filter(!stri_detect(variable, regex = '(DLCO|volume)$'))
+  
+  ## analysis table
+  
+  distr_long$analysis_tbl <- 
+    list(covild$ct, 
+         covild$lft %>% 
+           select(-follow_up), 
+         covild$baseline[c('ID', 'severity_class')]) %>% 
+    reduce(left_join, by = 'ID') %>% 
+    filter(follow_up != 'acute COVID-19') %>% 
+    mutate(follow_up = droplevels(follow_up), 
+           analysis_split = interaction(follow_up, severity_class)) %>% 
+    select(ID, follow_up, severity_class, analysis_split, 
+           all_of(distr_long$lexicon$variable))
+  
+# Shapiro test ------
+  
+  insert_msg('Shapiro test')
+  
+  distr_long$test <- distr_long$lexicon$variable %>% 
+    future_map(test_long_distribution, 
+               data = distr_long$analysis_tbl, 
+               split_factor = 'analysis_split', 
+               .options = furrr_options(seed = TRUE)) %>% 
+    set_names(distr_long$lexicon$variable) %>% 
+    compress(names_to = 'variable')
+  
+# Plotting the global Shapiro stats -------
+  
+  insert_msg('Plotting the global Shapiro test stats')
+  
+  distr_long$shapiro_plot <- distr_long$test %>% 
+    filter(analysis_split == 'global', 
+           best_transf == 'yes') %>% 
+    mutate(var_lab = exchange(variable, dict = distr_long$lexicon), 
+           var_lab  = paste(var_lab, transformation, sep = ', ')) %>% 
+    ggplot(aes(x = statistic, 
[…]

### 2023-06-14 exploration scripts/lft.R

@@ -1,0 +1,150 @@
+# Descriptive stats for the LFT parameters split be the timepoint 
+# and COVID-19 severity. No statistical testing is done - a 'proper' 
+# statistical analysis is done later on with participant-matched data.
+#
+# This analytic task is here to get an idea, how the variables change
+# in time and severity
+
+  insert_head()
+
+# container ------
+
+  cohort_lft <- list()
+
+# Analysis globals ------
+
+  insert_msg('Analysis globals')
+  
+  cohort_lft$lexicon <- covild$lft_lexicon %>% 
+    filter(!stri_detect(variable, regex = '(volume|DLCO)$'))
+
+  ## numeric and categorical variables
+
+  cohort_lft$variables <- cohort_lft$lexicon %>% 
+    blast(format) %>% 
+    map(~.x$variable)
+
+  ## analysis table, introducing the severity variable
+  ## removing the acute timepoint - there was no study CT but just 
+  ## a diagnostic scan for some participants
+
+  cohort_lft$analysis_tbl <- covild$lft %>% 
+    left_join(covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    mutate(cohort_split = severity_class)
+  
+  cohort_lft$analysis_tbl <- cohort_lft$analysis_tbl %>% 
+    full_rbind(covild$lft) %>% 
+    filter(follow_up != 'acute COVID-19') %>% 
+    mutate(cohort_split = ifelse(is.na(cohort_split), 
+                                 'cohort', as.character(cohort_split)), 
+           cohort_split = factor(cohort_split, 
+                                 c('cohort', 
+                                   levels(cohort_lft$analysis_tbl$severity_class))), 
+           follow_up = droplevels(follow_up), 
+           analysis_split = interaction(follow_up, cohort_split)) %>% 
+    relocate(cohort_split) %>% 
+    relocate(analysis_split) %>% 
+    relocate(ID)
+
+# Number of complete cases per follow-up ------
+
+  insert_msg('Number of complete cases per follow-up')
+  
+  cohort_lft$n_numbers <- cohort_lft$analysis_tbl %>% 
+    filter(cohort_split != 'cohort') %>% 
+    blast(severity_class, follow_up) %>%
+    map(~filter(.x, complete.cases(.x))) %>% 
+    map_dbl(nrow) %>% 
+    compress(names_to = 'strata', 
+             value = 'n') %>% 
[…]

### 2023-06-14 exploration.R

@@ -1,0 +1,37 @@
+# Exploratory data analysis: characteristic of the study variables
+# and the study cohort
+
+# tools -----
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+  
+  library(exda)
+  library(rstatix)
+  library(rcompanion)
+
+  library(soucer)
+  library(furrr)
+
+  library(ggrepel)
+  
+  c('./tools/globals.R', 
+    './tools/tools.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+  
+# analysis scripts -----
+
+  insert_msg('Analysis scripts')
+  
+  c('./exploration scripts/baseline_cohort.R', 
+    './exploration scripts/ct.R',
+    './exploration scripts/lft.R', 
+    './exploration scripts/distribution_longitudinal.R', 
+    './exploration scripts/distribution_baseline.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# END ------
+  
+  insert_tail()

### 2023-06-14 import.R

@@ -1,0 +1,176 @@
+# Import and formatting of the Radio CovILD study data
+#
+# I'm fetching them from the Warehouse Package CovILD
+
+# tools ------
+
+  library(tidyverse)
+  library(trafo)
+  library(stringi)
+  
+  library(covILD)
+  
+  library(soucer)
+
+  insert_head()
+
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+# container ------
+  
+  insert_msg('Data container')
+  
+  covild <- list()
+  
+# Data import -------
+  
+  insert_msg('Data import')
+  
+  covild[c('baseline', 
+           'baseline_lexicon', 
+           'ct', 
+           'ct_lexicon', 
+           'lft', 
+           'lft_lexicon')] <- list(covILD::baseline, 
+                                   covILD::baseline_lexicon, 
+                                   covILD::ct, 
+                                   covILD::ct_lexicon, 
+                                   covILD::lft, 
+                                   covILD::lft_lexicon)
+  
+# Restricting to the patients present in all datasets ------
+  
+  insert_msg('Resticting to the patients present in all datasets')
+  
+  covild$common_ids <- covild[c("baseline", "ct", "lft")] %>% 
+    map(~.x$ID) %>% 
+    map(unique) %>% 
+    reduce(intersect)
+  
+  covild[c("baseline", "ct", "lft")] <- 
+    covild[c("baseline", "ct", "lft")] %>% 
+    map(filter, ID %in% covild$common_ids)
+  
+# Baseline lexicon -----
+  
+  insert_msg('Baseline lexicon')
+  
+  base_lex <- 
+    c('sex' = 'sex', 
+      'age' = 'age at admission', 
+      'age_class' = 'age class', 
+      'smoking' = 'smoking history', 
+      'smoking_pack_years' = 'smoking', 
+      'BMI' = 'BMI', 
+      'body_mass_class' = 'body mass class', 
+      'cardiovascular_comorbidity' = 'cardiovascular illness', 
+      'hypertension_comorbidity' = 'hypertension', 
+      'pulmonary_comorbidity' = 'pulmonary illness', 
+      'COPD_comorbidity' = 'COPD', 
[…]

### 2023-06-14 kinetic scripts/factor.R

@@ -1,0 +1,161 @@
+# Analysis of kinetic of categorical CT and LFT variables
+# statistical significance determined by Cochran Q test 
+# (due to non-normality of e.g. AI opacity). 
+# Effect size estimated by Kendall's W
+#
+# The analysis is done for individuals with the complete datasets
+# W = chisq/(df * n)
+
+  insert_head()
+
+# container -----
+
+  kin_factor <- list()
+
+# parallel backend -------
+
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+
+# Analysis globals -------
+
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  kin_factor$variables <- kin_globals$variables$factor
+  
+  kin_factor$variable_class <- 
+    ifelse(kin_factor$variables %in% covild$ct_lexicon$variable, 
+           'CT', 'LFT') %>% 
+    tibble(variable = kin_factor$variables, 
+           class = .)
+  
+  kin_factor$variables <- 
+    set_names(kin_factor$variables, 
+              kin_factor$variables)
+  
+  ## analysis table
+  
+  kin_factor$analysis_tbl <- kin_globals$analysis_tbl %>% 
+    map(select, ID, follow_up, all_of(kin_factor$variables))
+  
+  ## formulas
+  
+  kin_factor$formulas <- kin_factor$variables %>% 
+    map(~paste(.x, '~ follow_up | ID')) %>% 
+    map(formula)
+  
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  kin_factor$stats <- kin_factor$analysis_tbl %>% 
+    future_map(function(data) kin_factor$variables %>% 
+                 map(~explore(data[c('ID', 'follow_up', .x)] %>% 
+                                complete_cases('ID'), 
+                              variables = .x, 
+                              split_factor = 'follow_up', 
+                              what = 'table', 
+                              pub_styled = TRUE)), 
+               .options = furrr_options(seed = TRUE))
+  
+  kin_factor$stats <- kin_factor$stats %>% 
+    map(map_dfr, reduce, left_join, by = 'variable') %>% 
+    compress(names_to = 'severity_class') %>% 
+    relocate(severity_class) %>% 
+    set_names(c('severity_class', 'variable', 
+                levels(kin_factor$analysis_tbl[[1]]$follow_up)))
+  
+# Testing ------
+  
+  insert_msg('Testing')
+  
+  kin_factor$test <- kin_factor$analysis_tbl %>% 
[…]

### 2023-06-14 kinetic scripts/numeric.R

@@ -1,0 +1,150 @@
+# Analysis of kinetic of numeric CT and LFT variables
+# statistical significance determined by Friedman test 
+# (due to non-normality of e.g. AI opacity). 
+# Effect size estimated by Kendall's W
+#
+# The analysis is done for individuals with the complete datasets
+
+  insert_head()
+  
+# container -----
+  
+  kin_numeric <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# Analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables
+  
+  kin_numeric$variables <- kin_globals$variables$numeric
+  
+  kin_numeric$variable_class <- 
+    ifelse(kin_numeric$variables %in% covild$ct_lexicon$variable, 
+           'CT', 'LFT') %>% 
+    tibble(variable = kin_numeric$variables, 
+           class = .)
+  
+  kin_numeric$variables <- 
+    set_names(kin_numeric$variables, 
+              kin_numeric$variables)
+
+  ## analysis table
+  
+  kin_numeric$analysis_tbl <- kin_globals$analysis_tbl %>% 
+    map(select, ID, follow_up, all_of(kin_numeric$variables))
+
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  kin_numeric$stats <- kin_numeric$analysis_tbl %>% 
+    future_map(function(data) kin_numeric$variables %>% 
+                 map(~explore(data[c('ID', 'follow_up', .x)] %>% 
+                                complete_cases('ID'), 
+                              variables = .x, 
+                              split_factor = 'follow_up', 
+                              what = 'table', 
+                              pub_styled = TRUE)), 
+               .options = furrr_options(seed = TRUE))
+  
+  kin_numeric$stats <- kin_numeric$stats %>% 
+    map(map_dfr, reduce, left_join, by = 'variable') %>% 
+    compress(names_to = 'severity_class') %>% 
+    relocate(severity_class) %>% 
+    set_names(c('severity_class', 'variable', 
+                levels(kin_numeric$analysis_tbl[[1]]$follow_up)))
+  
+# Testing ---------
+  
+  insert_msg('Testing')
+  
+  kin_numeric$test <- kin_numeric$analysis_tbl %>% 
+    future_map(function(data) kin_numeric$variables %>% 
+                 map_dfr(~compare_variables(data[c('ID', 'follow_up', .x)] %>% 
[…]

### 2023-06-14 kinetic.R

@@ -1,0 +1,84 @@
+# Analysis of kinetic of CT and LFT parameters, done in participants with 
+# complete longitudinal observation sets.
+#
+# The analysis is done for the entire cohort and the severity subsets
+# Statistical tools are Cochran Q test (binary responses) and Friedman test
+
+# tools ------
+
+  library(tidyverse)
+  library(rlang)
+  library(trafo)
+  library(stringi)
+  
+  library(exda)
+  library(rstatix)
+  library(rcompanion)
+  library(effectsize)
+
+  library(soucer)
+  library(furrr)
+  
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+# analysis globals -------
+  
+  insert_msg('Analysis globals')
+  
+  kin_globals <- list()
+  
+  ## study variables
+  ## 'crazy_paving_finding', and volume LFT parameters are excluded
+  
+  kin_globals$lexicon <- 
+    covild[c("ct_lexicon", "lft_lexicon")] %>% 
+    reduce(rbind) %>%
+    filter(!variable %in% c('crazy_paving_finding', 'DLCO')) %>%
+    filter(!stri_detect(variable, regex = '_volume$'))
+  
+  kin_globals$variables <- kin_globals$lexicon %>% 
+    blast(format) %>% 
+    map(~.x$variable)
+  
+  ## analysis tables for the entire cohort and severity groups
+
+  kin_globals$analysis_tbl <- covild[c("ct", "lft")] %>% 
+    reduce(left_join, by = c('ID', 'follow_up')) %>% 
+    left_join(covild$baseline[c('ID', 'severity_class')], 
+              by = 'ID') %>% 
+    select(ID, severity_class, follow_up, 
+           all_of(kin_globals$lexicon$variable))
+  
+  kin_globals$analysis_tbl <- 
+    kin_globals$analysis_tbl %>% 
+    mutate(severity_class = 'cohort') %>%
+    rbind(kin_globals$analysis_tbl) %>% 
+    mutate(severity_class = factor(severity_class, 
+                                   c('cohort', 
+                                     levels(kin_globals$analysis_tbl$severity_class)))) %>% 
+    filter(follow_up != 'acute COVID-19') %>% 
+    filter(!is.na(severity_class)) %>% 
+    mutate(follow_up = droplevels(follow_up))
+  
[…]

### 2023-06-14 lft.R

@@ -1,0 +1,287 @@
+# Modeling of the key LFT outcomes by CT findings and parameters 
+# (manually identified findings, CTSS, AI opacity and AI high opacity).
+# The responses are 
+# 1) presence of any LFT findings defined as parameters <80% or <70%
+# 2) presence of impaired DLCO (< 80% reference)
+# 3) DLCO expressed as percentage of the reference
+#
+# In addition , a 'univariate' analysis is done, i.e. comparison and 
+# correlation of the CT variables with the outcomes
+
+# tools -----
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+
+  library(exda)
+  library(rstatix)
+  library(caret)
+  library(caretExtra)
+  library(effectsize)
+  library(ranger)
+
+  library(soucer)
+  library(furrr)
+  library(doParallel)
+
+  library(OptimalCutpoints)
+  library(plotROC)
+
+  library(ggrepel)
+
+  explore <- exda::explore
+
+  c('./tools/tools.R', 
+    './tools/globals.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# globals container -------
+  
+  lft_globals <- list()
+  
+# analysis variables ------
+  
+  insert_msg('Analysis variables')
+  
+  lft_globals$responses <- 
+    c('LFT_findings', 
+      'DLCO_reduced', 'DLCO_percent', 
+      'FVC_reduced', 'FVC_percent', 
+      'FEV1_reduced', 'FEV1_percent')
+  
+  lft_globals$ct_variables <- 
+    c('follow_up', 'CT_findings', 'GGO_finding', 'reticulation_finding', 
+      'consolidation_finding', 'bronchiectasis_finding', 
+      'opacity_percent', 'high_opacity_percent', 'CTSS')
+  
+  ## clinical baseline variables and severity of acute COVID-19
+  ## 'pulmonary_comorbidity' is defined as presence of any pulmonary, COPD, 
+  # asthma, ILD, or chronic lung disease prior to COVID-19
+  ## 'anticoagulant_treatment' subsumes anti-coagulant and ant-platelet therapy
+  ## 'antiinfective_treatment' subsumes macrolide and othe antibiotic treatment
+  
+  lft_globals$baseline_variables <- 
+    c('sex', 'age', 'age_class', 
+      'smoking', 'smoking_pack_years', 
+      'BMI', 'body_mass_class', 
[…]

### 2023-06-14 report scripts/links.R

@@ -1,0 +1,28 @@
+# Handling of common links in the Rmarkdown project
+
+  insert_head()
+  
+# container ------
+  
+  proj_links <- list()
+  
+# Links to development packages --------
+  
+  insert_msg('Links to development packages')
+  
+  proj_links <- 
+    list('ExDA' = 'https://github.com/PiotrTymoszuk/ExDA', 
+         'trafo' = 'https://github.com/PiotrTymoszuk/trafo', 
+         'figur' = 'https://github.com/PiotrTymoszuk/figur', 
+         'caretExtra' = 'https://github.com/PiotrTymoszuk/caretExtra') %>% 
+    compress(names_to = 'obj_name', 
+             values_to = 'x') %>% 
+    mutate(ref_name = paste0('_', obj_name, '_'))
+  
+  proj_links <- proj_links[c('ref_name', 'x')] %>% 
+    pmap(mdlink) %>% 
+    set_names(proj_links$obj_name)
+  
+# END ------
+  
+  insert_tail()

### 2023-06-14 report scripts/render.R

@@ -1,0 +1,29 @@
+# Renders the report
+
+  insert_head()
+  
+# Bibliography -----
+  
+  insert_msg('Reading the bibliography')
+  
+  tesca_bib <- read_bib('./report/markdown/ct.bib') %>% 
+    as_mdbib
+  
+# Rendering the report ------
+  
+  insert_msg('Rendering the report')
+  
+  render('./report/markdown/report.Rmd', 
+         output_format = robobook(template = 'default', 
+                                  use_bookdown = TRUE, 
+                                  number_sections = FALSE, 
+                                  toc_depth = 3, 
+                                  lightbox = TRUE, 
+                                  thumbnails = TRUE, 
+                                  gallery = TRUE, 
+                                  css = 'styles.css'), 
+         output_dir = './report')
+  
+# END ----
+  
+  insert_tail()

### 2023-06-14 report scripts/tables.R

@@ -1,0 +1,388 @@
+# Report tables 
+
+  insert_head()
+  
+# container -----
+  
+  tables <- list()
+  
+# Table 1: characteristic of the cohort ------
+  
+  insert_msg('Table 1: characteristic of the cohort')
+  
+  tables$cohort <- cohort$result_tbl %>% 
+    filter(variable != 'COVID-19 severity') %>% 
+    set_names(c('Variable', 
+                'Cohort', 
+                'Ambulatory COVID-19', 
+                'Moderate COVID-19', 
+                'Severe COVID-19', 
+                'Significance', 
+                'Effect size')) %>% 
+    mdtable(label = 'table_1_cohort_characteristic', 
+            ref_name = 'cohort', 
+            caption = paste('Baseline characteristics and COVID-19 course', 
+                            'in the CovILD cohort.', 
+                            'Numeric variables are presented as medians with', 
+                            'interquartile ranges (IQR) and ranges.', 
+                            'Categorical variables are', 
+                            'presented as percentages', 
+                            'and counts within the complete observation set.'))
+  
+# Table 2: CT variables at follow-ups ------
+  
+  insert_msg('Table 2: CT variables at follow-ups')
+  
+  tables$cohort_ct <- cohort_ct$result_tbl %>% 
+    compress(names_to = 'Severity subset') %>% 
+    relocate(`Severity subset`) %>% 
+    mutate(`Severity subset` = globals$sev_labels[`Severity subset`])
+  
+  tables$cohort_ct <- tables$cohort_ct %>% 
+    set_colnames(capitalize_first_char(names(tables$cohort_ct))) %>% 
+    mdtable(label = 'table_2_cohort_ct_variables',
+            ref_name = 'cohort_ct', 
+            caption = paste('Chest computed tomography variables', 
+                            'at consecutive follow-ups.', 
+                            'Numeric variables are presented as medians with', 
+                            'interquartile ranges (IQR) and ranges.', 
+                            'Categorical variables are', 
+                            'presented as percentages', 
+                            'and counts within the complete observation set.'))
+  
+# Table 3: LFT variables at follow-ups ------
+  
+  insert_msg('Table 3: LFT variables at follow ups')
+  
+  tables$cohort_lft <- cohort_lft$result_tbl %>% 
+    compress(names_to = 'Severity subset') %>% 
+    relocate(`Severity subset`) %>% 
[…]

### 2023-06-14 report.R

@@ -1,0 +1,42 @@
+# Analysis report
+
+# tools -----
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+
+  library(flextable)
+  library(figur)
+  library(cowplot)
+  library(writexl)
+
+  library(rmarkdown)
+  library(bookdown)
+  library(knitr)
+  library(rmdformats)
+
+  library(soucer)
+
+  insert_head()
+
+  width <- flextable::width
+
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# report scripts -------
+  
+  insert_msg('Report scripts')
+  
+  c('./report scripts/links.R', 
+    './report scripts/html.R', 
+    './report scripts/tables.R', 
+    './report scripts/render.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# END ------
+  
+  insert_tail()

### 2023-06-14 tools/globals.R

@@ -1,0 +1,66 @@
+# This script contains project globals
+
+# data container ------
+
+  globals <- list()
+
+# graphics -----
+  
+  globals$common_text <- element_text(size = 8,
+                                      face = 'plain',
+                                      color = 'black')
+  
+  globals$common_margin <- ggplot2::margin(t = 5,
+                                           l = 4,
+                                           r = 2,
+                                           unit = 'mm')
+  
+  globals$common_theme <- theme_classic() + 
+    theme(axis.text = globals$common_text,
+          axis.title = globals$common_text,
+          plot.title = element_text(size = 8,
+                                    face = 'bold'),
+          plot.subtitle = globals$common_text,
+          plot.tag = element_text(size = 8,
+                                  face = 'plain',
+                                  color = 'black',
+                                  hjust = 0,
+                                  vjust = 1),
+          plot.tag.position = 'bottom',
+          legend.text = globals$common_text,
+          legend.title = globals$common_text,
+          strip.text = globals$common_text,
+          strip.background = element_rect(fill = 'gray95',
+                                          color = 'gray80'),
+          plot.margin = globals$common_margin,
+          panel.grid.major = element_line(color = 'gray90'))
+  
+# strata colors and labels ------
+
+  globals$sev_colors <- c('ambulatory mild' = 'steelblue', 
+                          'hospitalized moderate' = 'darkolivegreen3', 
+                          'hospitalized severe' = 'coral3', 
+                          'cohort' = 'plum4')
+  
+  globals$sev_labels <- c('ambulatory mild' = 'ambulatory', 
+                          'hospitalized moderate' = 'moderate', 
+                          'hospitalized severe' = 'severe', 
+                          'cohort' = 'CovILD cohort')
+  
+# follow-up colors and labels -----
+  
+  globals$fup_colors = c('acute COVID-19' = 'black', 
+                         '2-month follow-up' = 'coral4', 
+                         '3-month follow-up' = 'coral3', 
+                         '6-month follow-up' = 'steelblue2', 
+                         '12-month follow-up' = 'steelblue4', 
+                         'cohort' = 'plum4')
+  
+  globals$fup_labels = c('acute COVID-19' = 'acute CoV', 
+                         '2-month follow-up' = '2 mo', 
+                         '3-month follow-up' = '3 mo', 
+                         '6-month follow-up' = '6 mo', 
+                         '12-month follow-up' = '12 mo', 
+                         'cohort' = 'CovILD cohort')
+
+# END -----

### 2023-06-14 tools/tools.R

@@ -1,0 +1,1498 @@
+# Functions for the project
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+  library(OptimalCutpoints)
+  library(plotROC)
+  library(caret)
+  library(furrr)
+  library(coxed)
+  library(soucer)
+  
+# Lexicon handling ------
+  
+  make_lexicon <- function(label_vector, 
+                           unit_vector, 
+                           lexicon, 
+                           data) {
+    
+    ## merges user-provided labels and unit for variables
+    ## with an existing lexicon
+    
+    new_lex <- list(label_vector, 
+                    unit_vector) %>% 
+      map2(., c('label', 'unit'), 
+           ~compress(.x, 
+                     names_to = 'variable', 
+                     values_to = .y)) %>% 
+      reduce(left_join, by = 'variable') %>% 
+      inner_join(lexicon %>% 
+                   transmute(variable = Variable, 
+                             description = Description), 
+                 by = 'variable')
+    
+    ## information about the variable format
+    
+    new_lex$format <- 
+      data[new_lex$variable] %>% 
+      map_lgl(is.numeric) %>% 
+      ifelse('numeric', 'factor')
+    
+    ## table label
+    
+    new_lex %>% 
+      mutate(table_label = ifelse(is.na(unit), 
+                                  label, paste(label, unit, sep = ', ')))
+    
+  }
+  
+# Text functions -----
+  
+  capitalize_first_char <- function(x) {
+    
+    ## capitalizes the first character of a string
+    
+    stri_trans_totitle(stri_sub(x, 1, 1)) %s+% stri_sub(x, 2)
+    
+  }
+  
+# Formatting of result tables, multiple-testing adjustment ---------
+  
+  format_summ_tbl <- function(data, 
+                              rm_n = FALSE, 
+                              rm_mean = TRUE, 
+                              dict = NULL, 
+                              key = 'variable', 
+                              value = 'table_label') {
+    
+    ## formats a summary table with descriptive stats
+    
+    data <- data %>% 
+      map_dfc(stri_replace, regex = 'no:.*\\nyes:\\s{1}', replacement = '') %>% 
+      map_dfc(stri_replace, regex = '\\nno:.*$', replacement = '') %>% 
+      map_dfc(stri_replace_all, fixed = '% (', replacement = '% (n = ') %>% 
[…]

### 2023-09-12 report/markdown/report.Rmd
Figures, tables and report RMD

@@
                       message = FALSE, 
-                      dev = "CairoPDF", 
+                      dev = "CairoPNG", 
                       dpi = 600)
@@
 flextable::flextable(tables$cohort) %>% 
-  width(1, width = 3, unit = 'cm') %>% 
-  width(2:5, width = 3.0, unit = 'cm') %>% 
-  width(6, width = 2.7, unit = 'cm') %>% 
-  width(7, width = 2.5, unit = 'cm') %>% 
+  set_widths(c(3, rep(3.5, 4), 2.7, 2.5)) %>% 
   footnote(1, 1, 
@@
 
-## Chest computed tomography paramaters and findings
+## Chest computed tomography abnormalities
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+
+figures$ct_finding_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<br>
+Frequencies of any chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulations and consolidations at the consecutive follow-ups after COVID-19 in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stacked bar plots.
+Numbers of complete observations at the follow-up examinations are displayed above the data bars.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+
+figures$ct_numeric_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<br>
+Severity of chest computed tomography (CT) abnormalities was assessed by a human-determined CT severity score (CTSS), and AI-determined percentages of the lung tissue with opacity and high opacity.
+Time courses of these parameters are visualized in box plots.
+Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range.
+Single observations are depicted as points.
+Numbers of complete observations at the follow-up examinations are displayed above the data points.
+</p>
 
@@
 flextable::flextable(tables$cohort_ct) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 2.7, unit = 'cm') %>% 
-  width(3:6, width = 3.8, unit = 'cm') %>% 
+  set_widths(c(2.7, 2.7, rep(3.8, 4))) %>% 
   merge_v(1) %>% 
@@
 
+```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+
+figures$lft_finding_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<br>
+Frequency of any lung function testing (LFT) abnormalities, as well as percentage of patients with DLCO, FVC and FEV1 values below 80% of the reference values at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots.
+Numbers of complete observations at the follow-up examinations are displayed above the data bars.
+<br>
+LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second. 
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+
+figures$lft_numeric_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-lft-numeric-course).
+Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<br>
+Lung function was assessed by measurement of DLCO, FVC and FEV1 expressed as percentage of the age-, weight- and sex-specific reference.
+Time courses of these parameters are visualized in box plots.
+Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range.
+Single observations are depicted as points.
+Numbers of complete observations at the follow-up examinations are displayed above the data points.
+<br>
+LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
 ```{r tab-cohort-lft, tab.cap = 'Lung function testing variables at consecutive follow-ups.
@@
 flextable::flextable(tables$cohort_lft) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 2.7, unit = 'cm') %>% 
-  width(3:6, width = 3.8, unit = 'cm') %>% 
+  set_widths(c(2.7, 2.7, rep(3.8, 4))) %>% 
   merge_v(1) %>% 
   footnote(1, 2, 
-           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second; DLCO: diffusion capacity for CO; FEV1:FVC: FEV1 to FVC ratio'), 
+           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; DLCO: diffusion capacity for carbon monoxide; FEV1:FVC: FEV1 to FVC ratio'),
            part = 'header', 
@@
 
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ai-opacity-roc, fig.width = figur::convert(figures$ai_opacity_roc, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_roc, to = 'in')$h, fig.cap = 'Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung opacity.'}
+
+figures$ai_opacity_roc$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-opacity-roc). Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung opacity.</b>
+<br>
+The optimal cutoffs of the artificial intelligence (AI) determined lung opacity for detection of the human-recognizable chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulation, and consolidations were found by maximizing the Youden's J statistic (i.e.
+sensitivity + specificity - 1).
+Results of the optimal opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves.
+The optimal opacity cutoffs are labeled with points and the corresponding opacity values.
+Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ai-high-opacity-roc, fig.width = figur::convert(figures$ai_high_opacity_roc, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_roc, to = 'in')$h, fig.cap = 'Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung high opacity.'}
+
+figures$ai_high_opacity_roc$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-high-opacity-roc). Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung high opacity.</b>
+<br>
+The optimal cutoffs of the artificial intelligence (AI) determined lung high opacity for detection of the human-recognizable chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulation, and consolidations were found by maximizing the Youden's J statistic (i.e.
+sensitivity + specificity - 1).
+Results of the optimal high opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves.
+The optimal high opacity cutoffs are labeled with points and the corresponding opacity values.
+Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
 ```{r tab-ai-cutoffs, tab.cap = 'Optimal cutoffs for AI-determined lung opacity and high opacity at detection of chest computed tomography abnormalities.
@@
 flextable::flextable(tables$ai_cutoffs) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 3, unit = 'cm') %>% 
-  width(3, width = 4.8, unit = 'cm') %>% 
-  width(4:8, width = 2.5, unit = 'cm') %>% 
+  set_widths(c(2.7, 3, 4.8, rep(2.5, 5))) %>% 
   merge_v(1) %>% 
@@
   footnote(1, 7, 
-           value = as_paragraph('J: Youden J statistic'), 
+           value = as_paragraph("J: Youden's J statistic"), 
            part = 'header', 
@@
 
+```{r fig-ai-opacity-severity, fig.width = figur::convert(figures$ai_opacity_severity, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_severity, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.'}
+
+figures$ai_opacity_severity$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-opacity-severity).
+Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.</b>
+<br>
+The optimal cutoffs of artificial intelligence (AI) determined lung opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-opacity-roc).
+Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
+$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
+<br>
+CT: computed tomography; GGO: ground glass opacity.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ai-opacity-fup, fig.width = figur::convert(figures$ai_opacity_fup, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_fup, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.'}
+
+figures$ai_opacity_fup$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-opacity-fup).
+Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.</b>
+<br>
+The optimal cutoffs of artificial intelligence (AI) determined lung opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-opacity-roc).
+Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
+$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
+<br>
+CT: computed tomography; GGO: ground glass opacity.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ai-high-opacity-severity, fig.width = figur::convert(figures$ai_high_opacity_severity, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_severity, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.'}
+
+figures$ai_high_opacity_severity$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-high-opacity-severity).
+Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.</b>
+<br>
+The optimal cutoffs of artificial intelligence (AI) determined lung high opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-high-opacity-roc).
+Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
+$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
+<br>
+CT: computed tomography; GGO: ground glass opacity.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ai-high-opacity-fup, fig.width = figur::convert(figures$ai_high_opacity_fup, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_fup, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.'}
+
+figures$ai_high_opacity_fup$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ai-high-opacity-fup).
+Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.</b>
+<br>
+The optimal cutoffs of artificial intelligence (AI) determined lung high opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-high-opacity-roc).
+Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
+$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
+<br>
+CT: computed tomography; GGO: ground glass opacity.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
 ```{r tab-ai-cutoffs-severity, tab.cap = 'Performance of AI-determined lung opacity and high opacity at detection of chest computed tomography abnormalities in the COVID-19 severity subsets.
@@
 flextable::flextable(tables$ai_cutoffs_severity) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 3, unit = 'cm') %>% 
-  width(3, width = 3, unit = 'cm') %>% 
-  width(4:7, width = 5.2, unit = 'cm') %>% 
+  set_widths(c(2.7, 3, 3, rep(5.2, 4))) %>% 
   merge_v(1:2) %>% 
@@
 flextable::flextable(tables$ai_cutoffs_fup) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 3, unit = 'cm') %>% 
-  width(3, width = 3, unit = 'cm') %>% 
-  width(4:7, width = 5.2, unit = 'cm') %>% 
+  set_widths(c(2.7, 3, 3, rep(5.2, 4))) %>% 
   merge_v(1:2) %>% 
@@
 
+```{r fig-ctss-ild-cutoff, fig.width = figur::convert(figures$ctss_ild_cutoff, to = 'in')$w, fig.height = figur::convert(figures$ctss_ild_cutoff, to = 'in')$h, fig.cap = 'Detection of interstitial lung disease defined by opacity >5% of the lung by a human-determined computed tomography severity score.'}
+
+figures$ctss_ild_cutoff$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ctss-ild-cutoff). Detection of interstitial lung disease defined by opacity >5% of the lung by a human-determined computed tomography severity score.</b>
+<br>
+(A) The optimal cutoff of the radiologist's computed tomography severity score (CTSS) for diagnosis of interstitial lung disease (ILD) defined as opacity > 5% of the lung was found by maximizing the Youden's J statistic (i.e sensitivity +  specificity - 1).
+Results of the optimal CTSS cutoff finding are presented as a receiver-operating characteristic (ROC) curve.
+The optimal CTSS cutoff is labeled with a point and the corresponding CTSS value.
+Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plot.
+<br>
+(B) Concordance between the ILD diagnosis with CTSS > 10 points and ILD defined as opacity > 5% of the lung in the whole observation set ('cohort') and in observations stratified by the acute disease severity or post-COVID-19 follow-up was assessed by Cohen's $\kappa$ inter-rater reliability statistic.
+$\kappa$ values with 95% confidence intervals obtained by bootstrap with B = 2000 iterations are presented in Forest plots.
+Numbers of complete observations ('total') and cases of ILD defined by >5% lung opacity ('events') are indicated in the Y axes.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
 ```{r tab-ild-cutoff, tab.cap = 'Detecion of interstitial lung disease defined by lung opacity > 5% by chest computed tomography severity score (CTSS).
@@
 flextable::flextable(tables$ild_cutoff) %>% 
-  width(1, width = 3, unit = 'cm') %>% 
-  width(2, width = 4.8, unit = 'cm') %>% 
-  width(3:8, width = 2.5, unit = 'cm') %>% 
+  set_widths(c(3, 4.8, rep(2.6, 6))) %>% 
   footnote(1, 1, 
@@
 flextable::flextable(tables$ild_cutoff_severity) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2:5, width = 5, unit = 'cm') %>% 
+  set_widths(c(2.7, rep(5, 4))) %>% 
   footnote(1, 2, 
@@
 flextable::flextable(tables$ild_cutoff_fup) %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2:5, width = 5, unit = 'cm') %>% 
+  set_widths(c(2.7, rep(5, 4))) %>% 
   footnote(1, 2, 
@@
 
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ctss-ai-relation, fig.width = figur::convert(figures$ctss_ai_relation, to = 'in')$w, fig.height = figur::convert(figures$ctss_ai_relation, to = 'in')$h, fig.cap = "Mathematical relationship between the artificial intelligence-determined lung opacity and high opacity, and the radiologist's computed tomography severity score."}
+
+figures$ctss_ai_relation$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ctss-ai-relation).
+Mathematical relationship between the artificial intelligence-determined lung opacity and high opacity, and the radiologist's computed tomography severity score.</b>
+<br>
+Mathematical relationships (A) between the artificial intelligence's lung opacity and radiologist's CTSS (computed tomography severity score), and (B) between the artificial intelligence's high lung opacity and radiologist's CTSS were explored by ordinary least squares 4^th^ order polynomial regression, robust 4^th^ order polynomial regression or generalized additive modeling (GAM) with thin plate splines.
+The model fit quality was assessed by the pseudo-R^2^ metric of explained variance and median absolute error (MedAE).
+The relationships are presented in scatter plots with single observations visualized as points.
+The model fits with their 2 $\times$ standard error regions are represented by blue lines with gray ribbons.
+The R^2^ and MedAE values along with numbers of complete observations are displayed in the plot captions.
+</p>
+
 <hr style = "border-top: 5px solid #417a8b" />
@@
 
-## Development of multi-paramater Random Forest models of lung function variables and abnormalities
+## Development of multi-paramater machine learning models of lung function variables and abnormalities
 
@@
 
-```{r tab-rf-tuning, tab.cap = 'Optimal values of the mtry, splitrule and min.node.size paramaters for Random Forest models of lung function abnormalities and paramaters found by cross-validation tuning.'}
+```{r fig-dlco-red-tuning, fig.width = figur::convert(figures$dlco_red_tuning, to = 'in')$w, fig.height = figur::convert(figures$dlco_red_tuning, to = 'in')$h, fig.cap = 'Tuning of machine learning models of insufficient diffusion capacity for carbon monoxide.'}
 
-flextable::flextable(tables$rf_tuning) %>% 
-  width(1, width = 2.5, unit = 'cm') %>% 
-  width(2, width = 2.7, unit = 'cm') %>% 
-  width(3, width = 3, unit = 'cm') %>% 
-  width(4:6, width = 2.7, unit = 'cm') %>% 
-  merge_v(1:2) %>% 
-  footnote(1, 2, 
-           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
+figures$dlco_red_tuning$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-red-tuning). Tuning of machine learning models of insufficient diffusion capacity for carbon monoxide.</b>
+<br>
+Insufficient diffusion capacity for carbon monoxide (DLCO) was defined as values < 80% of the age-, sex-, and weight-specific reference. Probability of insufficient DLCO was modeled as a function of 
+`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The optimal sets of algorithm's parameters such as number of trees for GBM, number of randomly selected variables for random forests, size of the hidden layer for neuronal networks or cost penalty value for SVM were found by maximizing the sum of sensitivity and specificity, i.e.
+Youden's J criterion, in 10-repeats 10-fold cross-validation - in a process called 'model tuning'.
+Values of the J statistic for various combination of the algorithm's parameters are presented in plots.
+The optimal parameter combinations corresponding to the maximal J value are displayed in the plot captions.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-red-resamples, fig.width = figur::convert(figures$dlco_red_resamples, to = 'in')$w, fig.height = figur::convert(figures$dlco_red_resamples, to = 'in')$h, fig.cap = 'Performance of machine learning models of insufficient diffusion capacity for carbon monoxide in cross-validation folds.'}
+
+figures$dlco_red_resamples$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-red-resamples). Performance of machine learning models of insufficient diffusion capacity for carbon monoxide in cross-validation folds.</b>
+<br>
+Machine learning models of insufficient diffusion capacity for carbon monoxide (DLCO < 80% reference value) employing the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) were developed as presented in Figure \@ref(fig:fig-dlco-red-tuning).
+Their performance at detection of the reduced DLCO in 10-repeats 10-fold cross-validation was measured by Youden's J statistic, i.e.
+the sum of sensitivity and specificity.
+Sorted J values are presented in plots.
+Orange dashed lines visualize J values expected for a meaningless model.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-tuning, fig.width = figur::convert(figures$dlco_tuning, to = 'in')$w, fig.height = figur::convert(figures$dlco_tuning, to = 'in')$h, fig.cap = 'Tuning of machine learning regression models of diffusion capacity for carbon monoxide.'}
+
+figures$dlco_tuning$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-tuning). Tuning of machine learning regression models of diffusion capacity for carbon monoxide.</b>
+<br>
+Diffusion capacity for carbon monoxide (DLCO) expressed as percentage of he age-, sex-, and weight-specific reference was modeled as as a function of 
+`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The optimal sets of algorithm's parameters such as number of trees for GBM, number of randomly selected variables for random forests, size of the hidden layer for neuronal networks or cost penalty value for SVM were found by minimizing mean absolute error (MAE) in 10-repeats 10-fold cross-validation - in a process called 'model tuning'.
+Values of the MAE statistic for various combination of the algorithm's parameters are presented in plots.
+The optimal parameter combinations corresponding to the minimal MAE are displayed in the plot captions.
+</p>
+
+```{r fig-dlco-resamples, fig.width = figur::convert(figures$dlco_resamples, to = 'in')$w, fig.height = figur::convert(figures$dlco_resamples, to = 'in')$h, fig.cap = 'Performance of machine learning regression models of diffusion capacity for carbon monoxide in cross-validation folds.'}
+
+figures$dlco_resamples$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-resamples). Performance of machine learning regression models of diffusion capacity for carbon monoxide in cross-validation folds.</b>
+<br>
+Machine learning models of diffusion capacity for carbon monoxide (DLCO, percentage of the reference value) employing the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) were developed as presented in Figure \@ref(fig:fig-dlco-tuning).
+Their performance at detection of the reduced DLCO in 10-repeats 10-fold cross-validation was measured by mean absolute error (MAE).
+Sorted MAE values are presented in plots.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm paramaters by cross-validation-mediated tuning.'}
+
+flextable::flextable(tables$tuning) %>% 
+  set_widths(c(3, 6, 6)) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('LFT: lung function testing; DLCO: diffusion lung capacity for carbon monoxide; FCV: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
+          ref_symbols = 'a', 
+          part = 'header') %>% 
   theme_vanilla
@@
 
-## Performance of Random Forest models at prediction of lung function
+## Performance of machine learning models at prediction of lung function testing abnormalities
 
@@
 
-```{r tab-bin-models, tab.cap = 'Performance of Random Forest classifiers at prediction of lung function testing (LFT) abnormalities in the training dataset and cross-validation (CV).
-CV results for any LFT abnormalities and diffusion capacity for CO (DLCO) < 80% reference are presented.
-The entire table is available in a supplementary Excel file.'}
+```{r fig-bin-performance, fig.width = figur::convert(figures$bin_performance, to = 'in')$w, fig.height = figur::convert(figures$bin_performance, to = 'in')$h, fig.cap = 'Performance of machine learning models at prediction of abnormalities of lung function testing in the training data set and repeated cross-validation.'}
 
-tables$bin_models %>% 
-  filter(`LFT variable` %in% c('LFT findings', 'DLCO < 80%'), 
-         `Data type` == 'CV') %>% 
-  select(-`Data type`) %>% 
-  flextable %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 2.7, unit = 'cm') %>% 
-  width(3, width = 4, unit = 'cm') %>% 
-  width(4, width = 5.5, unit = 'cm') %>% 
+figures$bin_performance$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-bin-performance). Performance of machine learning models at prediction of abnormalities of lung function testing in the training data set and repeated cross-validation.</b>
+<br>
+Probability of any lung function testing (LFT) abnormalities, as well as probability of abnormalities of diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC) and forced expiratory volume in one second (FEV1) defined as values < 80% of the age-, sex-, and weight-specific reference were modeled with machine learning algorithms.
+The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+Parameters controlling the algorithm's behavior were optimized by tuning as presented in Figure \@ref(fig:fig-dlco-red-tuning).
+<br>
+Performance of the machine learning models at prediction of the LFT abnormalities described above was assessed by Cohen's $\kappa$ inter-rater reliability statistic (high value: good concordance between the model prediction and the observed abnormality), Brier scores (low values: good agreement between the observed and model-predicted probability of LFT abnormality) and overall accuracy.
+The performance metric values in the training data (open symbols) and 10-repeats 10-fold cross-validation (filled symbols) are presented in scatter plots.
+Point size codes for the overall accuracy value.
+Points are labeled with the corresponding overall accuracy values.
+Point color represents the machine learning algorithm.
+Dashed lines visualize performance metrics expected for a meaningless model.
+Numbers of complete observations ('total') and observations with the particular LFT abnormality ('events') are displayed in the plot captions.
+<br>
+Note superior performance of all investigated algorithms at prediction of DLCO as compared with other LFT abnormalities.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-red-performance, fig.width = figur::convert(figures$dlco_red_performance, to = 'in')$w, fig.height = figur::convert(figures$dlco_red_performance, to = 'in')$h, fig.cap = 'Quality of prediction of insufficient diffusion capacity for carbon monoxide with machine learning models.'}
+
+figures$dlco_red_performance$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-red-performance). Quality of prediction of insufficient diffusion capacity for carbon monoxide with machine learning models.</b>
+<br>
+Machine learning models of probability of insufficient diffusion capacity for carbon monoxide (DLCO < 80% of reference value) were developed as presented in Figure \@ref(fig:fig-dlco-red-tuning) and Figure \@ref(fig:fig-bin-performance) with gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+<br>
+(A) Performance of the model assessed by receiver-operating characteristic (ROC) in the training data set and 10-repeats 10-fold cross-validation (CV).
+ROC curve and text annotation color codes for the algorithm.
+Points represent cutoffs corresponding to 50% probability of insufficient DLCO.
+Values of the area under the ROC curves (AUC) as well as sensitivity (Se), specificity (Sp) and Youden's J statistic (Se - Sp - 1) at the 50% probability cutoff are presented in the plots.
+<br>
+(B) Observed and predicted, cross-validated frequency of insufficient DLCO at consecutive follow-ups after COVID-19.
+Lines represent mean frequency estimates, standard errors are visualized as tinted ribbons.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-bin-classifiers, tab.cap = 'Performance of binary machine learining classifiers at predicting lung function testing (LFT) abnormalities in the CovILD study participants.'}
+
+flextable::flextable(tables$bin_classifiers) %>% 
+  set_widths(c(2.5, 2.5, 2.7, rep(2, 7))) %>% 
   merge_v(1:2) %>% 
   footnote(1, 1, 
-           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
-  footnote(1, 3, 
-           value = as_paragraph('C-index: concordance index; \u03BA: Cohen \u03BA inter-rater relaibility statistic; AUC, ROC: area under the curve, receiver - operating characteristic'), 
-           part = 'header', 
-           ref_symbols = 'b') %>% 
+           value = as_paragraph('LFT: lung function testing; DLCO: diffusion lung capacity for carbon monoxide; FCV: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
+          ref_symbols = 'a', 
+          part = 'header') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('SVM radial: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
+           ref_symbols = 'b', 
+           part = 'header') %>% 
+  footnote(1, 5, 
+           value = as_paragraph("\u03BA: Cohen's \u03BA inter-rater reliability statistic."), 
+           ref_symbols = 'c', 
+           part = 'header') %>% 
+  footnote(1, 7, 
+           value = as_paragraph('AUC: Area under the receiver-operating characteristic curve'), 
+           ref_symbols = 'd', 
+           part = 'header') %>% 
+  footnote(1, 10, 
+           value = as_paragraph("J: Youden's J statistic"), 
+           ref_symbols = 'e', 
+           part = 'header') %>% 
   theme_vanilla
@@
 
+<hr style = "border-top: 5px solid DarkGray" />
+
+## Performance of machine learning models at prediction of lung function paramaters
+
 <hr style = "border-top: 1px solid DarkGray" />
 
-```{r tab-reg-models, tab.cap = 'Performance of Random Forest regression models at prediction of lung function testing (LFT) variables in the training dataset and cross-validation (CV).
-CV results for diffusion capacity for CO (DLCO), forced vital capacity (FVC) and forced expiratory volume in 1 second (FEV1) are presented.
-The entire table is available in a supplementary Excel file.'}
+```{r fig-reg-performance, fig.width = figur::convert(figures$reg_performance, to = 'in')$w, fig.height = figur::convert(figures$reg_performance, to = 'in')$h, fig.cap = 'Performance of machine learning regression models at prediction of lung function testing parameters in the training data set and repeated cross-validation.'}
 
-tables$reg_models %>% 
-  filter(`Data type` == 'CV') %>% 
-  select(-`Data type`) %>% 
-  flextable %>% 
-  width(1, width = 2.7, unit = 'cm') %>% 
-  width(2, width = 2.7, unit = 'cm') %>% 
-  width(3, width = 4, unit = 'cm') %>% 
-  width(4, width = 5.5, unit = 'cm') %>% 
+figures$reg_performance$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-reg-performance).
+Performance of machine learning regression models at prediction of lung function testing parameters in the training data set and repeated cross-validation.</b>
+<br>
+Lung function testing (LFT) parameters: diffusion diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC) and forced expiratory volume in one second (FEV1) were expressed as percentages of the respective age-, sex- and weight-specific reference values.
+The LFT parameters were modeled with the following machine learning algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+Parameters controlling the algorithm's behavior were optimized by tuning as presented in Figure \@ref(fig:fig-dlco-tuning).
+<br>
+Performance of the machine learning models at prediction of the LFT parameters was assessed by mean absolute error (MAE: low value indicate a good model fit), pseudo-R^2^ as a measure of data set variance explained by the model, and Spearman's $\rho$ coefficient of correlation between the observed and fitted parameter value.
+The performance metric values in the training data (open symbols) and 10-repeats 10-fold cross-validation (filled symbols) are presented in scatter plots.
+Point size codes for the overall accuracy value.
+Points are labeled with the corresponding overall accuracy values.
+Point color represents the machine learning algorithm.
+Dashed lines visualize the R^2^ values expected for a meaningless model.
+Numbers of complete observations are displayed in the plot captions.
+<br>
+Note the overall higher accuracy of the investigated models at prediction of DLCO as compared with the remaining LFT paramaters.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-performance, fig.width = figur::convert(figures$dlco_performance, to = 'in')$w, fig.height = figur::convert(figures$dlco_performance, to = 'in')$h, fig.cap = 'Observed and model-fitted values of diffusion capacity for carbon dioxide.'}
+
+figures$dlco_performance$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-performance). Observed and model-fitted values of diffusion capacity for carbon dioxide.</b>
+<br>
+Machine learning models of diffusion capacity for carbon monoxide (percentage of the reference values) were developed as presented in Figure \@ref(fig:fig-dlco-tuning) and Figure \@ref(fig:fig-reg-performance) with the gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) algorithms.
+Fitted and observed values of DLCO are presented in scatter plots for the training data set (blue) and 10-repeats 10-fold cross-validation (CV, orange).
+Observations are visualized as points.
+General additive model (GAM) trends with standard error regions are depicted as blue lines with gray ribbons.
+Dashed lines represent the slope = 1, intercept = 0 linear trends, expected for a model with ideal concordance with the observed data.
+Values of Spearman's correlation coefficients $\rho$ of correlation of the fitted and observed DLCO values are displayed in the plot captions along with the number of complete observations in the training data set.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-model-course, fig.width = figur::convert(figures$dlco_model_course, to = 'in')$w, fig.height = figur::convert(figures$dlco_model_course, to = 'in')$h, fig.cap = 'Observed and model-predicted time course of diffusion capacity for carbon monoxide at consecutive post-COVID-19 follow-up examinations.'}
+
+figures$dlco_model_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-model-course). Observed and model-predicted time course of diffusion capacity for carbon monoxide at consecutive post-COVID-19 follow-up examinations.</b>
+<br>
+Machine learning models of diffusion capacity for carbon monoxide (percentage of the reference values) were developed as presented in Figure \@ref(fig:fig-dlco-tuning) and Figure \@ref(fig:fig-reg-performance) with the gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) algorithms.
+Observed and predicted, cross-validated diffusion DLCO values at consecutive follow-ups after COVID-19. Lines represent mean DLCO estimates, standard errors are visualized as tinted ribbons.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-reg-models, tab.cap = 'Performance of regression machine learning models at predicting lung function testing parameters in the CovILD study participants.'}
+
+flextable::flextable(tables$reg_models) %>% 
+  set_widths(c(2.5, 2.5, 2.7, rep(2.2, 3))) %>% 
   merge_v(1:2) %>% 
   footnote(1, 1, 
-           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
-  footnote(1, 3, 
-           value = as_paragraph('MAE: mean absolute error; RMSE: root mean square error; Pearson r: Pearson correlation coefficient between fitter and observed values'), 
-           part = 'header', 
-           ref_symbols = 'b') %>% 
+           value = as_paragraph('DLCO: diffusion lung capacity for carbon monoxide; FCV: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
+          ref_symbols = 'a', 
+          part = 'header') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('SVM radial: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
+           ref_symbols = 'b', 
+           part = 'header') %>% 
+  footnote(1, 5, 
+           value = as_paragraph('MAE: mean absolute error'), 
+           ref_symbols = 'c', 
+           part = 'header') %>% 
   theme_vanilla
@@
 ## Severity score, and AI-determined opacity and high opacity are the key predictors of lung function
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-reduced-global-vim, fig.width = figur::convert(figures$dlco_reduced_global_vim, to = 'in')$w, fig.height = figur::convert(figures$dlco_reduced_global_vim, to = 'in')$h, fig.cap = 'Top most influential variables for prediction of abnormalities of diffusion capacity for carbon monoxide by machine learning models.'}
+
+figures$dlco_reduced_global_vim$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-reduced-global-vim). Top most influential variables for prediction of abnormalities of diffusion capacity for carbon monoxide by machine learning models.</b>
+<br>
+Abnormalities in diffusion capacity for carbon dioxide defined as values < 80% of the age-, sex-, and weight-specific reference were modeled with machine learning as presented in Figure \@ref(fig:fig-dlco-red-tuning) and Figure \@ref(fig:fig-bin-performance).
+The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+<br>
+Variable importance measures: permutation importance for random forests, input weight importance for neural network, area under the receiver-operating characteristic curve for SVM, and mean tree-split reduction of model deviance for GBM were extracted from the models.
+The variable 0 - 100 scaled importance metrics for the top ten most important explanator variables are presented in bar plots.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-dlco-regression-global-vim, fig.width = figur::convert(figures$dlco_regression_global_vim, to = 'in')$w, fig.height = figur::convert(figures$dlco_regression_global_vim, to = 'in')$h, fig.cap = 'Top most influential variables for prediction of diffusion capacity for carbon monoxide by machine learning models.'}
+
+figures$dlco_regression_global_vim$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-dlco-regression-global-vim). Top most influential variables for prediction of diffusion capacity for carbon monoxide by machine learning models.</b>
+<br>
+Diffusion capacity for carbon monoxide expressed as percentage of the age-, sex-, and weight-specific reference was modeled with machine learning as presented in Figure \@ref(fig:fig-dlco-tuning) and Figure \@ref(fig:fig-reg-performance).
+The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+<br>
+Variable importance measures: permutation importance for random forests, input weight importance for neural network, R^2^ for the LOESS trend between the variable and outcome for SVM, and mean tree-split reduction of model deviance for GBM were extracted from the models.
+The variable 0 - 100 scaled importance metrics for the top ten most important explanatory variables are presented in bar plots.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-univariable-dlco-red, fig.width = figur::convert(figures$univariable_dlco_red, to = 'in')$w, fig.height = figur::convert(figures$univariable_dlco_red, to = 'in')$h, fig.cap = 'Human- and artificial intelligence-derived measures of severity of radiological lung abnormality and anti-coagulation treatment are strong predictors of insufficient diffusion capacity for carbon monoxide.'}
+
+figures$univariable_dlco_red$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-univariable-dlco-red).
+Human- and artificial intelligence-derived measures of severity of radiological lung abnormality and anti-coagulation treatment are strong predictors of insufficient diffusion capacity for carbon monoxide.</b>
+<br>
+Differences in single demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography (CT) parameters between CovILD data set observations with and without insufficient diffusion capacity for carbon monoxide (DLCO < 80 of reference value) were investigated.
+Differences in numeric explanatory variables were assessed by Mann-Whitney test with r effect size statistic.
+Difference in categorical variables were investigated by $\chi^2$ test with Cramer's V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method (FDR).
+<br>
+(A) Effect size and significance for differences in explanatory variable levels presented in a scatted plot.
+Each point represents a single explanatory variable.
+Significant effects are color coded.
+The dashed line represents the significance cutoff (pFDR < 0.05).
+Numbers of complete observations ('total') and observations with DLCO insufficiency ('events') are displayed in the plot caption.
+<br>
+(B) Radiologist's CT severity score (CTSS), artificial intelligence (AI)-derived lung opacity and lug high opacity in observations with and without DLCO insufficiency were presented as box plots (medians with interquartile ranges) with whiskers spanning over 150% of the interquartile range and single observations visualized as points.
+Frequencies of observations with anti-coagulant treatment during acute COVID-19 between observations with and without DLCO insufficiency are depicted in stack plot.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations with and without DLCO insufficiency are indicated in the X axes.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-univariable-dlco-corr, fig.width = figur::convert(figures$univariable_dlco_corr, to = 'in')$w, fig.height = figur::convert(figures$univariable_dlco_corr, to = 'in')$h, fig.cap = "Radiologist's computed tomography severity score, artificial intelligence-computed lung opacity and lung high opacity, and length of hospital stay are significant correlates of diffusion capacity for carbon monoxide."}
+
+figures$univariable_dlco_corr$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-univariable-dlco-corr).
+Radiologist's computed tomography severity score, artificial intelligence-computed lung opacity and lung high opacity, and length of hospital stay are significant correlates of diffusion capacity for carbon monoxide.</b>
+<br>
+The human derived computed tomography severity score (CTSS), artificial intelligence (AI) calculated lung opacity and lung high opacity as well as the length of hospital stay were the strongest correlates of diffusion capacity for carbon monoxide (DLCO, percentage of the reference value) as assessed by Spearman's rank test corrected for multiple testing with the false discovery rate method.
+<br>
+Association of DLCO with the explanatory variables described above was visualized in scatter plots of ranks.
+Each point represents a single observation.
+Blue lines with gray ribbons represent fitted linear trends with standard error regions.
+Spearman's $\rho$ coefficients of correlation with 95% confidence intervals, p values and numbers of complete observations are displayed in the plot captions.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-lft-roc, fig.width = figur::convert(figures$dlco_roc, to = 'in')$w, fig.height = figur::convert(figures$dlco_roc, to = 'in')$h, fig.cap = 'Detection of insufficient diffusion capacity for carbon monoxide by human- and artificial intelligence- derived measures of radiological lung abnormality.'}
+
+figures$dlco_roc$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-lft-roc). Detection of insufficient diffusion capacity for carbon monoxide by human- and artificial intelligence- derived measures of radiological lung abnormality.</b>
+<br>
+The optimal cutoffs of the radiologist's computed tomography severity score (CTSS), and artificial intelligence (AI) determined lung opacity and high lung opacity for detection of insufficient diffusion capacity for carbon monoxide (DLCO < 80% of reference) were found by maximizing the Youden's J statistic (i.e.
+sensitivity + specificity - 1).
+Results of the optimal opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves.
+The optimal opacity cutoffs are labeled with points and the corresponding CTSS, opacity or high opacity values.
+Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots.
+</p>
 
@@
   footnote(1, 1, 
-           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
+           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second'), 
            part = 'header', 
@@
   footnote(1, 1, 
-           value = as_paragraph('DLCOP: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
+           value = as_paragraph('DLCOP: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second'), 
            part = 'header', 
@@
   footnote(1, 1, 
-           value = as_paragraph('LFT:: lung function testting; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
+           value = as_paragraph('LFT:: lung function testting; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second'), 
            part = 'header', 
@@
   footnote(1, 2, 
-           value = as_paragraph('CTSS: chest computed tomography severity score; AI: artificual intelligence'),
+           value = as_paragraph('CTSS: chest computed tomography severity score; AI: artifical intelligence'),
            part = 'header',

### 2023-09-29 report/markdown/report.Rmd
Figures, Tables and complete Analysis Report

@@
 
+* The current analysis of data of COVID-19 convalescents pursued three goals: (1) to investigate concordance between prediction of lung computed tomography (CT) abnormalities detected by a human radiologist and by an artificial intelligence (AI) powered software, (2) to establish a relationship between AI-determined lung opacity or high opacity and CT pathology severity assessed by the radiologist, and (3) to explore, how lung CT results, demographic and clinical features translate into impaired lung function in COVID-19 patients.
+
+* These goals were addressed by a re-analysis of the published observation [CovILD data set](https://clinicaltrials.gov/study/NCT04416100) consisting of baseline demographic and clinical information, data on the course and treatment of acute COVID-19 as well as longitudinal assessment of persistent symptoms, lung CT scans and lung function testing (LFT) at two, three, six and twelve months after COVID-19 diagnosis [@Sonnweber2022; @Luger2022; @Sonnweber2020; @Sahanic2023].
+
+* A total of `r nrow(lft_globals$analysis_tbl)` observations obtained from `r length(covild$complete_ids)` CovILD study participants with complete CT and LFT results were eligible for the analysis.
+Statistical tools used to address the analysis goals included receiver operating characteristic, bootstrapped inter-rater reliability metrics, linear, robust linear and generalized additive modeling as well as multi-parameter modeling with machine learning.
+
+* In a comparison with the human radiologist, AI-determined lung opacity demonstrated a good accuracy at detection of any CT abnormalities, ground glass opacity (GGO) and reticulations in the lungs of COVID-19 convalescents.
+By contrast its ability to detect lung consolidations was poor.
+Performance of AI-determined high opacity at detection of any CT findings, GGO, and reticulation was moderate, its accuracy at identification of consolidation was fair.
+
+* CT severity score (CTSS) was established as a tool to determine an overall grade of radiological lung abnormalities in COVID-19 patients by radiologists [@Luger2022; @Sonnweber2020].
+As investigated by receiver-operating characteristic analysis, values of CTSS > 10 points may be used to identify cases of interstitial lung disease (> 5% of lung opacity computed by AI) with a moderate reliability.
+AI-computed opacity and high opacity can be associated with the radiologist's CTSS with a simple polynomial formula.
+
+* Among major LFT readouts, diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC) and forced expiratory volume in one second (FEV1), only DLCO could be successfully modeled as a function of demographic, clinical and lung CT explanatory factors.
+This suggests that DLCO but not FVC or FEV1 may be impacted by SARS-CoV-2 infection-mediated lung damage.
+
+* In a more detailed receiver operating characteristic analysis, we could show that even mild structural lung abnormalities (CTSS > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'CTSS')$cutoff[1] %>% signif(2)`, opacity > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'opacity_percent')$cutoff[1] %>% signif(2)`%, high opacity > `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'high_opacity_percent')$cutoff[1] %>% signif(2)`) may translate to insufficient DLCO defined as values below 80% of the age-, sex- and weight-specific reference.
+
+* Small size of the cohort and unavailability of a validation data set were the main limitations of the analysis.
+In particular, external validation of the LFT modeling and the CT severity - LFT relationship would allow for an unbiased evaluation of the model predictions and reliability of CTSS, opacity and high opacity as a markers of functional lung deficiency.
+
 <hr style = "border-top: 5px solid #417a8b" />
 
+# Analysis goals
+
+The analysis pursued three main goals:
+
+* comparison of artificial intelligence (AI) determined lung density and lung high density with human-determined radiological lung abnormalities in COVID-19 convalescents
+
+* determination of a cutoff of the radiologist's computed tomography (CT) severity score (CTSS) to detect cases of post-COVID-19 interstitial lung disease defined as opacity in >5% of the lung tissue and development of a mathematical relationship between CTSS and AI-determined opacity and high opacity
+
+* prediction of outcomes and abnormalities of lung function testing (LFT) in COVID-19 convalescents by radiological readouts combined with data on demographics, clinical history, course of acute COVID-19, time afer COVID-19 diangosis, persistent respiratory symptoms and physical performance rating
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-analysis-strategy, fig.width = figur::convert(figures$analysis_strategy, to = 'in')$w, fig.height = figur::convert(figures$analysis_strategy, to = 'in')$h, fig.cap = 'Goals and strategy of the analysis.'}
+
+figures$analysis_strategy$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-analysis-strategy). Goals and strategy of the analysis.</b>
+<br>
+AI: artificial intelligence; CT: computed tomography; LFT: lung function testing; CTSS: radiologist's CT severity score; ROC: receiver-operating characteristic; ILD: interstitial lung disease; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; SHAP: Shapley additive explanations.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
 # Characteristic of the cohort and exploratory data analysis
@@
 ## Characteristic of the CovILD cohort
+
+Herein, baseline demographic and clinical characteristic, data on the course and treatment of acute COVID-19, longitudinally recorded lung CT and LFT measurements as well as longitudinally surveyed respiratory symptoms and physical performance rating in the published CovILD study were investigated [@Sahanic2023; @Luger2022; @Sonnweber2022; @Sonnweber2020].
+In brief, the longitudinal observation CovILD study cohort consisted of convalescents of symptomatic SARS-CoV-2 infection during the fist outbreak of the pandemic. 
+Initially, `r length(covild$common_ids)` individuals were enrolled between March 2023 and June 2023 from three clinical centers in Tyrol, Austria: the University Clinic of Innsbruck, St.
+Vinzenz Hospital in Zams and Karl Landsteiner Acute Rehabilitation Center in Münster.
+The study inclusion criteria were PCR-confirmed SARS-CoV-2 infection and presence of characteristic COVID-19 symptoms. 
+Follow-up visits were scheduled at two, three, six and twelve months after COVID-19 diagnosis. 
+The routine follow-up visit protocol included 
+general examination by a physician, 
+survey of symptoms (e.g. fever, cough, sleep problems, olfactory dysfunction), 
+assessment of dyspnea with the modified Medical Research Council scale (mMRC) and 
+rating of physical performance with the Eastern Cooperative Oncology Group scale (ECOG) 
+along with LFT, chest CT and trans-thoracic electrocardiography. 
+Information on the course, severity and treatment of acute COVID-19 was obtained at the two-month follow-up via an interview of the participant and screening of the hospital electronic records [@Sahanic2023; @Sonnweber2020].
+CT scans were evaluated by a team of experienced radiologists according to the Fleischner society guidelines. 
+CT lung opacity and high lung opacity expressed as percentages of the organ were computed with an AI - powered software (Syngo.via CT Pneumonia Analysis Software) [@Luger2022; @Sonnweber2020].
+
+In the current report, `r nrow(lft_globals$analysis_tbl)` observations from all follow-up examinations with the complete CT and LFT results recorded in `r length(covild$complete_ids)` participants were analyzed.
+Please note, that per participant, up to four observations were available. 
+Due to this participant-matching, the observations were not independent, which had consequences for the analysis procedures, e.g.
+for multi-parameter modeling (__Figure \@ref(fig:fig-consort)__, __Table \@ref(tab:tab-study-variables)__).
+
+The analyzed participants were grouped according to the severity of acute COVID-19 as 
+convalescents of ambulatory COVID-19 (`r get_percent(cohort$analysis_tbl, 'severity_class')['ambulatory mild']` of the cohort, home-isolated during infection, WHO ordinal scale for clinical improvement: 1 - 2),
+hospitalized moderate COVID-19 patients (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized moderate']`, without oxygen or with low-flow oxygen, WHO ordinal scale for clinical improvement: 3 - 4),
+and hospitalized severe COVID-19 patients (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized severe']`, high-flow oxygen or mechanical ventilation, WHO ordinal scale for clinical improvement: 5 - 7) (__Table \@ref(tab:tab-cohort)__).
+In the analyzed data set, `r get_percent(cohort$analysis_tbl, 'sex')['male']` of participants were male. 
+The percentage of males was significantly higher in moderate and severe COVID as compared with ambulatory COVID-19 individuals (`r get_test(cohort$test, 'sex')`, moderate). 
+The median age at COVID-19 diagnosis was `r signif(median(cohort$analysis_tbl$age), 2)` years, ambulatory COVID-19 patients were significantly younger than hospitalized COVID-19 survivors (`r get_test(cohort$test, 'age')`, moderate).
+Ex- or current smokers made up to `r sum(get_percent(cohort$analysis_tbl, 'smoking', percent_mark = FALSE)[2:3])`% of the analyzed cohort. 
+There were no significant differences in participants with smoking history between the acute COVID-19 severity subsets. 
+Yet, the pack-year number was the highest in hospitalized moderate COVID-19 convalescents (`r get_test(cohort$test, 'smoking_pack_years')`, weak). 
+Based on body mass index (BMI) > 25 kg/m^2^, 
+`r sum(get_percent(cohort$analysis_tbl, 'body_mass_class', percent_mark = FALSE)[2:3])`% of participants were overweight or obese; there were no significant differences in percentages of obese or overweight participants between the acute COVID-19 severity subsets.
+Endocrine or metabolic conditions (`r get_percent(cohort$analysis_tbl, 'endocrine_metabolic_comorbidity')['yes']` of analyzed participants), 
+cardiovascular diseases (`r get_percent(cohort$analysis_tbl, 'cardiovascular_comorbidity')['yes']`), 
+hypertension (`r get_percent(cohort$analysis_tbl, 'hypertension_comorbidity')['yes']`), 
+hypercholesterolemia (`r get_percent(cohort$analysis_tbl, 'hypercholesterolemia_comorbidity')['yes']`), 
+and type II diabetes (`r get_percent(cohort$analysis_tbl, 'diabetes_comorbidity')['yes']`) 
+were the most common comorbidities. 
+Endocrine or metabolic comorbidities (`r get_test(cohort$test, 'endocrine_metabolic_comorbidity')`), 
+cardiovascular conditions (`r get_test(cohort$test, 'cardiovascular_comorbidity')`), 
+hypercholesterolemia (`r get_test(cohort$test, 'hypercholesterolemia_comorbidity')`), 
+and diabetes (`r get_test(cohort$test, 'diabetes_comorbidity')`) 
+were more significantly common in survivors of moderate or severe COVID-19 as compared with ambulatory COVID-19 individuals. 
+Effects sizes of these differences were weak or moderate (__Table \@ref(tab:tab-cohort)__).
+
+The median length of hospital stay for analyzed participants was `r signif(median(cohort$analysis_tbl$hospital_stay_days), 2)` days and was longer for severe that moderate COVID-19 patients. 
+Antibiotic anti-infective treatment was applied to `r get_percent(cohort$analysis_tbl, 'antiinfective_treatment')['yes']`% of participants, this percentage was the highest in severe COVID-19 patients 
+(`r get_test(cohort$test, 'antiinfective_treatment')`, large). 
+Macrolides and anti-platelet drugs were administered to 
+`r get_percent(cohort$analysis_tbl, 'macrolide_treatment')['yes']` and 
+`r get_percent(cohort$analysis_tbl, 'antiplatelet_treatment')['yes']` of participants, respectively. 
+Their frequency was higher in hospitalized than in ambulatory patients 
+(macrolides: `r get_test(cohort$test, 'macrolide_treatment')`, weak; 
+anti-platelet: `r get_test(cohort$test, 'antiplatelet_treatment')`, weak). 
+Systematic steroid administration was not considered as a treatment of moderate or severe COVID-19 during the first European outbreak of SARS-CoV-2 in 2020. 
+Hence none of the study participants received corticosteroid therapy at the peak of acute COVID-19. 
+Corticosteroids were administered to 
+`r get_percent(cohort$analysis_tbl, 'steroid_treatment')['yes']`% of analyzed participants with persistent pneumonia beginning from the third week after diagnosis at the discretion of the physician. 
+Such sub-acute corticosteroid treatment was the most frequent in severe disease convalescents as compared with moderate or ambulatory COVID-19 (`r get_test(cohort$test, 'steroid_treatment')`, moderate).
+The median weight loss during acute COVID-19 was `r abs(median(cohort$analysis_tbl$weight_change_kg))` kg and was significantly higher in severe as compared with ambulatory or moderate COVID-19 patients
+(`r get_test(cohort$test, 'weight_change_kg')`, large) (__Table \@ref(tab:tab-cohort)__).
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-consort, fig.width = figur::convert(figures$consort, to = 'in')$w, fig.height = figur::convert(figures$consort, to = 'in')$h, fig.cap = 'Analysis inclusion scheme.'}
+
+figures$consort$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-consort). Analysis inclusion scheme.</b>
+<br>
+CT: computed tomography of the chest; LFT: lung function testing; GGO: ground glass opacity; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; ECG: echocardiography; AI: artificial intelligence.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-study-variables, tab.cap = 'Study variables'}
+
+flextable::flextable(tables$study_variables) %>% 
+  set_widths(c(2.2, 3, 2.5, 3, 5, 1.5, 1.5)) %>% 
+  merge_v(1:2) %>% 
+  theme_vanilla
+
+```
 
@@
 
-<hr style = "border-top: 1px solid DarkGray" />
-
-```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+Frequency of any radiological lung abnormalities, ground glass opacity (GGO) and reticulation diagnosed by a radiologist was substantially higher in hospitalized than in ambulatory COVID-19 individuals.
+Analogically, the pace of recovery from these lung lesions was the slowest in severe COVID-19 survivors: 
+among `r filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up') %>% nrow` severe COVID-19 convalescents investigated at the twelve-month follow-up, GGO were observed in
+`r get_percent(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up'), 'GGO_finding')['yes']` and reticulation was evident in `r get_percent(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up'), 'reticulation_finding')['yes']` of cases.
+Consolidation was diagnosed most commonly in hospitalized severe COVID-19 individuals at the two-month follow-up (__Figure \@ref(fig:fig-ct-finding-course)__, __Table \@ref(tab:tab-cohort-ct)__). 
+A more detailed descriptive analysis of lung lesions in the CovILD cohort is provided in our recent publications [@Sahanic2023; @Luger2022].
+
+Severity of radiological lung abnormalities was gauged with the CT severity score (CTSS) determined by the radiologist and, in parallel, by percentages of organ opacity and high opacity computed by an AI-powered software [@Luger2022; @Sonnweber2020].
+Independently of the severity readout, the most severe or widespread lesions were found in severe COVID-19 survivors followed by moderate COVID-19 patients. 
+A substantial recovery from structural lung damage could be inferred from decreasing CTSS, opacity and high opacity values at the consecutive follow-ups. 
+Still, low-grade residual lesions of unknown clinical relevance were observed in majority of severe COVID-19 survivors at one year after COVID-19 diagnosis (percentage of the lungs with opacity: 
+median `r median(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up')$opacity_percent) %>% signif(2)`%, 
+interquartile range: `r quantile(filter(cohort_ct$analysis_tbl, severity_class == 'hospitalized severe', follow_up == '12-month follow-up')$opacity_percent, c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' to ')`%) (__Figure \@ref(fig:fig-ct-numeric-course)__, __Table \@ref(tab:tab-cohort-ct)__).
+For a detailed longitudinal analysis of structural pulmonary recovery, please refer to our published reports [@Sahanic2023; @Luger2022].
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
 
@@
 <p class = "legend">
-<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
 <br>
@@
 
-```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
 
@@
 <p class = "legend">
-<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
 <br>
@@
 
-## Lung function testing paramaters and abnormalities
-
-<hr style = "border-top: 1px solid DarkGray" />
-
-```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+## Lung function testing parameters and abnormalities
+
+LFT abnormalities were defined as values of diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC), forced expiratory volume in one second (FEV1) or total lung capacity (TLC) below < 80% of the age-, sex- and weight-specific reference, or FEV1 to FVC ratio (FEV1:FVC) below 70% of the reference value [@Sahanic2023; @Sonnweber2022; @Sonnweber2020].
+Such LFT findings were discerned in more than one third of available observations at consecutive follow-ups in the entire analyzed data set
+(two months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '2-month follow-up'), 'LFT_findings')['yes']`, 
+twelve months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '12-month follow-up'), 'LFT_findings')['yes']` of observations). 
+Their percentage was the highest in severe COVID-19 survivors; in this subset, a clear recovery trend was observed 
+(two months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '2-month follow-up', severity_class == 'hospitalized severe'), 'LFT_findings')['yes']`, 
+twelve months: `r get_percent(filter(cohort_lft$analysis_tbl, follow_up == '12-month follow-up', severity_class == 'hospitalized severe'), 'LFT_findings')['yes']`). 
+This suggests, that in the severe COVID-19 subset, functional lung pathology was at least partly mediated by SARS-CoV-2 infection. 
+In turn, no clear recovery from LFT abnormalities was evident in ambulatory and moderate COVID-19 patients. 
+This may indicate that functional lung deficits in those severity subsets can be COVID-19-independent, e.g.
+as a result of a pre-existing diagnosed or undiagnosed pulmonary condition or represent a psychosomatic phenomenon [@Hufner2023].
+Similar observations were made for deficits of the major LFT readouts: DLCO, FVC and FEV1, as well as values of DLCO, FVC and FEV1 expressed as percentage of the respective references (__Figures \@ref(fig:fig-lft-finding-course)__ - __\@ref(fig:fig-lft-numeric-course)__, __Table \@ref(tab:tab-cohort-lft)__).
+A more detailed results of descriptive and longitudinal analysis of LFT measurements in the CovILD cohort can be found in our recent publications [@Sahanic2023; @Sonnweber2022].
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
 
@@
 <p class = "legend">
-<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
 <br>
@@
 
-```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}
+```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
 
@@
 <b>Figure \@ref(fig:fig-lft-numeric-course).
-Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
+Time course of numeric lung function testing readouts at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
 <br>
@@
 
+<hr style = "border-top: 5px solid DarkGray" />
+
+## Persistent respiratory symptoms and physical performance
+
+In the current analysis we explored persistent symptoms of COVID-19 of potential relevance for lung function and structural lung lesions: cough, dyspnea and physical performance impairment [@Sahanic2023; @Sonnweber2020].
+Dyspnea defined by modified Medical Research Council scale (mMRC) > 0 points affected 
+`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'dyspnea_symptom')['yes']` and 
+`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'dyspnea_symptom')['yes']` of participants at the two and twelve follow-up, respectively. 
+The highest frequency of dyspnea was found in severe COVID-19 survivors at the two-month follow-up. 
+Cough was reported by 
+`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'cough_symptom')['yes']` of study participants at the earliest follow-up and 
+by `r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'cough_symptom')['yes']` one year after COVID-19 diagnosis. 
+There were no clear differences in frequency of cough between the acute COVID-19 severity subsets. 
+Impaired physical performance defined as Eastern Cooperative Oncology Group scale (ECOG) > 0 points was reported by 
+`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '2-month follow-up'), 'impaired_performance_symptom')['yes']` of 
+participants at the two-month follow-up and by 
+`r get_percent(filter(cohort_sympt$analysis_tbl, follow_up == '12-month follow-up'), 'impaired_performance_symptom')['yes']` of participants one year after COVID-19. 
+Analogically, impaired physical performance was the most common in the severe acute COVID-19 subset (__Figures \@ref(fig:fig-symptom-course)__ - __\@ref(fig:fig-symptom-rating-course)
+__, __Table \@ref(tab:tab-cohort-symptoms)__). 
+For a more detailed analysis of the symptom kinetic, please refer to our recent publications [@Sahanic2023; @Sonnweber2022; @Hufner2023].
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-symptom-course, fig.width = figur::convert(figures$symptom_course, to = 'in')$w, fig.height = figur::convert(figures$symptom_course, to = 'in')$h, fig.cap = 'Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
+
+figures$symptom_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-symptom-course).
+Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
+<br>
+Intensity of self-reported dyspnea was assessed with the modified Medical Research Council scale (mMRC) and presence of dyspnea was defined as mMRC > 0. 
+Presence of cough was surveyed by a single 'yes'/'no' item. 
+Impairment of physical performance was assessed with the Eastern Cooperative Oncology Group scale (ECOG, high values indicate more profound impairment) and impaired physical performance was defined as ECOG > 0.
+Presence of dyspnea, cough and impairment of physical performance at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots.
+Numbers of complete observations at the follow-up examinations are displayed above the data bars.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-symptom-rating-course, fig.width = figur::convert(figures$symptom_rating_course, to = 'in')$w, fig.height = figur::convert(figures$symptom_rating_course, to = 'in')$h, fig.cap = 'Rating of dyspnea and physical performance loss at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.'}
+
+figures$symptom_rating_course$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-symptom-rating-course). Rating of dyspnea and physical performance loss at the consecutive follow-ups in CovILD study participants stratified by severity of acute COVID-19.</b>
+<br>
+Intensity of self-reported dyspnea and impairment of physical performance was assessed with the modified Medical Research Council (mMRC) and the Eastern Cooperative Oncology Group scales (ECOG, high values indicate more profound impairment), respectively.
+Dyspnea and physical performance impairment scores are presented in stack plots, with the score value coded by the bar color.
+Numbers of complete observations at the follow-up examinations are displayed above the data points.
+</p>
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-cohort-symptoms, tab.cap = 'Presence and rating of symptoms of relevance for lung function at consecutive follow-ups.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
+Categorical variables are presented as percentages and counts within the complete observation set.'}
+
+flextable::flextable(tables$cohort_symptoms) %>% 
+  set_widths(c(2.2, 2.5, rep(3.5, 4))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 2, 
+           value = as_paragraph('mMRC: modified Medical Research Council dyspnea scale; ECOG: Eastern Cooperative Oncology Group scale of physical performance impairment.'), 
+           ref_symbols = 'a', 
+           part = 'header') %>% 
+  theme_vanilla
+
+```
+
 <hr style = "border-top: 5px solid #417a8b" />
 
-# Detection of radiological lung findings with AI-determined opacity and high opacity
+# Detection of radiological lung findings by AI-determined opacity and high opacity
 
 ## Optimal cutoff, sensitivity and specificity of AI opacity and high opacity at detection of radiological findings
+
+The first main goal of the analysis was to compare radiological lung abnormalities detected by AI-determined percentage of the lung opacity and high opacity with diagnoses made by the human radiologist.
+In the initial step of this analysis task, we sought to find cutoffs of lung opacity and high lung opacity values which discriminate between observations with and without any CT findings, GGO, reticulation or consolidation discerned by the radiologist with the optimal sensitivity and specificity.
+Such optimal cutoffs were determined by an iterative screening of opacity and high opacity values and comparison of the respective Youden's J statistic values defined as $J = Sensitivity + Specificity - 1$.
+The opacity and high opacity cutoffs corresponding to the maximum of J were chosen for stratification of the observations [@Lopez-Raton2014] (__Table \@ref(tab:tab-ai-cutoffs)__).
+
+As presented in __Figure \@ref(fig:fig-ai-opacity-roc)__, values of AI-determined opacity ranging between 
+`r filter(cut_opacity$stats$optimal_cutoff, response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding', 'consolidation_finding'))$cutoff %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of the lungs were identified as the best cutoffs for detection of any CT abnormalities, GGO, reticulation and consolidation, respectively.
+Of note, the highest optimal detection cutoff of opacity was identified for consolidations. 
+All four types of CT findings were detected with a similar overall accuracy as evident from comparable values of J and area under the receiver operating characteristic curve (AUC). 
+Values of AI-determined high opacity ranging between 
+`r filter(cut_high$stats$optimal_cutoff, response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding', 'consolidation_finding'))$cutoff %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of the lungs were proposed as the optimal cutoffs for detection of any CT abnormalities, GGO, reticulation and consolidation.
+The highest optimal cutoff of high opacity was found for consolidation. 
+Interestingly, performance of high opacity at detection of consolidations was substantially better than performance of high opacity at detection of other CT findings as evident from higher sensitivity, J and AUC values (__Figure \@ref(fig:fig-ai-high-opacity-roc)__).
 
@@
 
+Canonical, unweighted Cohen's $\kappa$ metric is used to assess concordance between two raters at prediction of a categorical outcome. 
+Since $\kappa$ accounts for frequency of outcomes concordantly predicted by both raters by chance, it is considered a more robust measure of model's or cutoff's performance than accuracy, specificity or sensitivity, especially in data sets with highly frequent or rare outcomes [@Cohen1960; @McHugh2012].
+Herein we employed $\kappa$ to assess the concordance at prediction of CT findings between the human radiologist and AI-determined opacity/high opacity stratified by the optimal detection cutoffs characterized above.
+
+In such analysis performed for all available observations, lung opacity appeared to discriminate between observations with and without any CT findings, GGO and reticulation identified by the radiologist with good and comparable reliability
+($\kappa$: `r filter(op_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding'))$estimate %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+By contrast, reliability of AI-determined opacity at detection of relatively infrequent consolidations was poor 
+($\kappa$ = `r  filter(op_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response == 'consolidation_finding')$estimate %>% signif(2)`). 
+While reliability of opacity at identification of any radiological lung abnormalities was similar in ambulatory, moderate and severe COVID-19 survivors, the concordance at detection of GGO and reticulation was substantially better in moderate or severe than ambulatory COVID-19 patients (__Figure \@ref(fig:fig-ai-opacity-severity)__, __Table \@ref(tab:tab-ai-cutoffs-severity)__).
+Notably, reliability of AI-determined opacity at detection of any CT findings, GGO and reticulation tended to be better one-year after COVID-19 diagnosis as compared with the former follow-up examinations (__Figure \@ref(fig:fig-ai-opacity-fup)__, __Table \@ref(tab:tab-ai-cutoffs-fup)__).
+Curiously, a training effect of the human radiologists may explain this surprising phenomenon but also the possibility to consult CT scans obtained from a particular patient at earlier follow-ups.
+
+Reliability of AI-computed high lung opacity stratified by the respective optimal cutoffs at detection of any CT findings, GGO and reticulation identified by the radiologist in the entire analyzed data set was moderate
+($\kappa$: `r filter(high_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response %in% c('CT_findings', 'GGO_finding', 'reticulation_finding'))$estimate %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+As inferred from higher $\kappa$ values, performance of the high lung opacity at detection of GGO tended to be better in severe COVID-19 survivors than in ambulatory or moderate COVID-19 patients. 
+By contrast, the highest reliability at detection of reticulation was observed in ambulatory COVID-19 convalescents. 
+Reliability of high lung opacity at identification of consolidations in the entire data set in was fair 
+($\kappa$ = `r filter(high_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa', response == 'consolidation_finding')$estimate %>% signif(2)`) (__Figure \@ref(fig:fig-ai-high-opacity-severity)__, __Table \@ref(tab:tab-ai-cutoffs-severity)__).
+This illustrates the difficulty of robust detection of these rare structural lesions by software-determined tissue density.
+Reliability of detection of any CT abnormalities, GGO and reticulation by high opacity was comparable at the two, three and twelve month follow-up. 
+Yet, $\kappa$ values computed for the six-month follow-up measurements were remarkably lower, which may reflect an methodological inconsistency or a systematic quality issue at this particular time point (__Figure \@ref(fig:fig-ai-high-opacity-fup)__, __Table \@ref(tab:tab-ai-cutoffs-fup)__).
+
 <hr style = "border-top: 1px solid DarkGray" />
@@
 Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
-$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+$\kappa$ values with their 95% confidence intervals obtained by block bootstrapping with B = 2000 iterations are presented in Forest plots.
 Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
@@
 Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
-$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+$\kappa$ values with their 95% confidence intervals obtained by block bootstrapping with B = 2000 iterations are presented in Forest plots.
 Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
@@
 Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
-$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+$\kappa$ values with their 95% confidence intervals obtained by block bootstrapping with B = 2000 iterations are presented in Forest plots.
 Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
@@
 Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic.
-$\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots.
+$\kappa$ values with their 95% confidence intervals obtained by block bootstrapping with B = 2000 iterations are presented in Forest plots.
 Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
@@
 The opacity and high opacity values were stratified by their optimal cutoffs determined by Youden criterion in the entire CovILD cohort.
-Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}
+Statistic values are presented with 95% confidence intervals obtained by block bootstrap.'}
 
@@
   merge_v(1:2) %>% 
-  footnote(1, 3, 
+  footnote(1, 2, 
            value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
@@
 ## Chest computed tomography severity score and interstitial lung disease
+
+The second main goal of the analysis concerned determination of a threshold of the radiologist-determined CTSS that discriminates between cases of ILD defined by the AI-determined opacity in > 5% of the lungs and observations without CT lesions or bona-fide sub-clinical radiological lesions, i.e.
+without ILD.
+In addition, we sough to establish a mathematical relationship between CTSS, and AI-computed lung opacity and high lung opacity.
+
+At this point it has to be stressed that CTSS defined as a sum of pathology scores calculated separately by the radiologist for each lobe may prove an inconsistent measure of lung damage severity, especially in cases of localized severe lung damage.
+For instance, single subtle findings scattered in the entire organ are expected to yield CTSS of 5. 
+The same CTSS value is expected for a patient with a profound damage of a single lobe. 
+For this reason caution is recommended at treating CTSS as a potential diagnostic marker of ILD or as a numeric explanatory variable of lung opacity.
+
+Cases of ILD made up for 
+`r get_percent(filter(ild_freq$analysis_tbl, severity_class == 'cohort'), 'opacity_class')[2]` of all analyzed observations and were virtually restricted to hospitalized participants at the two- and three-month follow-ups.
+In particular, half of severe COVID-19 survivors at the earliest follow-up examination fulfilled the ILD definition (__Figure \@ref(fig:fig-ild-freq)__). 
+By applying the procedure of optimal cutoff finding based on the maximal Youden's J statistic, 
+CTSS of `r ild_cutoff$stats$optimal_cutoff$cutoff[1]` points was proposed as the ILD detection threshold characterized by an excellent 
+sensitivity (Se = `r signif(ild_cutoff$result_tbl$Sensitivity, 2)`) and 
+overall accuracy measured by AUC (AUC = `r stri_extract(ild_cutoff$result_tbl$AUC, regex = '\\d{1}\\.\\d{2}')`) (__Figure \@ref(fig:fig-ctss-ild-cutoff)A__, __Table \@ref(tab:tab-ild-cutoff)__). 
+The concordance between the stratified CTSS and ILD defined by opacity > 5% in the entire data set was, however, only moderate as measured by 
+$\kappa$ = `r filter(ild_strata$severity_stats, severity_class == 'cohort', statistic == 'kappa')$estimate %>% signif(2)`. 
+In line with the presence of ILD virtually only during early COVID-19 recovery, reliability of ILD detection by CTSS was remarkably better at the the two- and three-month follow-up as compared with six and twelve months after COVID-19 diagnosis (__Figure \@ref(fig:fig-ctss-ild-cutoff)B__, __Table \@ref(tab:tab-ild-cutoff-severity)__ - __\@ref(tab:tab-ild-cutoff-fup)__).
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-ild-freq, fig.width = figur::convert(figures$ild_freq, to = 'in')$w, fig.height = figur::convert(figures$ild_freq, to = 'in')$h, fig.cap = 'Frequency of interstitial lung disease defined as opacity in more than 5% of the organ in the CovILD data set.'}
+
+figures$ild_freq$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-ild-freq). Frequency of interstitial lung disease defined as opacity in more than 5% of the organ in the CovILD data set.</b>
+<br>
+Interstitial lung disease (ILD) was defined as presence of AI-computed opacity in > 5% of the lungs.
+Percentages of observations with ILD in the entire analyzed data set, and its strata defined by acute COVID-19 severity and follow-up examinations are presented in a stacked bar plot.
+Numbers of complete observations are displayed above the bars.
+</p>
 
@@
 (B) Concordance between the ILD diagnosis with CTSS > 10 points and ILD defined as opacity > 5% of the lung in the whole observation set ('cohort') and in observations stratified by the acute disease severity or post-COVID-19 follow-up was assessed by Cohen's $\kappa$ inter-rater reliability statistic.
-$\kappa$ values with 95% confidence intervals obtained by bootstrap with B = 2000 iterations are presented in Forest plots.
+$\kappa$ values with 95% confidence intervals obtained by block bootstrap with B = 2000 iterations are presented in Forest plots.
 Numbers of complete observations ('total') and cases of ILD defined by >5% lung opacity ('events') are indicated in the Y axes.
@@
 The optimal cutoff for CTSS was determined by Youden criterion in the entire CovILD cohort.
-Statistic values are presented with 95% confidence intervals obtained by bootstrap.'}
+Statistic values are presented with 95% confidence intervals obtained by block bootstrap.'}
 
@@
 ## Mathematical relationship of the severity score, and AI-determined opacity and high opacity
+
+To establish a mathematical relationship between CTSS, and AI-computed lung opacity and high lung opacity, three modeling approaches were employed:
+
+* ordinary least square linear modeling of square root opacity or square root high opacity as a function of fourth order polynomial of CTSS
+
+* robust linear modeling of square root opacity or square root high opacity as a function of fourth order polynomial of CTSS [@Huber2011; @Ripley2022]
+
+* generalized additive modeling of square root opacity or square root high opacity as a function of linear CTSS term and thin plate spline of CTSS [@Wood2023; @Wood2017]
+
+Goodness of fit of those models was assessed by median absolute error, pseudo-R^2^ as a metric of variance explained by the model was used as a measure of explanatory performance. 
+As shown in __Figure \@ref(fig:fig-ctss-ai-relation)A__, all three models of lung opacity demonstrated large and nearly identical explanatory power. 
+However, the robust polynomial model demonstrated the best fit as inferred from the lowest median absolute error. 
+Explanatory performance of the models of high lung opacity were characterized by a substantially worse but still very high fractions of explained variance measured with pseudo-R^2^ statistic. 
+Analogically, the robust polynomial model of high opacity displayed a lower fit error as compared with the least square linear or general additive model (__Figure \@ref(fig:fig-ctss-ai-relation)B__). 
+A visual analysis of the modeled opacity - CTSS and high opacity - CTSS trends reveals an interesting phenomenon. 
+Increases of CTSS in the 0 - 10 point range correspond to relatively minute changes in opacity (predominantly in the 0 - 5% range) and high opacity (predominantly 0 - 0.5% range). 
+This indicates a generally high granularity of CTSS and an almost linear relationship between CTSS and AI parameters for subtle lung lesions. 
+However, high CTSS values demonstrated no linear relationship with opacity or high opacity and rather poor ability to quantitate high-grade or widespread lesions. 
+In an illustrative example, opacity of CT scans rated with CTSS = 15 points ranged between `r filter(ild_cutoff$analysis_tbl, CTSS == 15)$opacity_percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of the lungs (median opacity: `r filter(ild_cutoff$analysis_tbl, CTSS == 15)$opacity_percent %>% median %>% signif(2)`%, interquartile range: `r filter(ild_cutoff$analysis_tbl, CTSS == 15)$opacity_percent %>% quantile(c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' to ')`).
+This inaccuracy may be traced back to the inability of CTSS to differentiate between mild but frequent radiological findings and localized but severe lesions described above.
+
+Collectively, we could link AI-determined opacity and high opacity with human-determined CTSS in lung CT measurements in COVID-19 convalescents using relatively simple mathematical models. 
+Still, due to inconsistency of CTSS expected for localized high-grade lung lesions and mild scattered findings, the relationship of CTSS and AI-computed opacity or high opacity may not be easily generalized.
 
@@
 
-## Development of multi-paramater machine learning models of lung function variables and abnormalities
+## Modeling strategy
+
+The final goal of the analysis was to investigate how lung function of COVID-19 convalescents is impacted by presence and severity of structural lung abnormalities detected by chest CT along with demographic, clinical history and COVID-19-related explanatory factors.
+To this end we resorted to multi-parameter modeling of the major LFT readouts:
+
+* presence of any LFT abnormalities defined as DLCO, FVC, FEV1 or TLC below 80% of the reference value or FEV1:FVC below 70% of the reference value
+
+* insufficiency of each DLCO, FVC and FEV1 defined as measurements below 80% of the reference value
+
+* DLCO, FVC and FEV1 expressed as percentages of the reference values
+
+The explanatory factors included the following:
+
+* baseline demographic and medical history variables like age, sex, body mass index, smoking history, smoking pack-years and comorbidity
+
+* variables related to the course and treatment of acute COVID-19 like acute disease severity, hospitalization length and treatment
+
+* follow-up: two, three, six and twelve months after COVID-19 diagnosis
+
+* longitudinally recorded persistent respiratory symptoms and their rating (dyspnea, cough) and impairment of physical performance (ECOG)
+
+* longitudinal results of lung CT including presence of any findings, GGO, reticulation, consolidation, CTSS as well as AI-determined lung opacity and high opacity
+
+The modeling responses and explanatory variables are listed in __Table \@ref(tab:tab-study-variables)__ and the modeling strategy scheme is presented in __Figure \@ref(fig:fig-mod-strategy)__. 
+Because of the large number of explanatory variables 
+(n = `r lft_globals[c("baseline_variables", "ct_variables", "symptom_variables")] %>% reduce(union) %>% length`) in relation to the participant number (n = `r lft_globals$analysis_tbl$ID %>% unique %>% length`)
+as well as possibly non-normally distributed explanatory features (e.g.
+CTSS, pack-years) and expected high level interactions between the explanatory factors, traditional linear modeling or regularized linear modeling were not regarded as feasible modeling algorithms.
+Instead we constructed multi-parameter models of lung function with four machine learning algorithms expected to be more robust to variable distribution, high dimensionality of the data set and explanatory variable interactions were used [@Strobl2009]:
+(1) canonical random forest algorithm implemented in R by the _ranger_ package [@Wright2017; @Breiman2001], 
+(2) gradient boosted machines (GBM) implemented by the _gbm_ package [@Natekin2013; @Greenwell2022; @Friedman2002; @Friedman2001], 
+(3) neural network with a single hidden layer provided by the _nnet_ package [@Ripley2014], and 
+(4) support vector machines (SVM) with radial kernel implemented by the R package _kernlab_ [@Karatzoglou2004; @Weston1998]. 
+Of importance, these algorithms predict the outcome based on diverse mathematical principles: 
+the random forests and GBM build and evaluate ensembles of simple tree models [@Natekin2013; @Strobl2009; @Breiman2001], 
+the neural network employs a layered structure of modeling units (neurons) that is recursively optimized to fit the outcome [@Ripley2014], and 
+the SVM aims at fitting hyperplanes to discriminate between the outcome categories with a small fraction of data points called support vectors [@Weston1998]. 
+The interested reader may refer to excellent review articles cited above for details of the modeling algorithms.
+
+Because of a small number of the study participants and data points, we abstained from splitting the data set into a training portion utilized for optimization (tuning) and training of the models and a test subset used solely for evaluation of the modeling results - a so-called 'holdout' strategy.
+Instead, we decided to optimize (tune) and evaluate the model performance in a single 10-repeats 10-fold cross-validation setting - a strategy similar to the approach used in an analysis of the six-month follow-up CovILD study results [@Sonnweber2022].
+However, it has to be stressed that such evaluation strategy does not allow for a fully unbiased assessment of the model performance. 
+Parameters controlling behavior of the modeling algorithms were optimized by testing performance of models constructed for parameter value combinations in 10-repeats 10-fold cross-validation  - in a process called 'tuning'.
+The parameter set corresponding to the best model performance assessed by Youden's J and mean absolute error for the classification and regression models, respectively, was chosen for construction of the final models subsequently subjected to a more detailed performance evaluation [@Kuhn2008].
+Since the data points were not independent, our cross-validation approach had to account for the participant matching of the observations. 
+Thus, we resorted to 'participant-wise' or 'block' cross-validation by placing all observations obtained from a particular participant either in the training or the test portion of the data (__Figure \@ref(fig:fig-mod-strategy)__).
+
+Performance of the final tuned models was evaluated by comparing the observed outcomes with predictions made in the genuine training data set and predictions made in the test portions of the cross-validation folds (so called 'out-of-fold' or OOF predictions).
+The classification models predicted presence or absence of particular LFT abnormalities by returning probability of the pathological outcome. 
+By default, the pathological outcome, e.g. reduced DLCO or any LFT findings, was predicted for probability > 50%. 
+Quality of predictions of classification models was assessed by 
+Cohen's $\kappa$ as a measure of concordance between the predicted and observed outcome [@Cohen1960; @McHugh2012], 
+Brier score as a readout of model credibility, i.e. average probability with which the pathological outcome is correctly identified [@Goldstein-Greenwood2021; @Brier1950], 
+and by standard receiver operating characteristic measures: overall accuracy, area under the curve (AUC), Youden's J, sensitivity and specificity of detection of the LFT abnormality. 
+The regression models predicted numeric values of DLCO, FVC and FEV1. 
+Their performance was investigated by mean absolute error as a measure of goodness of the model fit, pseudo-R^2^ as a metric of outcome variance explained by the model and correlation between the predicted and observed values assessed by Spearman's $\rho$ coefficient (__Figure \@ref(fig:fig-mod-strategy)__).
+
+Finally, results of the modeling were interpreted by computing mean SHAP values (Shapley additive explanations) for explanatory variables. 
+The basic principle of SHAP is to compare performance of the models refitted for all possible subsets of explanatory variables with the variable of interest with the models refitted for all possible subsets of explanatory variables without the variable of interest.
+By this means, the global contribution of the particular explanatory variable to the model performance, so called global variable performance, can be estimated. 
+Of note, the SHAP procedure is model-agnostic, i.e. works the same way for any machine learning algorithm [@Lundberg2017]. 
+As an complementary approach to model interpretation, we compared levels and distribution of the most influential explanatory variables between the observations with and without selected LFT abnormalities and assessed their ability to detect those LFT findings in a receiver-operating characteristic analysis.
+Yet, this 'univariable' interpretation technique is limited as it ignores interactions between explanatory variables which may be crucial for prediction of insufficient lung function. 
+In addition, it does not account for matching of observations originating from the same participants which poses a possible violation of the independence assumption made by most statistical tests. 
+For this reason, interpretation of the univariable analysis results has to be made with caution (__Figure \@ref(fig:fig-mod-strategy)__).
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r fig-mod-strategy, fig.width = figur::convert(figures$mod_strategy, to = 'in')$w, fig.height = figur::convert(figures$mod_strategy, to = 'in')$h, fig.cap = 'Strategy for modeling of lung function outcomes.'}
+
+figures$mod_strategy$plot
+
+```
+
+<p class = "legend">
+<b>Figure \@ref(fig:fig-mod-strategy). Strategy for modeling of lung function outcomes.</b>
+<br>
+LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; CT: computed tomography; GGO: ground glass opacity; CTSS: radiologist's lung CT severity score; AI: artificial intelligence.
+</p>
+
+<hr style = "border-top: 5px solid DarkGray" />
+
+## Development of multi-parameter machine learning models of lung function
+
+As outlined above, choice of parameters controlling behavior of modeling algorithms such as number of random trees for the random forest or cost penalty for SVM were found by tuning with 10-repeats 10-fold cross-validation [@Kuhn2008] (__Table \@ref(tab:tab-tuning)__).
+Representative tuning results for insufficiency in DLCO and DLCO values expressed as percentages of the reference are presented in __Figures \@ref(fig:fig-dlco-red-tuning)__ - __\@ref(fig:fig-dlco-resamples)__.
 
@@
 Insufficient diffusion capacity for carbon monoxide (DLCO) was defined as values < 80% of the age-, sex-, and weight-specific reference. Probability of insufficient DLCO was modeled as a function of 
-`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time, longitudinally evaluated persistent dyspnea, cough and physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
 The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
@@
 Diffusion capacity for carbon monoxide (DLCO) expressed as percentage of he age-, sex-, and weight-specific reference was modeled as as a function of 
-`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time, longitudinally evaluated dyspnea, cough and physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
 The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
@@
 
-```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm paramaters by cross-validation-mediated tuning.'}
+```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm parameters by cross-validation-mediated tuning.'}
 
@@
 
+Among the investigated LFT abnormalities, solely the insufficient DLCO could be modeled with satisfactory quality as evident from moderate accuracy assessed by Cohen's $\kappa$ inter-rater reliability statistic in cross-validation.
+Notably, performance of DLCO models was basically independent of the machine learning algorithms (out-of-fold predictions $\kappa$: `r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+The cross-validated Brier score values ranged from 
+`r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, indicating that 
+the correct predictions of reduced DLCO were made with probability ranging from 
+`r range(1 - sqrt(filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$brier_score)) %>% signif(2) %>% paste(collapse = ' to ')`. 
+By contrast, reliability of models predicting 
+any CT abnormalities ($\kappa$: `r filter(bin_models$stats$LFT_findings, dataset == 'cv')$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`), 
+reduced FVC ($\kappa$: `r filter(bin_models$stats$FVC_reduced, dataset == 'cv')$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`), and 
+reduced FEV1 ($\kappa$: `r filter(bin_models$stats$FEV1_reduced, dataset == 'cv')$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`) in cross-validation was only poor-to-fair (__Figure \@ref(fig:fig-bin-performance)__, __Table \@ref(tab:tab-bin-classifiers)__).
+
+Next, we had a closer look at predictions made by the models of DLCO insufficiency. 
+In the training data set, observations with DLCO < 80% were detected with excellent sensitivity, specificity and overall accuracy gauged by AUC 
+(training, AUC: `r filter(bin_models$stats$DLCO_reduced, dataset == 'train')$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
+The overall accuracy in receiver operating characteristic for the out-of-fold predictions was moderate-to-good and similar for all machine learning algorithms 
+(cross-validated AUC: `r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
+Remarkably, sensitivity of the models at the cutoff corresponding to 50% predicted probability of insufficient DLCO was quite low (`r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$Se %>% range %>% signif(2) %>% paste(collapse = ' to ')`), while the specificity was good-to-excellent (`r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$Sp %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+This suggests that the models of reduced DLCO may potentially benefit from an additional calibration step or a choice of another probability cutoff for detection of DLCO insufficiency (__Figure \@ref(fig:fig-dlco-red-performance)A__).
+As described before, there was a tendency towards decreasing frequency of DLCO insufficiency at consecutive follow-ups suggestive of a recovery from a COVID-19-mediated structural lung damage, in particular in the severe COVID-19 subset (__Figure \@ref(fig:fig-lft-finding-course)A__, __Table \@ref(tab:tab-bin-classifiers)__).
+Hence, we investigated how the models predict the time course of reduced insufficiency in the CovILD cohort. 
+For out-of-fold predictions, the neural network could emulate the recovery from DLCO findings with the highest reliability (__Figure \@ref(fig:fig-bin-performance)B__).
+
+In summary, among the investigated LFT abnormalities, only insufficient DLCO defined as values < 80% of the reference could be successfully modeled by a rich set of explanatory variables including features of demographic and clinical background, acute COVID-19 course and treatment, follow-up time, respiratory symptoms, physical performance and chest CT measurements in COVID-19 convalescents.
+This may indicate that impaired DLCO rather than low FVC or reduced FEV1 reflects structural lung lesions such as fibrosis caused by SARS-CoV-2 infection.
+
 <hr style = "border-top: 1px solid DarkGray" />
@@
 The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
-The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time, longitudinally evaluated dyspnea, cough and physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
 Parameters controlling the algorithm's behavior were optimized by tuning as presented in Figure \@ref(fig:fig-dlco-red-tuning).
@@
 
-## Performance of machine learning models at prediction of lung function paramaters
+## Performance of machine learning models at prediction of lung function parameters
+
+As expected from the results of modeling of LFT abnormalities, only DLCO expressed as a percentage of the reference value could be modeled with a satisfactory accuracy.  
+In more detail, models of DLCO constructed with the GBM, random forest and SVM algorithms were characterized with moderate-to-good explanatory performance measured in out-of-fold predictions by pseudo-R^2^, i.e.
+a direct measure of variance explained by the model
+(pseudo-R^2^: `r filter(reg_models$stats$DLCO_percent, dataset == 'cv', algorithm %in% c('ranger', 'gbm', 'svmRadial'))$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
+Both in terms of the explained variance and model fit error, the GBM algorithm showed the paramount performance. 
+In turn, the neural network model basically failed at predicting DLCO 
+(cross-validated pseudo-R^2^ = `r filter(reg_models$stats$DLCO_percent, dataset == 'cv', algorithm == 'nnet')$rsq %>% signif(2)`). 
+Independent of the algorithm, models of 
+FVC (pseudo-R^2^: `r filter(reg_models$stats$FVC_percent, dataset == 'cv', algorithm %in% c('ranger', 'gbm', 'svmRadial'))$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and 
+FEV1 (pseudo-R^2^: `r filter(reg_models$stats$FEV1_percent, dataset == 'cv', algorithm %in% c('ranger', 'gbm', 'svmRadial'))$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`) 
+demonstrated virtually no explanatory value in out-of-fold predictions (__Figure \@ref(fig:fig-reg-performance)__, __Table \@ref(tab:tab-reg-models)__).
+
+In a more detailed analysis of the DLCO models, the GBM and random forest models yielded the best correlation between the predicted and observed DLCO values in the cross-validation setting 
+(GBM: $\rho$ = `r filter(reg_models$stats$DLCO_percent, dataset == 'cv', algorithm == 'gbm')$spearman %>% signif(2)`, 
+random forest: $\rho$ = `r filter(reg_models$stats$DLCO_percent, dataset == 'cv', algorithm == 'ranger')$spearman %>% signif(2)`). 
+Still, both models tended to overestimate low DLCO values and underestimate high DLCO values, and may profit from an additional calibration step, e.g.
+by general additive modeling of the predictions [@Hufner2022; @Fasiolo2020] (__Figure \@ref(fig:fig-dlco-performance)__).
+Hence time after COVID-19 diagnosis was an important factor influencing DLCO measurements in the CovILD data set (__Figure \@ref(fig:fig-lft-numeric-course)__), we investigated how the DLCO models recapitulate increasing DLCO with consecutive follow-ups in out-fold-predictions.
+As shown in __Figure \@ref(fig:fig-dlco-model-course)__, the best performing GBM and random forest models could reliably recapitulate DLCO recovery at two, three and six months after COVID-19. 
+However, both algorithms clearly overestimated DLCO values at the one year follow-up (__Figure \@ref(fig:fig-dlco-model-course)__).
+
+Collectively, in line with the results of modeling of LFT abnormalities, we could construct meaningful machine learning models of DLCO but not of FVC and FEV1. 
+This supports our previous notion, that among the investigated LFT parameters DLCO may serve as the best readout of post-COVID-19 lung damage e.g. fibrosis.
 
@@
 The LFT parameters were modeled with the following machine learning algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
-The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time, longitudinally evaluated dyspnea, cough and physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
 Parameters controlling the algorithm's behavior were optimized by tuning as presented in Figure \@ref(fig:fig-dlco-tuning).
@@
 <br>
-Note the overall higher accuracy of the investigated models at prediction of DLCO as compared with the remaining LFT paramaters.
+Note the overall higher accuracy of the investigated models at prediction of DLCO as compared with the remaining LFT parameters.
 </p>
@@
 
+Our modeling results put forward DLCO as the key lung function parameter associated with COVID-19-mediated lung damage. 
+In order to get a more thorough insight into key factors influencing probability of insufficient DLCO and DLCO values in COVID-19 convalescents, we resorted to an analysis of global importance of explanatory variables with the SHAP technique [@Mayer2023a; @Mayer2023; @Lundberg2017; @Covert2020].
+
+In __Figure \@ref(fig:fig-dlco-reduced-global-vim)__ and __Figure \@ref(fig:fig-dlco-regression-global-vim)__ we presented the top 15 most important explanatory variables for prediction of reduced DLCO and DLCO, respectively, by the GBM, random forest, SVM and neural net models.
+Interestingly, although all modeling algorithms returned meaningful models of insufficient DLCO with comparable predictive performance, they differed in the key explanatory variables. 
+In more detail, while the random forest and GBM models relied on AI-determined high lung opacity, smoking intensity (pack-years), AI-determined opacity, radiologist's CTSS and age as the most influential variables, the neural network model's performance was impacted by patient's age, smoking intensity, length of hospitalization, CTSS, weight loss and body mass index to the largest extent.
+In turn, smoking intensity and anti-coagulant therapy during acute COVID-19 were the most important explanatory factors for the SVM model of insufficient DLCO (__Figure \@ref(fig:fig-dlco-reduced-global-vim)__).
+A similar phenomenon, i.e. differing sets of the key explanatory variables for models of similar performance, was observed for DLCO expressed as percentage of the reference value. 
+Whereas hospital stay, high lung opacity computed by AI, body mass index and age were the most influential variables for DLCO predictions made by the GBM and random forest models, the SVM forecasts were most extensively affected by acute COVID-19 severity measured by the WHO ordinal scale of clinical improvement, patient's age class (young adult, middle-aged or elderly), age at COVID-19 diagnosis and body mass index (__Figure \@ref(fig:fig-dlco-regression-global-vim)__).
+
+In an 'univariable' comparison of explanatory variable levels between observations with and without DLCO < 80% of the reference value the largest differences were observed for 
+high lung opacity (`r get_test(dlco_red_uni$test, 'high_opacity_percent')`), 
+lung opacity (`r get_test(dlco_red_uni$test, 'opacity_percent')`), 
+CTSS (`r get_test(dlco_red_uni$test, 'CTSS')`) and 
+and anti-coagulation therapy during acute COVID-19 (`r get_test(dlco_red_uni$test, 'anticoagulant_treatment')`). 
+Those features were significantly higher or significantly more frequent in observations with reduced DLCO, the effect size of the differences was moderate (__Figure \@ref(fig:fig-univariable-dlco-red)__, __Table \@ref(tab:tab-lft-binary-uni)__).
+CTSS (`r get_test(dlco_uni$correlation$test, 'CTSS')`), 
+AI-computed opacity (`r get_test(dlco_uni$correlation$test, 'opacity_percent')`) and 
+high opacity (`r get_test(dlco_uni$correlation$test, 'high_opacity_percent')`), and 
+length of hospital stay (`r get_test(dlco_uni$correlation$test, 'hospital_stay_days')`) were the strongest negative correlates of DLCO values expressed as percentages of the reference in a similar 'univariable' analysis (__Figure \@ref(fig:fig-univariable-dlco-corr)__, __Table \@ref(tab:tab-lft-correlation-uni)__).
+
+The readouts of lung CT abnormality severity, CTSS, opacity and lung opacity, were identified as highly influential explanatory variable for most models of reduced DLCO and DLCO values and were found to be strongly associated with DLCO insufficiency and DLCO measurements in the 'univariable' statistical testing analysis.
+For this reason, we set off to check how impaired DLCO can be predicted by each of CTSS, opacity or high opacity alone by means of optimal cutoff finding with Youden's J criterion, receiver operating characteristic and inter-rater reliability analysis [@Cohen1960; @Lopez-Raton2014; @McHugh2012].
+By maximizing the Youden's J statistic for detection of observations with DLCO < 80% of the reference, the cutoffs of 
+`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'CTSS')$cutoff %>% signif(2)` points for CTSS, 
+`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'opacity_percent')$cutoff %>% signif(2)`% of the lungs for AI-determined opacity, 
+and `r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'high_opacity_percent')$cutoff %>% signif(2)`% of the lungs for AI-computed high opacity were proposed. 
+The accuracy of detection of DLCO insufficiency measured by AUC was moderate and ranged from `r lft_roc$stats$optimal_cutoff$DLCO_reduced$auc %>% range %>% signif(2) %>% paste(collapse = ' to ')`. 
+Sensitivity ranged from `r lft_roc$stats$optimal_cutoff$DLCO_reduced$Se %>% range %>% signif(2) %>% paste(collapse = ' to ')` and specificity was within the `r lft_roc$stats$optimal_cutoff$DLCO_reduced$Sp %>% range %>% signif(2) %>% paste(collapse = ' to ')` range (__Figure \@ref(fig:fig-lft-roc)A__, __Table \@ref(tab:tab-lft-roc)__).
+Yet, the inter-rater reliability between the observed DLCO insufficiency and low DLCO predicted by each of CTSS, lung opacity or high lung opacity was only fair 
+($\kappa$: `r filter(lft_roc$bootstrap$stats, statistic == 'kappa', response == 'DLCO_reduced')$estimate %>% range %>% signif(2) %>% paste(collapse = ' to ')`) (__Figure \@ref(fig:fig-lft-roc)B__, __Table \@ref(tab:tab-lft-interrater)__).
+
+Collectively, CT severity readouts such as CTSS determined by the radiologist, or lung opacity or high lung opacity calculated by AI, are crucial predictors of DLCO insufficiency and DLCO values in COVID-19 patients.
+Even low-grade radiological findings assessed by CTSS, lung opacity and high lung opacity may translate into insufficient DLCO. 
+Yet, clinical applicability of CTSS, opacity and high opacity as single markers of lung function deficits may bear insufficient accuracy.
+
 <hr style = "border-top: 1px solid DarkGray" />
@@
 The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
-The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
-<br>
-Variable importance measures: permutation importance for random forests, input weight importance for neural network, area under the receiver-operating characteristic curve for SVM, and mean tree-split reduction of model deviance for GBM were extracted from the models.
-The variable 0 - 100 scaled importance metrics for the top ten most important explanator variables are presented in bar plots.
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time, longitudinally evaluated physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
+<br>
+Global explanatory variable importance was computed with the Shapley additive explanation (SHAP) method, which compares contribution variable combinations to the overall model performance.
+Mean SHAP statistic values for the top 15 most influential explanatory variables are presented in bar plots.
 </p>
@@
 The following algorithms were employed: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial).
-The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits.
-<br>
-Variable importance measures: permutation importance for random forests, input weight importance for neural network, R^2^ for the LOESS trend between the variable and outcome for SVM, and mean tree-split reduction of model deviance for GBM were extracted from the models.
-The variable 0 - 100 scaled importance metrics for the top ten most important explanatory variables are presented in bar plots.
+The explanatory variables included demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time, longitudinally evaluated dyspnea, cough and physical performance impairment as well as lung computed tomography results obtained at consecutive follow-up visits.
+<br>
+Global explanatory variable importance was computed with the Shapley additive explanation (SHAP) method, which compares contribution variable combinations to the overall model performance.
+Mean SHAP statistic values for the top 15 most influential explanatory variables are presented in bar plots.
 </p>
@@
 <br>
-Differences in single demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time as well as lung computed tomography (CT) parameters between CovILD data set observations with and without insufficient diffusion capacity for carbon monoxide (DLCO < 80 of reference value) were investigated.
+Differences in single demographic and clinical history variables, readouts of course of acute COVID-19, follow-up time, presence and intensity of dyspnea, cough and impaired physical performance as well as lung computed tomography (CT) parameters between CovILD data set observations with and without insufficient diffusion capacity for carbon monoxide (DLCO < 80 of reference value) were investigated.
 Differences in numeric explanatory variables were assessed by Mann-Whitney test with r effect size statistic.
@@
 sensitivity + specificity - 1).
-Results of the optimal opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves.
+<br>
+(A) Results of the optimal opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves.
 The optimal opacity cutoffs are labeled with points and the corresponding CTSS, opacity or high opacity values.
 Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots.
+<br>
+(B) Concordance between the prediction of insufficient DLCO by CTSS, lung opacity and high lung opacity stratified by their optimal cutoffs and the observed DLCO insufficiency was assessed by Cohen's $\kappa$ inter-rater reliability statistic.
+$\kappa$ values with 95% confidence intervals obtained by block bootstrap with B = 2000 iterations are presented in a Forest plot.
+Numbers of complete observations ('total') and observations with DLCO < 80% of the reference ('events') are indicated in the plot caption.
 </p>
@@
   footnote(1, 1, 
-           value = as_paragraph('DLCOP: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second'), 
+           value = as_paragraph('DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second'), 
            part = 'header', 
@@
 
+<hr style = "border-top: 1px solid DarkGray" />
+
+```{r tab-lft-interrater, tab.cap = 'Inter-rater reliability and receiver operator characteristic metrics for detection of lung function testing abnormalities with lung computed tomography severity markers.
+Statistic values are presented with 95% confidence intervals obtained by block bootstrap.'}
+
+flextable::flextable(tables$lft_interrater) %>% 
+  set_widths(c(2.5, 2.5, rep(5.2, 3))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
+           ref_symbols = 'a', 
+           part = 'header') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CT: lung computed tomography; AI: artificial intelligence; CTSS: CT severity score.'),
+           ref_symbols = 'a', 
+           part = 'header') %>% 
+  footnote(1, 3, 
+           value = as_paragraph("Cohen's kappa inter-rater reliability statistic."), 
+           ref_symbols = 'c', 
+           part = 'header') %>% 
+  theme_vanilla
+
+```
+
 <hr style = "border-top: 5px solid #417a8b" />
 
+# Discusion, limitations and conclusions
+
+In the CovILD data set, we could observe an overall good concordance between each of lung CT findings, GGO or reticulations identified by the radiologist and lung opacity determined by AI. 
+The same abnormalities were detected by AI-computed high lung opacity with a moderate reliability. 
+Yet, detection of consolidations, which were rare in the investigated data, proved to be a challenging task for both AI-determined density parameters. 
+Still, consolidations could be detected with quite high sensitivity and specificity by high opacity that rather than opacity. 
+Interestingly, we could observe a clear trend towards a higher concordance between the opacity and findings detected by the radiologist at one year after COVID-19 diagnosis as compared with earlier follow-ups.
+This may be attributed to a training effect of the radiologist or availability of CT scans from previous follow-up visits during evaluation by the radiologist. 
+In turn, quite low concordance between high opacity and human assessment results specifically at the six-month follow up as compared with the remaining follow-ups may suggest a systematic technical problem.
+
+The CTSS > `r ild_cutoff$stats$optimal_cutoff$cutoff` was proposed as a diagnostic criterion of ILD defined as > 5% of the lungs with opacity identified by the software. 
+This cutoff was characterized by an excellent sensitivity, specificity and overall diagnostic accuracy assessed by AUC but only moderate reliability gauged by Cohen's $\kappa$. 
+This paradox may be explained by the fact that ILD was a rare event in the study cohorts and virtually restricted to hospitalized COVID-19 patients and the two- and three-month follow-up. 
+Additionally, we were able to establish relatively simple mathematical relationships between CTSS, and square root of opacity and high opacity computed by AI. 
+A visual analysis of the opacity - CTSS ad high opacity - CTSS trends let us conclude that CTSS is a reliable and quantitative tool at assessment of severity and number of low-grade lesions. 
+Yet it could not rate high-grade or widespread radiological lung findings with a sufficient granularity. 
+This phenomenon may be explained by the fact that CTSS is computed as a sum of pathological scores for all lobes and may be thus unable to discriminate between frequent subtle lesions and localized severe damage.
+
+Our machine learning modeling results show that among the most important LFT parameters, only DLCO can be predicated by a rich set of explanatory variables related to demography, clinical history, acute COVID, time after diagnosis, persistent respiratory symptoms, physical performance rating and lung CT results with a sufficient accuracy.
+Hence, DLCO, but not FVC or FEV1, may be proposed as the LFT readout of COVID-19-mediated lung damage. 
+Its tempting to speculate if and how FVC and FEV1 depend on pre-existing undiagnosed respiratory deficits, or improper bronchial and respiratory muscular function as a consequence of COVID-19. 
+Alternatively, those two LFT readouts may be partly dependent on psychosomatic phenomena, such as persistent somatic symptoms which were also evident in the CovILD cohort [@Hufner2023; @Sahanic2023]. 
+Interestingly, performance of the DLCO models, except for the neural network used to predict numeric DLCO value, was virtually independent from the modeling algorithm. 
+Moreover, equally well performing models of DLCO or insufficient DLCO relied obviously on a differing sets of highly influential explanatory variables. 
+Hence, an ensemble of the machine learning models may improve the prediction quality even further. 
+However, a validation data set would be required for to reliably evaluate accuracy of such approach.
+
+CTSS, AI-computed opacity and high opacity were found to be influential explanatory variables for the majority of machine learning models of DLCO and reduced DLCO. 
+Additionally, those readouts of severity and quantity of structural lung lesions were shown to be significantly associated with presence of DLCO insufficiency and DLCO expressed as a percentage of the reference.
+The subsequent receiver operating characteristic analysis revealed that each of CTSS, opacity and high opacity could detect cases of insufficient DLCO with at least `r signif(min(lft_roc$stats$optimal_cutoff$DLCO_reduced$Se) * 100, 2)`% sensitivity and `r signif(min(lft_roc$stats$optimal_cutoff$DLCO_reduced$Sp) * 100, 2)`% specificity.
+However use of CTSS, opacity or high opacity as single makers of reduced DLCO may bear insufficient reliability, which was only fair in the CovILD data set as assessed by Cohen's $\kappa$ metric. 
+Quite surprisingly, the optimal cutoffs of 
+CTSS (`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'CTSS')$cutoff[1] %>% signif(2)` points), 
+opacity (`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'opacity_percent')$cutoff[1] %>% signif(2)`% of the lungs), 
+and high opacity (`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'high_opacity_percent')$cutoff[1] %>% signif(2)`% of the lungs) were quite low. 
+This suggests that even mild, low-grade or singular radiological lung abnormalities may translate to improper lung function and, as a consequence, interfere with performance and quality of life of the patient.
+
+Our analysis has limitations. 
+First, the size of the cohort was small which likely impacted on quality of the modeling. 
+A larger data set used for model training is expected to improve model accuracy and reduce the fit error. 
+Second, no validation data set was available. 
+A possible external validation would only be possible with a similar study cohort with a similar set of explanatory variable. 
+As discussed by us previously, such study collectives are not known [@Sonnweber2022]. 
+Additionally, due to the small size of the CovILD data set, we abstained from internal validation of the models with a test subset not used for model tuning (so called 'holdout' strategy). 
+Hence, model performance estimates obtained with out-of-fold predictions are expected to suffer from some bias due to the fact that the models were tuned and concomitantly evaluated with the same cross-validation setting.
+Third, multiple data points were obtained from the same participant, and the observations were hence not independent. 
+While this problem was addressed in the inter-rater reliability analysis and modeling by resorting to block bootstrap and participant-wise cross-validation design, respectively, it may have affected results of statistical hypothesis testing, e.g.
+by increase of type II error.
+
+<hr style = "border-top: 5px solid #417a8b" />
+
 # Methods
 
+## Ethics
+
+The CovILD study was conducted in accordance with the Declaration of Helsinki and the European Data policy. 
+All participants gave written informed consent to participate and to process their data. 
+The study data were processed, stored and analyzed in anonymized form. 
+The study protocol was approved by the ethics committee of the Medical University of Innsbruck, Austria (approval number: 1103/2020). 
+The study was registered at [ClinicalTrials.gov](https://classic.clinicaltrials.gov/ct2/show/NCT04416100) (NCT04416100).
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Study procedures, data sources, analysis inclusion and handling of missing values
+
+The longitudinal observation CovILD study aimed at investigation of symptom, cardiopulmonary and mental health recovery of COVID-19 patients. 
+Initially, n = `r length(covild$common_ids)` convalescent COVID-19 survivors were recruited among patients of the University Hospital of Innsbruck, St.
+Vinzenz Hospital in Zams and Karl-Landsteiner Rehabilitation Center in Münster (all in Austria) between March and June 2020.
+The inclusion criteria were SARS-CoV-2 positivity confirmed by PCR and presence of COVID-19 symptoms. 
+All participants were infected with the wild-type form of SARS-CoV-2. 
+The study visits were scheduled at two, three, six and twelve months after COVID-19 diagnosis. 
+Information on demography (age, sex), smoking status, routine medication, pre-existing conditions, as well as the course and treatment of acute COVID-19 were obtained in an interview and retrieved from electronic patient's record at the two-month follow-up visit.
+The standard follow-up visit protocol included a survey of symptoms and self-reported physical performance, examination and interview by a physician, determination of standard blood markers (hemoglobin, iron turnover, complete blood count, markers of inflammation and vascular pathology), trans-thoracic echocardiography, lung function testing (LFT), and computed tomography of the chest (CT).
+The details of the study design and procedures are provided in our recent publications [@Luger2022; @Sonnweber2020; @Sahanic2023]. 
+The complete list of study variables with their format and descriptions is provided in __Table \@ref(tab:tab-study-variables)__.
+
+Regarding the severity of acute COVID-19, the study participants were classified as 'ambulatory' (home isolated, WHO ordinal scale for clinical improvement: 1 - 2), 'hospitalized moderate' (hospitalized, without oxygen therapy or oxygen by mask or nasal prongs, WHO: 3 - 4), 'hospitalized severe' (hospitalized, high flow oxygen or mechanical ventilation, WHO: 5 - 7).
+
+Chest CT was done with a 128 slice multi-detector SOMATOM Definition Flash device (Siemens Healthineers, Erlangen, Germany) with a 128 x 0.6 mm collimation and spiral pitch factor of 1.1. 
+Scans were obtained in craniocaudal direction without iodine contrast agent and in low-dose setting (100 kVp tube potential). 
+Artificial intelligence (AI) supported analysis of the lung CT scans was done with the Syngo.via CT Pneumonia Analysis Software (Siemens Healthineers, Erlangen, Germany), which returned percentage of the lungs with opacity and high opacity.
+Evaluation of the CT scans by the radiologists was done in accordance with the Fleischner society guidelines as described before [@Luger2022; @Sonnweber2020]. 
+In brief, the scans were examined for presence of ground glass opacity (GGO), reticulation, consolidations, and bronchiectasis.
+Severity of lung abnormalities was rated by the radiologist with the CT severity score (CTSS) described before.
+Abnormalities were scored separately for each lobe with the following scheme:
+
+* 0: no abnormalities
+
+* 1: minimal, subtle GGO
+
+* 2: mild, several GGO and subtle reticulations
+
+* 3: moderate, multiple GGO, reticulation, small consolidation
+
+* 4: severe, extensive GGO, consolidation, reticulation with distortion
+
+* 5: massive findings, parenchymal destruction
+
+CTSS was calculated as a sum of the scores over all lobes and ranged from 0 to 25 [@Luger2022; @Sonnweber2020].
+
+The following LFT parameters were recorded: diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC), forced expiratory volume in one second (FEV1), total lung capacity (TLC), and FEV1 to FVC ratio (FEV1:FVC).
+These variables were expressed as percentage of the patient's individual age-, sex- and weight-specific reference. 
+Insufficiency of DLCO, FVC, FEV1 or TLC were defined as values below 80% of the reference. 
+Reduced FEV1:FVC was defined as a value below 70% of the reference. 
+Presence of LFT abnormalities ('any LFT abnormalities') was defined as insufficient DLCO, FVC, FEV1, TLC or FEV1:FVC [@Sonnweber2022; @Sonnweber2020; @Sahanic2023].
+
+The following longitudinally recorded symptoms of potential relevance for lung function were evaluated: 
+dyspnea measured by Modified Medical Research Council scale (mMRC, presence of dyspnea defined as mMRC > 0), 
+self-reported cough, and 
+self-rated physical performance measured by Eastern Cooperative Oncology Group (ECOG, impaired physical performance defined as ECOG > 0) [@Sahanic2023]. 
+Missing information on the symptoms or their intensity in the patient's symptom survey were interpreted as absence of the complaint.
+
+The source study dataset was fetched from the [CovILD study data repository](https://github.com/PiotrTymoszuk/covILD) available as an R package to authorized users. 
+The analysis inclusion criteria were availability of both LFT and CT results. 
+Follow-up observations with missing CT or LFT data were excluded. 
+The CovILD data set analyzed here consisted of n = `r nrow(lft_globals$analysis_tbl)` observations obtained from 
+`r length(unique(lft_globals$analysis_tbl$ID))` participants. 
+For the patients included in the analysis, the set of explanatory variables concerning demographic and clinical background as well as the course and treatment of acute COVID-19 was complete.
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Software
+
+The analysis was done with R version 4.2.3. 
+Tabular data and code pipelines were handled by tools provided by the packages _tidyverse_ [@Wickham2016], _rlang_ [@Henry2022]
+and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Text data were handled with _stringi_ [@Gagolewski2021]. 
+Parallelization was accomplished with _furrr_ [@Vaughan2022] and _doParallel_ [@Folashade2022].
+
+For explorative data analysis, statistical hypothesis testing and analysis of correlations, the packages _rstatix_ [@Kassambara2021], _rcompanion_ [@Mangiafico2022] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) were employed.
+Receiver-operating characteristic (ROC), confusion matrix analysis and inter-rater reliability analysis was done with _caret_ [@Kuhn2008], _OptimalCutpoints_ [@Lopez-Raton2014] and [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).
+For modeling of the CTSS - AI-determined opacity and CTSS - AI-determined high opacity relationships, the packages [_lmqc_](https://github.com/PiotrTymoszuk/lmqc) and _MASS_ were used [@Ripley2022].
+
+For machine learning modeling of lung function testing outcomes, the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) providing wrappers around the R's implementations of the random forest algorithm (_ranger_) [@Wright2017; @Breiman2001],
+neural network (_nnet_) [@Ripley2014], 
+support vector machines (SVM, _kernlab_) [@Weston1998; @Karatzoglou2004], 
+and gradient boosted machines (GBM, _gbm_) [@Greenwell2022; @Friedman2002; @Friedman2001]. 
+Model diagnostic and performance metrics were computed with the development package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra). 
+Mean SHAP (Shapley additive explanations) as a metric of global variance importance in machine learning models was calculated with the package _kernelshap_ [@Covert2020; @Lundberg2017; @Mayer2023].
+
+Analysis results were visualized with _ggplot_ [@Wickham2016] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (scatter, stack and box plots), [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) (plots for machine learning model diagnostic), _plotROC_ [@Sachs2017] (ROC curves) and _shapviz_ [@Mayer2023a; @Mayer2023] (bar plots of mean SHAP values).
+Result tables were created with _flextable_ [@Gohel2022]. 
+Report figures were generated with _cowplot_ [@Wilke2019]. 
+The analysis report was written in the _rmarkdown_ environment [@Allaire2022] with the _bookdown_ package [@Xie2016]. 
+Figures, tables, links and R expressions in the Rmarkdown document were managed with [_figur_](https://github.com/PiotrTymoszuk/figur). 
+The HTML presentation with the analysis report was rendered with _bookdown_ [@Xie2016], _knitr_ [@Xie2022] and _rmdformats_ [@Barnier2022].
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Descriptive statistic, statistical tests and effect sizes
+
+If not stated otherwise, numeric values are presented as medians with interquartile ranges and ranges within the analysis data set or analysis strata. 
+Categorical variables are presented as percentages and counts within the analysis data set or analysis strata. 
+Normality of distribution of numeric variables upon identity, logarithm and square root transformation was checked with Shapiro-Wilk test (function `shapiro_test`, package _rstatix_). 
+Since multiple numeric variables were not normally distributed, Mann-Whitney test or Kruskal-Wallis test were used for comparison of numeric variables between analysis groups. 
+Differences in distribution of categorical variables between the study groups were assessed with $\chi^2$ test. 
+Correlations were investigated with Spearman's rank test. 
+P values were adjusted for multiple testing with the false discovery rate method separately for each analysis task (e.g.
+comparison of explanatory variables between observations with and without LFT findings) [@Benjamini1995].
+Effects with p < 0.05 were considered significant.
+
+The following effect size metrics were used to assess differences between analysis groups and in performance of machine learning models [@McHugh2012; @Cohen2013; @Cohen1960]:
+
+* r was used for two-group comparisons of numeric variables: < 0.3: weak, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size
+
+* $\eta^2$ was used for multi-group comparison of numeric variables: < 0.13: weak, 0.13 - 0.26: moderate, $\geq$ 0.26: large effect size
+
+* Cramer's V was used for comparisons of categorical variables: < 0.3: weak, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size
+
+* Spearman's $\rho$ coefficient of correlation: < 0.3: weak, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size
+
+* pseudo-R^2^ was used as a measure of explained variance of a model: < 0.13: weak, 0.13 - 0.26: moderate, $\geq$ 0.26: large effect size
+
+* Cohen's $\kappa$ inter-rater reliability statistic: < 0.20: no effect, 0.2 - 0.4: fair, 0.4 - 0.6: moderate, 0.6 - 0.8: good, $\geq$ 0.8: excellent concordance
+
+Statistical significance and effect sizes for comparisons and correlations was assessed with the functions `compare_variables()` and `correlate_variables()` (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Cutoff finding
+
+Optimal cutoffs were found for the following study variables: 
+(1) the AI-determined lung opacity and high opacity corresponding to lung CT findings found by the radiologist, 
+(2) the human-determined CTSS corresponding to interstitial lung disease defined as > 5% of AI-measured lung opacity, 
+(3) the AI-determined lung opacity, high opacity, and human - estimated CTSS for detection of any LFT abnormalities, DLCO < 80% of the reference, FVC < 80% of the reference and FEV1 < 80% of the reference.
+The cutoffs were computed based on the maximum of the Youden's J statistic ($J = Sensitivity + Specificity - 1$) in the entire analysis data set (function `optimal.cutpoints()`, package _OptimalCutpoints_) [@Lopez-Raton2014].
+The cutoff finding results are presented in __Tables \@ref(tab:tab-ai-cutoffs)__, __\@ref(tab:tab-ild-cutoff)__ and __\@ref(tab:tab-lft-roc)__.
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Inter-rater reliability and receiver-operating characteristic analysis
+
+Lung opacity, high opacity and CTSS were stratified by the optimal cutoffs for detection of radiologist-observed radiological lung findings, ILD or lung function abnormalities as described above. 
+Concordance between these stratified marker variables and the observed outcomes of interest was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@McHugh2012; @Cohen1960] as well as sensitivity, specificity and area under the curve metrics in the entire dataset and its subsets defined by acute COVID-19 severity and follow-up visits.
+The statistics were computed with the function `multiClassSummary()` from the _caret_ package [@Kuhn2008]. 
+Boundaries of the statistic's 95% confidence intervals were obtained by block bootstrap with 2000 iterations and were defined by the 2.5 and 97.5 percentiles of the bootstrap estimates. 
+Because of non-independently distributed, participant-matched data points, use of canonical bootstrap was not appropriate and, possibly, compromised by an excessive error. 
+To overcome this problem, we resorted to block bootstrap for determination of confidence intervals. 
+Briefly, the block bootstrap scheme involved re-sampling with repetition of the participants instead of single observations.
+Bootstrapping was accomplished by a functional interface implemented in the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).
+Bootstrapped inter-rater reliability, sensitivity and specificity metrics for AI-determined CT opacity, high opacity and CTSS at detection of CT abnormalities, ILD and LFT findings are presented in __Tables \@ref(tab:tab-ai-cutoffs-severity)__, __\@ref(tab:tab-ai-cutoffs-fup)__, __\@ref(tab:tab-ild-cutoff-severity)__, __\@ref(tab:tab-ild-cutoff-fup)__, and __\@ref(tab:tab-lft-interrater)__.
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Modeling of opacity and high opacity with computed tomography severity score
+
+In order to establish a mathematical relationship between the AI-determined lung opacity or high lung opacity and CTSS, three types of models were built:
+
+* ordinary least squares linear models of square root opacity or square root high opacity as a function of fourth order polynomial of CTSS (general formula: $sqrt(AImeasure) \sim CTSS + CTSS^2 + CTSS^3 + CTSS^4$, function `lm()`, package _stats_)
+
+* Huber's robust linear models of square root opacity or square root high opacity as a function of fourth order polynomial of CTSS (general formula: $sqrt(AImeasure) \sim CTSS + CTSS^2 + CTSS^3 + CTSS^4$) (general formula: $AImeasure \sim CTSS + CTSS^2 + CTSS^3 + CTSS^4$, function `rlm()`, package _MASS_) [@Huber2011; @Ripley2022]
+
+* generalized additive models of square root opacity or square root high opacity as a function of linear CTSS term and 8-knot thin plate spline of CTSS (general formula: $sqrt(AImeasure) \sim CTSS + s(CTSS, bs = "tp", k = 8)$, function `gam()`, package _mgcv_) [@Wood2023; @Wood2017]
+
+The models were constructed using the `make_lm()` wrapper provided by the [_lmqc_](https://github.com/PiotrTymoszuk/lmqc) package. 
+The model diagnostic involved visual analysis of plots of model residuals (residuals vs fitted, quantile-quantile plots, method `plot()`, package [_lmqc_](https://github.com/PiotrTymoszuk/lmqc)). 
+Performance of the models was investigated by median absolute error as a readout of the model fit and pseudo-R^2^ as a measure of variance explained by the model.
+
+<hr style = "border-top: 1px solid DarkGray" />
+
+## Multi-parameter modeling of lung function
+
+Presence of any LFT abnormalities, reduced DLCO, FVC and FEV1, as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables including:
+
+* baseline demographic and clinical characteristic (e.g. age, sex, smoking, comorbidities)
+
+* explanatory variables referring to the course of acute COVID-19 (e.g. WHO ordinal scale for clinical improvement, hospitalization status and length, treatment)
+
+* follow-up after COVID-19 diagnosis (two, three, six and twelve months)
+
+* presence and rating of longitudinally recorded symptoms of relevance for lung function (dyspnea, cough, impaired physical performance)
+
+* presence and severity of lung CT findings rated by the radiologist (e.g. GGO, consolidations, CTSS) and AI (opacity and high opacity)
+
+To reduce the number of explanatory variables, some qualitative variables with rare categories were re-coded. 
+This concerned:
+
+* _`r  exchange('pulmonary_comorbidity', lft_globals$lexicon)`_ which subsumed pre-existing chronic obstructive lung disease (COPD), bronchial asthma, interstitial lung disease, chronic lung diseases and other pulmonary conditions
+
+* _`r exchange('anticoagulant_treatment', lft_globals$lexicon)`_ which subsumed anti-coagulation and anti-platelet treatment during acute COVID-19
+
+* _`r exchange('antiinfective_treatment', lft_globals$lexicon)`_ which subsumed anti-infective and anti-macrolide treatment during acute COVID-19
+
+* _`r exchange('smoking', lft_globals$lexicon)`_, where ex- and active smokers were grouped together
+
+Severity of acute COVID-19 measured by the WHO ordinal scale of clinical improvement was re-coded as categorical explanatory variable. 
+The response and explanatory variables are listed in __Table \@ref(tab:tab-study-variables)__.
+
+Multi-parameter models were constructed with four machine learning algorithms employing diverse mathematical principles: 
+canonical random forest (implemented in R by the _ranger_ package, caret's method name: 'ranger') [@Wright2017; @Breiman2001], 
+GBM (_gbm_ package, caret's method name: 'gbm') [@Greenwell2022; @Friedman2002; @Friedman2001], 
+neural network with a single hidden layer (_nnet_ package, caret's method name: 'nnet') [@Ripley2014], 
+and SVM with radial kernel (_kernlab_ package, caret's method name: 'svmRadial') [@Weston1998; @Karatzoglou2004]. 
+For random forest models, 1000 random trees per model were generated. 
+Bernoulli and Gaussian loss functions were implemented in classification and regression GBM models, respectively. 
+Optimal values of parameters controlling the algorithm's behavior such as number of observations in terminal nodes of tree models in the random forest models or cost penalty in the SVM models were found by maximizing the Youden's J statistic
+($J = Sensitivity + Specificity - 1$, classification models of binary LFT outcomes) or by minimizing mean absolute error (MAE, regression models of numeric LFT outcomes) in 10-repeats 10-fold cross-validation.
+To account for the participant matching of the observations, a cross-validation workflow was designed to keep all observations obtained from a particular participant either in the training or in the test portion of the cross-validation split.
+Construction of such 'participant-wise' block cross-validation folds was done with an in-house-developed script employing the `createMultiFolds()` function (package _caret_). 
+For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed [@Kuhn2008]. 
+The optimal values of the tuning parameters are listed in __Table \@ref(tab:tab-tuning)__.
+
+Performance of the machine learning models was evaluated by comparison of the predictions in the genuine training dataset and in 10-repeats 10-fold cross-validation with the observed outcomes. 
+Performance of the classification models was assessed by Cohen's $\kappa$ as a measure of concordance between the predicted and observed outcome [@McHugh2012; @Cohen1960], 
+standard receiver operating characteristic metrics (AUC, overall accuracy, sensitivity, specificity, Youden's J) to assess the ability of the model to detect the pathological outcome [@Kuhn2008], and with
+Brier score to investigate credibility of the predictions [@Brier1950; @Goldstein-Greenwood2021] (__Table \@ref(tab:tab-bin-classifiers)__). 
+Brier score was computed with the following formula:
+
+$$BS = \frac{1}{N}\sum_{i = 1}^N (o_i - p_i)^2$$
+where $N$ is the total observation number, $i$ is an observation index, $p_i$ is the probability of the pathological outcome predicted by the model for the i-th observation, and $o_i$ is the observed outcome for the i-th observation coded with 1 for presence and 0 for absence of the pathology.
+Performance of the regression models was assessed with 
+mean absolute error as a measure of fit quality, 
+pseudo-R^2^ as a metric of explained variance and 
+Spearman's $\rho$ for correlation between the predicted and observed outcome [@Kuhn2008] (__Table \@ref(tab:tab-reg-models)__). 
+Pseudo-R^2^ was computed with the following formulas:
+
+$$pseudo R^2 = 1 - \frac{MSE(y)}{Var(y)}$$
+$$MSE(y) = \frac{1}{N} \sum_i^N (y_i - \hat{y}_i)^2$$$
+where $MSE(y)$ stands for mean squared error of the outcome variable $y$, $Var(y)$ is the variance of the outcome variable $y$, $N$ is the total observation number, $i$ represents the observation index, $y_i$ is the observed outcome value for the i-th observation, and $\hat{y}_i$ is the predicted outcome value for the i-th observation.
+Performance statistics were retrieved from the `caret` models with the `summary()` method from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package. 
+Visualization of the model tuning and evaluation results was accomplished with the method `plot()` from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package and in-house developed functions.
+
+Shapley additive explanations (SHAP) were used as a global importance statistic for the model's explanatory variables. 
+In brief, SHAP measures contribution of the given variable to the model fit by comparing a goodness-of-fit statistic between models re-fitted with subsets of explanatory variables including the variable of interest and models re-fitted with subsets of explanatory variables without the variable of interest [@Covert2020; @Lundberg2017].
+Matrices of SHAP values for observations and variables were computed for the machine learning models of lung function outcomes with the function `kernelshap()` provided by the _kernelshap_ package [@Mayer2023a].
+As a background for SHAP calculation, a single observation data frame was used with numeric explanatory variables set to the 25^th^ percentile of the values in the entire data set and with qualitative explanatory variables set to the baseline category [@Mayer2023a].
+Visualization of the mean SHAP values for the top most influential explanatory variables was accomplished with the `shapviz()` and `sv_importance()` functions (package _shapviz_) [@Mayer2023].
+
 <hr style = "border-top: 5px solid #417a8b" />
@@
 The entire R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/CovILD_AI). 
+The [CovILD data set](https://github.com/PiotrTymoszuk/covILD) is available to the study team and collaborators as an R package and a bunch of Excel datasheets.
 

### 2024-03-22 LFT scripts/bootstrap_tests.R
Analyses for the manuscript
Correspondence and correlation analysis. Statistical hypothesis testing with blocked boostrap tests. Analysis of model residuals for CoV severity strata and follow-ups.

@@ -1,0 +1,231 @@
+# Bootstrap tests for the following problems:
+#
+# 1) Comparison of CT severity variables (CTSS, opacity, high opacity) between 
+# patients with and without LFT abnormalities
+#
+# 2) Correlation of the CT severity variables with LFT parameters
+#
+# Those problems are addressed by blocked bootstrap tests for difference in 
+# medians between the observations with and without the LFT abnormality 
+# and by bootstrapped Spearman's correlation
+
+  insert_head()
+  
+# container ------
+  
+  lft_uniboot <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+
+  plan('multisession')
+    
+# analysis data -------
+  
+  insert_msg('Analysis data')
+  
+  ## LFT variables
+  
+  lft_uniboot$lft_binary <- 
+    c('LFT_findings', 'DLCO_reduced', 'FVC_reduced', 'FEV1_reduced')
+  
+  lft_uniboot$lft_numeric <- 
+    c('DLCO_percent', 'FVC_percent', 'FEV1_percent')
+  
+  ## CT variables
+  
+  lft_uniboot$ct_variables <- 
+    c('CTSS', 'opacity_percent', 'high_opacity_percent')
+  
+  ## analysis table
+  
+  lft_uniboot$analysis_tbl <- lft_globals$analysis_tbl %>% 
+    select(ID, 
+           all_of(lft_uniboot$lft_binary), 
+           all_of(lft_uniboot$lft_numeric), 
+           all_of(lft_uniboot$ct_variables)) %>% 
+    as_tibble
+  
+# Descriptive stats for the binary LFT variables ------
+  
+  insert_msg('Descriptive stats')
+  
+  lft_uniboot$stats <- lft_uniboot$lft_binary %>% 
+    set_names(lft_uniboot$lft_binary) %>% 
+    map(~explore(lft_uniboot$analysis_tbl, 
+                 variables = lft_uniboot$ct_variables, 
+                 split_factor = .x, 
+                 what = 'table', 
+                 pub_styled = TRUE)) %>% 
+    map(reduce, left_join, by = 'variable') %>% 
+    map(set_names, c('variable', 'no', 'yes')) %>% 
+    compress(names_to = 'lft_abnormality')
+  
+# Bootstrap tests for the binary LFT variables -------
+  
+  insert_msg('Bootstrap tests')
+  
[…]

### 2024-03-22 LFT scripts/residuals.R

@@ -1,0 +1,262 @@
+# Analysis of fit (residuals and Brier's squares) of the machine learning models
+# in the severity strata and at the study visits. Included aare only successful
+# models
+
+  insert_head()
+  
+# container -------
+  
+  dlco_resid <- list()
+  
+# analysis data -------
+  
+  insert_msg('Analysis data')
+  
+  ## Brier's squares for the classifiers
+  
+  dlco_resid$data$DLCO_reduced <- 
+    bin_models$predictions$cv$DLCO_reduced %>% 
+    map(squared) %>% 
+    map(transmute, 
+        .observation = .observation, 
+        error = square_dist)
+  
+  ## error rate for the classifiers
+  
+  dlco_resid$data$DLCO_accuracy <- 
+    bin_models$predictions$cv$DLCO_reduced %>% 
+    map(model.frame) %>% 
+    map(transmute, 
+        .observation = .observation, 
+        error = as.numeric(.outcome != .fitted))
+
+  ## absolute residuals for the regression models
+  
+  dlco_resid$data$DLCO_percent <- 
+    reg_models$predictions$cv$DLCO_percent %>% 
+    map(residuals) %>% 
+    map(transmute, 
+        .observation = .observation, 
+        error = abs(.resid))
+  
+  ## residuals for the regression models
+  
+  dlco_resid$data$DLCO_residuals <- 
+    reg_models$predictions$cv$DLCO_percent %>% 
+    map(residuals) %>% 
+    map(transmute, 
+        .observation = .observation, 
+        error = .resid)
+  
+  ## collapsing of the resamples
+  
+  dlco_resid$data <- dlco_resid$data %>% 
+    map(map, 
+        group_by, 
+        .observation) %>% 
+    map(map, 
+        summarise, 
+        error = mean(error, na.rm = TRUE))
+  
+  ## appending with the visit and severity information
+  
+  dlco_resid$assignment <- lft_globals$analysis_tbl %>% 
+    transmute(.observation = 1:nrow(.), 
+              follow_up = follow_up, 
+              severity_class = severity_class) %>% 
+    as_tibble
+  
+  dlco_resid$data <- dlco_resid$data %>% 
+    map(map, 
+        ~left_join(dlco_resid$assignment, 
+                   .x, 
+                   by = '.observation'))
+  
+# Statistics for the strata --------
+  
[…]

### 2024-03-22 LFT scripts/roc_ct.R

@@ -120,28 +120,6 @@
     
   }
   
-# ROC plots ------
-  
-  insert_msg('ROC plots')
-  
-  for(i in names(lft_roc$stats$all_cutoffs)) {
-    
-    lft_roc$roc_plots[[i]] <- 
-      list(cutoff_stats = blast(lft_roc$stats$optimal_cutoff[[i]], variable), 
-           plot_title = names(lft_roc$stats$all_cutoffs[[i]]) %>% 
-             exchange(lft_globals$lexicon) %>% 
-             capitalize_first_char %>% 
-             paste(exchange(i, lft_globals$lexicon), sep = ' and '), 
-           plot_subtitle = lft_roc$n_caps[[i]], 
-           m_variable = as.character(lft_roc$stats$optimal_cutoff[[i]]$variable), 
-           color = lft_roc$variable_colors) %>% 
-      pmap(plot_roc,
-           data = lft_roc$analysis_tbl %>% 
-             map_dfc(function(x) if(is.factor(x)) as.numeric(x) - 1 else x), 
-           d_variable = i)
-    
-  }
-  
 # Summary ROC plots for each LFT abnormality ------
   
   insert_msg('Summary ROC plots')
@@ -286,10 +264,56 @@
            marker = exchange(marker, 
                              lft_globals$lexicon), 
            response = exchange(response, 
-                              lft_globals$lexicon)) %>% 
+                               lft_globals$lexicon)) %>% 
     select(response, marker, statistic, table_lab) %>% 
     pivot_wider(names_from = 'statistic', 
-                values_from = 'table_lab')
+                values_from = 'table_lab') %>% 
+    set_names(c('LFT abnormality', 
+                'CT variable', 
+                'Kappa', 
+                'Sensitivity', 
+                'Specificity'))
+  
+  ## A single summary table
+  
+  lft_roc$summary_result_tbl <- 
+    lft_roc[c("result_tbl", "bootstrap_result_tbl")] %>% 
+    map2(., 
+         list(c('Cutoff', 'AUC'), 
+              c('Kappa', 'Sensitivity', 'Specificity')), 
+         ~select(.x, 
+                 `LFT abnormality`, 
+                 `CT variable`, 
+                 all_of(.y))) %>% 
+    reduce(left_join, by = c('LFT abnormality', 'CT variable')) %>% 
[…]

### 2024-03-22 LFT scripts/shap_importance.R

@@ -15,7 +15,11 @@
   shap_imp$lexicon <- lft_globals$lexicon %>% 
     mutate(label = ifelse(!is.na(unit), 
                           paste(label, unit, sep = ', '), 
-                          label))
+                          label), 
+           label = ifelse(variable %in% lft_globals$ct_variables & 
+                            variable != 'follow_up', 
+                          paste0('<b>', label, '</b>'), 
+                          label))  
   
 # SHAP objects ------
   
@@ -25,7 +29,7 @@
   
   shap_imp$shap_viz_objects <- shap_imp$shap_objects %>% 
     map(map, shapviz)
-  
+
 # Bar plots with the mean importance measures -------  
   
   insert_msg('Bar plots')
@@ -41,10 +45,8 @@
                                          kind = 'bar', 
                                          fill = z, 
                                          color = 'black') + 
-           scale_y_discrete(labels = function(x) exchange(x, dict = shap_imp$lexicon)) + 
            labs(title = y,
-                subtitle = 'Importance for probability of DLCO < 80%') + 
-           globals$common_theme)
+                subtitle = 'Importance for probability of DLCO < 80%'))
   
   ## regression
   
@@ -57,10 +59,18 @@
                                          kind = 'bar', 
                                          fill = z, 
                                          color = 'black') + 
-           scale_y_discrete(labels = function(x) exchange(x, dict = shap_imp$lexicon)) + 
            labs(title = y,
-                subtitle = 'Importance for DLCO, % of reference') + 
-           globals$common_theme)
+                subtitle = 'Importance for DLCO, % of reference'))
+  
+  ## common styling
+  
+  shap_imp$bar_plots <- shap_imp$bar_plots %>% 
+    map(map, 
+        ~.x + 
+          scale_y_discrete(labels = function(x) exchange(x, dict = shap_imp$lexicon)) + 
+          globals$common_theme + 
+          theme(axis.text.y = element_markdown()))
+  
   
 # Bee swarm plots ------
   
@@ -80,10 +90,8 @@
                                          kind = 'bee', 
                                          size = 2, 
                                          alpha = 0.5) + 
-           scale_y_discrete(labels = function(x) exchange(x, dict = shap_imp$lexicon)) + 
            labs(title = y,
[…]

### 2024-03-22 exploration scripts/correlation.R

@@ -1,0 +1,98 @@
+# Correlation of numeric LFT readouts and CT readouts. Because the observations 
+# are not independent, we're resorting to a blocked bootstrap Spearman's test.
+
+  insert_head()
+  
+# container -------
+  
+  corr_lft <- list()
+  
+# parallel backend -------
+  
+  insert_msg('Parallel backend')
+  
+  plan('multisession')
+  
+# analysis data ------
+  
+  insert_msg('Analysis data')
+  
+  corr_lft$analysis_tbl$lft <- covild$lft %>% 
+    select(ID, DLCO_percent, FVC_percent, FEV1_percent)
+  
+  corr_lft$analysis_tbl$ct <- covild$ct %>% 
+    select(ID, CTSS, opacity_percent, high_opacity_percent)
+  
+  ## variable lexicon 
+  
+  corr_lft$lexicon <- covild[c("lft_lexicon", "ct_lexicon")] %>% 
+    map_dfr(filter, 
+            variable %in% c('DLCO_percent', 'FVC_percent', 'FEV1_percent', 
+                            'CTSS', 'opacity_percent', 'high_opacity_percent'))
+
+  ## variable pairs and plot titles
+  
+  corr_lft$pairs <- corr_lft$analysis_tbl %>% 
+    map(names) %>% 
+    map(~.x[-1]) %>% 
+    map(combn, 
+        m = 2, 
+        simplify = FALSE)
+
+  corr_lft$titles <- corr_lft$pairs %>% 
+    map(map, exchange, corr_lft$lexicon) %>% 
+    map(map, paste, collapse = ' and ')
+  
+# Correlation coefficients ---------
+  
+  insert_msg('Correlation coefficients')
+  
+  for(i in names(corr_lft$pairs)) {
+    
+    corr_lft$test[[i]] <- corr_lft$pairs[[i]] %>% 
+      map_dfr(spearman_rho, 
+              data = corr_lft$analysis_tbl[[i]], 
+              B = 2000, 
+              positive = TRUE) %>% 
+      re_adjust %>% 
+      mutate(est_lab = paste0('\u03C1 = ', signif(rho, 2), 
+                              ' [', signif(lower_ci, 2), 
+                              ' - ', signif(upper_ci, 2), ']'), 
+             n = nrow(corr_lft$analysis_tbl[[i]]), 
+             plot_cap = paste(est_lab, significance, sep = ', '), 
[…]

### 2024-03-22 lft.R

@@ -20,6 +20,7 @@
   library(caretExtra)
   library(effectsize)
   library(sciplot)
+  library(rcompanion)
 
   library(ranger)
   library(nnet)
@@ -33,10 +34,11 @@
   library(furrr)
   library(doParallel)
 
+  library(ggrepel)
+  library(ggtext)
+
   library(OptimalCutpoints)
   library(plotROC)
-
-  library(ggrepel)
 
   explore <- exda::explore
   train <- caret::train
@@ -65,7 +67,8 @@
     './LFT scripts/univariable_FVC_findings.R', 
     './LFT scripts/univariable_FVC.R',
     './LFT scripts/univariable_FEV1_findings.R', 
-    './LFT scripts/univariable_FEV1.R') %>% 
+    './LFT scripts/univariable_FEV1.R', 
+    './LFT scripts/bootstrap_tests.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
 # Tuning of the machine learining models --------
@@ -86,7 +89,7 @@
                    'Loading cached tuning results for DLCO findings', 
                    'Loading cached tuning results for FVC findings', 
                    'Loading cached tuning results for FEV1 findings')) %>% 
-    pwalk(access_cache)
+    pwalk(soucer::access_cache)
   
   ## for regression
   
@@ -99,14 +102,16 @@
        message = c('Loading cached tuning results for DLCO percentage',
                    'Loading cached tuning results for FVC percentage', 
                    'Loading cached tuning results for FEV1 percentage')) %>% 
-    pwalk(access_cache)
+    pwalk(soucer::access_cache)
   
 # Assessment of the model performance ------
   
   insert_msg('Assessment of the model performance')
   
   c('./LFT scripts/binary_performance.R', 
-    './LFT scripts/regression_performance.R') %>% 
+    './LFT scripts/regression_performance.R', 
+    './LFT scripts/residuals.R', 
+    './LFT scripts/ct_predictions.R') %>% 
     source_all(message = TRUE, crash = TRUE)
 
 # Variable importance ------
@@ -127,7 +132,6 @@
   
   c('./LFT scripts/roc_ct.R') %>% 
     source_all(message = TRUE, crash = TRUE)
-  
 
 # END -----
   

### 2024-03-22 paper scripts/supplementary_tables.R

@@ -1,0 +1,284 @@
+# Supplementary tables
+
+  insert_head()
+  
+# container ------
+  
+  suppl_tabs <- list()
+  
+# Table S1: numbers of samples and participants per time point -------
+  
+  insert_msg('Table s1: numbers of samples and participants')
+  
+  suppl_tabs$n_numbers <- cohort_ct$analysis_tbl %>% 
+    blast(cohort_split) %>% 
+    map(count, follow_up) %>% 
+    reduce(left_join, by = 'follow_up') %>% 
+    mutate(follow_up = stri_extract(follow_up, regex = '\\d{1}-month')) %>% 
+    set_names(c('Follow-up, months', 
+                'Cohort', 
+                'Ambulatory, mild COVID-19', 
+                'Hospitalized, moderate COVID-19', 
+                'Hospitalized, severe COVID-19')) %>%
+    mdtable(label = 'table_s1_observation_n_numbers', 
+            ref_name = 'n_numbers', 
+            caption = paste('Numbers of observations included in the analysis', 
+                            'split by acute COVID-19 severity and the follow-up', 
+                            'examination.'))
+
+# Table S2: study variables --------
+  
+  insert_msg('Table S2: Study variables')
+  
+  suppl_tabs$study_variables <- 
+    list(lft = tibble(variable = lft_globals$responses), 
+         ct = tibble(variable = lft_globals$ct_variables), 
+         symptoms = tibble(variable = lft_globals$symptom_variables), 
+         baseline = tibble(variable = lft_globals$baseline_variables)) %>% 
+    compress(names_to = 'category') %>% 
+    mutate(modeling = ifelse(category == 'lft', 
+                             'response variable', 
+                             'explanatory variable'), 
+           category = car::recode(category, 
+                                  "'lft' = 'Lung function testing'; 
+                                   'ct' = 'Chest computed tomography'; 
+                                   'symptoms' = 'Persistent symptoms';
+                                   'baseline' = 'Baseline characteristic and acute COVID-19'")) %>% 
+    left_join(lft_globals$lexicon[c('variable', 'label', 'format', 
+                                    'unit', 'description')] %>% 
+                filter(!duplicated(variable)), 
+              by = 'variable') %>% 
+    select(modeling, 
+           category, 
+           variable, 
+           label, 
+           description, 
+           format, 
+           unit) %>% 
+    set_names(c('Use in modeling', 
+                'Category', 
+                'Name in R', 
+                'Name in the report', 
[…]

### 2024-03-22 paper scripts/tables.R

@@ -1,0 +1,122 @@
+# Main paper_tabs
+
+  insert_head()
+  
+# container ------
+  
+  paper_tabs <- list()
+  
+# Table 1: characteristic of the study cohort ------
+  
+  insert_msg('Table 1: characteristic of the study cohort')
+  
+  paper_tabs$cohort <- cohort$result_tbl %>% 
+    filter(variable != 'COVID-19 severity') %>% 
+    set_names(c('Variable', 
+                'Cohort', 
+                'Ambulatory COVID-19', 
+                'Moderate COVID-19', 
+                'Severe COVID-19', 
+                'Significance', 
+                'Effect size')) %>% 
+    mdtable(label = 'table_2_cohort_characteristic', 
+            ref_name = 'cohort', 
+            caption = paste('Baseline characteristics and COVID-19 course', 
+                            'in the CovILD cohort.', 
+                            'Numeric variables are presented as medians with', 
+                            'interquartile ranges (IQR) and ranges.', 
+                            'Categorical variables are', 
+                            'presented as percentages', 
+                            'and counts within the complete observation set.'))
+  
+# Table 2: performance of the machine learning classifiers ------
+  
+  insert_msg('Table 2: performance of machine learning classifiers, CV')
+  
+  paper_tabs$bin_classifiers <- bin_models$stats %>% 
+    compress(names_to = 'response') %>% 
+    filter(dataset == 'cv', 
+           response != 'LFT_findings') %>% 
+    mutate(response = exchange(response, lft_globals$lexicon), 
+           response = factor(response, 
+                             exchange(names(bin_models$stats), 
+                                      lft_globals$lexicon)), 
+           algorithm = globals$algo_labs[algorithm], 
+           algorithm = factor(algorithm, globals$algo_labs)) %>% 
+    select(response, algorithm,  
+           correct_rate, kappa, brier_score, 
+           AUC, Se, Sp) %>% 
+    arrange(response, algorithm) %>% 
+    set_names(c('Response', 'Algorithm', 
+                'Overall accuracy', "\u03BA", 
+                'Brier score', 'AUC', 
+                'Sensitivity', 'Specificity')) %>% 
+    map_dfc(function(x) if(is.numeric(x)) signif(x, 2) else x) %>%
+    mdtable(label = 'table_2_binary_classifier_performance', 
+            ref_name = 'bin_classifiers', 
[…]

### 2024-03-22 paper.R

@@ -1,0 +1,46 @@
+# Parts of the manuscript
+
+# tools -----
+
+  library(tidyverse)
+  library(trafo)
+  library(rlang)
+  library(stringi)
+  
+  library(flextable)
+  library(figur)
+  library(cowplot)
+  
+  library(caretExtra)
+  library(ggtext)
+  
+  library(rmarkdown)
+  library(bookdown)
+  library(knitr)
+  library(writexl)
+
+  library(soucer)
+  
+  insert_head()
+  
+  width <- flextable::width
+  nobs <- stats::nobs
+  
+  c('./tools/globals.R', 
+    './tools/tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# Scripts generating paper parts -------
+  
+  insert_msg('Scripts for paper parts')
+  
+  c('./paper scripts/tables.R', 
+    './paper scripts/supplementary_tables.R', 
+    './paper scripts/figures.R', 
+    './paper scripts/supplementary_figures.R', 
+    './paper scripts/render.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+  
+# END ------
+  
+  insert_tail()

### 2024-03-22 report scripts/tables.R

@@ -266,7 +266,7 @@
                                     replacement = 'Support vector machines, radial kernel'), 
            algorithm = stri_replace(algorithm, 
                                     fixed = 'GBM', 
-                                    replacement = 'Gradiant boosted machines')) %>% 
+                                    replacement = 'Gradient boosted machines')) %>% 
     select(response, algorithm, parameter) %>% 
     set_names(c('Response', 'Algorithm', 'Parameters')) %>% 
     mdtable(label = 'table_12_ml_tuning',

### 2024-03-22 report/markdown/report.Rmd
Report update

@@
 ---
-title: "Artificial inteligence-assisted analysis of CT abnormalities during COVID-19 recovery"
+title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
 subtitle: "Analysis report"

### 2024-03-22 paper/markdown/manuscript.Rmd
Manuscript draft

---
title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Manuscript"
author: "Department of Radiology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Abstract

\newpage

# Introduction

\newpage

# Methods

## Ethics

The CovILD study was conducted in accordance with the Declaration of Helsinki and the European Data policy. 
All participants gave written informed consent to participate and to process their data. 
The study data were processed, stored and analyzed in anonymized form. 
The study protocol was approved by the ethics committee of the Medical University of Innsbruck, Austria (approval number: 1103/2020). 
The study was registered at [ClinicalTrials.gov](https://classic.clinicaltrials.gov/ct2/show/NCT04416100) (NCT04416100).

## Study design and participants

Details on the study design and the cohort are provided in our recent publications [@Sahanic2023; @Sonnweber2020; @Luger2022; @Sonnweber2022]. 
In brief, the longitudinal observation CovILD study aimed at investigation of symptom, cardiopulmonary and mental health recovery in a cohort of `r length(covild$common_ids)` COVID-19 survivors at two, three, six, and twelve months after COVID-19 diagnosis. 
The participants were recruited between March and June 2020 among patients of three clinical centers in Tyrol, Austria (Medical University of Innsbruck, St. Vinzenz Hospital in Zams, and Karl-Landsteiner Rehabilitation Center in Münster). 
The study inclusion criteria were age $\geq$ 18 years, SARS-CoV-2 positivity confirmed by PCR and presence of COVID-19 symptoms. 
All participants were infected with the wild-type form of SARS-CoV-2. 
Herein, n = `r nrow(covild$ct)` longitudinal observations from `r length(covild$complete_ids)` participants were analyzed, with complete computed tomography (CT) of the chest and lung function testing (LFT) as analysis inclusion criteria (__Figure \@ref(fig:fig-consort)__, __Supplementary Table S1__).

## Procedures

Details of study procedures and variables are provided in __Supplementary Methods__, __Supplementary Table S2__, and in our recent publications [@Sahanic2023; @Sonnweber2020; @Luger2022; @Sonnweber2022]. 
In brief, baseline demographic and clinical data, and data concerning acute COVID-19 course were gathered in an interview and retrieved from electronic patient's record at the two-month follow-up visit. 
The standard follow-up visit protocol included a survey of symptoms and self-reported physical performance, examination and interview by a physician, determination of standard blood markers (hemoglobin, iron turnover, complete blood count, markers of inflammation and vascular pathology), trans-thoracic echocardiography, LFT, and CT of the chest.

Acute COVID-19 severity was classified as 
'ambulatory' (home isolated, WHO ordinal scale for clinical improvement: 1 - 2), 
'hospitalized moderate' (hospitalized, without oxygen therapy or oxygen by mask or nasal prongs, WHO: 3 - 4), and 
'hospitalized severe' (hospitalized, high flow oxygen or mechanical ventilation, WHO: 5 - 7). 
Chest CT scans were evaluated by experienced radiologists blinded to the study design according to the Fleischner society guidelines [@Hansell2008] for presence of ground glass opacity (GGO),  reticulation, consolidations, and bronchiectasis. 
Severity of lung lesions was scored separately for each lobe with a 0 - 5 scale and CT severity score (CTSS) was calculated as a sum of the scores for the entire organ, as described before [@Sonnweber2020; @Luger2022]. 
Additionally, percentages of the lung with opacity and high opacity were determined for the chest CT scans with an artificial intelligence (AI) based software (Syngo.via CT Pneumonia Analysis Software, Siemens Healthineers, Erlangen, Germany). 
In the current report, the following LFT parameters were analyzed: 
diffusion capacity for carbon monoxide (DLCO), 
forced vital capacity (FVC), and 
forced expiratory volume in one second (FEV1). 
The LFT variables were expressed as percentages of the patient's reference value. 
Insufficiency was assumed for the LFT parameter value < 80% of reference [@Sahanic2023; @Sonnweber2020; @Sonnweber2022]. 
The following self-reported COVID-19 related symptoms of potential relevance for LFT results were analyzed: 
dyspnea (modified Medical Research Council [mMRC] scale > 0), 
cough (yes/no item), 
impaired physical performance (Eastern Cooperative Oncology Group [ECOG] scale > 0).

## Analysis endpoints

The primary analysis endpoint was construction and evaluation of multi-parameter models of the most common LFT abnormalities (each of DLCO < 80%, FVC < 80%, FEV1 < 80% of the patient's reference) and of numeric LFT readouts (DLCO, FVC, FEV1) during COVID-19 convalescence. 
Such models employed common machine learning algorithms with a wide range of demographic and clinical features, as well as human-determined CT abnormalities, and AI-measured lung opacity and high opacity as explanatory variables. 
The secondary analysis endpoint was prediction of LFT abnormalities and numeric LFT readouts during COVID-19 recovery by human-determined CTSS and AI-measured lung opacity and high opacity. 
This endpoint was addressed by statistical hypothesis testing, correlation and receiver-operating characteristic (ROC) analysis.

## Statistical analysis

Details of statistical analysis are provided in __Supplementary Methods__.

Statistical analysis was performed with R version 4.2.3 (R Foundation for Statistical Computing). 
Numeric variables were presented as medians with interquartile ranges and ranges. 
Qualitative variables were presented as percentages and counts of the categories within the complete observations set. 
Differences in independently distributed numeric variables were analyzed by Mann-Whitney and Kruskal-Wallis test with, respectively, biseral r and $\eta^2$ effect size statistic. 
Statistical significance for differences in distribution of qualitative variables was determined by $\chi^2$ test with Cramer's V effect size statistic. 
Co-occurrence of each of LFT findings, CT abnormalities, and symptoms was investigated by two-dimensional correspondence analysis [@Ripley2022].

Differences in medians of non-independently distributed, participant matched numeric variables between observations with and without LFT abnormalities were assessed by a blocked bootstrap test with blocks defined by the participant's identifier, and effect size measured by biseral r effect size statistic [@Mangiafico2022]. 
Correlations of non-independently distributed CT and LFT readouts were assessed by blocked bootstrap Spearman's rank test. 
Cutoffs of CTSS, opacity and high opacity for detection of LFT abnormalities were found by maximizing the Youden's J statistic. 
ROC analysis statistics (area under the curve [AUC], sensitivity, specificity, Cohen's $\kappa$) for those optimal cutoffs were computed and their 95% confidence intervals were obtained by blocked bootstrap [@McHugh2012; @Lopez-Raton2014].

Presence of reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables. 
The explanatory variables included 
demographic features (e.g. age, sex, body mass index, smoking, comorbidity), 
characteristic of acute COVID-19 (severity, medication) and recovery (e.g. weight loss, symptoms of relevance for lung function, time after diagnosis), 
and presence of human- and AI-rated structural lung abnormalities in CT scans (e.g. GGO, CTSS, opacity and high opacity). 
The modeling responses and explanatory variables are listed in __Supplementary Table S2__. 
The models were constructed with four common machine learning algorithms: 
Random Forest [@Wright2017; @Breiman2001], 
gradient boosted machines (GBM) [@Greenwell2022; @Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], 
and support vector machines (SVM) with radial kernel [@Weston1998; @Karatzoglou2004]. 
Selection of the optimal values of parameters controlling model behavior such as number of random trees, neurons in the hidden layer, or cost parameter was motivated by the maximum of Youden's J statistic (classification models of LFT abnormalities) or minimum mean absolute error (MAE, regression models of LFT readouts) in 10-repeats 10-fold cross-validation [@Kuhn2008]. 
Because of the presence of participant-matched observation, blocked cross-validation design was used both in the model selection and model evaluation, with blocks defined by participant's identifier. 
Model predictions were evaluated both in the training data and blocked 10-repeats 10-fold cross-validation. 
Concordance between the model-predicted and observed outcomes for classification models was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@McHugh2012; @Cohen1960]. 
Accuracy, AUC, specificity and sensitivity of the classification model were investigated by ROC. 
Calibration of the classification models was assessed by Brier scores [@Brier1950]. 
Fraction of explained variance in predictions of the regression models was measured by pseudo-R^2^, the regression model error was expressed as MAE. 
Spearman's $\rho$ coefficients of correlation between the predicted and observed values were used to gauge calibration of the regression models.
Importance of explanatory variables was estimated by absolute values of SHAP statistics (Shapley additive explanations) [@Covert2020; @Lundberg2017; @Mayer2023a].

\newpage

# Results

## Baseline Characteristic of the cohort and course of acute COVID-19

The longitudinal observation CovILD study cohort consisted of n = `r length(covild$common_ids)` convalescents of symptomatic SARS-CoV-2 infection initially enrolled during the fist outbreak of the pandemic (March - June 2020) in three clinical centers in western Austria [@Sahanic2023; @Luger2022; @Sonnweber2022; @Sonnweber2020]. 
Out of them, n = `r length(covild$complete_ids)` participants with n = `r nrow(covild$ct)` observations with complete CT and LFT data obtained at follow-up visits at the two, three, six and twelve months after COVID-19 diagnosis were included in the current analysis. 
Please note, that per participant, up to four observations were available. 
Due to this participant-matching, the observations were not independent, which had consequences for our analysis strategy (__Figure \@ref(fig:fig-consort)__). 
The majority of analyzed data points were obtained at the two- (n = `r covild$ct %>% filter(follow_up == '2-month follow-up') %>% nrow`) and three-month follow-up (n = `r covild$ct %>% filter(follow_up == '3-month follow-up') %>% nrow`, __Supplementary Table S1__ - __S2__).

The analyzed participants were grouped according to the severity of acute COVID-19 as 
convalescents of ambulatory COVID-19 (`r get_percent(cohort$analysis_tbl, 'severity_class')['ambulatory mild']` of the cohort, home-isolated during infection, WHO ordinal scale for clinical improvement: 1 - 2), 
hospitalized moderate COVID-19 (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized moderate']`, without oxygen or with low-flow oxygen, WHO ordinal scale for clinical improvement: 3 - 4), 
and hospitalized severe COVID-19 patients (`r get_percent(cohort$analysis_tbl, 'severity_class')['hospitalized severe']`, high-flow oxygen or mechanical ventilation, WHO ordinal scale for clinical improvement: 5 - 7, __Table \@ref(tab:tab-cohort)__). 
In the analyzed data set, `r get_percent(cohort$analysis_tbl, 'sex')['male']` of participants were male. 
The percentage of males was significantly higher in moderate and severe COVID as compared with ambulatory COVID-19 individuals (`r get_test(cohort$test, 'sex')`, moderate). 
The median age at COVID-19 diagnosis was `r signif(median(cohort$analysis_tbl$age), 2)` years, ambulatory COVID-19 patients were significantly younger than hospitalized COVID-19 survivors (`r get_test(cohort$test, 'age')`, moderate). 
Ex- or current smokers made up to `r sum(get_percent(cohort$analysis_tbl, 'smoking', percent_mark = FALSE)[2:3])`% of the analyzed cohort. 
There were no significant differences in participants with smoking history between the acute COVID-19 severity subsets. 
Yet, the pack-year number was the highest in hospitalized moderate COVID-19 convalescents (`r get_test(cohort$test, 'smoking_pack_years')`, small). 
Based on body mass index (BMI) > 25 kg/m^2^, 
`r sum(get_percent(cohort$analysis_tbl, 'body_mass_class', percent_mark = FALSE)[2:3])`% of participants were overweight or obese; there were no significant differences in percentages of obese or overweight participants between the acute COVID-19 severity subsets. 
Endocrine or metabolic conditions (`r get_percent(cohort$analysis_tbl, 'endocrine_metabolic_comorbidity')['yes']` of analyzed participants), 
cardiovascular diseases (`r get_percent(cohort$analysis_tbl, 'cardiovascular_comorbidity')['yes']`), 
hypertension (`r get_percent(cohort$analysis_tbl, 'hypertension_comorbidity')['yes']`), 
hypercholesterolemia (`r get_percent(cohort$analysis_tbl, 'hypercholesterolemia_comorbidity')['yes']`), 
and type II diabetes (`r get_percent(cohort$analysis_tbl, 'diabetes_comorbidity')['yes']`) 
were the most common comorbidities at COVID-19 diagnosis. 
Frequencies of endocrine or metabolic comorbidities (`r get_test(cohort$test, 'endocrine_metabolic_comorbidity')`, small), 
cardiovascular conditions (`r get_test(cohort$test, 'cardiovascular_comorbidity')`, moderate), 
hypercholesterolemia (`r get_test(cohort$test, 'hypercholesterolemia_comorbidity')`, small), 
and diabetes (`r get_test(cohort$test, 'diabetes_comorbidity')`, small) were significantly more common in hospitalized than in ambulatory COVID-19 patients (__Table \@ref(tab:tab-cohort)__).

The median length of hospital stay for analyzed participants was `r signif(median(cohort$analysis_tbl$hospital_stay_days), 2)` days and was significantly longer for severe that moderate COVID-19 patients (`r get_test(cohort$test, 'hospital_stay_days')`, large). 
Antibiotic anti-infective treatment was applied to `r get_percent(cohort$analysis_tbl, 'antiinfective_treatment')['yes']` of participants, this percentage was the highest in severe COVID-19 patients 
(`r get_test(cohort$test, 'antiinfective_treatment')`, large). 
Macrolides and anti-platelet drugs were administered to 
`r get_percent(cohort$analysis_tbl, 'macrolide_treatment')['yes']` and 
`r get_percent(cohort$analysis_tbl, 'antiplatelet_treatment')['yes']` of participants, respectively. 
Frequency of this treatment was significantly higher in hospitalized than in ambulatory patients 
(macrolides: `r get_test(cohort$test, 'macrolide_treatment')`, small; 
anti-platelet: `r get_test(cohort$test, 'antiplatelet_treatment')`, small). 
Systematic steroid administration was not considered as a treatment of moderate or severe COVID-19 during the first European outbreak of SARS-CoV-2 in 2020. 
Hence none of the study participants received corticosteroid therapy at the peak of acute COVID-19. 
Corticosteroids were administered to 
`r get_percent(cohort$analysis_tbl, 'steroid_treatment')['yes']` of analyzed participants with persistent pneumonia beginning from the third week after diagnosis at the discretion of the physician. 
Such sub-acute corticosteroid treatment was the most frequent in severe COVID-19 convalescents (`r get_test(cohort$test, 'steroid_treatment')`, moderate). 
The median weight loss during acute COVID-19 was `r abs(median(cohort$analysis_tbl$weight_change_kg))` kg and was significantly higher in severe as compared with ambulatory or moderate COVID-19 patients 
(`r get_test(cohort$test, 'weight_change_kg')`, large, __Table \@ref(tab:tab-cohort)__).

## Pulmonary recovery

Frequency of any radiological lung abnormalities, GGO, reticulation, consolidation, and bronchiectasis diagnosed by a human radiologist was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients. 
GGO and reticulation were the most frequent lesion types, independently of the follow-up and acute disease severity. 
Resolution of radiological lung findings. 
Consolidation and bronchiectasis lesions were detected primarily in moderate and severe COVID-19 survivors at the two-month follow-up examination. 
The increased frequency and severity of structural lung lesions in hospitalized patients as well as gradual resolution of the findings was also evident from analysis of the time course of human-determined CTSS, and AI-measured lung opacity and high opacity (__Supplementary Figures S1__ and __S2__, __Supplementary Table S3__). 
As demonstrated by correspondence analysis, in the majority of observations with GGO, reticulations were also present (`r covild$ct %>% filter(GGO_finding == 'yes', reticulation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with GGO). 
consolidations co-occurred with GGO lesions only in a minute fractions of observations (`r covild$ct %>% filter(GGO_finding == 'yes', consolidation_finding == 'yes') %>% nrow` of `r covild$ct %>% filter(GGO_finding == 'yes') %>% nrow` observations with  GGO), most frequently at the two-month follow-up (__Supplementary Figure S3A__). 
A high degree of correlation was observed between human-determined CTSS, and AI-determined lung opacity and high opacity (Spearman's $\rho$: `r corr_lft$test$ct$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S3B__).

Insufficient DLCO, FVC, and FEV1 defined as values below < 80% of the patient's reference value were the most common abnormalities of lung function. 
As with the CT findings, their frequency was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients, independently of the follow-up time point. 
Additionally, lung function recovery reflected by decreasing rates of insufficient DLCO and FVC, and by an improvement of DLCO and FVC values expressed as percentages of the patient's reference was the most evident in severe COVID-19 survivors (__Supplementary Figures S4__ and __S5__, __Supplementary Table S4__). 
In most of observations with insufficient FVC, insufficient FEV1 was found (`r covild$lft %>% filter(FVC_reduced == 'yes', FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(FVC_reduced == 'yes') %>% nrow` observations with reduced FVC). 
By contrast, in approximately half of observations with insufficient DLCO reduced FVC or FEV1 was diagnosed (`r covild$lft %>% filter(DLCO_reduced == 'yes') %>% filter(FVC_reduced == 'yes' | FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(DLCO_reduced == 'yes') %>% nrow` observations with insufficient DLCO, __Supplementary Figure S6A__). 
Accordingly, numeric values of FVC and FEV1 were strongly correlated (Spearman's $\rho$ = `r corr_lft$test$lft %>% filter(variable1 == 'FVC_percent', variable2 == 'FEV1_percent') %>% .$rho %>% signif(2)`), while correlation of DLCO with FVC or FEV1 was only moderate (Spearman's $\rho$: `r corr_lft$test$lft %>% filter(variable1 == 'DLCO_percent' | variable2 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S6B__). 
This may suggest that reduced DLCO in COVID-19 survivors reflects a type of functional lung pathology that is distinct from abnormalities measured by FVC or FEV1.

Impaired physical performance defined by ECOG > 0 and exertional dyspnea defined by mMRC > 0 were the most frequent persistent COVID-19-related symptoms of potential relevance for pulmonary function, and their rates declined during convalescence in all acute COVID-19 severity subsets.  
Notably, we could not observe any substantial differences in frequency of physical performance impairment, dyspnea, and persistent cough between severe, moderate and ambulatory COVID-19 convalescents (__Supplementary Figure S7__, __Supplementary Table S5__).

## Modeling of lung function impairment in COVID-19 convalescents

Prediction of lung function deficits following COVID-19 with demographic, clinical and biochemical explanatory factors and characteristic of acute SARS-CoV-2 infection suffers from low accuracy [@Sonnweber2022]. 
Additionally, relevance of residual structural lung lesions and severity of lung damage in COVD-19 convalescents for pulmonary function is not entirely clear [@Luger2022; @Sahanic2023]. 
Herein, we sought to model the major LFT abnormalities, i.e. insufficient DLCO, FVC, and FEV1, and numeric readouts, i.e DLCO, FVC, FEV1 in the CovILD cohort with a broad panel of `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` independent variables including baseline demographic and clinical characteristic (e.g. gender, age, smoking history, comorbidity), characteristic of acute COVID-19 (e.g. severity, medication, hospitalization) and convalescence (follow-up time point, persistent symptoms), as well as presence and severity CT abnormalities (e.g. GGO, CTSS, AI-determined opacity and high opacity, __Supplementary Table S2__). 
To this end, we employed four popular machine learning algorithms, 
Random Forest [@Breiman2001], 
gradient boosted machines (GBM) [@Friedman2002; @Friedman2001], 
neural network with a single hidden layer [@Ripley2014], and 
support vector machines (SVM) [@Weston1998]. 
Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for model optimization (tuning, __Supplementary Table S6__). 
Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).

Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference could be modeled with satisfactory accuracy in the cross-validation setting independently of the machine learning algorithm (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Low Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__). 
In turn, models of insufficient FVC and FEV1 exhibited poor predictive performance in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).

As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with sufficiently low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__). 
Analogically, no meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).

Concerning the models of insufficient DLCO, the highest concordance between the predicted and observed DLCO < 80% of the patient's reference was achieved for moderate COVID-19 convalescents at the two- to six-month follow-up examinations (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'hospitalized moderate', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`). 
The accuracy was in turn the poorest for ambulatory patients at the six- and twelve-month follow-up (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('6-month follow-up', '12-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`, __Supplementary Figure S11A__). 
The Random Forest, SVM and GBM models of DLCO expressed as percentage of the patient's reference exhibited the lowest average errors for moderate COVID-19 survivors (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized moderate') %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
In turn, DLCO values were systematically overestimated for severe COVID-19 patients at the two- to six-month follow-up visits (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized severe', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and underestimated for ambulatory COVID-19 convalescents at the same time points (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S11B__). 
The best performance in the hospitalized moderate subset is likely attributed to the large number of observations and frequent DLCO deficits in this group of participants available for training of the models (n = `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate') %>% nrow` observations from moderate COVID-19 patients). 
Conversely, the numbers of observations from ambulatory COVID-19 patients and numbers of observations DLCO insufficiency in this group was low (`r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'ambulatory mild') %>% nrow` observations from ambulatory patients) which resulted in relatively poor performance of the machine learning models.

As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO. 
Other influential explanatory variables were 
well characterized risk factors of severe COVID-19 (age, male sex, body mass index, pre-existing malignancy and cardiovascular disease), 
readouts of acute COVID-19 severity (severity class, WHO ordinal severity scale, hospitalization length, ICU treatment, anti-coagulant treatment, and weight change), 
smoking intensity (pack-years), as well as 
rating of physical performance impairment (ECOG) and 
exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S12__, __Supplementary Table S9__).

## Rating of structural lung damage by human radiologist and AI, and lung function deficiency

Because CTSS, lung opacity, and high opacity were identified to be crucial for prediction of DLCO insufficiency and DLCO by machine learning, we explored the association of the human- and AI-determined readouts of lung damage with DLCO in more detail.

Human-determined CTSS (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'CTSS')$eff_size`), 
AI-determined opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of the lung, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'opacity_percent')$eff_size`) and 
high opacity (difference of medians: `r blast(lft_uniboot$analysis_tbl, DLCO_reduced) %>% map(~.x$high_opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$significance`% of the lung, effect size: `r filter(lft_uniboot$test, split_factor == 'DLCO_reduced', variable == 'high_opacity_percent')$eff_size`) 
were significantly higher in observations with DLCO < 80% of the patient's reference than in the remaining data points. 
The effect size of those differences was large (__Figure \@ref(fig:fig-dlco-red-detection)A__). 
AI-measured opacity of the lung, but not CTSS or high opacity, was found to be significantly higher in observations with insufficient 
FVC (difference of medians: `r blast(lft_uniboot$analysis_tbl, FVC_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$eff_size`) and 
insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S10__). 
Furthermore, CTSS, opacity and high opacity correlated negatively with moderate effect size with DLCO expressed as percentage of the patient's reference (p < 0.001, Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S11__).

Finally, in a univariable ROC analysis, 
human-determined CTSS (AUC = `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'CTSS') %>% .$auc %>% signif(2)`, 95% CI: `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'CTSS') %>% .[c('lower_ci', 'upper_ci')] %>% unlist %>% signif(2) %>% paste(collapse = ' to ')`), and 
AI-determined lung opacity (AUC = `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'opacity_percent') %>% .$auc %>% signif(2)`, 95% CI: `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'opacity_percent') %>% .[c('lower_ci', 'upper_ci')] %>% unlist %>% signif(2) %>% paste(collapse = ' to ')`) and 
high opacity (AUC = `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'high_opacity_percent') %>% .$auc %>% signif(2)`, 95% CI: `r lft_roc$stats$optimal_cutoff$DLCO_reduced %>% filter(variable == 'high_opacity_percent') %>% .[c('lower_ci', 'upper_ci')] %>% unlist %>% signif(2) %>% paste(collapse = ' to ')`) 
were identified as standalone markers of insufficient DLCO in COVID-19 patients. 
The optimal cutoffs of CTSS, opacity, and high opacity for detection of DLCO < 80% of the patient's reference were, respectively, 
`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'CTSS')$cutoff %>% signif(2)` points, 
`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'opacity_percent')$cutoff %>% signif(2)`% of the lung, and 
`r filter(lft_roc$stats$optimal_cutoff$DLCO_reduced, variable == 'high_opacity_percent')$cutoff %>% signif(2)`% of the lung, and 
allowed for identification of insufficient DLCO with moderate sensitivity and specificity (sensitivity: `r lft_roc$stats$optimal_cutoff$DLCO_reduced$Se %>% range %>% signif(2) %>% paste(collapse = ' to ')`, specificity: `r lft_roc$stats$optimal_cutoff$DLCO_reduced$Sp %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-dlco-red-detection)B__). 
In case of the AI measures of lung opacity and high opacity, the low values of the optimal cutoffs let us infer that event low-grade radiological abnormalities can be associated with clinically relevant functional lung impairment. 
Yet, concordance between the predicted observed DLCO insufficiency assessed by Cohen's $\kappa$ for CTSS, opacity, and high opacity as standalone markers (blocked bootstrap $\kappa$: `r filter(lft_roc$bootstrap$stats, response == 'DLCO_reduced', statistic == 'kappa')$estimate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Table \@ref(tab:tab-ct-roc)__) 
was substantially lower that 
in the multi-parameter machine learning models (cross-validated $\kappa$: `r filter(bin_models$stats$DLCO_reduced, dataset == 'cv')$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`). 
This underlines the importance of other CT-unrelated explanatory factors such acute disease course for reliable prediction of functional lung deficits in COVID-19 convalescents.

\newpage

# Discussion

\newpage

# Limitations

Our study has limitations. 
First, the overall patient and observation number was low, in particular for ambulatory COVID-19 convalescents. 
Analogically, the study cohort was enriched in hospitalized individuals with in constitute a minute fraction of COVID-19 patients in the real world setting. 
Second, complete sets of longitudinal CT and LFT measurements at the two-, three-, six- and twelve-month follow-up examinations were available solely for `r covild$ct %>% blast(ID) %>% map_dbl(nrow) %>% map_lgl(~.x == 4) %>% sum` patients. 
Especially the number of observations obtained with ambulatory and moderate COVID-19 convalescents at the six- and twelve-month follow-up were substantially lower as compared with earlier time points. 
The incompleteness of the longitudinal data may have hence compromised performance of the machine learning models in particular for ambulatory COVID-19 cases and at the later time points. 
Third, because of the limited number of participants and observations, we abstained from definition of a test subset of the data used solely for bias-free model evaluation (hold-out strategy). 
Instead, both model selection and evaluation was done with blocked repeated cross-validation, which may have overestimated performance of the models. 
Hence, external validation of our findings in an independent cohort is recommended. 
Finally, the analyzed cohort was recruited in the initial phase of the pandemic and consisted of individuals infected with the wild-type variant of the SARS-CoV-2 virus. 
For this reason, it is not completely clear, how our findings translate to the recent variants of the pathogen and how the pulmonary recovery is affected by anti-SARS-CoV-2 immunity, improved treatment and care. 
However, it is feasible, that the CT severity readouts, human-determined CTSS as well as AI-determined opacity and high opacity, are equally applicable in the post-pandemic setting as standalone markers of functional lung impairment during recovery from COVID-19 and other respiratory infections.

\newpage

# Data and code availability

The entire R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/CovILD_AI). 
The [CovILD data set](https://github.com/PiotrTymoszuk/covILD) will be made available on reasonable request to the corresponding author.

\newpage

# Tables

```{r tab-cohort, tab.cap = 'Baseline characteristics and COVID-19 course in the CovILD cohort. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(paper_tabs$cohort) %>% 
  set_widths(c(2.3, rep(3.3, 4), 2.7, 2.5)) %>% 
  footnote(1, 1, 
           value = as_paragraph('age class, young adult: 16 - 40: young adult; 41 - 65: middle aged; 66 and older: elderly; BMI: body mass index; body mass class, normal: BMI \u2264 25, overweight: 25 - 30, obesity: > 30 kg/m\u00B2; COPD: chronic obstructive pulmonary disease; CKD: chronic kidney disease; ICU: intensive care unit; steroids during COVID-19: in cases of non-improving pneumonia, \u2265 3 weeks after diagnosis.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:7, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header',
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-bin-classifiers, tab.cap = 'Cross-validated performance of binary machine learning classifiers at predicting lung function testing (LFT) abnormalities.'}

flextable::flextable(paper_tabs$bin_classifiers) %>% 
  set_widths(c(2.7, 3.1, rep(2, 6))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing, DLCO: diffusion capacity for CO, FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('SVM: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3, 
           value = as_paragraph('Ratio of correct predictions to the total observation number.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 4, 
           value = as_paragraph('Cohen \u03BA statistic of inter-rater reliability between the predicted and observed outcome.'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 6, 
           value = as_paragraph('AUC: are under the receiver-operating characteristic curve.'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

```{r tab-reg-models, tab.cap = 'Cross-validated performance of regression machine learning models at predicting values of lung function testing parameters.'}

flextable::flextable(paper_tabs$reg_models) %>% 
  set_widths(c(2.5, 3.1, rep(2.1, 3))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCO: diffusion capacity for CO, FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('SVM: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3, 
           value = as_paragraph('Defined as 1 - ratio of mean squared error and variance.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 4, 
           value = as_paragraph('MAE: mean absolute error.'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 5, 
           value = as_paragraph('\u03A1: Spearman coefficient of correlation between the predicted and observed response values.'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

```{r tab-ct-roc, tab.cap = 'Detection of reduced diffusion capacity for CO DLCO < 80% reference value) by single CT-derived parameters: AI-determined opacity and high opacity, and human-determined CT severity score.'}

flextable::flextable(paper_tabs$ct_roc) %>% 
  set_widths(c(3, 2.5, 2.7, 3.8)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('CTSS: human-determined CT severity score, sum for al lung lobes; high opacity and opacity, AI: percentage of the lungs with high opacity and opacity determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('Cutoff of the CT variable corresponding to the maximum of Jouden Y statistic.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3,
           value = as_paragraph('AUC: area under the curve of receiver-operating characteristic; \u03BA: Cohen \u03BA statistic of inter-rater reliability between the predicted and observed outcome, computed for the CT variable cutoff; sensitivity and specificity: sensitivity and specificity computed at the CT variable cutoff.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-consort, fig.width = figur::convert(paper_figs$consort, to = 'in')$w, fig.height = figur::convert(paper_figs$consort, to = 'in')$h, fig.cap = 'Analysis inclusion scheme.'}

paper_figs$consort$plot

```

__Figure \@ref(fig:fig-consort). Analysis inclusion scheme.__

_The analysis inclusion criterion for participants of the longitudinal observation CovILD study was completeness of visit-matched computed tomography and lung function testing results._

_CT: compute tomography; LFT: lung function testing; GGO: ground glass opacity; AI: artificial intelligence; DLCO: diffusion capacity for carbon monoxide (CO); FVC: forced vital capacity; FEV1: forced expiratory volume in one second._

\newpage

```{r fig-ml-performance, fig.width = figur::convert(paper_figs$ml_performance, to = 'in')$w, fig.height = figur::convert(paper_figs$ml_performance, to = 'in')$h, fig.cap = 'Evaluation of performance of machine learning models of diffusion capacity for CO during COVID-19 convalescence.'}

paper_figs$ml_performance$plot

```

__Figure \@ref(fig:fig-ml-performance). Evaluation of performance of machine learning models of diffusion capacity for CO during COVID-19 convalescence.__

_(A) Four machine learning classification models of insufficient diffusion capacity for carbon monoxide (< 80% of reference value: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome %>% sum`) employing computed tomography readouts, demographic and clinical explanatory variables were trained. Their performance was evaluated in the entire data set and 10-repeats 10-fold cross-validation with overall accuracy metric, Cohen's $\kappa$ as a measure of concordance between predicted and observed outcome, and Brier score as a measure of model's calibration. Left: numeric performance measures of the models (open circles: the entire data set, filled circles: cross-validation); point sizes and point labels represent overall model accuracy, the dashed lines visualize values of Cohen's $\kappa$ and Brier score expected for prediction of insufficient DLCO be chance. Right: receiver-operating characteristic curves for predictions in cross-validation folds, numeric statistics are displayed in the plot. Numbers of complete observations and observations witj DLCO insufficiency ('events') are displayed in the plot captions._

_(B) Four machine learning regression models of diffusion capacity for carbon monoxide (percentage of reference values, total observations: n = `r nobs(reg_models$models$DLCO_percent[[1]])`) employing computed tomography readouts, demographic and clinical explanatory variables were trained. Their performance was evaluated in th entire data set and 10-repeats 10-fold cross-validation with $R^2$ as a measure of explained variation, mean absolute error, and $\rho$ Spearman's coefficient of correlation between the predicted and observed values. Bubble plot: numeric performance measures of the model (open circles: the entire data set, filled circles: cross-validation); point sizes and point labels represent values of $\rho$ correlation coefficient, the dashed line visualizes $R^2$ value expected for a meaningless model. Scatter plots: observed and predicted values of diffusion capacity for carbon monoxide in cross-validation folds; the blue dashed lines with slope 1 and intercept 0 represent absolutely accurate predictions, general additive model trends with standard errors are visualized as the solid blue lines with gray ribbons. Numbers of complete observations and Spearman's $\rho$ coeffficients of correlation between the predicted and observed values are displayed in the plot captions._

_DLCO: diffusion capacity for carbon monoxide, CV: cross-validation; AUC: are under the receiver-operating characteristic curve; Se: sensitivity; Sp: specificity; MAE: mean absolute error; GBM: gradient boosted machines; SVM radial: support vector machines with radial kernel._

\newpage

```{r fig-varimp-dlco-red, fig.width = figur::convert(paper_figs$varimp_dlco_red, to = 'in')$w, fig.height = figur::convert(paper_figs$varimp_dlco_red, to = 'in')$h, fig.cap = 'Explanatory variable importance for models of insufficient diffusion capacity for carbon monoxide measured by Shapley additive explanations.'}

paper_figs$varimp_dlco_red$plot

```

__Figure \@ref(fig:fig-varimp-dlco-red). Explanatory variable importance for models of insufficient diffusion capacity for carbon monoxide measured by Shapley additive explanations.__

_Importance of explanatory variables for the machine learning models of insufficient diffusion capacity for carbon monoxide (< 80% of reference, Figure \@ref(fig:fig-ml-performance)) was investigated by Shapley additive explanations (SHAP). Absolute SHAP values for explanatory variables with the 15 largest mean SHAP values are presented in violin plots. Points represent single observations, point colors code for minimum/maximum scaled value of the explanatory variable. Explanatory variables obtained via computed tomography are highlighted with bold font in the Y axes._

_CT: computed tomography; DLCO: diffusion capacity for carbon monoxide; opacity and high opacity, AI: opacity and high opacity of the lung determined by artificial intelligence; BMI: body mass index; CTSS: human-determined CT severity score, sum for all lobes; ECOG: Eastern Cooperative Oncology Group physical performance score; mMRC: modified Medical Research Council dyspnea scale; ICU: intensive care unit._

\newpage

```{r fig-dlco-red-detection, fig.width = figur::convert(paper_figs$dlco_red_detection, to = 'in')$w, fig.height = figur::convert(paper_figs$dlco_red_detection, to = 'in')$h, fig.cap = 'Detection of DLCO insufficiency by human- and artificial intelligence-determined CT readouts of severity of structural lung damage.'}

paper_figs$dlco_red_detection$plot

```

__Figure \@ref(fig:fig-dlco-red-detection). Detection of DLCO insufficiency by human- and artificial intelligence-determined CT readouts of severity of structural lung damage.__

_Human- and artificial intelligence-determined computed tomography readouts of structural lung damage were identified as influential explanatory variables at prediction of insufficiency of diffusion capacity for carbon monoxide (< 80%) by machine learning._

_(A) Values of the radiological readouts of lung damage severity were compared between data points with and without insufficient diffusion capacity for carbon monoxide by Mann-Whitney test with r effect size statistic. Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations are indicated in the X axis._

_(B) Quality of detection of insufficient diffusion capacity for carbon monoxide with the radiological readouts of lung damage severity was assessed by receiver-operating characteristic (ROC) analysis. ROC curves are shown, the optimal cutoffs of the severity readuts determined by Youden's criterion are represented by points with numbers. Sensitivity, specificity at the optimal cutoff, and area under the curve statistic with 95% confidence interval are displayed in the plots._

_CT: computed tomography; DLCO: diffusion capacity for carbon monoxide;  CTSS: human-determined CT severity score, sum for all lobes; opacity and high opacity, AI: opacity and high opacity of the lung determined by artificial intelligence; AUC: are under the curve of receiver-operating characteristic; Se: sensitivity; Sp: specificity._

\newpage

# References

### 2024-03-22 paper/markdown/supplementary_material.Rmd
Manuscript draft

---
title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Supplementary Material"
author: "Department of Radiology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Supplementary Methods

## Ethics

The CovILD study was conducted in accordance with the Declaration of Helsinki and the European Data policy. 
All participants gave written informed consent to participate and to process their data. 
The study data were processed, stored and analyzed in anonymized form. 
The study protocol was approved by the ethics committee of the Medical University of Innsbruck, Austria (approval number: 1103/2020). 
The study was registered at [ClinicalTrials.gov](https://classic.clinicaltrials.gov/ct2/show/NCT04416100) (NCT04416100).

## Study design and participants, handling of missing data

The longitudinal observation CovILD study aimed at investigation of symptom, cardiopulmonary and mental health recovery of COVID-19 patients. 
The details of the study design and the cohort are provided in our recent publications [@Sonnweber2020; @Sonnweber2022; @Luger2022; @Sahanic2023]. 
In brief, n = `r length(covild$common_ids)` convalescent COVID-19 survivors were recruited among patients of the University Hospital of Innsbruck, St. Vinzenz Hospital in Zams and Karl-Landsteiner Rehabilitation Center in Münster (all in Austria) between March and June 2020. 
The inclusion criteria were age $\geq$ 18 years, SARS-CoV-2 positivity confirmed by PCR and presence of COVID-19 symptoms. 
All participants were infected with the wild-type form of SARS-CoV-2. 
The study visits were scheduled at two, three, six, and twelve months after COVID-19 diagnosis. 
In the current analysis, participant-matched longitudinal observations were utilized. 
The analysis inclusion criterion was availability of data from computed tomography (CT) of the chest and lung function testing (LFT). 
This criterion was fulfilled by `r nrow(covild$ct)` observations obtained from `r length(covild$complete_ids)` CovILD study participants. 
For the patients included in the analysis, the set of explanatory variables concerning demographic and clinical background as well as the course and treatment of acute COVID-19 was complete.

The study design and analysis inclusion process are summarized in a flow chart in __Figure 1__. 
Numbers of observations at study follow-up examinations are presented in __Supplementary Table S\@ref(tab:tab-n-numbers)__.
Analyzed variables are listed in __Supplementary Table S\@ref(tab:tab-study-variables)__. 
Baseline demographic and clinical characteristic of the study cohort and characteristic of acute COVID-19 are provided in __Table 1__. 
Frequency of CT and LFT abnormalities, scoring of CT lesions and values of LFT readouts are listed in __Supplementary Tables S\@ref(tab:tab-cohort-ct)__ and __S\@ref(tab:tab-cohort-lft)__. 
Frequencies of COVID-19 -related symptoms of relevance for lung function are presented in __Supplementary Table S\@ref(tab:tab-cohort-symptoms)__.

## Study procedures and data sources

Information on demography (age, sex), smoking status, routine medication, pre-existing conditions, as well as the course and treatment of acute COVID-19 were obtained in an interview and retrieved from electronic patient's record at the two-month follow-up visit. 
The standard follow-up visit protocol included a survey of symptoms and self-reported physical performance, examination and interview by a physician, determination of standard blood markers (hemoglobin, iron turnover, complete blood count, markers of inflammation and vascular pathology), trans-thoracic echocardiography, LFT, and CT of the chest. 
The details of the study procedures are provided in our recent publications [@Luger2022; @Sonnweber2020; @Sonnweber2022; @Sahanic2023]. 
The complete list of study variables with their format and descriptions is provided in __Supplementary Table S\@ref(tab:tab-study-variables)__.

Regarding the severity of acute COVID-19, the study participants were classified as 
'ambulatory' (home isolated, WHO ordinal scale for clinical improvement: 1 - 2), 
'hospitalized moderate' (hospitalized, without oxygen therapy or oxygen by mask or nasal prongs, WHO: 3 - 4), and 
'hospitalized severe' (hospitalized, high flow oxygen or mechanical ventilation, WHO: 5 - 7).

Chest CT was done with a 128 slice multi-detector SOMATOM Definition Flash device (Siemens Healthineers, Erlangen, Germany) with a 128 $\times$ 0.6 mm collimation and spiral pitch factor of 1.1. 
Scans were obtained in craniocaudal direction without iodine contrast agent and in low-dose setting (100 kVp tube potential). 
Artificial intelligence (AI) supported analysis of the lung CT scans was done with the Syngo.via CT Pneumonia Analysis Software (Siemens Healthineers, Erlangen, Germany), which returned percentage of the lungs with opacity and high opacity. 
Evaluation of the CT scans by the radiologists was done in accordance with the Fleischner society guidelines as described before [@Luger2022; @Sonnweber2020; @Hansell2008]. 
In brief, the scans were examined for presence of ground glass opacity (GGO), reticulation, consolidations, and bronchiectasis. 
Severity of lung abnormalities was rated by the radiologist with the CT severity score (CTSS) described before [@Luger2022; @Sonnweber2020]. 
For computation of CTSS, radiological abnormalities were scored separately for each lobe with the following scheme:

* 0: no abnormalities

* 1: minimal, subtle GGO

* 2: mild, several GGO and subtle reticulations

* 3: moderate, multiple GGO, reticulation, small consolidation

* 4: severe, extensive GGO, consolidation, reticulation with distortion

* 5: massive findings, parenchymal destruction

CTSS was calculated as a sum of the scores over all lobes and ranged from 0 to 25 [@Luger2022; @Sonnweber2020].

The following LFT parameters were recorded: 
diffusion capacity for carbon monoxide (DLCO), 
forced vital capacity (FVC), 
forced expiratory volume in one second (FEV1), 
total lung capacity (TLC), and 
FEV1 to FVC ratio (FEV1:FVC). 
Among them, DLCO, FVC, and FEV1, whose abnormalities were the most frequent in the CovILD cohort, were analyzed in the current report. 
These variables were expressed as percentage of the patient's individual age-, sex- and weight-specific reference. 
Insufficiency of DLCO, FVC, or FEV1 were defined as values below 80% of the reference [@Sonnweber2022; @Sonnweber2020; @Sahanic2023].

The following longitudinally recorded symptoms of potential relevance for lung function were evaluated: 
dyspnea measured by Modified Medical Research Council scale (mMRC, presence of dyspnea defined as mMRC > 0), 
self-reported cough, and 
self-rated physical performance measured by Eastern Cooperative Oncology Group (ECOG, impaired physical performance defined as ECOG > 0) [@Sahanic2023]. 
Missing information on the symptoms or their intensity in the patient's symptom survey were interpreted as absence of the complaint.

The source study data set was fetched from the [CovILD study data repository](https://github.com/PiotrTymoszuk/covILD) available as an R package to authorized users.

## Software

The analysis was done with R version 4.2.3. 
Tabular data and code pipelines were handled by tools provided by the packages _tidyverse_ [@Wickham2016], _rlang_ [@Henry2022], and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
Text data were handled with _stringi_ [@Gagolewski2021]. 
Parallelization was accomplished with _furrr_ [@Vaughan2022] and _doParallel_ [@Folashade2022].

In exploratory data analysis, statistical hypothesis testing and analysis of correspondence and correlations, the packages _rstatix_ [@Kassambara2021], _rcompanion_ [@Mangiafico2022], _MASS_ [@Ripley2022], [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) were used. 
Receiver-operating characteristic (ROC), confusion matrix analysis and inter-rater reliability analyses were done with _caret_ [@Kuhn2008], _OptimalCutpoints_ [@Lopez-Raton2014] and [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).

For machine learning modeling of lung function testing outcomes, the packages _caret_ [@Kuhn2008] and [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) were utilized, which provide wrappers around the R implementations of the random forest algorithm (_ranger_) [@Wright2017; @Breiman2001], 
neural network (_nnet_) [@Ripley2014], 
support vector machines (SVM, _kernlab_) [@Weston1998; @Karatzoglou2004], 
and gradient boosted machines (GBM, _gbm_) [@Greenwell2022; @Friedman2002; @Friedman2001]. 
Model diagnostic statistics and performance evaluation metrics were computed with the development package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra). 
SHAP (Shapley additive explanations) as a metric of global variance importance in machine learning models was calculated with the package _kernelshap_ [@Covert2020; @Lundberg2017; @Mayer2023].

Analysis results were visualized with _ggplot_ [@Wickham2016] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (scatter, stack and box plots), [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (visualization of results of correspondence analysis), _ggvenn_ [@Yan2021] (Venn plots), [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) (plots for machine learning model diagnostic and evaluation), _plotROC_ [@Sachs2017] (ROC curves) and _shapviz_ [@Mayer2023a; @Mayer2023] (violin/scatter plots of absolute SHAP values). 
Result tables were created with _flextable_ [@Gohel2022]. 
Figures were generated with _cowplot_ [@Wilke2019]. 
Parts of the manuscript and supplementary material were written in the _rmarkdown_ environment [@Allaire2022] with the _bookdown_ package [@Xie2016]. 
Figures, tables, links and R expressions in the Rmarkdown document were managed with [_figur_](https://github.com/PiotrTymoszuk/figur).

## Descriptive statistic, statistical tests and effect sizes, correspondence analysis

If not stated otherwise, numeric values are presented as medians with interquartile ranges and ranges within the analysis data set or analysis strata. 
Qualitative variables are presented as percentages and counts of the category within the analysis data set or analysis strata. 
Normality of distribution of numeric variables upon identity, logarithm and square root transformation was checked with Shapiro-Wilk test (function `shapiro_test`, package _rstatix_). 
Since multiple numeric variables were not normally distributed, Mann-Whitney test or Kruskal-Wallis test were routinely used for comparison of independently distributed numeric variables between analysis groups. 
Differences in distribution of categorical variables between the study groups were assessed with $\chi^2$ test. 
P values were adjusted for multiple testing with the false discovery rate method separately for each analysis task (e.g. comparison of explanatory variables between observations with and without LFT findings) [@Benjamini1995]. 
Effects with p < 0.05 were considered significant.

The following effect size metrics were used to assess differences between analysis groups and to evaluate performance of machine learning models [@McHugh2012; @Cohen2013; @Cohen1960]:

* biseral r was used for two-group comparisons of numeric variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size

* $\eta^2$ was used for multi-group comparison of numeric variables: < 0.13: small, 0.13 - 0.26: moderate, $\geq$ 0.26: large effect size

* Cramer's V was used for comparisons of categorical variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size

* Spearman's $\rho$ coefficient of correlation: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size

* pseudo-R^2^ was used as a measure of explained variance of a model: < 0.13: small, 0.13 - 0.26: moderate, $\geq$ 0.26: large effect size

* Cohen's $\kappa$ inter-rater reliability statistic: < 0.20: no effect, 0.2 - 0.4: fair, 0.4 - 0.6: moderate, 0.6 - 0.8: good, $\geq$ 0.8: excellent concordance

Statistical significance and effect sizes for comparisons and correlations of independent observations was assessed with the functions `compare_variables()` and `correlate_variables()` (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
Because of non-independent observations included in the analysis, pairwise correlations of numeric LFT or CT variables were investigated with blocked bootstrap Spearman's rank test described below. 
Analogically, for comparison of medians between two subsets of non-independent observations, blocked bootstrap test was utilized. 
Please refer to __Comparison of CT readouts between observations with and without LFT abnormality, correlation analysis__ for details.

Co-occurrence of LFT abnormalities (DLCO < 80% reference, FVC < 80% of reference, and FEV1 < 780% of reference) was investigated by two-dimensional correspondence analysis (function `mca()`, package _MASS_) [@Ripley2022] and whose column factors were visualized in a scatter plot with functions from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
Additionally, counts of observations with co-occurring LFT abnormalities were visualized in Venn plot (function `ggvenn()`, package _ggvenn_).

## Multi-parameter modeling of lung function

Presence of reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables including:

* baseline demographic and clinical characteristic (e.g. age, sex, smoking, comorbidities)

* explanatory variables referring to the course of acute COVID-19 (e.g. WHO ordinal scale for clinical improvement, hospitalization status and length, treatment)

* follow-up after COVID-19 diagnosis (two, three, six and twelve months)

* presence and rating of longitudinally recorded symptoms of relevance for lung function (dyspnea, cough, impaired physical performance)

* presence and severity of lung CT findings rated by the radiologist (e.g. GGO, consolidations, CTSS) and AI (opacity and high opacity)

To reduce the number of explanatory variables, some qualitative variables with rare categories were re-coded. 
This concerned:

* _`r  exchange('pulmonary_comorbidity', lft_globals$lexicon)`_ which subsumed pre-existing chronic obstructive lung disease (COPD), bronchial asthma, interstitial lung disease, chronic lung diseases and other pulmonary conditions

* _`r exchange('anticoagulant_treatment', lft_globals$lexicon)`_ which subsumed anti-coagulation and anti-platelet treatment during acute COVID-19

* _`r exchange('antiinfective_treatment', lft_globals$lexicon)`_ which subsumed anti-infective and anti-macrolide treatment during acute COVID-19

* _`r exchange('smoking', lft_globals$lexicon)`_, where ex- and active smokers were grouped together

Severity of acute COVID-19 measured by the WHO ordinal scale of clinical improvement was re-coded as categorical explanatory variable. 
The response and explanatory variables are listed in __Supplementary Table S\@ref(tab:tab-study-variables)__, the modeling strategy is schematically summarized in __Supplementary Figure S\@ref(fig:fig-modeling-strategy)__.

Multi-parameter models were constructed with four machine learning algorithms employing diverse mathematical principles: 
canonical random forest (implemented in R by the _ranger_ package, caret's method name: 'ranger') [@Wright2017; @Breiman2001], 
GBM (_gbm_ package, caret's method name: 'gbm') [@Greenwell2022; @Friedman2002; @Friedman2001], 
neural network with a single hidden layer (_nnet_ package, caret's method name: 'nnet') [@Ripley2014], 
and SVM with radial kernel (_kernlab_ package, caret's method name: 'svmRadial') [@Weston1998; @Karatzoglou2004]. 
For random forest models, 1000 random trees per model were generated. 
Bernoulli and Gaussian loss functions were implemented in classification and regression GBM models, respectively. 
Optimal values of parameters controlling the algorithm's behavior such as number of observations in terminal nodes of tree models in the random forest models or cost penalty in the SVM models were found by maximizing the Youden's J statistic 
($J = Sensitivity + Specificity - 1$, classification models of binary LFT outcomes) or by minimizing mean absolute error (MAE, regression models of numeric LFT outcomes) in 10-repeats 10-fold cross-validation. 
To account for the participant matching of the observations, a blocked cross-validation workflow was designed to keep all observations obtained from a particular participant either in the training or in the test portion of the cross-validation split. 
Construction of such 'participant-wise' blocked cross-validation folds was done with an in-house-developed script employing the `createMultiFolds()` function (package _caret_). 
For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed [@Kuhn2008]. 
The optimal values of the tuning parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.

Performance of the machine learning models was evaluated by comparison of the model predictions with the observed outcome in the genuine training data set and in 10-repeats 10-fold cross-validation with the observed outcomes. 
Performance of the classification models was assessed by Cohen's $\kappa$ as a measure of concordance between the predicted and observed outcome [@McHugh2012; @Cohen1960], 
standard receiver operating characteristic metrics (AUC, overall accuracy, sensitivity, specificity) to assess the ability of the model to detect the pathological outcome [@Kuhn2008], and with 
Brier score to investigate credibility of the predictions and model calibration [@Brier1950; @Goldstein-Greenwood2021] (__Table 2__ and __Supplementary Table S\@ref(tab:tab-bin-classifiers)__). 
Performance of the regression models was assessed with 
mean absolute error (MAE) as a measure of fit quality, 
pseudo-R^2^ as a metric of explained variance and 
Spearman's $\rho$ for correlation between the predicted and observed outcome as as measure of overall model calibration [@Kuhn2008] (__Table 3__ and __Table \@ref(tab:tab-reg-models)__). 
Pseudo-R^2^ was computed with the following formulas:

$$pseudo R^2 = 1 - \frac{MSE(y)}{Var(y)}$$
$$MSE(y) = \frac{1}{N} \sum_i^N (y_i - \hat{y}_i)^2$$

where $MSE(y)$ stands for mean squared error of the outcome variable $y$, $Var(y)$ is the variance of the outcome variable $y$, $N$ is the total observation number, $i$ represents the observation index, $y_i$ is the observed outcome value for the i-th observation, and $\hat{y}_i$ is the predicted outcome value for the i-th observation. 
Performance statistics were retrieved from the `caret` models with the `summary()` method from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package. 
Visualization of the model tuning and evaluation results was accomplished with the method `plot()` from the [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) package and in-house developed, project-specific functions.

In order to assess the model accuracy as a function of acute COVID-19 severity and time after diagnosis, we computed Cohen's $\kappa$ (classification models) and MAE (regression models) of out-of-fold predictions for observations stratified by acute COVID-19 severity (mild, moderate, severe) and follow-up visit (2-, 3-, 6- and 12-month follow-up). 
The error values were subsequently visualized as heat maps.

## Variable importance for multi-parameter models of lung function

Shapley additive explanations (SHAP) were used as a global importance statistic for the model's explanatory variables. 
In brief, SHAP measures contribution of the given variable to the model fit by comparing a goodness-of-fit statistic between models re-fitted with subsets of explanatory variables including the variable of interest and models re-fitted with subsets of explanatory variables without the variable of interest [@Covert2020; @Lundberg2017]. 
Matrices of SHAP values for observations and variables were computed for the machine learning models of lung function outcomes with the function `kernelshap()` provided by the _kernelshap_ package [@Mayer2023a]. 
As a background for SHAP calculation, a single observation data frame was used with numeric explanatory variables set to the 25^th^ percentile of the values in the entire data set and with qualitative explanatory variables set to the baseline category [@Mayer2023a]. 
For mean absolute SHAP values of the most influential explanatory variables in the Random Forest, neural network, SVM, and GBM models, see: __Table \@ref(tab:tab-mean-shap)__. 
Visualization of the mean SHAP values for the top most influential explanatory variables was accomplished with the `shapviz()`, `sv_importance()` (package _shapviz_) [@Mayer2023], and in-house developed functions.

## Comparison of CT readouts between observations with and without LFT abnormality, correlation analysis

Because observations were non-independent, i.e. participant-matched, comparison of CTSS, opacity and high opacity of the lung between data points with and without LFT abnormalities could not be investigated by standard statistical tests (e.g. Mann-Whitney test) and required instead a special approach. 
Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biseral r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality). 
The blocked bootstrap scheme involved re-sampling with repetition of the participants instead of single observations (B = 2000 re-samples). 
P values were obtained by counting re-samples in which the test hypothesis H1 was not met and dividing the count by the total re-sample number. 
The effect size statistic was computed as an arithmetic average of effect sizes in the re-samples.

A similar blocked bootstrap approach was applied in a correlation analysis of the CT (CTSS, opacity, high opacity) and LFT readouts (DLCO, FVC, FEV1, expressed as percentages of reference values). 
The H0 null hypothesis of such correlation test was no or positive correlation of a CT and LFT readout. 
The H1 hypothesis was the presence of negative correlation between the CT and LFT readout. 
Spearman's $\rho$ correlation coefficients were computed for the bootstrap re-samples (B = 2000). 
The p values were obtained by counting re-samples with $\rho \geq$ 0 and dividing the count by the total re-sample number. 
The $\rho$ values were averaged and their confidence values were calculated as the 2.5 - 97.5 percentile range.

Bootstrap tests were performed with a functional interface implemented in the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat). 
The analysis results are listed in __Supplementary Table S\@ref(tab:tab-lft-binary-uni)__ and __\@ref(tab:tab-lft-correlation-uni)__.

## Inter-rater reliability and receiver-operating characteristic analysis

The optimal cutoffs of human-determined CTSS, and AI-determined lung opacity and high opacity for detection of insufficient DLCO (< 80% of reference) were computed with the maximum of the Youden's J statistic ($J = Sensitivity + Specificity - 1$) in the entire analysis data set (function `optimal.cutpoints()`, package _OptimalCutpoints_) [@Lopez-Raton2014]. 
Those optimal cutoffs were subsequently used to dichotomize the analyzed observations. 
Concordance between the CTSS^high^, opacity^high^ and high opacity^high^ strata assignment and presence of LFT insufficiency of reference was assessed by Cohen's $\kappa$ inter-rater reliability statistic [@McHugh2012; @Cohen1960] as well as sensitivity, specificity and AUC metrics. 
Those statistics were computed with the function `multiClassSummary()` from the _caret_ package [@Kuhn2008]. 
Boundaries of the statistic's 95% confidence intervals were obtained by blocked bootstrap with 2000 iterations and were defined by the 2.5 and 97.5 percentiles of the blocked bootstrap estimates. 
Bootstrapping was accomplished by a functional interface implemented in the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat). 
The analysis results are listed in __Table 4__.

## Data and code availability

The entire R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/CovILD_AI). 
The [CovILD data set](https://github.com/PiotrTymoszuk/covILD) will be made available on reasonable request to the corresponding author.

\newpage

# Supplementary Tables

```{r tab-n-numbers, tab.cap = 'Numbers of observations included in the analysis split by acute COVID-19 severity and the follow-up examination.'}

flextable::flextable(suppl_tabs$n_numbers) %>% 
  set_widths(c(3, rep(3.5, 4))) %>% 
  theme_vanilla

```

\newpage

```{r tab-study-variables, tab.cap = 'Study variables. The table is available as a supplementary Excel file.'}

suppl_tabs$study_variables[1, 1] %>% 
  mutate(`Use in modeling` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-cohort-ct, tab.cap = 'Chest computed tomography variables at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tabs$cohort_ct) %>% 
  set_widths(c(2.3, 2.7, rep(3.9, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('any CT findings: any CT findings diagnosed by a human pathologist; GGO: ground glass opacity; CTSS: human-determined CT severity score; reticulation, consolidation, bronchiectasis: diagnosed by a human radiologist; opacity and high opacity, AI: opacity and high opacity determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-cohort-lft, tab.cap = 'Lung function testing variables at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tabs$cohort_lft) %>% 
  set_widths(c(2.3, 2.7, rep(3.7, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; DLCO: diffusion capacity for carbon monoxide.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

```

\newpage

```{r tab-cohort-symptoms, tab.cap = 'Presence and rating of symptoms of relevance for lung function at consecutive follow-ups. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tabs$cohort_symptoms) %>% 
  set_widths(c(2.3, 3.5, rep(3.7, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('mMRC: modified Medical Research Council scale; ECOG: Eastern Cooperative Oncology Group performance scale.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

```

\newpage

```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm parameters by cross-validation-mediated tuning.'}

flextable::flextable(suppl_tabs$tuning) %>% 
  set_widths(c(3, 4, 4.5)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-bin-classifiers, tab.cap = 'Performance of binary machine learning classifiers at predicting lung function testing (LFT) abnormalities in the entire data set.'}

flextable::flextable(suppl_tabs$bin_classifiers) %>% 
   set_widths(c(2.7, 3.1, rep(2, 6))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing, DLCO: diffusion capacity for carbon monoxide, FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('SVM: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3, 
           value = as_paragraph('Ratio of correct predictions to the total observation number.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 4, 
           value = as_paragraph('Cohen \u03BA statistic of inter-rater reliability between the predicted and observed outcome.'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 6, 
           value = as_paragraph('AUC: are under the receiver-operating characteristic curve.'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

```{r tab-reg-models, tab.cap = 'Performance of regression machine learning models at predicting values of lung function testing parameters in the entire data set.'}

flextable::flextable(suppl_tabs$reg_models) %>% 
  set_widths(c(2.5, 3.1, rep(2.3, 3))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCO: diffusion capacity for carbon monoxide, FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('SVM: support vector machines with radial kernel; GBM: gradient boosted machines.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 3, 
           value = as_paragraph('Defined as 1 - ratio of mean squared error and variance.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 4, 
           value = as_paragraph('MAE: mean absolute error.'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 5, 
           value = as_paragraph('\u03A1: Spearman coefficient of correlation between the predicted and observed response values.'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

```

\newpage

```{r tab-mean-shap, tab.cap = 'Mean values of SHAP (Shapley additive explanations) variable importance statistic for the models of reduced diffusion capacity for carbon monoxide and of diffusion capacity for carbon monoxide < 80% of reference. The table is available as a supplementary Excel file.'}

suppl_tabs$mean_shap[1, 1] %>% 
  mutate(Response = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-lft-binary-uni, tab.cap = 'Differences in chest computed tomography severity score (CTSS), and AI-determined lung opacity and high opacity in CovILD study participants with and without lung function testing (LFT) abnormalities. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.'}

flextable::flextable(suppl_tabs$lft_binary_uni) %>% 
  set_widths(c(2.5, 3, rep(4, 2), 2.7, 2.5)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for all lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Blocked bootstrap test with biseral r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

\newpage

```{r tab-lft-correlation-uni, tab.cap = 'Correlation of LFT variables with chest computed tomography severity score, and AI-determined opacity and high opacity.'}

flextable::flextable(suppl_tabs$lft_correlation_uni) %>% 
  set_widths(c(2.5, 2.5, 1.8, 4, 2.7)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for al lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Blocked bootstrap Spearman correlation test; \u03C1 correlation coefficients with 95% confidence intervals. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

```

\newpage

# Supplementary Figures

```{r fig-ct-finding-course, fig.width = figur::convert(suppl_figs$ct_finding_course, to = 'in')$w, fig.height = figur::convert(suppl_figs$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of the most common abnormalities in computed tomography of the chest at the consecutive follow-ups in participants stratified by severity of acute COVID-19.'}

suppl_figs$ct_finding_course$plot

```

__Supplementary Figure S\@ref(fig:fig-ct-finding-course). Frequency of the most common abnormalities in computed tomography of the chest at the consecutive follow-ups in participants stratified by severity of acute COVID-19.__

_Frequencies of any chest computed tomography (CT) abnormalities, ground glass opacity (GGO), and reticulations identified by a human radiologist at the consecutive follow-ups after COVID-19 in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stack plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars._

\newpage

```{r fig-ct-numeric-course, fig.width = figur::convert(suppl_figs$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(suppl_figs$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in participants stratified by severity of acute COVID-19.'}

suppl_figs$ct_numeric_course$plot

```

__Supplementary Figure S\@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in participants stratified by severity of acute COVID-19.__

_Severity of chest computed tomography (CT) abnormalities was assessed by a human-determined CT severity score (CTSS), and AI-determined percentages of the lung tissue with opacity and high opacity. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points._

\newpage

```{r fig-ct-corr, fig.width = figur::convert(suppl_figs$ct_corr, to = 'in')$w, fig.height = figur::convert(suppl_figs$ct_corr, to = 'in')$h, fig.cap = 'Co-occurrence of abnormalities of chest computed tomography and correlation of computed tomography readouts in COVID-19 convalescents.'}

suppl_figs$ct_corr$plot

```

__Supplementary Figure S\@ref(fig:fig-ct-corr). Co-occurrence of abnormalities of chest computed tomography and correlation of computed tomography readouts in COVID-19 convalescents.__

_(A) Co-occurrence of ground glass opacity (GGO), reticulation, and consolidation diagnosed by a human radiologist was investigated by two-dimensional correspondence analysis. Left: plots of the column factors, proximity of points reflects co-occurrence of normal or pathological values of lung function testing. Right: Venn plot of frequencies of GGO, reticulation, and consolidation; the total observation number is displayed in the plot caption._

_(B) Correlation between human-determined (CTSS: computed tomography severity score, sum of points form the entire lung) and artificial intelligence determined (AI, opacity and high opacity, percentage of the lung) readouts of severity of structural lung lesions was assessed by blocked bootstrap Spearman's test and visualized in scatter plots of observation ranks. Each point represents a single observation, linear trends with standard errors are depicted as blue lines with gray ribbons. Values of correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._

\newpage

```{r fig-lft-finding-course, fig.width = figur::convert(suppl_figs$lft_finding_course, to = 'in')$w, fig.height = figur::convert(suppl_figs$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in participants stratified by severity of acute COVID-19.'}

suppl_figs$lft_finding_course$plot

```

__Supplementary Figure S\@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in participants stratified by severity of acute COVID-19.__

_Insufficient diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC), and forced expiratory volume in one second (FEV1) was defined as values < 80% of the individual's reference. Frequencies of these abnormalities in in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stack plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars._

\newpage

```{r fig-lft-numeric-course, fig.width = figur::convert(suppl_figs$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(suppl_figs$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in participants stratified by severity of acute COVID-19.'}

suppl_figs$lft_numeric_course$plot

```

__Supplementary Figure S\@ref(fig:fig-lft-numeric-course). Time course of numeric lung function testing readouts at the consecutive follow-ups in participants stratified by severity of acute COVID-19.__

_Diffusion capacity for carbon monoxide (DLCO), forced vital capacity (FVC), forced expiratory volume in one second (FEV1) were recorded a percentages of the patient's reference values. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points._

\newpage

```{r fig-lft-corr, fig.width = figur::convert(suppl_figs$lft_corr, to = 'in')$w, fig.height = figur::convert(suppl_figs$lft_corr, to = 'in')$h, fig.cap = 'Co-occurrence of abnormalities of lung function testing and correlation of lung function readouts in COVID-19 convalescents.'}

suppl_figs$lft_corr$plot

```

__Supplementary Figure S\@ref(fig:fig-lft-corr). Co-occurrence of abnormalities of lung function testing and correlation of lung function readouts in COVID-19 convalescents.__

_(A) Co-occurrence of insufficient diffusion capacity for carbon monoxide (DLCO < 80% of reference value), insufficient forced vital capacity (FVC < 80% of reference), and insufficient forced expiratory volume in one second (FEV1 < 80% of reference) was investigated by two-dimensional correspondence analysis. Left: plots of the column factors, proximity of points reflects co-occurrence of normal or pathological values of lung function testing. Right: Venn plot of frequencies of insufficient DLCO, FVC, and FEV1; the total observation number is displayed in the plot caption._

_(B) Correlation between the lung function testing parameters was assessed by blocked bootstrap Spearman's test and visualized in scatter plots of observation ranks. Each point represents a single observation, linear trends with standard errors are depicted as blue lines with gray ribbons. Values of correlation coefficients with 95% confidence intervals and p values are displayed in the plot captions._

\newpage

```{r fig-symptom-course, fig.width = figur::convert(suppl_figs$symptom_course, to = 'in')$w, fig.height = figur::convert(suppl_figs$symptom_course, to = 'in')$h, fig.cap = 'Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in participants stratified by severity of acute COVID-19.'}

suppl_figs$symptom_course$plot

```

__Supplementary Figure S\@ref(fig:fig-symptom-course). Frequency of persistent symptoms of relevance for lung function at the consecutive follow-ups in participants stratified by severity of acute COVID-19.__

_Intensity of self-reported dyspnea was assessed with the modified Medical Research Council scale (mMRC) and presence of dyspnea was defined as mMRC > 0. Presence of cough was surveyed by a single 'yes'/'no' item. Impairment of physical performance was assessed with the Eastern Cooperative Oncology Group scale (ECOG, high values indicate more profound impairment) and impaired physical performance was defined as ECOG > 0. Frequency of dyspnea, cough and impairment of physical performance at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars._

\newpage

```{r fig-modeling-strategy, fig.width = figur::convert(suppl_figs$modeling_strategy, to = 'in')$w, fig.height = figur::convert(suppl_figs$modeling_strategy, to = 'in')$h, fig.cap = 'Modeling strategy.'}

suppl_figs$modeling_strategy$plot

```

__Supplementary Figure S\@ref(fig:fig-modeling-strategy). Modeling strategy.__

_DLCO: diffusion capacity for carbon dioxide; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second; CT: computed tomography; GGO: ground glass opacity; CTSS: human-determined CT severity score; AI opacity and high opacity: artificial intelligence-determined opacity and high opacity of the lung; AUC: area under the receiver-operating characteristic curve._

\newpage

```{r fig-fvc-fev-reduced-models, fig.width = figur::convert(suppl_figs$fvc_fev_reduced_models, to = 'in')$w, fig.height = figur::convert(suppl_figs$fvc_fev_reduced_models, to = 'in')$h, fig.cap = 'Evaluation of performance of machine learning classification models at prediction of insufficient forced vital capacity and forced expiratory volume in one second.'}

suppl_figs$fvc_fev_reduced_models$plot

```

__Supplementary Figure S\@ref(fig:fig-fvc-fev-reduced-models). Evaluation of performance of machine learning classification models at prediction of insufficient forced vital capacity and forced expiratory volume in one second.__

_Machine learning classification models of insufficient forced vital capacity (A, FVC < 80% of reference value: n = `r ngroups(bin_models$models$FVC_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$FVC_reduced[[1]])$train$n_outcome %>% sum`) and of insufficient forced expiratory volume in one second (B, FEV1 < 80% of reference value: n = `r ngroups(bin_models$models$FEV1_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$FEV1_reduced[[1]])$train$n_outcome %>% sum`) employing computed tomography readouts, demographic and clinical explanatory variables were trained. The model performance was evaluated in the entire data set and 10-repeats 10-fold cross-validation with overall accuracy metric, Cohen's $\kappa$ as a measure of concordance between predicted and observed outcome, and Brier score as a measure of model's calibration. Left: numeric performance measures of the models (open circles: the entire data set, filled circles: cross-validation); point sizes and point labels represent overall model accuracy, the dashed lines visualize values of Cohen's $\kappa$ and Brier score expected for prediction of insufficient DLCO be chance. Right: receiver-operating characteristic curves for predictions in cross-validation folds, numeric statistics are displayed in the plot. Numbers of complete observations and observations with the particular LFT insufficiency ('events') are displayed in the plot captions._

_FVC: forced vital capacity; FEV1: forced expiratory volume in one second, CV: cross-validation; AUC: are under the receiver-operating characteristic curve; Se: sensitivity; Sp: specificity; GBM: gradient boosted machines; SVM radial: support vector machines with radial kernel._

\newpage

```{r fig-fvc-fev-models, fig.width = figur::convert(suppl_figs$fvc_fev_models, to = 'in')$w, fig.height = figur::convert(suppl_figs$fvc_fev_models, to = 'in')$h, fig.cap = 'Evaluation of performance of machine learning regression models at prediction of forced vital capacity and forced expiratory volume in one second.'}

suppl_figs$fvc_fev_models$plot

```

__Supplementary Figure S\@ref(fig:fig-fvc-fev-models). Evaluation of performance of machine learning regression models at prediction of forced vital capacity and forced expiratory volume in one second.__

_Machine learning regression models of forced vital capacity (FVC, total observations: n = `r nobs(reg_models$models$FVC_percent[[1]])`) and of forced expiratory volume in one second (FEV1,  total observations: n = `r nobs(reg_models$models$FVC_percent[[1]])`) employing computed tomography readouts, demographic and clinical explanatory variables were trained. Their performance was evaluated in th entire data set and 10-repeats 10-fold cross-validation with $R^2$ as a measure of explained variation, mean absolute error, and $\rho$ Spearman's coefficient of correlation between the predicted and observed values. Numeric performance measures of the model are presented in bubble plots (open circles: the entire data set, filled circles: cross-validation); point sizes and point labels represent values of $\rho$ correlation coefficient, the dashed line visualizes $R^2$ value expected for a meaningless model. Numbers of complete observations are shown in the plot captions._

_FVC: forced vital capacity; FEV1: forced expiratory volume in one second, CV: cross-validation; MAE: mean absolute error; GBM: gradient boosted machines; SVM radial: support vector machines with radial kernel._

\newpage

```{r fig-residuals, fig.width = figur::convert(suppl_figs$residuals, to = 'in')$w, fig.height = figur::convert(suppl_figs$residuals, to = 'in')$h, fig.cap = "Cohen's kappa and mean absolute error of the machine learning models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide."}

suppl_figs$residuals$plot

```

__Supplementary Figure S\@ref(fig:fig-residuals). Cohen's $\kappa$ and mean error of the machine learning models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide in observations stratified by acute COVID-19 severity and follow-up visit.__

_Classification machine learning models of insufficient diffusion capacity for carbon monoxide (A, DLCO < 80% of reference value: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome %>% sum`) and regression models of diffusion capacity for carbon monoxide (DLCO, % of reference value, total observations: n = `r nobs(reg_models$models$DLCO_percent[[1]])`) were trained and evaluated as presented in Figure 2. Cohen's $\kappa$ (A) and mean errors (B) of cross-validated model predictions for observations stratified by severity of acute COVID-19 and the consecutive follow-up visits were investigated. Cohen's $\kappa$ and error values are visualized in heat maps; tiles are labeled with the error values. Global $\kappa$ and mean absolute model errors are indicated in the plot captions._

_DLCO: diffusion capacity for carbon monoxide; SVM radial: support vector machines with radial kernel; GBM: gradient boosted machines; MAE: mean absolute error; A: ambulatory COVID-19; HM: hospitalized, moderate COVID-19; HS: hospitalized, severe COVID-19.._

\newpage

```{r fig-varimp-dlco, fig.width = figur::convert(suppl_figs$varimp_dlco, to = 'in')$w, fig.height = figur::convert(suppl_figs$varimp_dlco, to = 'in')$h, fig.cap = 'Explanatory variable importance for models of diffusion capacity for carbon monoxide measured by Shapley additive explanations.'}

suppl_figs$varimp_dlco$plot

```

__Supplementary Figure S\@ref(fig:fig-varimp-dlco). Explanatory variable importance for models of diffusion capacity for carbon monoxide measured by Shapley additive explanations.__

_Importance of explanatory variables for the machine learning models of diffusion capacity for carbon monoxide (% of reference, Figure 2) was investigated by Shapley additive explanations (SHAP). Absolute SHAP values for explanatory variables with the 15 largest mean SHAP values are presented in violin plots. Points represent single observations, point colors code for minimum/maximum scaled value of the explanatory variable. Explanatory variables obtained via computed tomography are highlighted with bold font in the Y axes._

_CT: computed tomography; DLCO: diffusion capacity for carbon monoxide; BMI: body mass index; opacity and high opacity, AI: opacity and high opacity of the lung determined by artificial intelligence; CTSS: human-determined CT severity score, sum for all lobe; ECOG: Eastern Cooperative Oncology Group physical performance score; mMRC: modified Medical Research Council dyspnea scale; ICU: intensive care unit._

\newpage

# References

### 2025-01-18 paper/markdown/manuscript.Rmd
Changes in Figure legends

@@
 
-_(A) Four machine learning classification models of insufficient diffusion capacity for carbon monoxide (< 80% of reference value: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome %>% sum`) employing computed tomography readouts, demographic and clinical explanatory variables were trained.
+_(A) Four machine learning classification models of insufficient diffusion capacity for carbon monoxide (< 80% of reference value: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome[2]`, total observations: n = `r ngroups(bin_models$models$DLCO_reduced[[1]])$train$n_outcome %>% sum`) employing time after COVID-19 diagnosis, computed tomography readouts, demographic and clinical explanatory variables were trained.
 Their performance was evaluated in the entire data set and 10-repeats 10-fold cross-validation with overall accuracy metric, Cohen's $\kappa$ as a measure of concordance between predicted and observed outcome, and Brier score as a measure of model's calibration.
@@
 
-_(B) Four machine learning regression models of diffusion capacity for carbon monoxide (percentage of reference values, total observations: n = `r nobs(reg_models$models$DLCO_percent[[1]])`) employing computed tomography readouts, demographic and clinical explanatory variables were trained.
+_(B) Four machine learning regression models of diffusion capacity for carbon monoxide (percentage of reference values, total observations: n = `r nobs(reg_models$models$DLCO_percent[[1]])`) employing time after COVID-19 diagnosis, computed tomography readouts, demographic and clinical explanatory variables were trained.
 Their performance was evaluated in th entire data set and 10-repeats 10-fold cross-validation with $R^2$ as a measure of explained variation, mean absolute error, and $\rho$ Spearman's coefficient of correlation between the predicted and observed values.
@@
 Scatter plots: observed and predicted values of diffusion capacity for carbon monoxide in cross-validation folds; the blue dashed lines with slope 1 and intercept 0 represent absolutely accurate predictions, general additive model trends with standard errors are visualized as the solid blue lines with gray ribbons.
-Numbers of complete observations and Spearman's $\rho$ coeffficients of correlation between the predicted and observed values are displayed in the plot captions._
+Numbers of complete observations and Spearman's $\rho$ coefficients of correlation between the predicted and observed values are displayed in the plot captions._
 
@@
 
-_(A) Values of the radiological readouts of lung damage severity were compared between data points with and without insufficient diffusion capacity for carbon monoxide by Mann-Whitney test with r effect size statistic.
+_(A) Values of the radiological readouts of lung damage severity were compared between data points with and without insufficient diffusion capacity for carbon monoxide by blocked bootstrap test with r effect size statistic.
 Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points.
@@
 _(B) Quality of detection of insufficient diffusion capacity for carbon monoxide with the radiological readouts of lung damage severity was assessed by receiver-operating characteristic (ROC) analysis.
-ROC curves are shown, the optimal cutoffs of the severity readuts determined by Youden's criterion are represented by points with numbers.
+ROC curves are shown, the optimal cutoffs of the severity readouts determined by Youden's criterion are represented by points with numbers.
 Sensitivity, specificity at the optimal cutoff, and area under the curve statistic with 95% confidence interval are displayed in the plots._

### 2025-03-04 paper/markdown/supplementary_material.Rmd
Revision analyses
Learning curves for ML models of DLCO and insufficient DLCO. Correlation graphs for the top explanatory variables for predictions of DLCO.

@@
 
-* biseral r was used for two-group comparisons of numeric variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size
+* biserial r was used for two-group comparisons of numeric variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size
 
@@
 Mann-Whitney test) and required instead a special approach.
-Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biseral r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality).
+Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biserial r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality).
 The blocked bootstrap scheme involved re-sampling with repetition of the participants instead of single observations (B = 2000 re-samples). 
@@
 
-```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm parameters by cross-validation-mediated tuning.'}
+```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm hyper-parameters by cross-validation-mediated tuning.'}
 
@@
   footnote(1, 5:6, 
-           value = as_paragraph('Blocked bootstrap test with biseral r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
+           value = as_paragraph('Blocked bootstrap test with biserial r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
            part = 'header', 
@@
 
+```{r fig-learn-curves, fig.width = figur::convert(suppl_figs$learn_curves, to = 'in')$w, fig.height = figur::convert(suppl_figs$learn_curves, to = 'in')$h, fig.cap = 'Learning curves of the gradient boosted machines models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.'}
+
+suppl_figs$learn_curves$plot
+
+```
+
+__Figure \@ref(fig:fig-learn-curves). Learning curves of the gradient boosted machines models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.__
+
+_Gradient boosted machines (GBM) was the best performing algorithm at modeling of insufficiency of diffusion capacity for carbon monoxide (DLCO < 80% of reference) and diffusion capacity for carbon monoxide expressed as percentage of the reference value (DLCO, Figure 2)._
+_Over-fitting behavior of the GBM models was investigated by examination of learning curves, i.e.
+curves of model performance metrics (A, DLCO < 80%: accuracy and Cohen's $\kappa$, B, DLCO: mean absolute error [MAE] and $R^2$) for models trained in subsets of observations with increasing sizes._
+_The performance metrics were computed for the training subsets, test subsets (size of each corresponded to 25% of the training subsets) and 10-repeats 10-fold cross-validation, and were visualized in scatter plots.
+Each point represents a single prediction in the training or test subset, or an out-of-fold prediction in the cross-validation.
+Locally estimated scatterplot smoothing (LOESS) trends with standard error regions are visualized as solid lines with gray ribbons._
+
+\newpage
+
 ```{r fig-varimp-dlco, fig.width = figur::convert(suppl_figs$varimp_dlco, to = 'in')$w, fig.height = figur::convert(suppl_figs$varimp_dlco, to = 'in')$h, fig.cap = 'Explanatory variable importance for models of diffusion capacity for carbon monoxide measured by Shapley additive explanations.'}
@@
 
+```{r fig-top-corr, fig.width = figur::convert(suppl_figs$top_corr, to = 'in')$w, fig.height = figur::convert(suppl_figs$top_corr, to = 'in')$h, fig.cap = 'Co-linearity of the most influential computed tomography-related variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.'}
+
+suppl_figs$top_corr$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-top-corr).
+Co-linearity of the most influential computed tomography-related variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.__
+
+_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco).__
+
+_(A) Weighted graph of correlations between the most influential explanatory variables.
+The graph edges are defined by correlations with Kendall's $\tau \geq$ 0.3, the edge weights correspond to the $\tau$ values.
+Isolated vertices, i.e.
+explanatory variables without any correlations with $\tau \geq$ 0.3, were removed.
+The graph was visualized with the Fruchterman-Reingold algorithm.
+The explanatory variables, i.e.
+vertices of the graph, are depicted as points; point color codes for classification of the explanatory variable.
+Edges are represented by lines; line width and color codes for $\tau$ value._
+
+_(B) Correlations between computed tomography (CT) readouts of lung damage, the human-determined CT severity score and software-determined lung opacity and high opacity, were assessed by blocked bootstrap Kendall's correlation test.
+Ranks of the CT readouts are presented in scatter plots.
+Each point represents a single observation, blue straight lines are visualizations of linear trends.
+Correlation coefficients $\tau$ with bootstrap confidence intervals, p values, and numbers of observations are displayed in the plot captions._
+
+_(C) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations.
+Statistical significance was determined by blocked bootstrap test with r effect size statistic.
+Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of observations are indicated in the X axes._
+
+\newpage
+
 # References

### 2025-03-04 LFT scripts/bootstrap_tests.R

@@ -198,7 +198,7 @@
     format_summ_tbl(dict = lft_globals$lexicon) %>% 
     mutate(variable = ifelse(is.na(variable), 
                              'Observations, n', variable), 
-           lft_abnormality = exchange(lft_abnormality, 
+           lft_abnormality = exchange(as.character(lft_abnormality), 
                                       lft_globals$lexicon)) %>% 
     select(lft_abnormality, 
            variable, 

### 2025-03-04 LFT scripts/globals.R

@@ -147,6 +147,33 @@
   lft_globals$lexicon[lft_globals$lexicon$variable == 'severity_WHO', 'format'] <- 
     'factor'
 
+  ## adding classification of the explanatory variables
+  
+  lft_globals$lexicon <- lft_globals$lexicon %>% 
+    mutate(class = ifelse(variable %in% lft_globals$responses, 
+                          'response', 
+                          ifelse(variable %in% lft_globals$ct_variables, 
+                                 'CT readouts and findings', 
+                                 ifelse(variable %in% lft_globals$symptom_variables, 
+                                        'symptoms', 
+                                        ifelse(variable %in% lft_globals$baseline_variables, 
+                                               'risk factors and CoV severity', 
+                                               'time after CoV')))), 
+           class = factor(class, 
+                          c('risk factors and CoV severity', 
+                            'symptoms', 
+                            'time after CoV', 
+                            'CT readouts and findings')))
+  
+  ## colors of the variable classes
+  
+  lft_globals$class_colors <- 
+    c('risk factors and CoV severity' = 'steelblue', 
+      'symptoms' = 'plum4', 
+      'time after CoV' = 'aquamarine3', 
+      'CT readouts and findings' = 'orangered3')
+  
+  
 # CV folds -------
 
   insert_msg('CV folds')

### 2025-03-04 LFT scripts/learning_curves.R

@@ -1,0 +1,97 @@
+# Learning curves for models of reduced and DLCO and DLCO percent of reference. 
+# 
+# The general principle: testing the quality of model fit for increasing sizes 
+# of the training data; the cost function values are computed for the training 
+# data sets, hold-oiut (25% of the training data size), and 10-repeats 10-fold 
+# cross-validation. 
+#
+# Working with the `caret`-native `learning_curve_dat()` function. 
+# Unfortunately, the computation crashes for some algorithms (like SVM). 
+# Additionally, there's no way to implement the patient-blocked hold-out and 
+# and cross-validation designs. Hence, the cost values are likely too optimistic. 
+
+  insert_head()
+  
+# container --------
+  
+  lft_learn <- list()
+  
+# analysis globals ---------
+  
+  insert_msg('Analysis globals')
+
+  ## the train control objects
+  
+  lft_learn$train_control$DLCO_reduced <- 
+    trainControl(method = 'repeatedcv', 
+                 number = 10, 
+                 repeats = 10, 
+                 summaryFunction = multiClassSummary)
+  
+  lft_learn$train_control$DLCO_percent <- 
+    trainControl(method = 'repeatedcv', 
+                 number = 10, 
+                 repeats = 10)
+  
+  ## the models of DLCO < 80% of the reference, 
+  ## and DLCO as reference's percentage
+  
+  lft_learn$models <- 
+    list(DLCO_reduced = bin_models$models$DLCO_reduced, 
+         DLCO_percent = reg_models$models$DLCO_percent)
+  
+  ## the best tunes and analysis data
+  
+  lft_learn$best_tunes <- lft_learn$models %>% 
+    map(map, ~.x$bestTune)
+  
+  lft_learn$data <- lft_learn$models %>% 
+    map(map, ~.x$trainingData)
+  
+  ## algorithm names
+  
+  lft_learn$algos <- lft_learn$models %>% 
+    map(names)
+  
+  lft_learn$algos <- lft_learn$algos %>% 
+    map(~set_names(.x, .x))
+
+# computation of the cross-validated learning curves ------
+  
+  insert_msg('computation of the learning curves')
+  
+  for(i in names(lft_learn$data)) {
+    
+    registerDoParallel(cores = 8)
+    
+    ## safely: computation crashes for some algorithms
+    
+    lft_learn$curves[[i]] <- lft_learn$algos[[i]] %>% 
+      map(~safely(learning_curve_dat)(dat = lft_learn$data[[i]][[.x]], 
[…]

### 2025-03-04 LFT scripts/learning_plots.R

@@ -1,0 +1,76 @@
+# Plots of the learning curves: statistics of performance of machine learning 
+# models as a function of size of the training data set. Performance in the 
+# training data portion, a hold-out test subset and 10-repeats 10 fold CV 
+# is evaluated.
+
+  insert_head()
+  
+# container -------
+  
+  lft_lcurves <- list()
+  
+# analysis globals: learning curve data -------
+  
+  insert_msg('Analysis globals')
+
+  ## metrics of performance and evaluation data sets
+  
+  lft_lcurves$metrics <- 
+    list(DLCO_reduced = c('Accuracy', 'Kappa'), 
+         DLCO_percent = c('MAE', 'Rsquared'))
+  
+  lft_lcurves$metric_labels <- 
+    c(Accuracy = 'Accuracy', 
+      Kappa = "Cohen's \u03BA", 
+      MAE = 'MAE', 
+      Rsquared = 'R\u00B2')
+  
+  lft_lcurves$sets <- c('Training', 'Testing', 'Resampling')
+  
+  lft_lcurves$set_labels <- 
+    c(Training = 'training subset', 
+      Testing = 'test subset', 
+      Resampling = 'cross-validation')
+  
+  lft_lcurves$set_colors <- 
+    c(Training = globals$dataset_colors[["train"]], 
+      Testing = 'aquamarine3', 
+      Resampling = globals$dataset_colors[["cv"]])
+  
+  ## learning curve data
+  
+  lft_lcurves$analysis_tbl <- lft_learn$curves %>% 
+    map(map, 
+        mutate, 
+        Data = factor(Data, lft_lcurves$sets))
+  
+# The plots --------
+  
+  insert_msg('The plots')
+  
+  for(i in names(lft_lcurves$analysis_tbl)) {
+    
+    lft_lcurves$plots[[i]] <- 
+      list(data = lft_lcurves$analysis_tbl[[i]], 
+           plot_title = exchange(i, lft_globals$lexicon) %>% 
+             paste(globals$algo_labs[names(lft_lcurves$analysis_tbl[[i]])], 
+                   sep = ', ')) %>% 
+      pmap(plot_learn_curves, 
+           metrics = lft_lcurves$metrics[[i]], 
+           metric_labels = lft_lcurves$metric_labels,
+           data_labels = lft_lcurves$set_labels, 
+           data_colors = lft_lcurves$set_colors, 
+           method = 'loess', 
[…]

### 2025-03-04 LFT scripts/shap_importance.R

@@ -119,12 +119,24 @@
           globals$common_theme + 
           theme(axis.text.y = element_markdown()))
   
-# A table with mean SHAP values ---------
+# A table with mean absolute SHAP values for the top most influential variables ---------
   
-  insert_msg('A table with mean absolute SHAP values')
+  insert_msg('A table with mean absolute SHAP values, top most influential')
+  
+  ## the top 15 most important explanatory variables, 
+  ## as defined by mean absolute SHAP values
   
   shap_imp$mean_table <- shap_imp$bar_plots %>% 
-    map(map, ~.x$data) %>% 
+    map(map, ~.x$data)
+  
+  shap_imp$top_features <- shap_imp$mean_table %>% 
+    map(map, ~.x$feature) %>% 
+    map(map, as.character)
+  
+  ## mean absolute SHAP values for the top 15 most
+  ## important features
+  
+  shap_imp$mean_table <- shap_imp$mean_table %>% 
     map(~map2(., names(.), 
               ~set_names(.x, c('Variable', globals$algo_labs[.y])))) %>% 
     map(reduce, full_join, by = 'Variable') %>% 

### 2025-03-04 LFT scripts/shap_pca.R

@@ -1,0 +1,97 @@
+# Principal component analysis for SHAP importance measures of 
+# the explanatory factors for numeric DLCO and DLCO. 
+# For the sake of simplicity, we're analysing just the top 15 most important 
+# features as gauged by mean absolute SHAP.
+# Pre-processing 
+
+  insert_head()
+  
+# container -------
+  
+  shap_pca <- list()
+  
+# analysis globals --------
+  
+  insert_msg('Analysis globals')
+  
+  ## the top most influential explanatory factors
+  
+  shap_pca$top_features <- shap_imp$top_features
+  
+  ## analysis data: SHAP values for the variables and observations, 
+  ## row names set for smooth identification of observations 
+  ## by tools for `clust_tools`. The data are median centered
+  
+  shap_pca$analysis_tbl <- shap_imp$shap_objects %>% 
+    map(map, ~.x$S) %>% 
+    map(map, as.data.frame) %>% 
+    map(map, 
+        ~set_rownames(.x, paste0('obs_', 1:nrow(.x)))) %>% 
+    map(map, center_data, type = 'median')
+  
+  shap_pca$analysis_tbl <- 
+    map2(shap_pca$analysis_tbl, 
+         shap_pca$top_features, 
+         function(x, y) map2(x, y, 
+                             ~select(.x, all_of(.y))))
+  
+# PCA objects -------
+  
+  insert_msg('PCA objects')
+  
+  ## predictions of DLCO findings: 2 - 3 components that explain the majority
+  ## of the variance. 
+  ## analogically 3 - 4 principal components for SHAPs for prediction of 
+  ## DLCO percentages
+
+  shap_pca$pca_obj <- shap_pca$analysis_tbl %>%
+    map(map, 
+        reduce_data, 
+        kdim = 6, 
+        red_fun = 'pca')
+
+# Plots for the PCA objects -------
+  
+  insert_msg('Plots for the PCA objects')
+  
+  ## scree plots
+  
+  for(i in names(shap_pca$pca_obj)) {
+    
+    shap_pca$scree_plots[[i]] <- 
+      list(x = shap_pca$pca_obj[[i]]) %>% 
+      pmap(plot, 
+           type = 'scree', 
+           cust_theme = globals$common_theme) %>% 
+      map(tag2subtitle)
+    
+    ## styling
+    
+    shap_pca$scree_plots[[i]] <- 
[…]

### 2025-03-04 LFT scripts/variable_networks.R

@@ -1,0 +1,118 @@
+# Principal component analysis for the top explanatory variables for predictions 
+# of reduced DLCO and of DLCO percentage. 
+# Pre-processing: median centering after conversion of factors to integers
+
+  insert_head()
+  
+# container ------
+  
+  lft_net <- list()
+  
+# variables of interest and analysis table --------
+  
+  insert_msg('Top variables')
+  
+  ## the top variables: top 15 most important variables (mean absolute SHAP)
+  
+  lft_net$top_variables <- shap_imp$top_features %>% 
+    map(reduce, union)
+  
+  lft_net$top_variables$all <- reduce(lft_net$top_variables, union)
+  
+  lft_net$analysis_tbl <- lft_net$top_variables %>% 
+    map(~select(lft_globals$analysis_tbl, all_of(.x))) %>%
+    map(map_dfc, function(x) if(is.factor(x)) as.integer(x) else x) %>% 
+    map(set_rownames, 
+        paste0('obs_', 1:nrow(lft_globals$analysis_tbl)))
+  
+  ## names of the graphs
+  
+  lft_net$graph_titles <- 
+    paste('Top explanatory factors,', 
+          c(exchange(names(lft_net$analysis_tbl)[1:2], 
+                     lft_globals$lexicon), 
+            'DLCO < 80% and DLCO'))
+  
+# Correlation graphs --------
+  
+  insert_msg('Correlation graphs')
+  
+  lft_net$graph_obj <- lft_net$analysis_tbl %>% 
+    map(as_iGraph, 
+        fun = cor, 
+        method = 'kendall', 
+        cutoff = 0.3, 
+        diag = FALSE, 
+        weighted = TRUE) %>% 
+    map(prune_degree, cutoff = 0)
+  
+  ## setting the attributes: 
+  ## reader-friendly labels of the vertices, and 
+  ## classification of the explanatory factors
+  
+  lft_net$attr_tbl <- lft_net$graph_obj %>% 
+    map(get_vertex_attributes) %>% 
+    map(mutate, 
+        variable_label = exchange(name, lft_globals$lexicon), 
+        variable_class = exchange(name, lft_globals$lexicon, value = 'class')) %>% 
+    map(select, -index)
+
+  lft_net$graph_obj <- 
+    map2(lft_net$graph_obj, 
+         lft_net$attr_tbl, 
+         set_vertex_attributes)
+  
+# Fruchterman-Reingold network plots --------
+  
[…]

### 2025-03-04 LFT scripts/variable_pca.R

@@ -1,0 +1,77 @@
+# Principal component analysis for the top explanatory variables for predictions 
+# of reduced DLCO and of DLCO percentage. 
+# Pre-processing: median centering after conversion of factors to integers
+
+  insert_head()
+  
+# container ------
+  
+  lft_pca <- list()
+  
+# variables of interest and analysis table --------
+  
+  insert_msg('Top variables')
+  
+  ## the top variables: top 15 most important variables (mean absolute SHAP)
+  
+  lft_pca$top_variables <- shap_imp$top_features %>% 
+    map(reduce, union)
+  
+  lft_pca$analysis_tbl <- lft_pca$top_variables %>% 
+    map(~select(lft_globals$analysis_tbl, all_of(.x))) %>%
+    map(map_dfc, function(x) if(is.factor(x)) as.integer(x) else x) %>% 
+    map(center_data, 'median') %>% 
+    map(set_rownames, 
+        paste0('obs_', 1:nrow(lft_globals$analysis_tbl)))
+  
+# PCA objects -------
+  
+  insert_msg('PCA objects')
+  
+  ## there are 2 - 3 components that explain the majority of the variance
+  
+  lft_pca$pca_obj <- lft_pca$analysis_tbl %>%
+    map(reduce_data, 
+        kdim = 6, 
+        red_fun = 'pca')
+  
+# Plots --------
+  
+  insert_msg('Plots')
+
+  ## scree plots 
+  
+  lft_pca$scree_plots <- 
+    list(x = lft_pca$pca_obj) %>% 
+    pmap(plot, 
+         type = 'scree', 
+         cust_theme = globals$common_theme) %>% 
+    map(tag2subtitle)
+  
+  lft_pca$scree_plots <- 
+    list(x = lft_pca$scree_plots, 
+         y = paste('Top explanatory variables,', 
+                   exchange(names(lft_pca$scree_plots), 
+                            dict = lft_globals$lexicon))) %>% 
+    pmap(function(x, y) x + 
+           labs(title = y))
+  
+  ## plots of the loadings
+  
+  lft_pca$loading_plots <- 
+    list(x = lft_pca$pca_obj, 
+         plot_title = paste('Top explanatory variables,', 
+                            exchange(names(lft_pca$scree_plots), 
+                                     dict = lft_globals$lexicon)), 
+         point_color = c('orangered3', 'orangered4')) %>% 
[…]

### 2025-03-04 exploration scripts/comparison.R

@@ -1,0 +1,165 @@
+# Bootstrap tests for comparison of opacity, high opacity and CTSS between 
+# observations with and without lung CT findings
+
+  insert_head()
+  
+# container ------
+  
+  comp_ct <- list()
+  
+# parallel backend --------
+  
+  insert_msg('Parallel backend')
+
+  plan('multisession')
+    
+# analysis globals: variables and their values ------
+  
+  insert_msg('Analysis globals')
+  
+  ## variables of interest
+  
+  comp_ct$ct_findings <- 
+    c('GGO_finding', 'reticulation_finding', 'consolidation_finding')
+  
+  comp_ct$ct_severity <- 
+    c('CTSS', 'opacity_percent', 'high_opacity_percent')
+  
+  ## and the corresponding data
+  
+  comp_ct$analysis_tbl <- covild$ct %>% 
+    select(ID, 
+           all_of(comp_ct$ct_findings), 
+           all_of(comp_ct$ct_severity))
+  
+  ## variable lexicon 
+  
+  comp_ct$lexicon <- covild$ct_lexicon %>% 
+    filter(variable %in% c(comp_ct$ct_findings, comp_ct$ct_severity))
+  
+# Descriptive stats -------
+  
+  insert_msg('Descriptive stats')
+  
+  comp_ct$stats <- comp_ct$ct_findings %>% 
+    set_names(comp_ct$ct_findings) %>% 
+    map(~explore(comp_ct$analysis_tbl, 
+                 variables = comp_ct$ct_severity, 
+                 split_factor = .x, 
+                 what = 'table', 
+                 pub_styled = TRUE)) %>% 
+    map(reduce, left_join, by = 'variable') %>% 
+    map(set_names, c('variable', 'no', 'yes')) %>% 
+    compress(names_to = 'ct_abnormality')
+  
+# Bootstrap tests -------
+  
+  insert_msg('Bootstrap tests')
+  
+  set.seed(12345)
+  
+  for(i in comp_ct$ct_findings) {
+    
+    comp_ct$test[[i]] <- comp_ct$ct_severity %>% 
+      map_dfr(~delta_median(data = comp_ct$analysis_tbl, 
+                            variable = .x, 
+                            split_factor = i, 
+                            B = 2000))
+    
+  }
+  
+  ## multiple testing adjustment and plot_captions
+  
+  comp_ct$test <- comp_ct$test %>%
[…]

### 2025-03-04 exploration scripts/correlation.R

@@ -1,5 +1,5 @@
 # Correlation of numeric LFT readouts and CT readouts. Because the observations 
-# are not independent, we're resorting to a blocked bootstrap Spearman's test.
+# are not independent, we're resorting to a blocked bootstrap Kendall's test.
 
   insert_head()
   
@@ -41,7 +41,8 @@
 
   corr_lft$titles <- corr_lft$pairs %>% 
     map(map, exchange, corr_lft$lexicon) %>% 
-    map(map, paste, collapse = ' and ')
+    map(map, paste, collapse = ' and ') %>% 
+    map(map, capitalize_first_char)
   
 # Correlation coefficients ---------
   
@@ -50,12 +51,13 @@
   for(i in names(corr_lft$pairs)) {
     
     corr_lft$test[[i]] <- corr_lft$pairs[[i]] %>% 
-      map_dfr(spearman_rho, 
+      map_dfr(corr_test, 
               data = corr_lft$analysis_tbl[[i]], 
               B = 2000, 
-              positive = TRUE) %>% 
+              positive = TRUE, 
+              method = 'kendall') %>% 
       re_adjust %>% 
-      mutate(est_lab = paste0('\u03C1 = ', signif(rho, 2), 
+      mutate(est_lab = paste0('\u03C4 = ', signif(rho, 2), 
                               ' [', signif(lower_ci, 2), 
                               ' - ', signif(upper_ci, 2), ']'), 
              n = nrow(corr_lft$analysis_tbl[[i]]), 

### 2025-03-04 exploration.R

@@ -11,6 +11,7 @@
   library(exda)
   library(rstatix)
   library(rcompanion)
+  library(fastTest)
 
   library(clustTools)
   library(ggvenn)
@@ -37,7 +38,8 @@
     './exploration scripts/distribution_longitudinal.R', 
     './exploration scripts/distribution_baseline.R', 
     './exploration scripts/correspondence.R', 
-    './exploration scripts/correlation.R') %>% 
+    './exploration scripts/correlation.R', 
+    './exploration scripts/comparison.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
 # END ------

### 2025-03-04 lft.R

@@ -16,11 +16,15 @@
 
   library(exda)
   library(rstatix)
-  library(caret)
-  library(caretExtra)
   library(effectsize)
   library(sciplot)
   library(rcompanion)
+  library(clustTools)
+  library(graphExtra)
+  library(igraph)
+
+  library(caret)
+  library(caretExtra)
 
   library(ranger)
   library(nnet)
@@ -39,6 +43,8 @@
 
   library(OptimalCutpoints)
   library(plotROC)
+
+  library(figur)
 
   explore <- exda::explore
   train <- caret::train
@@ -108,22 +114,44 @@
   
   insert_msg('Assessment of the model performance')
   
+  ## assessment of predictions in the training and CV data
+  
   c('./LFT scripts/binary_performance.R', 
     './LFT scripts/regression_performance.R', 
     './LFT scripts/residuals.R', 
     './LFT scripts/ct_predictions.R') %>% 
     source_all(message = TRUE, crash = TRUE)
 
+  ## assessment of over-fitting by learning curves
+  
+ access_cache(cache_path = './cache/lft_learn.RData',
+              script_path = './LFT scripts/learning_curves.R', 
+              message = 'Assessment of over-fitting by learning curves')
+ 
+ c('./LFT scripts/learning_plots.R') %>% 
+   source_all(message = TRUE, crash = TRUE)
+ 
 # Variable importance ------
   
   insert_msg('Variable importance')
+  
+  ## importance determined by SHAP and algorithm-specific algorithms 
+  ## by caret.
+  ## For the SHAP measures, we're analyzing the co-linearity by PCA
+  ## and investigation of correlation graphs
+  ##
+  ## Co-regulation of the top explanatory variables: 
+  ## PCA and correlation graphs
 
   access_cache(cache_path = './cache/shap_devel.RData', 
                script_path = './LFT scripts/shap_development.R', 
                message = 'Loading cached SHAP importance data')
   
   c('./LFT scripts/shap_importance.R', 
-    './LFT scripts/caret_importance.R') %>% 
+    './LFT scripts/shap_pca.R', 
+    './LFT scripts/caret_importance.R', 
+    './LFT scripts/variable_pca.R', 
+    './LFT scripts/variable_networks.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
 # ROC analysis for the CT parameters -------

### 2025-03-04 paper scripts/supplementary_figures.R

@@ -6,9 +6,9 @@
   
   suppl_figs <- list()
   
-# Figure S1: time course of CT findings -------
+# time course of CT findings -------
   
-  insert_msg('Figure S1: time course of the CT findings')
+  insert_msg('Time course of the CT findings')
   
   suppl_figs$ct_finding_course <- 
     cohort_ct$plots$factor[c("GGO_finding", 
@@ -25,7 +25,7 @@
                            labs(fill = 'Abnormality')), 
               ncol = 2, 
               rel_widths = c(0.9, 0.1)) %>% 
-    as_figure(label = 'figure_s1_ct_abnormality_time_course', 
+    as_figure(label = 'ct_abnormality_time_course', 
               ref_name = 'ct_finding_course', 
               caption = paste('Frequency of the most common abnormalities in', 
                               'computed tomography of the chest', 
@@ -35,9 +35,9 @@
               w = 180, 
               h = 210)
   
-# Figure S2: time course of numeric CT variables -----
+# time course of numeric CT variables -----
   
-  insert_msg('Figure S2: CT pathology scoring at the follow-ups')
+  insert_msg('CT pathology scoring at the follow-ups')
   
   suppl_figs$ct_numeric_course <- 
     cohort_ct$plots$numeric[c("CTSS", 
@@ -61,9 +61,9 @@
               w = 180, 
               h = 210)
   
-# Figure S3: CT feature overlap --------
+# CT feature overlap --------
   
-  insert_msg('Figure S3: overlap of the CT features')
+  insert_msg('Overlap of the CT features')
   
   ## upper 
   
@@ -91,7 +91,7 @@
               rel_heights = c(1, 2), 
               labels = LETTERS, 
               label_size = 10) %>% 
-    as_figure(label = 'figure_s3_cooccurrence_ct_findings', 
+    as_figure(label = 'cooccurrence_ct_findings', 
               ref_name = 'ct_corr', 
               caption = paste('Co-occurrence of abnormalities of chest', 
                               'computed tomography and correlation of computed', 
@@ -99,11 +99,9 @@
               w = 180, 
               h = 210)
   
+# time course of LFT abnormalities -------
   
-  
-# Figure S4: time course of LFT abnormalities -------
-  
-  insert_msg('Figure S4: time course of LFT abnormalities')
+  insert_msg('Time course of LFT abnormalities')
   
   suppl_figs$lft_finding_course <- 
     cohort_lft$plots$factor[c("DLCO_reduced", 
@@ -120,7 +118,7 @@
[…]

### 2025-03-04 paper scripts/supplementary_tables.R

@@ -158,7 +158,7 @@
     mdtable(label = 'table_S6_ml_tuning',
             ref_name = 'tuning', 
             caption = paste('Selection of machine learning algorithm', 
-                            'parameters by cross-validation-mediated tuning.'))
+                            'hyper-parameters by cross-validation-mediated tuning.'))
   
 # Table S7: performance of the machine learning classifiers, training ------
   

### 2025-03-04 tools/tools.R

@@ -134,7 +134,7 @@
     
   }
   
-# Plots for explorative analysis ------
+# Plots for exploratory analysis ------
   
   plot_explo_numeric <- function(data, 
                                  outcome, 
@@ -944,16 +944,17 @@
     
   }
   
-  spearman_rho_ <- function(data, 
-                            variables, 
-                            positive = FALSE) {
+  corr_test_ <- function(data, 
+                         variables, 
+                         positive = FALSE, 
+                         method = 'spearman') {
     
-    ## computes Spearman's rho correlation coefficient
+    ## computes a correlation coefficient
     ## used in a bootstrap correlation test
     
     rho <- cor(data[[variables[1]]], 
                data[[variables[2]]], 
-               method = 'spearman')
+               method = method)
     
     if(positive) {
       
@@ -970,20 +971,22 @@
     
   }
   
-  spearman_rho <- function(data, 
-                           variables, 
-                           id_variable = 'ID', 
-                           B = 1, 
-                           positive = FALSE) {
+  corr_test <- function(data, 
+                        variables, 
+                        id_variable = 'ID', 
+                        B = 1, 
+                        positive = FALSE, 
+                        method = 'spearman') {
     
     ## implements a block bootstrap correlation test
     
     boots <- sboot.data.frame(data, B = B, .by = id_variable)
     
     boot_stats <- boots %>% 
-      future_map_dfr(spearman_rho_, 
+      future_map_dfr(corr_test_, 
                      variables = variables, 
                      positive = positive, 
+                     method = method, 
                      .options = furrr_options(seed = TRUE))
     
     tibble(rho = mean(boot_stats$rho), 
@@ -2245,6 +2248,60 @@
     
   }
   
+  plot_learn_curves <- function(data, 
+                                metrics = lft_lcurves$metrics$DLCO_reduced, 
+                                plot_title = NULL, 
+                                metric_labels = lft_lcurves$metric_labels,
+                                data_labels = lft_lcurves$set_labels, 
+                                data_colors = lft_lcurves$set_colors, 
+                                point_size = 2, 
+                                point_shape = 16, 
+                                point_wjitter = 1, 
+                                point_hjitter = 0, 
+                                point_alphas = c(1, 1, 0.25), ...) {
+    
[…]

### 2025-03-05 paper/markdown/manuscript.Rmd
Revision analyses
Learning curves for ML models of DLCO and insufficient DLCO. Correlation graphs for the top explanatory variables for predictions of DLCO.
Supplementary Matherial and Manuscript update
Update of Supplementary Methods, Supplementary Tables, and Methods. Results for analysis of learning curves included.

@@
 
-Differences in medians of non-independently distributed, participant matched numeric variables between observations with and without LFT abnormalities were assessed by a blocked bootstrap test with blocks defined by the participant's identifier, and effect size measured by biseral r effect size statistic [@Mangiafico2022].
+Differences in medians of non-independently distributed, participant matched numeric variables between observations with and without LFT and CT abnormalities were assessed by a blocked bootstrap test with blocks defined by the participant's identifier, and effect size measured by biseral r effect size statistic [@Mangiafico2022].
 Correlations of non-independently distributed CT and LFT readouts were assessed by blocked bootstrap Spearman's rank test. 
@@
 
-Presence of reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables.
+Reduced DLCO, FVC and FEV1 (each < 80% of reference), as well as values of DLCO, FVC and FEV1 expressed as percentages of the reference value were modeled with `r lft_globals[c("ct_variables", "symptom_variables", "baseline_variables")] %>% reduce(union) %>% length` explanatory variables.
 The explanatory variables included 
@@
 Spearman's $\rho$ coefficients of correlation between the predicted and observed values were used to gauge calibration of the regression models. 
+Over- and under-fitting was assessed by learning curves [@Viering2023]. 
 Importance of explanatory variables was estimated by absolute values of SHAP statistics (Shapley additive explanations) [@Covert2020; @Lundberg2017; @Mayer2023a]. 
+Co-linearity of the most influential explanatory variables was assessed by weighted correlation graph, whose edge weights corresponded to Kendall's $\tau$ coefficients of pairwise correlations [@Csardi2006].
 
@@
 
+To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO.
+With this approach, we evaluated performance of the models re-trained in subsets of the modeling data set of varying sizes.
+The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
+The learning curves for the best-performing GBM models are presented in __Supplementary Figure S12__, for the remaining algorithms comparable results were obtained (not shown). 
+As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e.
+poor generalizability for unseen data.
+The generalizability of the models of DLCO expressed as percentage of the patient#s reference was substantially better.
+
 As investigated by absolute values of SHAP variable importance metrics [@Covert2020; @Lundberg2017], human- and AI-derived ratings of structural lung damage, CTSS, opacity and high opacity, belonged to the most influential explanatory variables of the models of DLCO insufficiency and the meaningful models of DLCO.
@@
 rating of physical performance impairment (ECOG) and 
-exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S12__, __Supplementary Table S9__).
+exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S13__, __Supplementary Table S9__).
+
+***placeholder for colinearity of the explanatory varibales***
 
@@
 FVC (difference of medians: `r blast(lft_uniboot$analysis_tbl, FVC_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$eff_size`) and
-insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S10__).
+insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S12__).
 Furthermore, CTSS, opacity and high opacity correlated negatively with moderate effect size with DLCO expressed as percentage of the patient's reference (p < 0.001, Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
-Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S11__).
+Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S13__).
 
@@
 Hence, external validation of our findings in an independent cohort is recommended. 
+Fourth, as inferred from the analysis of learning curves, the models of insufficient DLCO suffered for substantial over-fitting. 
+While this problem can be partially traced back to an unsharp distinction between sufficient and reduced DLCO with the 80% reference cutoff, a modeling approach employing regularized machine learning algorithm like XGBoost or regularized neural networks may further improve quality of predictions.
+Furthermore, most of the highly influential explanatory variables for predictions of DLCO and reduced DLCO, and in particular the CT readouts of lung damage were strongly inter-correlated, which raises questions about redundancy of explanatory variables.
 Finally, the analyzed cohort was recruited in the initial phase of the pandemic and consisted of individuals infected with the wild-type variant of the SARS-CoV-2 virus. 
@@
 Effect sizes and p values are displayed in the plot captions.
-Numbers of observations are indicated in the X axis._
+Numbers of observations are indicated in the X axes._
 

### 2025-03-05 paper/markdown/supplementary_material.Rmd
Supplementary Matherial and Manuscript update
Update of Supplementary Methods, Supplementary Tables, and Methods. Results for analysis of learning curves included.

@@
 In exploratory data analysis, statistical hypothesis testing and analysis of correspondence and correlations, the packages _rstatix_ [@Kassambara2021], _rcompanion_ [@Mangiafico2022], _MASS_ [@Ripley2022], [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) were used.
+Graph analyses were done with _igraph_ [@Csardi2006] and development package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra).
+
 Receiver-operating characteristic (ROC), confusion matrix analysis and inter-rater reliability analyses were done with _caret_ [@Kuhn2008], _OptimalCutpoints_ [@Lopez-Raton2014] and [_bootStat_](https://github.com/PiotrTymoszuk/bootStat).
@@
 
-Analysis results were visualized with _ggplot_ [@Wickham2016] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (scatter, stack and box plots), [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (visualization of results of correspondence analysis), _ggvenn_ [@Yan2021] (Venn plots), [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) (plots for machine learning model diagnostic and evaluation), _plotROC_ [@Sachs2017] (ROC curves) and _shapviz_ [@Mayer2023a; @Mayer2023] (violin/scatter plots of absolute SHAP values).
+Analysis results were visualized with _ggplot_ [@Wickham2016] and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (scatter, stack and box plots), [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) (visualization of results of correspondence analysis), _ggvenn_ [@Yan2021] (Venn plots), [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra) (plots for machine learning model diagnostic and evaluation), _plotROC_ [@Sachs2017] (ROC curves), _shapviz_ [@Mayer2023a; @Mayer2023] (violin/scatter plots of absolute SHAP values), and [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) (graph visualizations, re-implementation of tools from _igraph_ and _ggnetwork_) [@Csardi2006; @Briatte2021].
 Result tables were created with _flextable_ [@Gohel2022]. 
@@
 
+* Kendall's $\tau$ coefficient of correlation: < 0.1: small, 0.1 - 0.3: moderate, $\geq$ 0.3: large effect size
+
 * pseudo-R^2^ was used as a measure of explained variance of a model: < 0.13: small, 0.13 - 0.26: moderate, $\geq$ 0.26: large effect size
@@
 Statistical significance and effect sizes for comparisons and correlations of independent observations was assessed with the functions `compare_variables()` and `correlate_variables()` (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
-Because of non-independent observations included in the analysis, pairwise correlations of numeric LFT or CT variables were investigated with blocked bootstrap Spearman's rank test described below. 
+Because of non-independent observations included in the analysis, pairwise correlations of numeric LFT or CT variables were investigated with blocked bootstrap Spearman's rank and Kendall' tests described below.
 Analogically, for comparison of medians between two subsets of non-independent observations, blocked bootstrap test was utilized. 
@@
 
-* _`r exchange('antiinfective_treatment', lft_globals$lexicon)`_ which subsumed anti-infective and anti-macrolide treatment during acute COVID-19
+* _`r exchange('antiinfective_treatment', lft_globals$lexicon)`_ which subsumed anti-infective and macrolide treatment during acute COVID-19
 
@@
 canonical random forest (implemented in R by the _ranger_ package, caret's method name: 'ranger') [@Wright2017; @Breiman2001], 
-GBM (_gbm_ package, caret's method name: 'gbm') [@Greenwell2022; @Friedman2002; @Friedman2001], 
+GBM (gradient boosted machines, _gbm_ package, caret's method name: 'gbm') [@Greenwell2022; @Friedman2002; @Friedman2001], 
 neural network with a single hidden layer (_nnet_ package, caret's method name: 'nnet') [@Ripley2014], 
-and SVM with radial kernel (_kernlab_ package, caret's method name: 'svmRadial') [@Weston1998; @Karatzoglou2004]. 
+and SVM with radial kernel (support vector machines, _kernlab_ package, caret's method name: 'svmRadial') [@Weston1998; @Karatzoglou2004]. 
 For random forest models, 1000 random trees per model were generated. 
 Bernoulli and Gaussian loss functions were implemented in classification and regression GBM models, respectively.
-Optimal values of parameters controlling the algorithm's behavior such as number of observations in terminal nodes of tree models in the random forest models or cost penalty in the SVM models were found by maximizing the Youden's J statistic
-($J = Sensitivity + Specificity - 1$, classification models of binary LFT outcomes) or by minimizing mean absolute error (MAE, regression models of numeric LFT outcomes) in 10-repeats 10-fold cross-validation.
-To account for the participant matching of the observations, a blocked cross-validation workflow was designed to keep all observations obtained from a particular participant either in the training or in the test portion of the cross-validation split.
-Construction of such 'participant-wise' blocked cross-validation folds was done with an in-house-developed script employing the `createMultiFolds()` function (package _caret_). 
+
+Hyper-parameters that control the modeling algorithm's behavior were found by maximizing the Youden's J statistic 
+($J = Sensitivity + Specificity - 1$, classification models of binary LFT outcomes) or by minimizing mean absolute error (MAE, regression models of numeric LFT outcomes) in 10-repeats 10-fold cross-validation, in a process called tuning.
+The tunable hyper-parameters were:
+
+* for random forest, `mtry`: number of randomly chosen explanatory variables pre decision tree in the random forest ensemble, `splitrule`: criterion for splits of the explanatory variables at branches of the decision trees, `min.node.size`: minimal size of the node (leaf) of the decision tree
+
+* for GBM, `n.trees`: number of decision trees in the GBM ensemble, `interaction.depth`: number of splits in the decision trees, `shrinkage`: regularization penalty applied to predictions made by single decision trees, `n.minobsinnode`: minimal size of the node (leaf) of the decision tree
+
+* for neural network, `size`: number of neurons in the hidden layer, `decay`: reduction of weights during algorithm iteration
+
+* for SVM, `sigma`: width of the radial kernel, `C`: cost/penalty applied to misclassified observations
+
+To account for the participant matching of the observations in the cross-validated tuning of hyper-parameters, a blocked cross-validation design was applied to keep all observations obtained from a particular participant either in the training or in the test portion of the cross-validation split.
+Construction of such 'participant-wise' blocked cross-validation folds was done with an in-house-developed wrapper around `createMultiFolds()` function (package _caret_). 
 For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed [@Kuhn2008]. 
@@
 
-Performance of the machine learning models was evaluated by comparison of the model predictions with the observed outcome in the genuine training data set and in 10-repeats 10-fold cross-validation with the observed outcomes.
-Performance of the classification models was assessed by Cohen's $\kappa$ as a measure of concordance between the predicted and observed outcome [@McHugh2012; @Cohen1960], 
+Performance of the machine learning models was evaluated by comparison of the model predictions with the observed outcome in the genuine training data set and in participant-wise blocked 10-repeats 10-fold cross-validation.
+Performance of the classification models was assessed by 
+accuracy statistic defined as a fraction of correct predictions, 
+Cohen's $\kappa$ as a measure of concordance between the predicted and observed outcome [@McHugh2012; @Cohen1960], 
 standard receiver operating characteristic metrics (AUC, overall accuracy, sensitivity, specificity) to assess the ability of the model to detect the pathological outcome [@Kuhn2008], and with 
@@
 pseudo-R^2^ as a metric of explained variance and 
-Spearman's $\rho$ for correlation between the predicted and observed outcome as as measure of overall model calibration [@Kuhn2008] (__Table 3__ and __Table \@ref(tab:tab-reg-models)__). 
+Spearman's $\rho$ for correlation between the predicted and observed outcome as as measure of overall model calibration [@Kuhn2008] (__Table 3__ and __Supplementary Table S\@ref(tab:tab-reg-models)__).
 Pseudo-R^2^ was computed with the following formulas:
@@
 
+For assessment of potential over- and under-fitting of predictions made by the machine learning models, we resorted to visualizations of learning curves [@Viering2023]. 
+The learning curves were constructed as values of model performance metrics (classification models: accuracy and Cohen's $\kappa$,regression models: mean absolute error and pseudo-R^2^) as functions of increasing size of the training data derived from the genuine COVILD cohort data set used for model construction.
+The performance metrics were computed for the training subset, a test subset (1/4 of observations not used for training), and out-of-fold predictions in 10-repeats 10-fold cross-validation. 
+The learning curves were computed with `learning_curve_dat()` function from _caret_ package and visualized as LOESS trends (Locally Estimated/weighted Scatterplot Smoothing).
+
 ## Variable importance for multi-parameter models of lung function
@@
 
-## Comparison of CT readouts between observations with and without LFT abnormality, correlation analysis
+## Co-linearity of the most influential explanatory variables
+
+The most influential explanatory factors for predictions of DLCO (algorithms: random forest, GBM, neuronal network, and SVM radial) and DLCO < 80% of reference (algorithms: random forest, GBM, SVM radial) were identified as top 15 variables with the largest mean absolute SHAP importance values for each of the modeling algorithms.
+Associations between these highly influential variables were assessed with a correlation graph [@Csardi2006]. 
+For construction of the correlation graph, ordered categorical variables (i.e. factors in R) were converted to integers. 
+The graph edges were defined by pairwise correlations between the variables with Kendall's $\tau \geq 0.3$. 
+The edge weights corresponded to values of the $\tau$ correlation coefficients. 
+The graph was constructed with `as_iGraph()` function; isolated vertices, i.e.
+variables without any correlations with $\tau \geq$ 0.3 were removed from the graph with `prune_degree()` (both functions from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).
+The graph was visualized with Fruchterman-Reingold algorithm (method `plot()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).
+
+## Comparison and correlation of CT readouts
 
 Because observations were non-independent, i.e.
-participant-matched, comparison of CTSS, opacity and high opacity of the lung between data points with and without LFT abnormalities could not be investigated by standard statistical tests (e.g.
+participant-matched, comparison of CTSS, opacity and high opacity of the lung between data points with and without LFT and CT abnormalities could not be investigated by standard statistical tests (e.g.
 Mann-Whitney test) and required instead a special approach.
-Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biserial r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality).
+Differences in median of the CT parameters between observations with and without LFT or CT abnormalities were investigated with blocked bootstrap test with biserial r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality).
 The blocked bootstrap scheme involved re-sampling with repetition of the participants instead of single observations (B = 2000 re-samples). 
@@
 
-Bootstrap tests were performed with a functional interface implemented in the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat). 
-The analysis results are listed in __Supplementary Table S\@ref(tab:tab-lft-binary-uni)__ and __\@ref(tab:tab-lft-correlation-uni)__.
+Bootstrap tests were performed with a functional programming interface implemented in the development package [_bootStat_](https://github.com/PiotrTymoszuk/bootStat). 
+The analysis results are listed in __Supplementary Tables S\@ref(tab:tab-ct-correlations)__ and __S\@ref(tab:tab-lft-correlation-uni)__.
 
@@
 flextable::flextable(suppl_tabs$tuning) %>% 
-  set_widths(c(3, 4, 4.5)) %>% 
+  set_widths(c(3, 7, 5.5)) %>% 
   merge_v(1) %>% 
@@
            ref_symbols = 'a') %>% 
+  footnote(1, 3, 
+           value = as_paragraph('mtry: number of randomly choosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node; size: number of neurons in the hidden layer of the neuronal network; decay: decay of weights in the the neuronal network; sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
   theme_vanilla
@@
   set_widths(16)
+
+```
+
+\newpage
+
+```{r tab-ct-correlations, tab.cap = 'Correlations between chest computed tomography severity score, and AI-determined opacity and high opacity.'}
+
+flextable::flextable(suppl_tabs$ct_correlations) %>% 
+  set_widths(c(3.5, 3.5, 1.8, 4, 2.7)) %>% 
+  footnote(1, 1:2, 
+           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for al lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4:5, 
+           value = as_paragraph('Blocked bootstrap Spearman correlation test; \u03C1 correlation coefficients with 95% confidence intervals.
+P values were corrected for multiple testing with the false discovery rate method.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-ct-comparisons, tab.cap = 'Differences in chest computed tomography severity score (CTSS), and AI-determined lung opacity and high opacity in CovILD study participants with and without ground glass opacity (GGO) and reticulation.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.'}
+
+flextable::flextable(suppl_tabs$ct_comparisons) %>% 
+  set_widths(c(2.5, 3, rep(4, 2), 2.7, 2.5)) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for all lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'),
+           part = 'header', 
+           ref_symbols = 'b') %>% 
+  footnote(1, 5:6, 
+           value = as_paragraph('Blocked bootstrap test with biserial r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  theme_vanilla
 

### 2025-03-05 LFT scripts/bootstrap_tests.R

@@ -120,7 +120,7 @@
     unlist(recursive = FALSE)
   
   lft_uniboot$cor_test <- lft_uniboot$cor_pairs %>% 
-    map_dfr(spearman_rho, 
+    map_dfr(corr_test, 
             data = lft_uniboot$analysis_tbl, 
             B = 2000)
   

### 2025-03-05 LFT scripts/variable_networks.R

@@ -62,6 +62,25 @@
          lft_net$attr_tbl, 
          set_vertex_attributes)
   
+# Graph communities -------
+  
+  insert_msg('Graph communities')
+  
+  ## communities with single features are lumped
+  
+  set.seed(12345687)
+  
+  lft_net$communities <- lft_net$graph_obj %>% 
+    map(cluster_leiden, 
+        objective_function = 'modularity', 
+        n_iterations = 100) %>% 
+    map(comm_lump, cutoff = 1)
+  
+  lft_net$graph_obj <- 
+    map2(lft_net$graph_obj, 
+         lft_net$communities, 
+         add_communities)
+  
 # Fruchterman-Reingold network plots --------
   
   insert_msg('Network plots')
@@ -91,6 +110,30 @@
          show.legend = FALSE, 
          seed = 12345)
   
+  ## vertex color codes for assignment to the communities
+  
+  lft_net$community_plots <- 
+    list(x = lft_net$graph_obj, 
+         plot_title = lft_net$graph_titles) %>% 
+    pmap(plot, 
+         layout = layout.fruchterman.reingold, 
+         vertex_fill_variable = 'community_id', 
+         vertex_color_variable = 'community_id', 
+         weighting_order = 1, 
+         vertex_size = 2, 
+         label_vertices = TRUE, 
+         vertex_label_variable = 'variable_label', 
+         vertex_txt_color_variable = 'community_id', 
+         vertex_txt_size = 2.5, 
+         cust_theme = theme_void() + 
+           theme(plot.title = element_text(size = 8, face = 'bold'), 
+                 plot.subtitle = globals$common_text, 
+                 legend.text = globals$common_text, 
+                 legend.title = globals$common_text, 
+                 plot.margin = globals$common_margin), 
+         show.legend = FALSE, 
+         seed = 12345)
+  
   ## additional styling by setting color and line width scales
   
   lft_net$plots <- lft_net$plots %>% 
@@ -101,6 +144,15 @@
           scale_color_manual(values = lft_globals$class_colors, 
                              name = 'variable\nclassifcation', 
                              drop = FALSE) + 
+          scale_linewidth(range = c(0.25, 1.5), 
+                          limits = c(0.3, 1), 
+                          name = "Kendall's \u03C4") + 
+          scale_alpha_continuous(range = c(0.15, 0.6), 
+                                 limits = c(0.3, 1), 
+                                 name = "Kendall's \u03C4"))
+  
[…]

### 2025-03-05 exploration scripts/correlation.R

@@ -1,5 +1,5 @@
 # Correlation of numeric LFT readouts and CT readouts. Because the observations 
-# are not independent, we're resorting to a blocked bootstrap Kendall's test.
+# are not independent, we're resorting to a blocked bootstrap Spearman's test.
 
   insert_head()
   
@@ -55,9 +55,9 @@
               data = corr_lft$analysis_tbl[[i]], 
               B = 2000, 
               positive = TRUE, 
-              method = 'kendall') %>% 
+              method = 'spearman') %>% 
       re_adjust %>% 
-      mutate(est_lab = paste0('\u03C4 = ', signif(rho, 2), 
+      mutate(est_lab = paste0('\u03C1 = ', signif(rho, 2), 
                               ' [', signif(lower_ci, 2), 
                               ' - ', signif(upper_ci, 2), ']'), 
              n = nrow(corr_lft$analysis_tbl[[i]]), 

### 2025-03-05 paper scripts/supplementary_tables.R

@@ -6,9 +6,9 @@
   
   suppl_tabs <- list()
   
-# Table S1: numbers of samples and participants per time point -------
+# Numbers of samples and participants per time point -------
   
-  insert_msg('Table s1: numbers of samples and participants')
+  insert_msg('Numbers of samples and participants')
   
   suppl_tabs$n_numbers <- cohort_ct$analysis_tbl %>% 
     blast(cohort_split) %>% 
@@ -20,15 +20,15 @@
                 'Ambulatory, mild COVID-19', 
                 'Hospitalized, moderate COVID-19', 
                 'Hospitalized, severe COVID-19')) %>%
-    mdtable(label = 'table_s1_observation_n_numbers', 
+    mdtable(label = 'observation_n_numbers', 
             ref_name = 'n_numbers', 
             caption = paste('Numbers of observations included in the analysis', 
                             'split by acute COVID-19 severity and the follow-up', 
                             'examination.'))
 
-# Table S2: study variables --------
+# Study variables --------
   
-  insert_msg('Table S2: Study variables')
+  insert_msg('Study variables')
   
   suppl_tabs$study_variables <- 
     list(lft = tibble(variable = lft_globals$responses), 
@@ -64,14 +64,14 @@
                 'Unit'))
   
   suppl_tabs$study_variables <- suppl_tabs$study_variables %>% 
-    mdtable(label = 'table_S2_study_variables', 
+    mdtable(label = 'study_variables', 
             ref_name = 'study_variables', 
             caption = paste('Study variables. The table is available in a', 
                             'supplementary Excel file.'))
   
-# Tables S3 - 5: CT, LFT and symptom variables at follow-ups -------
+# CT, LFT and symptom variables at follow-ups -------
   
-  insert_msg('Tables S3 - S5: CT, LFT and symptom variables at follow-ups')
+  insert_msg('CT, LFT and symptom variables at follow-ups')
   
   suppl_tabs[c("cohort_ct", 
                "cohort_lft", 
@@ -106,9 +106,9 @@
                  "cohort_lft", 
                  "cohort_symptoms")] %>% 
     list(x = ., 
-         label = c('table_S3_cohort_ct_variables', 
-                   'table_S4_cohort_lft_variables', 
-                   'table_S5_cohort_symptom_variables'), 
+         label = c('cohort_ct_variables', 
+                   'cohort_lft_variables', 
+                   'cohort_symptom_variables'), 
          ref_name = names(.), 
          caption = c(paste('Chest computed tomography variables', 
[…]

### 2025-03-06 paper/markdown/manuscript.Rmd
Variable redundancy
Inclusion of variable redundancy analysis data in Results of manuscript. modification of Supplementary Figure S14

@@
 Importance of explanatory variables was estimated by absolute values of SHAP statistics (Shapley additive explanations) [@Covert2020; @Lundberg2017; @Mayer2023a]. 
-Co-linearity of the most influential explanatory variables was assessed by weighted correlation graph, whose edge weights corresponded to Kendall's $\tau$ coefficients of pairwise correlations [@Csardi2006].
+Co-linearity of the most influential explanatory variables (top 15 mean absolute SHAP for each of the models of DLCO and DLCO < 80%) was assessed by soft-threshold weighted graph of correlations [@Csardi2006; @Zhang2005].
+The graph edges were defined by pairwise correlations with Kendall's $\tau \geq$ 0.3 and edge weights corresponded to $\tau$ coefficient values.
 
@@
 poor generalizability for unseen data.
-The generalizability of the models of DLCO expressed as percentage of the patient#s reference was substantially better.
+The generalizability of the models of DLCO expressed as percentage of the patient's reference was substantially better.
 
@@
 
-***placeholder for colinearity of the explanatory varibales***
+Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S14A__).
+In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated.
+CTSS values were significantly higher in observations with than without 
+GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and
+reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`) (__Supplementary Figure S3B__, __Supplementary Table S10__).
+Analogically, significantly higher opacity was detected in observations with than without 
+GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'GGO_finding') %>% .$eff_size`) and
+reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`).
+Effect sizes of these differences were large (__Supplementary Figure S14B__, __Supplementary Tables S11__).
+Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
+Tackling this redundancy e.g. by regularization may control the over-fitting and improve accuracy of predictions of DLCO and reduced DLCO by machine learning algorithms.
 

### 2025-03-06 paper/markdown/supplementary_material.Rmd
Variable redundancy
Inclusion of variable redundancy analysis data in Results of manuscript. modification of Supplementary Figure S14

@@
 
-## Co-linearity of the most influential explanatory variables
+## Co-linearity of explanatory variables
 
 The most influential explanatory factors for predictions of DLCO (algorithms: random forest, GBM, neuronal network, and SVM radial) and DLCO < 80% of reference (algorithms: random forest, GBM, SVM radial) were identified as top 15 variables with the largest mean absolute SHAP importance values for each of the modeling algorithms.
-Associations between these highly influential variables were assessed with a correlation graph [@Csardi2006]. 
+Associations between these highly influential variables were assessed with a soft-threshold weighted graph of correlations, a method widely used in omics and system biology [@Csardi2006; @Zhang2005]. 
 For construction of the correlation graph, ordered categorical variables (i.e. factors in R) were converted to integers. 
@@
 flextable::flextable(suppl_tabs$lft_correlation_uni) %>% 
-  set_widths(c(2.5, 2.5, 1.8, 4, 2.7)) %>% 
+  set_widths(c(3.5, 3.5, 1.8, 5, 2.7)) %>% 
   merge_v(1) %>% 
@@
 
-_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco).__
+_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco)._
 
@@
 
-_(B) Correlations between computed tomography (CT) readouts of lung damage, the human-determined CT severity score and software-determined lung opacity and high opacity, were assessed by blocked bootstrap Kendall's correlation test.
-Ranks of the CT readouts are presented in scatter plots.
-Each point represents a single observation, blue straight lines are visualizations of linear trends.
-Correlation coefficients $\tau$ with bootstrap confidence intervals, p values, and numbers of observations are displayed in the plot captions._
-
-_(C) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations.
+_(B) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations.
 Statistical significance was determined by blocked bootstrap test with r effect size statistic.

### 2025-03-06 paper scripts/supplementary_figures.R

@@ -387,14 +387,15 @@
   
   suppl_figs$top_corr$upper <- lft_net$plots$all
   
-  ## middle panel: scatter plots for visualization of correlations
+  ## middle panel: scatter plots for visualization of correlations, 
+  ## redundant, already present in Supplemetary Figure S3B
   
-  suppl_figs$top_corr$middle <- corr_lft$plots$ct %>% 
-    map(~.x + theme(plot.title.position = 'plot')) %>% 
-    plot_grid(plotlist = .,
-              ncol = 3, 
-              align = 'hv', 
-              axis = 'tblr')
+  #suppl_figs$top_corr$middle <- corr_lft$plots$ct %>% 
+   # map(~.x + theme(plot.title.position = 'plot')) %>% 
+    #plot_grid(plotlist = .,
+     #         ncol = 3, 
+      #        align = 'hv', 
+       #       axis = 'tblr')
   
   ## bottom panel: comparison of CTSS, and AI measures between observations 
   ## with and without GGO and reticulation
@@ -422,11 +423,13 @@
   
   suppl_figs$top_corr <- 
     plot_grid(suppl_figs$top_corr$upper, 
-              suppl_figs$top_corr$middle, 
+              ggdraw(), #suppl_figs$top_corr$middle, 
               suppl_figs$top_corr$bottom, 
               nrow = 3, 
-              rel_heights = c(1.25, 1, 1), 
-              labels = LETTERS, 
+              rel_heights = c(1.4, 
+                              0.1, 
+                              1), 
+              labels = c('A', '', 'B'), #LETTERS, 
               label_size = 10) %>% 
     as_figure(label = 'top_variables_co_linearity', 
               ref_name = 'top_corr', 
@@ -435,7 +438,7 @@
                              'insufficiency and percentage of reference of', 
                              'diffusion capacity for carbon monoxide.'), 
               w = 190, 
-              h = 230)
+              h = 160)
     
 # Saving figures on the disc --------
   

### 2025-03-07 paper/markdown/rebuttal_letter.Rmd
DLCO <80% predictions and DLCO values
Analysis of predictions of DLCO < 80% in observations with increasing DLCO values expressed as percentages of the patient's reference. Drop-out rates and frequency of LFT and CT abnormalities included in the manuscript text.

---
title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Rebuttal Letter"
author: "Department of Radiology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Reviewer 1

## Point 1

### Issue

Add results in the abstract as well.

### Response

## Point 2

### Issue

The introduction is not sufficient.

### Response

## Point 3

### Issue

The literature review part needs to be added.

### Response

## Point 4

### Issue

After adding the literature review, add a summary table and clearly mention this study's research gaps and contributions.

### Response

## Point 5

### Issue

More details, such as features, are required in the study data.

### Response

## Point 6

### Issue

The paper lacks details about hyper-parameters for machine learning models. 
For example, what was the C and class weight value for the SVM model?

### Response

## Point 7

### Issue

Have you considered the GridSearchCV for model hyper-parameters or set it manually?

### Response

## Point 8

### Issue

Comparison with state-of-the-art studies is missing.

### Response

## Point 9

### Issue

The Practical Implication of the study is missing.

### Response

## Point 10

### Issue

Conclusion also not well presented.

### Response

# Reviewer 2

## Point 11

### Issue

The introduction does not sufficiently contextualize how other published multi-parameter or machine learning approaches have succeeded or failed. What specific gaps remain? A more detailed table is needed to perform the literature review.

### Response

## Point 12

### Issue

What is new or improved about your multi-parameter modeling approach compared to existing frameworks?

### Response

## Point 13

### Issue

More details about selection and potential biases (e.g., dropouts, missing visits) are needed for the data set.

### Response

## Point 14

### Issue

It remains unclear how each scoring system aligns, differs, or complements the other.

### Response

Overemphasis on “satisfactory accuracy” without deeper clinical context.

## Point 15

### Issue

The neural network model performed poorly for DLCO.

### Response

## Point 16

### Issue

The SHAP analysis identifies “top 15” predictors. However, the manuscript does not address correlations among those predictors or discuss whether certain features might be redundant.

### Response

## Point 17

### Issue

With “opacity cutoff: 0.12%” or “high opacity cutoff: 0.002%” appear extremely small. The clinical significance of such minimal changes in lung volume being “abnormal” is unclear—are these thresholds truly meaningful in practice?

### Response

## Point 18

### Issue

The paper concedes “model overfitting” could be an issue. However, it lacks any systematic attempt (e.g., a learning-curve analysis or thorough regularization strategy) to demonstrate that overfitting is minimal or under control.

### Response

# References

### 2025-03-07 paper/markdown/manuscript.Rmd
DLCO <80% predictions and DLCO values
Analysis of predictions of DLCO < 80% in observations with increasing DLCO values expressed as percentages of the patient's reference. Drop-out rates and frequency of LFT and CT abnormalities included in the manuscript text.

@@
 The majority of analyzed data points were obtained at the two- (n = `r covild$ct %>% filter(follow_up == '2-month follow-up') %>% nrow`) and three-month follow-up (n = `r covild$ct %>% filter(follow_up == '3-month follow-up') %>% nrow`, __Supplementary Table S1__ - __S2__).
+As compared with the initial set of patients = 145 with acute infection in the CovILD study by Sonnweber et al.
+[@Sonnweber2020], the dropout rates were `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['2-month follow-up']`%, `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['3-month follow-up']`%, `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['6-month follow-up']`% and  `r signif((1 - table(covild$ct$follow_up)/145) * 100, 2)['12-month follow-up']`% at the two-.
+three-, six- and twelve-month follow-up.
 
@@
 
+Radiological lung abnormalities were present in 
+`r signif(table(lft_globals$analysis_tbl$CT_findings)['yes']/sum(table(lft_globals$analysis_tbl$CT_findings)), 2) * 100`% of observations (n = `r table(lft_globals$analysis_tbl$CT_findings)['yes']`); 
+with GGO (`r signif(table(lft_globals$analysis_tbl$GGO_finding)['yes']/sum(table(lft_globals$analysis_tbl$GGO_finding)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$GGO_finding)['yes']`)
+and reticulation (`r signif(table(lft_globals$analysis_tbl$reticulation_finding)['yes']/sum(table(lft_globals$analysis_tbl$reticulation_finding)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$reticulation_finding)['yes']`) as the most common lesions.
 Frequency of any radiological lung abnormalities, GGO, reticulation, consolidation, and bronchiectasis diagnosed by a human radiologist was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients.
 GGO and reticulation were the most frequent lesion types, independently of the follow-up and acute disease severity. 
-Resolution of radiological lung findings. 
 Consolidation and bronchiectasis lesions were detected primarily in moderate and severe COVID-19 survivors at the two-month follow-up examination. 
@@
 
-Insufficient DLCO, FVC, and FEV1 defined as values below < 80% of the patient's reference value were the most common abnormalities of lung function. 
+Insufficient DLCO (`r signif(table(lft_globals$analysis_tbl$DLCO_reduced)['yes']/sum(table(lft_globals$analysis_tbl$DLCO_reduced)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$DLCO_reduced)['yes']`),
+FVC (`r signif(table(lft_globals$analysis_tbl$FVC_reduced)['yes']/sum(table(lft_globals$analysis_tbl$FVC_reduced)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$FVC_reduced)['yes']`), and 
+FEV1 (`r signif(table(lft_globals$analysis_tbl$FEV1_reduced)['yes']/sum(table(lft_globals$analysis_tbl$FEV1_reduced)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$FEV1_reduced)['yes']`) defined as values below < 80% of the patient's reference value were the most common abnormalities of lung function.
 As with the CT findings, their frequency was the highest in severe COVID-19 survivors followed by moderate COVID-19 patients, independently of the follow-up time point. 
-Additionally, lung function recovery reflected by decreasing rates of insufficient DLCO and FVC, and by an improvement of DLCO and FVC values expressed as percentages of the patient's reference was the most evident in severe COVID-19 survivors (__Supplementary Figures S4__ and __S5__, __Supplementary Table S4__).
+Additionally, lung function recovery was the most pronounced in severe COVID-19 survivors as reflected by decreasing rates of insufficient DLCO and FVC, and by an improvement of DLCO and FVC values expressed as percentages of the patient's reference (__Supplementary Figures S4__ and __S5__, __Supplementary Table S4__).
 In most of observations with insufficient FVC, insufficient FEV1 was found (`r covild$lft %>% filter(FVC_reduced == 'yes', FEV1_reduced == 'yes') %>% nrow` of `r covild$lft %>% filter(FVC_reduced == 'yes') %>% nrow` observations with reduced FVC).
@@
 
-Impaired physical performance defined by ECOG > 0 and exertional dyspnea defined by mMRC > 0 were the most frequent persistent COVID-19-related symptoms of potential relevance for pulmonary function, and their rates declined during convalescence in all acute COVID-19 severity subsets.
+Impaired physical performance defined by ECOG > 0 (`r signif(table(lft_globals$analysis_tbl$impaired_performance_symptom)['yes']/sum(table(lft_globals$analysis_tbl$impaired_performance_symptom)), 2) * 100`% of observations, n = `r table(lft_globals$analysis_tbl$impaired_performance_symptom)['yes']`) and
+exertional dyspnea defined by mMRC > 0 (`r signif(table(lft_globals$analysis_tbl$dyspnea_symptom)['yes']/sum(table(lft_globals$analysis_tbl$dyspnea_symptom)), 2) * 100`%, n = `r table(lft_globals$analysis_tbl$dyspnea_symptom)['yes']`) were the most frequent persistent COVID-19-related symptoms of potential relevance for pulmonary function, and their rates declined during convalescence in all acute COVID-19 severity subsets.
 Notably, we could not observe any substantial differences in frequency of physical performance impairment, dyspnea, and persistent cough between severe, moderate and ambulatory COVID-19 convalescents (__Supplementary Figure S7__, __Supplementary Table S5__).
@@
 neural network with a single hidden layer [@Ripley2014], and 
-support vector machines (SVM) [@Weston1998]. 
-Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for model optimization (tuning, __Supplementary Table S6__).
+support vector machines (SVM) with radial kernel [@Weston1998]. 
+Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Table S6__).
 Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).
 
-Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference could be modeled with satisfactory accuracy in the cross-validation setting independently of the machine learning algorithm (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
-Low Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__).
-In turn, models of insufficient FVC and FEV1 exhibited poor predictive performance in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).
+Among the investigated LFT abnormalities, solely insufficient DLCO defined as values < 80% of the patient's reference yielded meaningful models as evaluated by cross-validation; accuracy of the models of DLCO < 80% was reproducible between the algorithms (overall accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+Low cross-validated Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__).
+In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).
 
-As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with sufficiently low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+As in case of the LFT abnormalities, solely DLCO expressed as percentages of the patient's reference was modeled with low error and substantial explanatory power in cross-validation with the Random Forest, GBM, and SVM algorithms (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$MAE %>% range %>% signif(3) %>% paste(collapse = ' to ')`, pseudo-R^2^: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$rsq %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
 Those models were also characterized by good calibration assessed by Spearman's $\rho$ coefficients of correlation between the predicted and observed DLCO (Spearman's $\rho$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>%  .$spearman %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
 By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__).
-Analogically, no meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).
+No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).
 
@@
 
+Prediction of reduced DLCO < 80% was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference by moving averages of accuracy and error metered as squared distance to the 0/1-coded outcome.
+As presented in __Supplementary Figure S12__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference.
+An analogical phenomenon was observed for the remaining machine learning algorithms (not shown). 
+This illustrates that while predictions of highly compromised and normal-to-high DLCO are accurate, forecasts for borderline cases of reduced DLCO suffer from high error. 
+This may question biological and clinical relevance of the 80% cutoff widely used for diagnosis of insufficient DLCO in COVID-19 survivors [@Sonnweber2020; @Zhang2022a].
+
 To assess under- and over-fitting, we resorted to analyses of learning curves of the models of DLCO and reduced DLCO.
@@
 The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
-The learning curves for the best-performing GBM models are presented in __Supplementary Figure S12__, for the remaining algorithms comparable results were obtained (not shown). 
+The learning curves for the best-performing GBM models are presented in __Supplementary Figure S13__, for the remaining algorithms comparable results were obtained (not shown). 
 As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e.
@@
 rating of physical performance impairment (ECOG) and 
-exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S13__, __Supplementary Table S9__).
+exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S14__, __Supplementary Table S9__).
 
-Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S14A__).
+Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S15A__).
 In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated.
@@
 reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`).
-Effect sizes of these differences were large (__Supplementary Figure S14B__, __Supplementary Tables S11__).
+Effect sizes of these differences were large (__Supplementary Figure S15B__, __Supplementary Tables S11__).
 Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 

### 2025-03-07 paper/markdown/supplementary_material.Rmd
DLCO <80% predictions and DLCO values
Analysis of predictions of DLCO < 80% in observations with increasing DLCO values expressed as percentages of the patient's reference. Drop-out rates and frequency of LFT and CT abnormalities included in the manuscript text.

@@
 set_flextable_defaults(font.family = "Cambria", 
-                       font.size = 10)
+                       font.size = 9)
 
@@
 SHAP (Shapley additive explanations) as a metric of global variance importance in machine learning models was calculated with the package _kernelshap_ [@Covert2020; @Lundberg2017; @Mayer2023]. 
+Moving averages of statistics of performance of machine learning models were computed with _zoo_ [@Zeileis2022].
 
@@
 
+Performance of the machine learning models at prediction of reduced DLCO < 80% of the patient's reference was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference.
+To this end, we sorted the observations by increasing DLCO values and computed moving averages of accuracy and squared distance to 0/1-coded DLCO < 80% outcome. 
+The moving averages were computed with the sliding window of k = 20 observations (function `rollmean()` package _zoo_). 
+The squared distance to the outcome were calculated with the following formula:
+
+$$SqDist = (p - o)^2$$
+
+where $p$ stands for probability of DLCO < 80%, and $o$ equals to 1 for DLCO < 80% and 0 for DLCO $\geq$ 80%.
+
 For assessment of potential over- and under-fitting of predictions made by the machine learning models, we resorted to visualizations of learning curves [@Viering2023]. 
@@
 flextable::flextable(suppl_tabs$cohort_ct) %>% 
-  set_widths(c(2.3, 2.7, rep(3.9, 4))) %>% 
+  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
   merge_v(1) %>% 
@@
 flextable::flextable(suppl_tabs$cohort_lft) %>% 
-  set_widths(c(2.3, 2.7, rep(3.7, 4))) %>% 
+  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
   merge_v(1) %>% 
@@
 flextable::flextable(suppl_tabs$cohort_symptoms) %>% 
-  set_widths(c(2.3, 3.5, rep(3.7, 4))) %>% 
+  set_widths(c(2, 2.4, rep(3.2, 5))) %>% 
   merge_v(1) %>% 
@@
 
+```{r fig-dlco-class, fig.width = figur::convert(suppl_figs$dlco_class, to = 'in')$w, fig.height = figur::convert(suppl_figs$dlco_class, to = 'in')$h, fig.cap = "Accuracy of prediction of insufficient diffusion capacity for carbon monoxide by the GBM model as a function of diffusion capacity for carbon monoxide expressed as percentage of patient's reference."}
+
+suppl_figs$dlco_class$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-dlco-class).
+Accuracy of prediction of insufficient diffusion capacity for carbon monoxide by the GBM model as a function of diffusion capacity for carbon monoxide expressed as percentage of patient's reference.__
+
+_Predictions made by the best performing model of reduced diffusion capacity for carbon monoxide (DLCO < 80% of the patient's reference) were investigated in observations with increasing DLCO values expressed as percentage of the patient's reference._
+
+_(A) The actual DLCO values expressed as percentages of the patient's reference are displayed in violin plots.
+Single observations are depicted as points, point color codes for the predicted reduced DLCO <80% in the training data and 10-repeats 10-fold cross-validation (CV).
+The dashed line represents the 80% cutoff used for definition of reduced DLCO,_
+
+_(B) Moving averages of accuracy (fraction of correct predictions) at prediction of DLCO < 80% and squared distances to the 0/1-coded DLCO < 80% in the training data set and 10-repeats 10-fold CV.
+The moving averages calculated with the sliding window of 20 observations are displayed as solid lines.
+The dashed lines represent the 80% cutoff used for definition of reduced DLCO._
+
+\newpage
+
 ```{r fig-learn-curves, fig.width = figur::convert(suppl_figs$learn_curves, to = 'in')$w, fig.height = figur::convert(suppl_figs$learn_curves, to = 'in')$h, fig.cap = 'Learning curves of the gradient boosted machines models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.'}

### 2025-03-07 LFT scripts/dlco_classification.R

@@ -1,0 +1,268 @@
+# Analysis of accuracy and confidence (Brier's squares) of prediction of 
+# DLCO < 80% as a function of DLCO values expressed as percentage of the 
+# patient's reference
+
+  insert_head()
+  
+# container -------
+  
+  dlco_class <- list()
+  
+# analysis globals: DLCO, observed and predicted DLCO insufficiency -------
+  
+  insert_msg('Analysis globals')
+  
+  ## models and predictions in the training and cross-validation 
+  ## setting. 
+  ## for cross-validation, the resamples are collapsed
+  
+  dlco_class$data <- bin_models$models$DLCO_reduced %>% 
+    map(predict) %>% 
+    map(compact) %>% 
+    map(map, components.predx, 'data') %>% 
+    transpose
+  
+  dlco_class$data$cv <- dlco_class$data$cv %>% 
+    map(group_by, .observation) %>% 
+    map(summarise, 
+        .outcome = .outcome[1], 
+        yes = mean(yes)) %>% 
+    map(mutate, 
+        .fitted = ifelse(yes > 0.5, 'yes', 'no'), 
+        .fitted = factor(.fitted, c('no', 'yes')))
+  
+  dlco_class$data <- dlco_class$data %>% 
+    transpose %>% 
+    map(map, 
+        transmute, 
+        index = .observation, 
+        DLCO_reduced = .fitted, 
+        p_value = yes) %>% 
+    map(compress, names_to = 'data_set')
+  
+  ## observed DLCO and reduced DLCO
+  
+  dlco_class$observed <- lft_globals$analysis_tbl %>% 
+    transmute(index = 1:nrow(.), 
+              DLCO_percent = DLCO_percent, 
+              DLCO_reduced = DLCO_reduced, 
+              p_value = 1, 
+              data_set = 'observed')
+  
+  dlco_class$data <- dlco_class$data %>% 
+    map(left_join, 
+        dlco_class$observed[c('index', 'DLCO_percent')], 
+        by = 'index') %>% 
+    map(rbind, dlco_class$observed) %>% 
+    map(mutate, 
+        data_set = factor(data_set, c('observed', 'train', 'cv')))
+  
+  ## DLCO cuts: used later for construction of confusion matrices
+  ## for intervals of DLCO
+  
+  dlco_class$data <- dlco_class$data %>% 
+    map(mutate, 
+        dlco_interval = cut(DLCO_percent, 
+                            c(-Inf, 60, 80, 100, 120, Inf), 
[…]

### 2025-03-07 exploration scripts/comparison.R

@@ -155,11 +155,7 @@
                 'Effect size'))
 
 # END -------
-  
-  comp_ct$analysis_tbl <- NULL
-  
-  comp_ct <- compact(comp_ct)
-  
+
   plan('sequential')
   
   insert_tail()

### 2025-03-07 exploration scripts/ct.R

@@ -16,11 +16,7 @@
   insert_msg('Analysis globals')
   
   cohort_ct$lexicon <- covild$ct_lexicon
-  
-  cohort_ct$short_fup_labs <- globals$fup_labels %>% 
-    stri_replace(fixed = ' mo', replacement = '') %>% 
-    set_names(names(globals$fup_labels))
-  
+
   ## numeric and categorical variables
   
   cohort_ct$variables <- cohort_ct$lexicon %>% 
@@ -44,8 +40,22 @@
            cohort_split = factor(cohort_split, 
                                  c('cohort', 
                                    levels(cohort_ct$analysis_tbl$severity_class))), 
-           follow_up = droplevels(follow_up), 
-           analysis_split = interaction(follow_up, cohort_split)) %>% 
+           follow_up = droplevels(follow_up))
+  
+  ## appending the table with the dummy follow-up for all time points
+  
+  cohort_ct$analysis_tbl <- 
+    rbind(cohort_ct$analysis_tbl, 
+          mutate(cohort_ct$analysis_tbl, 
+                 follow_up = 'all time points')) %>% 
+    mutate(follow_up = factor(follow_up, 
+                              c('all time points', 
+                                levels(cohort_ct$analysis_tbl$follow_up))))
+  
+  ## defining the splitting factor
+  
+  cohort_ct$analysis_tbl <- cohort_ct$analysis_tbl %>% 
+    mutate(analysis_split = interaction(follow_up, cohort_split)) %>% 
     relocate(cohort_split) %>% 
     relocate(analysis_split) %>% 
     relocate(ID)
@@ -123,7 +133,7 @@
     map(~.x + 
           scale_fill_manual(values = globals$sev_colors) + 
           scale_color_manual(values = globals$sev_colors) + 
-          scale_x_discrete(labels = cohort_ct$short_fup_labs) + 
+          scale_x_discrete(labels = globals$short_fup_labs) + 
           theme(legend.position = 'none')) %>% 
     set_names(cohort_ct$variables$numeric)
   
@@ -146,7 +156,7 @@
          txt_color = 'white') %>% 
     map(~.x + 
           scale_fill_manual(values = c('steelblue', 'coral3')) + 
-          scale_x_discrete(labels = cohort_ct$short_fup_labs)) %>% 
+          scale_x_discrete(labels = globals$short_fup_labs)) %>% 
     set_names(cohort_ct$variables$factor)
 
 # END -----

### 2025-03-07 exploration scripts/lft.R

@@ -18,10 +18,6 @@
   cohort_lft$lexicon <- covild$lft_lexicon %>% 
     filter(!stri_detect(variable, regex = '(volume|DLCO)$'))
 
-  cohort_lft$short_fup_labs <- globals$fup_labels %>% 
-    stri_replace(fixed = ' mo', replacement = '') %>% 
-    set_names(names(globals$fup_labels))
-  
   ## numeric and categorical variables
 
   cohort_lft$variables <- cohort_lft$lexicon %>% 
@@ -45,8 +41,22 @@
            cohort_split = factor(cohort_split, 
                                  c('cohort', 
                                    levels(cohort_lft$analysis_tbl$severity_class))), 
-           follow_up = droplevels(follow_up), 
-           analysis_split = interaction(follow_up, cohort_split)) %>% 
+           follow_up = droplevels(follow_up))
+  
+  ## appending with the dummy level for all time points
+  
+  cohort_lft$analysis_tbl <- 
+    rbind(cohort_lft$analysis_tbl, 
+          mutate(cohort_lft$analysis_tbl, 
+                 follow_up = 'all time points')) %>% 
+    mutate(follow_up = factor(follow_up, 
+                              c('all time points', 
+                                levels(cohort_lft$analysis_tbl$follow_up))))
+  
+  ## specification of the splitting factor
+  
+  cohort_lft$analysis_tbl <- cohort_lft$analysis_tbl %>% 
+    mutate(analysis_split = interaction(follow_up, cohort_split)) %>% 
     relocate(cohort_split) %>% 
     relocate(analysis_split) %>% 
     relocate(ID)
@@ -124,7 +134,7 @@
     map(~.x + 
           scale_fill_manual(values = globals$sev_colors) + 
           scale_color_manual(values = globals$sev_colors) + 
-          scale_x_discrete(labels = cohort_lft$short_fup_labs) + 
+          scale_x_discrete(labels = globals$short_fup_labs) + 
           theme(legend.position = 'none')) %>% 
     set_names(cohort_lft$variables$numeric)
   
@@ -147,7 +157,7 @@
          txt_color = 'white') %>% 
     map(~.x + 
           scale_fill_manual(values = c('steelblue', 'coral3')) + 
-          scale_x_discrete(labels = cohort_lft$short_fup_labs)) %>% 
+          scale_x_discrete(labels = globals$short_fup_labs)) %>% 
     set_names(cohort_lft$variables$factor)
 
 # END -----

### 2025-03-07 exploration scripts/symptoms.R

@@ -18,10 +18,6 @@
   cohort_sympt$lexicon <- covild$symptoms_lexicon %>% 
     filter(variable != 'follow_up')
 
-  cohort_sympt$short_fup_labs <- globals$fup_labels %>% 
-    stri_replace(fixed = ' mo', replacement = '') %>% 
-    set_names(names(globals$fup_labels))
-  
   ## numeric and categorical variables
 
   cohort_sympt$variables <- cohort_sympt$lexicon %>% 
@@ -29,7 +25,7 @@
     map(~.x$variable)
 
   ## analysis table, introducing the severity variable
-  ## removing the acute timepoint - there was no study CT but just 
+  ## removing the acute time point - there was no study CT but just 
   ## a diagnostic scan for some participants
 
   cohort_sympt$analysis_tbl <- covild$symptoms %>% 
@@ -45,8 +41,22 @@
            cohort_split = factor(cohort_split, 
                                  c('cohort', 
                                    levels(cohort_sympt$analysis_tbl$severity_class))), 
-           follow_up = droplevels(follow_up), 
-           analysis_split = interaction(follow_up, cohort_split)) %>% 
+           follow_up = droplevels(follow_up))
+  
+  ## dummy factor for all follow-up points: all time points
+  
+  cohort_sympt$analysis_tbl <- cohort_sympt$analysis_tbl %>% 
+    rbind(cohort_sympt$analysis_tbl, 
+          mutate(cohort_sympt$analysis_tbl, 
+                 follow_up = 'all time points')) %>% 
+    mutate(follow_up = factor(follow_up,
+                              c('all time points', 
+                                levels(cohort_sympt$analysis_tbl$follow_up))))
+  
+  ## specification of the analysis splits
+  
+  cohort_sympt$analysis_tbl <- cohort_sympt$analysis_tbl %>% 
+    mutate(analysis_split = interaction(follow_up, cohort_split)) %>% 
     relocate(cohort_split) %>% 
     relocate(analysis_split) %>% 
     relocate(ID)
@@ -123,7 +133,7 @@
     map2(c('mMRC', 'ECOG'), 
          ~.x + 
           scale_fill_brewer(palette = 'Purples') + 
-          scale_x_discrete(labels = cohort_sympt$short_fup_labs)) %>% 
+          scale_x_discrete(labels = globals$short_fup_labs)) %>% 
     set_names(cohort_sympt$variables$numeric)
   
[…]

### 2025-03-07 lft.R

@@ -20,6 +20,8 @@
   library(sciplot)
   library(rcompanion)
   library(clustTools)
+  library(zoo)
+
   library(graphExtra)
   library(igraph)
 
@@ -48,6 +50,7 @@
 
   explore <- exda::explore
   train <- caret::train
+  components <- generics::components
 
   c('./tools/tools.R', 
     './tools/globals.R') %>% 
@@ -119,7 +122,8 @@
   c('./LFT scripts/binary_performance.R', 
     './LFT scripts/regression_performance.R', 
     './LFT scripts/residuals.R', 
-    './LFT scripts/ct_predictions.R') %>% 
+    './LFT scripts/ct_predictions.R', 
+    './LFT scripts/dlco_classification.R') %>% 
     source_all(message = TRUE, crash = TRUE)
 
   ## assessment of over-fitting by learning curves

### 2025-03-07 paper scripts/render.R

@@ -23,6 +23,15 @@
                                         reference_docx = 'ms_template.docx'), 
          output_dir = './paper')
   
+# Rendering the Rebuttal Letter -------
+  
+  insert_msg('Rendering the rebuttal letter')
+  
+  render('./paper/markdown/rebuttal_letter.Rmd', 
+         output_format = word_document2(number_sections = FALSE, 
+                                        reference_docx = 'ms_template.docx'), 
+         output_dir = './paper')
+  
 # END ----
   
   insert_tail()

### 2025-03-07 paper scripts/supplementary_figures.R

@@ -321,6 +321,55 @@
               w = 180, 
               h = 180)
   
+# Performance of the classification models of DLCO for DLCO percentages ------
+  
+  insert_msg('Performnce of DLCO classifiers for DLCO percentages')
+  
+  ## upper panel: class assignment and DLCO values
+  
+  suppl_figs$dlco_class$upper <- 
+    plot_grid(dlco_class$class_plots$class$gbm + 
+                labs(title = 'DLCO < 80%, GBM predictions'), 
+              ncol = 2, 
+              rel_widths = c(0.8, 0.2))
+  
+  ## bottom panel: moving accuracy and moving Brier square
+  
+  suppl_figs$dlco_class$bottom <- 
+    map2(dlco_class$moving_plots$gbm, 
+         c('DLCO < 80%, GBM prediction accuracy', ''), 
+         ~.x + 
+           labs(title = .y) + 
+           theme(legend.position = 'none')) %>% 
+    plot_grid(plotlist = ., 
+              ncol = 2, 
+              align = 'hv', 
+              axis = 'tblr') %>% 
+    plot_grid(get_legend(dlco_class$moving_plots$gbm[[1]] + 
+                           theme(legend.position = 'bottom')), 
+              nrow = 2, 
+              rel_heights = c(0.85, 0.15))
+  
+  ## the entire figure
+  
+  suppl_figs$dlco_class <- 
+    plot_grid(suppl_figs$dlco_class$upper, 
+              suppl_figs$dlco_class$bottom, 
+              nrow = 2, 
+              rel_heights = c(1, 1.25), 
+              labels = LETTERS, 
+              label_size = 10) %>% 
+    as_figure(label = 'reduced_dlco_predictions_dlco_percentage', 
+              ref_name = 'dlco_class', 
+              caption = paste('Accuracy of prediction of insufficient', 
+                              'diffusion capacity for carbon monoxide', 
+                              'by the GBM model as a function of', 
+                              'diffusion capacity for carbon monoxide', 
+                              'expressed as percentage of', 
+                              "patient's reference."), 
+              w = 180, 
+              h = 160)
+  
 # Learning curves for the models of DLCO < 80% and DLCO -------
   
   insert_msg('Learning curves for the models of reduced and percentage DLCO')

### 2025-03-07 paper scripts/supplementary_tables.R

@@ -9,12 +9,15 @@
 # Numbers of samples and participants per time point -------
   
   insert_msg('Numbers of samples and participants')
-  
+
   suppl_tabs$n_numbers <- cohort_ct$analysis_tbl %>% 
     blast(cohort_split) %>% 
     map(count, follow_up) %>% 
     reduce(left_join, by = 'follow_up') %>% 
-    mutate(follow_up = stri_extract(follow_up, regex = '\\d{1}-month')) %>% 
+    mutate(follow_up = stri_extract(follow_up, regex = '\\d+-month'), 
+           follow_up = ifelse(is.na(follow_up), 
+                              'all time points', 
+                              follow_up)) %>% 
     set_names(c('Follow-up, months', 
                 'Cohort', 
                 'Ambulatory, mild COVID-19', 

### 2025-03-07 report scripts/links.R

@@ -17,7 +17,8 @@
          'caretExtra' = 'https://github.com/PiotrTymoszuk/caretExtra', 
          'lmqc' = 'https://github.com/PiotrTymoszuk/lmqc', 
          'bootStat' = 'https://github.com/PiotrTymoszuk/bootStat', 
-         'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools') %>% 
+         'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools', 
+         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra') %>% 
     compress(names_to = 'obj_name', 
              values_to = 'x') %>% 
     mutate(ref_name = paste0('_', obj_name, '_'))

### 2025-03-07 tools/globals.R

@@ -86,14 +86,20 @@
                          '3-month follow-up' = 'coral3', 
                          '6-month follow-up' = 'steelblue2', 
                          '12-month follow-up' = 'steelblue4', 
-                         'cohort' = 'plum4')
+                         'cohort' = 'plum4', 
+                         'all time points' = 'plum4')
   
   globals$fup_labels = c('acute COVID-19' = 'acute CoV', 
                          '2-month follow-up' = '2 mo', 
                          '3-month follow-up' = '3 mo', 
                          '6-month follow-up' = '6 mo', 
                          '12-month follow-up' = '12 mo', 
-                         'cohort' = 'CovILD cohort')
+                         'cohort' = 'CovILD cohort', 
+                         'all time points' = 'all')
+  
+  globals$short_fup_labs <- globals$fup_labels %>% 
+    stri_replace(fixed = ' mo', replacement = '') %>% 
+    set_names(names(globals$fup_labels))
   
 # Algorithm labels and colors ------
   

### 2025-03-09 paper/markdown/rebuttal_letter.Rmd
Rebuttal letter
Backbone of the rebuttal letter, figure for the reviewers

@@
 
+_Anna and Gerlig: I think, we can stress that we were able to model DLCO but not other LFT responses and include some key characteristics of the machine learning models of DLCO < 80% and DLCO.
+My proposal for the response below._
+
+As suggested, we state now in Abstract, that we were able to establish meaningful models of DLCO and DLCO < 80% but not for other LFT responses. 
+We also included key cross-validated performance statistics of meaningful machine learning models at predicting 
+DLCO < 80% (accuracy: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$correct_rate %>% range %>% signif(2) %>% paste(collapse = ' - ')`, Cohen's $\kappa$: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$kappa %>% range %>% signif(2) %>% paste(collapse = ' - ')`, AUC: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$AUC %>% range %>% signif(2) %>% paste(collapse = ' - ')`) and
+DLCO expressed as percentage of the patient's reference (mean absolute error [MAE]: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>% .$MAE %>% range %>% signif(3) %>% paste(collapse = ' - ')`%, $R^2$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm != 'nnet') %>% .$rsq %>% range %>% signif(2) %>% paste(collapse = ' - ')`).
+Please note that the later performance statistics of DLCO as % of the patient's reference, we provide the performance metric range for the random forest, GBM, and SVM algorithm, as the neural network model yielded a poor fit (MAE: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>% .$MAE %>% signif(2)`%, $R^2$: `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>% .$rsq %>% signif(2)`).
+
 ## Point 2
@@
 ### Response
+
+_Anna and Gerlig: please answer, thanks!_
 
@@
 
+_Anna and Gerlig: as above. Below some papersa I found interesting._ 
+_https://pubs.rsna.org/doi/full/10.1148/radiol.222888: no modeling but reduced DLCO in COVID-19 survivors without or delayed radiological resolution._ 
+_https://aapm.onlinelibrary.wiley.com/doi/10.1002/mp.16915: a deep learning approach to modeling LFT, including DLCO with CT data in large data sets (s8K+ for training) of CT scans, allows for DLO predictions for single lobes; please have a look at Figure 3, where they report correlations between the observed and predicted DLCO - they are not so much better than our best GBM model of DLCO._
+_https://pubs.rsna.org/doi/full/10.1148/radiol.221488: they succeeded at predictions of FEV1, FVC, and FEV1:FVC using deep learning (neuronal network) and a large data set of CT images._
+_https://pmc.ncbi.nlm.nih.gov/articles/PMC8264499/: a nice review with a short description of algorithms._
+
 ## Point 4
@@
 
+_Anna and Gerlig, here some my thoughts: As I see our results as a bigger picture: we were limited by the number of patients and by the fact that we worked with processed CT data: scores and values of opacity/high opacity, and human-identified findings - for this limited information, we established quite nice models of DLCO and reduced DLCO < 80%._
+_There are already some papers on machine learning and lung function._ 
+_What looks pretty novel to me: we are among the first groups who used not only the CT information for modeling but also clinical and demographic data._ 
+_We also integrate information derived from CT by a software and a human radiologist, which is also a quite novel approach._
+
 ## Point 5
@@
 ### Response
+
+Our manuscript is accompanied by a thorough information on patients, procedures, and study variables provided in __Supplementary Methods__. 
+Please refer to __Supplementary Table S2__ for a detailed listing of modeling response and explanatory variables with their format and descriptions.
 
@@
 
+Thanks for this point. 
+We agree that the information on hyper-parameters in the initial submission was hard to understand for many clinical readers and too specific. 
+In the revised manuscript, we provide short descriptions of the hyper-parameters in __Supplementary Methods/Multi-parameter modeling of lung function__. 
+These descriptions were also added as footnotes of __Supplementary Table S6__.
+
 ## Point 7
@@
 ### Response
+
+Thanks for this question. 
+The hyper-parameters were found by searching grids of values with the maximum Youden's J statistic for the classification models, and minimum mean absolute error (MAE) for the regression models in 10-repeats 10-fold patient-blocked cross-validation as the selection criteria.
+This tuning procedure was executed by calling `train()` function from _caret_ package [@Kuhn2008], which, together with our open-source package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra), belongs to our standard toolbox for explainable machine learning.
+This information is provided in __Supplementary Methods/Multi-parameter modeling of lung function__.
 
@@
 
+_Anna and Gerlig: you may consider examples of papers above._
+
 ## Point 9
@@
 
+_Anna and Gerlig: please respod, thanks!_
+
 ## Point 10
@@
 ### Response
+
+_Anna and Gerlig: I have few thoughts on conclusions:)_ 
+_First of all: we demonstrate the feasibility of a machine learning approach combining CT readouts with demographic and clinical information on acute COVID-19 and convalescence at predicting DLCO._ 
+_Second: this modeling strategy merges information derived from CT images by a software and a human radiologist._
+_Third: while we constructed meaningful models of DLCO, our machine learning models failed to predict FEV1 and FVC.
+This lets us reason that DLCO but not FEV1 or FVC is the prime LFT readout of post-COVID-19 structural lung damage.
+By univariable analyses we show that DLCO is tightly associated with CT readouts of severity of lung damage such as CTSS, opacity, or high opacity._
 
@@
 
+_Anna and Gerlig: please respond.
+From my quite cursory search of the literature, I've an impression that there are indeed reports with more or less successful trials of prediction of LFT based on CT data - mostly raw images._
+_In turn, I was not able to find a single modeling papers, where someone tried to combine the demographic, clinical, and rariological information to predict LFT outcomes._ 
+_One can say: that the clinic and demography is irrelevant, once you have CT data._ 
+_To me it is not, our combination approach let us understand the clinical background and identify some broad risk hactors such as patient's sex and comorbidities._
+
 ## Point 12
@@
 ### Response
+
+_Anna and Gerlig: see my thoughts to Point 11._
 
@@
 
+We appreciate this point. 
+While the CovILD cohort has been already extensively described in our previous papers [@Luger2022; @Sonnweber2022; @Sonnweber2020; @Sahanic2023], we provided in the revised manuscript information on the drop-ou rates at particular time points as compared with the initial collective of n = 145 patients.
+We also references numbers and percentages for the categorical modeling responses (DLCO < 80%, FEV1 < 80%, and FVC < 80%).
+
 ## Point 14
@@
 
+_Anna and Gerlig: I'd suggest 1 - 2 sentences in Discussion with emphasis that CTSS is a human-derived discretized severity score averaged over the entire lung, and opacity and high opacity are continuous readouts of lesion extent computed by the software.
+CTSS is expected to be more coarse the opacity or high opacity and cannot accurately capture severe, highly localized cases such as severe damage restricted to only one lobe (it is scored with 5, the same as minimal damage in all lobes).
+In our data set, CTSS, opacity, and high opacity are strongly inter-correlated._
+
+## Point 15
+
+### Issue
+
 Overemphasis on “satisfactory accuracy” without deeper clinical context.
 
-## Point 15
+### Response
+
+We apologize for this vague term. 
+In the revised manuscript, avoid suggestive phrases on model performance and stick to teh performance metrics and evaluations.
+
+## Point 16
 
@@
 
-## Point 16
+In __Review Figure \@ref(fig:fig-dlco-correlations)__ we present correlations between the observed DLCO expressed as percentages of the patient's reference and the out-of-fold predictions of DLCO in participant-blocked 10-repeats 10-fold cross-validation for all four machine learning algorithm.
+To facilitate the comparison, we set ranges of the X and Y axes to the range of observed DLCO values - we applied this transformation to the respective scatter plots in __Figure 2__ as well. 
+From this visualization it is that predictions by the neural network algorithm were shrunk to the 75% - 100% range and the neural network algorithm failed to predict cases with very low and very high values of DLCO.
+We can only speculate on the reasons for this behavior, yet, inclusion of more observations with at the lowest and the highest tail of DLCO is expected to improve performance of not only the neural net but also of the remaining models.
+
+## Point 17
 
@@
 
-The SHAP analysis identifies “top 15” predictors. However, the manuscript does not address correlations among those predictors or discuss whether certain features might be redundant.
+The SHAP analysis identifies “top 15” predictors. 
+However, the manuscript does not address correlations among those predictors or discuss whether certain features might be redundant.
 
@@
 
-## Point 17
+Thanks for this is a very good point. 
+In the initial manuscript, we presented results of exploratory analysis of the modeling data set including the overlap between radiological lung findings, and correlations between opacity, high opacity and CTSS (__Supplementary Figure S3__) - which were also identified as highly influential explanatory factors for prediction of DLCOP < 80% and DLCO (__Figure 3__ and __Supplementary Figure S14__).
+In the revised manuscript, we specifically addressed associations between the most influential explanatory variables with an analysis of correlations graph [@Csardi2006; @Zhang2005] (__Supplementary Figure S15__).
+The graph edges were defined by pairwise correlations between the explanatory factors with Kendall's $\tau \geq$ 0.3 and weighted by $\tau$ values. 
+For computation of the correlations, ordinal factors such as presence of GGS of severity of acute COVID-19 were transformed to integers. 
+This analysis supports the findings of the initial manuscript version that the CT readouts: CTSS, opacity, high opacity, GGO and reticulations are tightly inter-correlated. 
+Another cluster of tightly associated features was formed by readouts of severity of ACUTE COVID-19 such as acute COVID-19 severity, hospital and ICU stay, and anti-infective treatment during acute COVID-19.
+Another community of the influential features was made up by known risk factors of severe COVID-19 course (age, cardiovascular disease, hypercholesterolemia, and endocrine/metabolic disease). 
+This leads us to the conclusion that there was indeed some redundancy between the modeling variables and a potential for pruning e.g. by regularization. 
+These new results were also briefly discussed.
+
+## Point 18
 
@@
 
-## Point 18
+_Anna and Gerlig: please respond._
+
+## Point 19
 
@@
 
-The paper concedes “model overfitting” could be an issue.
-However, it lacks any systematic attempt (e.g., a learning-curve analysis or thorough regularization strategy) to demonstrate that overfitting is minimal or under control.
+The paper concedes “model over-fitting” could be an issue. 
+However, it lacks any systematic attempt (e.g., a learning-curve analysis or thorough regularization strategy) to demonstrate that over-fitting is minimal or under control.
 
@@
 
+Thanks for raising this important issue. 
+As suggested, we tacked it by a learning curve analysis [@Viering2023] for the predictions of DLCO < 80% and DLCO by the best performing GBMm models presented in __Supplementary Figure S13__. 
+The over-fitting was particularly evident for the predictions of DLCO < 80% as visualized by a slow convergence of curves of accuracy and Cohen's $\kappa$ for training and test subsets of increasing sizes derived from the modeling data and cross-validation.
+By contrast, we can argument that over-fitting of for predictions of DLCO expressed as percentages of the patient's reference was under control. 
+We list over-fitting explicitly as a limitation in the revised manuscript and propose regularized algorithms such as XGBoost as a possible remedy. 
+We we will certainly consider such algorithms in future studies.
+
+In order to investigate the over-fitting behavior of the models of DLCO < 80%, we resorted to an analysis of the cross-validated model's performance as a function of increasing DLCO percentages. 
+As shown in __Supplementary Figure S12__, we found out that the models made uncertain/erroneous predictions - and hence generalized poorly - for observations with DLCO ranging from 70% - 80% of the patient's reference.
+This phenomenon can be explained by the observation that the 80% DLCO cutoff - used in multiple COVID-19 papers [@Zhang2022a; @Sonnweber2020] did not yielded two clearly separated groups of observations in our modeling data set and may not correspond to differences in demographic and clinical background.
+In addition, we believe that expansion observations with DLCO < 80%, either be increasing the cohort size or augmentation of the data set, e.g.
+by SMOTE [@Chawla2002], may at least partly correct this behavior.
+
+# Review Figures
+
+```{r fig-dlco-correlations, fig.width = figur::convert(rev_figs$dlco_correlations, to = 'in')$w, fig.height = figur::convert(rev_figs$dlco_correlations, to = 'in')$h, fig.cap = 'Correlation of predicted and observed diffusion capacity for carbon monoxide during COVID-19 convalescence'}
+
+rev_figs$dlco_correlations$plot
+
+```
+
+__Review Figure \@ref(fig:fig-dlco-correlations). Correlation of predicted and observed diffusion capacity for carbon monoxide during COVID-19 convalescence__
+
+_Machine learning models of diffusion capacity for carbon monoxide (DLCO) expressed as percentage of the patient's reference were developed with the random forest, gradient boosted machines (GBM), support vector machines with radial kernel (SVM), and single hidden layer neural network as presented in Figure 2 of the manuscript._
+_Correlations between the predicted and observed DLCO´values were metered by Spearman's correlation coefficient $\rho$._ 
+_The observed DLCO values and out-of-fold DLCO predictions (participant-blocked 10-repeats 10-fold cross-validation [CV]) are visualized with scatter plots.
+Each point represents a single observation.
+The blue dashed lines with slope 1 and intercept 0 represent absolutely accurate predictions, general additive model trends with standard errors are visualized as the solid blue lines with gray ribbons.
+Numbers of complete observations and $\rho$ coefficients are displayed in the plot captions.
+Ranges of displayed DLCO values were set to the range of observed DLCO._
+
+\newpage
+
 # References

### 2025-03-09 paper/markdown/manuscript.Rmd
Rebuttal letter
Backbone of the rebuttal letter, figure for the reviewers

@@
 Fourth, as inferred from the analysis of learning curves, the models of insufficient DLCO suffered for substantial over-fitting. 
-While this problem can be partially traced back to an unsharp distinction between sufficient and reduced DLCO with the 80% reference cutoff, a modeling approach employing regularized machine learning algorithm like XGBoost or regularized neural networks may further improve quality of predictions.
-Furthermore, most of the highly influential explanatory variables for predictions of DLCO and reduced DLCO, and in particular the CT readouts of lung damage were strongly inter-correlated, which raises questions about redundancy of explanatory variables.
+While this problem can be partially traced back to an unsharp distinction between sufficient and reduced DLCO with the 80% patient's reference cutoff, a modeling approach employing regularized machine learning algorithm like XGBoost or regularized neural networks may further improve quality of predictions.
+Fifth, most of the highly influential explanatory variables for predictions of DLCO and reduced DLCO, and in particular the CT readouts of lung damage were strongly inter-correlated, which raises questions about redundancy of explanatory variables.
+Furthermore, CT-related explanatory variables were already processed and available in form of software- and human-derived severity metrics and findings without any spatial information. 
+Inclusion of raw CT image information in the machine learning models would likely enhance their accuracy, as demonstrated by literature reports. 
 Finally, the analyzed cohort was recruited in the initial phase of the pandemic and consisted of individuals infected with the wild-type variant of the SARS-CoV-2 virus. 
@@
 
-__Figure \@ref(fig:fig-ml-performance). Evaluation of performance of machine learning models of diffusion capacity for CO during COVID-19 convalescence.__
+__Figure \@ref(fig:fig-ml-performance). Evaluation of performance of machine learning models of diffusion capacity for carbomn monoxide during COVID-19 convalescence.__
 
@@
 Scatter plots: observed and predicted values of diffusion capacity for carbon monoxide in cross-validation folds; the blue dashed lines with slope 1 and intercept 0 represent absolutely accurate predictions, general additive model trends with standard errors are visualized as the solid blue lines with gray ribbons.
-Numbers of complete observations and Spearman's $\rho$ coefficients of correlation between the predicted and observed values are displayed in the plot captions._
+Numbers of complete observations and Spearman's $\rho$ coefficients of correlation between the predicted and observed values are displayed in the plot captions.
+Ranges of the displayed DLCO values were set to the range of observed DLCO._
 

### 2025-03-09 paper/markdown/supplementary_material.Rmd
Rebuttal letter
Backbone of the rebuttal letter, figure for the reviewers

@@
 
-Hyper-parameters that control the modeling algorithm's behavior were found by maximizing the Youden's J statistic 
+Hyper-parameters that control the modeling algorithm's behavior were found by searching grids of hyper-parameter values with maximizing the Youden's J statistic 
 ($J = Sensitivity + Specificity - 1$, classification models of binary LFT outcomes) or by minimizing mean absolute error (MAE, regression models of numeric LFT outcomes) in 10-repeats 10-fold cross-validation, in a process called tuning.
@@
 Construction of such 'participant-wise' blocked cross-validation folds was done with an in-house-developed wrapper around `createMultiFolds()` function (package _caret_). 
-For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed [@Kuhn2008]. 
+For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed; the tuning/training setting such as participant-blocked indexes for cross-validation folds and Youden's J cost function was specified by via a `trainControl()` object provided to the `train()` function [@Kuhn2008].
 The optimal values of the tuning parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.
@@
 For construction of the correlation graph, ordered categorical variables (i.e. factors in R) were converted to integers. 
-The graph edges were defined by pairwise correlations between the variables with Kendall's $\tau \geq 0.3$. 
+The graph edges were defined by pairwise correlations between the variables with tie-adjusted Kendall's $\tau \geq 0.3$. 
 The edge weights corresponded to values of the $\tau$ correlation coefficients. 
-The graph was constructed with `as_iGraph()` function; isolated vertices, i.e.
+The graph was constructed with `as_iGraph()` function, the correlation matrix was computed with `cor()` function from _stats_ package passed to`as_iGraph()`. 
+Isolated vertices, i.e.
 variables without any correlations with $\tau \geq$ 0.3 were removed from the graph with `prune_degree()` (both functions from [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra)).

### 2025-03-09 paper scripts/figures.R

@@ -6,6 +6,12 @@
   
   paper_figs <- list()
   
+# Figure globals --------
+  
+  insert_msg('Figure globals')
+  
+  dlco_range <- range(lft_globals$analysis_tbl$DLCO_percent)
+  
 # Figure 1: analysis inclusion scheme -------
   
   insert_msg('Figure 1: analysis inclusion scheme')
@@ -24,12 +30,15 @@
   insert_msg('Figure 2: model performance')
   
   paper_figs$ml_performance <- 
-    list(bin_models$performance_plots$DLCO_reduced, 
-         bin_models$roc_plots$DLCO_reduced$cv, 
-         reg_models$performance_plots$DLCO_percent, 
-         reg_models$scatter_plots$DLCO_percent$cv$gbm,
-         reg_models$scatter_plots$DLCO_percent$cv$ranger, 
-         reg_models$scatter_plots$DLCO_percent$cv$svmRadial) %>% 
+    c(list(bin_models$performance_plots$DLCO_reduced, 
+           bin_models$roc_plots$DLCO_reduced$cv, 
+           reg_models$performance_plots$DLCO_percent), 
+      reg_models$scatter_plots$DLCO_percent$cv[c("gbm", 
+                                                 "ranger", 
+                                                 "svmRadial")] %>% 
+        map(~.x + 
+              scale_x_continuous(limits = dlco_range) + 
+              scale_y_continuous(limits = dlco_range))) %>% 
     map(~.x + 
           theme(legend.position = 'none', 
                 legend.title = element_blank())) %>% 
@@ -154,5 +163,7 @@
          device = cairo_pdf)
   
 # END ------
+  
+  rm(dlco_range)
   
   insert_tail()

### 2025-03-09 paper scripts/reviewer_figures.R

@@ -1,0 +1,53 @@
+# Figures for the reviewers
+
+  insert_head()
+  
+# container --------
+  
+  rev_figs <- list()
+  
+# analysis globals -------
+  
+  insert_msg('Analyis globals')
+  
+  ## range of observed DLCO: to be used as axis limits
+  
+  dlco_range <- range(lft_globals$analysis_tbl$DLCO_percent)
+
+# Correlation of the observed and predicted DLCO for all models --------
+  
+  insert_msg('Correlation of the predicted and observed DLCO, all models')
+  
+  ## and common scales
+  
+  rev_figs$dlco_correlations <- 
+    reg_models$scatter_plots$DLCO_percent$cv %>% 
+    map(~.x + 
+          scale_x_continuous(limits = dlco_range) + 
+          scale_y_continuous(limits = dlco_range)) %>% 
+    plot_grid(plotlist = .,
+              ncol = 2, 
+              align = 'hv') %>% 
+    as_figure(label = 'dlco_predicted_observed_correlations', 
+              ref_name = 'dlco_correlations', 
+              caption = paste('Correlation of predicted and observed diffusion', 
+                              'capacity for carbon monoxide during COVID-19', 
+                              'convalescence'), 
+              w = 180, 
+              h = 180)
+  
+# Saving figures on the disc -------
+  
+  insert_msg('Saving figures on the disc')
+  
+  rev_figs %>% 
+    number_figures(prefix = 'reviewer_figure_') %>% 
+    walk(pickle, 
+         path = './paper/reviewer figures', 
+         device = cairo_pdf)
+  
+# END ------
+  
+  rm(dlco_range)
+  
+  insert_tail()

### 2025-03-09 paper.R

@@ -38,6 +38,7 @@
     './paper scripts/supplementary_tables.R', 
     './paper scripts/figures.R', 
     './paper scripts/supplementary_figures.R', 
+    './paper scripts/reviewer_figures.R', 
     './paper scripts/render.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   

### 2025-03-11 paper/markdown/rebuttal_letter.Rmd
Proofreading and author's feedback

@@
 
-_Anna and Gerlig: as above. Below some papersa I found interesting._ 
+_Anna and Gerlig: as above. Below some papers I found interesting._ 
 _https://pubs.rsna.org/doi/full/10.1148/radiol.222888: no modeling but reduced DLCO in COVID-19 survivors without or delayed radiological resolution._ 
-_https://aapm.onlinelibrary.wiley.com/doi/10.1002/mp.16915: a deep learning approach to modeling LFT, including DLCO with CT data in large data sets (s8K+ for training) of CT scans, allows for DLO predictions for single lobes; please have a look at Figure 3, where they report correlations between the observed and predicted DLCO - they are not so much better than our best GBM model of DLCO._
+_https://aapm.onlinelibrary.wiley.com/doi/10.1002/mp.16915: a deep learning approach to modeling LFT, including DLCO with CT data in large data sets (8K+ for training) of CT scans, allows for DLCO predictions for single lobes; please have a look at Figure 3, where they report correlations between the observed and predicted DLCO - they are not so much better than our best GBM model of DLCO._
 _https://pubs.rsna.org/doi/full/10.1148/radiol.221488: they succeeded at predictions of FEV1, FVC, and FEV1:FVC using deep learning (neuronal network) and a large data set of CT images._
@@
 Our manuscript is accompanied by a thorough information on patients, procedures, and study variables provided in __Supplementary Methods__. 
-Please refer to __Supplementary Table S2__ for a detailed listing of modeling response and explanatory variables with their format and descriptions.
+Please refer to __Supplementary Table S2__ for a detailed listing of modeling responses and explanatory variables with their format and descriptions.
 
@@
 We appreciate this point. 
-While the CovILD cohort has been already extensively described in our previous papers [@Luger2022; @Sonnweber2022; @Sonnweber2020; @Sahanic2023], we provided in the revised manuscript information on the drop-ou rates at particular time points as compared with the initial collective of n = 145 patients.
+While the CovILD cohort has been already extensively described in our previous papers [@Luger2022; @Sonnweber2022; @Sonnweber2020; @Sahanic2023], we provided in the revised manuscript information on the drop-out rates at particular time points as compared with the initial collective of n = 145 patients.
 We also references numbers and percentages for the categorical modeling responses (DLCO < 80%, FEV1 < 80%, and FVC < 80%).
@@
 To facilitate the comparison, we set ranges of the X and Y axes to the range of observed DLCO values - we applied this transformation to the respective scatter plots in __Figure 2__ as well. 
-From this visualization it is that predictions by the neural network algorithm were shrunk to the 75% - 100% range and the neural network algorithm failed to predict cases with very low and very high values of DLCO.
-We can only speculate on the reasons for this behavior, yet, inclusion of more observations with at the lowest and the highest tail of DLCO is expected to improve performance of not only the neural net but also of the remaining models.
+In this visualization it is clear that predictions by the neural network algorithm were shrunk to the 75% - 100% range and the neural network algorithm failed to predict cases with very low and very high values of DLCO.
+We can only speculate on the reasons for this behavior, yet, inclusion of more observations with at the lowest and the highest tail of DLCO is expected would likely improve performance of not only the neural net but also of the remaining models.
 
@@
 Thanks for this is a very good point. 
-In the initial manuscript, we presented results of exploratory analysis of the modeling data set including the overlap between radiological lung findings, and correlations between opacity, high opacity and CTSS (__Supplementary Figure S3__) - which were also identified as highly influential explanatory factors for prediction of DLCOP < 80% and DLCO (__Figure 3__ and __Supplementary Figure S14__).
-In the revised manuscript, we specifically addressed associations between the most influential explanatory variables with an analysis of correlations graph [@Csardi2006; @Zhang2005] (__Supplementary Figure S15__).
+In the initial manuscript, we presented results of exploratory analysis of the modeling data set including the overlap between radiological lung findings, and correlations between opacity, high opacity and CTSS (__Supplementary Figure S3__) - which were also identified as highly influential explanatory factors for prediction of DLCO < 80% and DLCO (__Figure 3__ and __Supplementary Figure S14__).
+In the revised manuscript, we specifically addressed associations between the most influential explanatory variables with an analysis of a correlation graph [@Csardi2006; @Zhang2005] (__Supplementary Figure S15__).
 The graph edges were defined by pairwise correlations between the explanatory factors with Kendall's $\tau \geq$ 0.3 and weighted by $\tau$ values. 
-For computation of the correlations, ordinal factors such as presence of GGS of severity of acute COVID-19 were transformed to integers. 
-This analysis supports the findings of the initial manuscript version that the CT readouts: CTSS, opacity, high opacity, GGO and reticulations are tightly inter-correlated. 
-Another cluster of tightly associated features was formed by readouts of severity of ACUTE COVID-19 such as acute COVID-19 severity, hospital and ICU stay, and anti-infective treatment during acute COVID-19.
+For computation of the correlations, ordinal factors such as presence of GGO or severity of acute COVID-19 were transformed to integers. 
+This analysis supports the findings of the initial manuscript version that the CT readouts, CTSS, opacity, high opacity, GGO and reticulations, are tightly inter-correlated. 
+Another cluster of tightly associated features was formed by readouts of severity of acute COVID-19 such as acute COVID-19 severity, hospital and ICU stay, and anti-infective treatment during acute COVID-19.
 Another community of the influential features was made up by known risk factors of severe COVID-19 course (age, cardiovascular disease, hypercholesterolemia, and endocrine/metabolic disease). 
@@
 As shown in __Supplementary Figure S12__, we found out that the models made uncertain/erroneous predictions - and hence generalized poorly - for observations with DLCO ranging from 70% - 80% of the patient's reference.
-This phenomenon can be explained by the observation that the 80% DLCO cutoff - used in multiple COVID-19 papers [@Zhang2022a; @Sonnweber2020] did not yielded two clearly separated groups of observations in our modeling data set and may not correspond to differences in demographic and clinical background.
-In addition, we believe that expansion observations with DLCO < 80%, either be increasing the cohort size or augmentation of the data set, e.g.
-by SMOTE [@Chawla2002], may at least partly correct this behavior.
+This phenomenon can be explained by the observation that the 80% DLCO cutoff used in multiple COVID-19 papers [@Zhang2022a; @Sonnweber2020] did not yielded two clearly separated groups of observations in our modeling data set and may not correspond to substantial differences in demographic and clinical background.
+In addition, we believe that expansion of observations with DLCO < 80%, either be increasing the cohort size or augmentation of the data set, e.g.
+by SMOTE [@Chawla2002], would at least partly correct this behavior.
 

### 2025-03-11 paper/markdown/manuscript.Rmd
Proofreading and author's feedback

@@
 Such models employed common machine learning algorithms with a wide range of demographic and clinical features, as well as human-determined CT abnormalities, and AI-measured lung opacity and high opacity as explanatory variables.
-The secondary analysis endpoint was prediction of LFT abnormalities and numeric LFT readouts during COVID-19 recovery by human-determined CTSS and AI-measured lung opacity and high opacity. 
-This endpoint was addressed by statistical hypothesis testing, correlation and receiver-operating characteristic (ROC) analysis.
+The secondary analysis endpoints were assessment of importance of explanatory variables for the machine learning predictions of impaired lung function and prediction of LFT abnormalities and numeric LFT readouts during COVID-19 recovery by human-determined CTSS and AI-measured lung opacity and high opacity.
+These endpoints were addressed by Shapley additive explanations (SHAP) as well as statistical hypothesis testing, correlation and receiver-operating characteristic (ROC) analysis.
 
@@
 
-__Figure \@ref(fig:fig-ml-performance). Evaluation of performance of machine learning models of diffusion capacity for carbomn monoxide during COVID-19 convalescence.__
+__Figure \@ref(fig:fig-ml-performance). Evaluation of performance of machine learning models of diffusion capacity for carbon monoxide during COVID-19 convalescence.__
 

### 2025-03-14 paper/markdown/manuscript.Rmd
Tuning grids in Supplementary Material
Showing the complete tuning grids in Supplementary Tables. Supplementary Figures for tuning of the models of DLCO < 80% and DLCO

@@
 support vector machines (SVM) with radial kernel [@Weston1998]. 
-Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Table S6__).
-Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S7__ and __S8__).
+Because the CovILD data set consisted of non-independent, participant-matched observations, blocked cross-validation with participant's identifier serving as the grouping variable was used for optimization of the model hyper-parameters (tuning, __Supplementary Tables S6__ and __S7__, __Supplementary Figures S8__ - __S10__).
+Predictions of the models were finally evaluated by comparison with the observed outcome in the training data and the blocked cross-validation folds [@Kuhn2008] (__Supplementary Figure S8__, __Table \@ref(tab:tab-bin-classifiers)__ and __\@ref(tab:tab-reg-models)__,  __Supplementary Tables S8__ and __S9__).
 
@@
 Low cross-validated Brier score values for the models of insufficient DLCO indicated good overall calibration (Brier score: `r bin_models$stats$DLCO_reduced %>% filter(dataset == 'cv') %>% .$brier_score %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Figure \@ref(fig:fig-ml-performance)A__).
-In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S9__).
+In turn, models of insufficient FVC and FEV1 performed poorly in the cross-validation setting (overall accuracy: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$correct_rate) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, AUC: `r bin_models$stats[c("FVC_reduced", "FEV1_reduced")] %>% map(filter, dataset == 'cv') %>% map(~.x$AUC) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S11__).
 
@@
 By contrast, the neural network model of DLCO suffered from large error and low fraction of explained variance in the cross-validation setting (MAE = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$MAE %>% signif(3)`, pseudo-R^2^ = `r reg_models$stats$DLCO_percent %>% filter(dataset == 'cv', algorithm == 'nnet') %>%  .$rsq %>% signif(2)`, __Figure \@ref(fig:fig-ml-performance)B__).
-No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S10__).
+No meaningful models of FVC or FEV1 could be developed with any of the machine learning algorithms (cross-validation, R^2^: `r reg_models$stats[c("FVC_percent", "FEV1_percent")] %>% map(filter, dataset == 'cv') %>% map(~.x$rsq) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S12__).
 
 Concerning the models of insufficient DLCO, the highest concordance between the predicted and observed DLCO < 80% of the patient's reference was achieved for moderate COVID-19 convalescents at the two- to six-month follow-up examinations (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'hospitalized moderate', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`).
-The accuracy was in turn the poorest for ambulatory patients at the six- and twelve-month follow-up (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('6-month follow-up', '12-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`, __Supplementary Figure S11A__).
+The accuracy was in turn the poorest for ambulatory patients at the six- and twelve-month follow-up (all algorithms, $\kappa$: `r dlco_resid$stats$kappa %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('6-month follow-up', '12-month follow-up')) %>% map(~.x$kappa) %>% reduce(c) %>% range %>% signif(2)`, __Supplementary Figure S13A__).
 The Random Forest, SVM and GBM models of DLCO expressed as percentage of the patient's reference exhibited the lowest average errors for moderate COVID-19 survivors (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized moderate') %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
-In turn, DLCO values were systematically overestimated for severe COVID-19 patients at the two- to six-month follow-up visits (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized severe', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and underestimated for ambulatory COVID-19 convalescents at the same time points (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S11B__).
+In turn, DLCO values were systematically overestimated for severe COVID-19 patients at the two- to six-month follow-up visits (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'hospitalized severe', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and underestimated for ambulatory COVID-19 convalescents at the same time points (mean error: `r dlco_resid$stats$DLCO_residuals[c("ranger", "svmRadial", "gbm")] %>% map(filter, severity_class == 'ambulatory mild', follow_up %in% c('2-month follow-up', '3-month follow-up', '6-month follow-up')) %>% map(~.x$mean) %>% reduce(c) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, __Supplementary Figure S13B__).
 The best performance in the hospitalized moderate subset is likely attributed to the large number of observations and frequent DLCO deficits in this group of participants available for training of the models (n = `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate', DLCO_reduced == 'yes') %>% nrow` observations of DLCO < 80% in `r lft_globals$analysis_tbl %>% filter(severity_class == 'hospitalized moderate') %>% nrow` observations from moderate COVID-19 patients).
@@
 Prediction of reduced DLCO < 80% was assessed in observations with increasing values of DLCO expressed as percentage of the patient's reference by moving averages of accuracy and error metered as squared distance to the 0/1-coded outcome.
-As presented in __Supplementary Figure S12__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference.
+As presented in __Supplementary Figure S14__ for the best performing GBM model, accuracy was the lowest and the error peaked for observations with DLCO ranging between 70% and 80% of the patient's reference.
 An analogical phenomenon was observed for the remaining machine learning algorithms (not shown). 
@@
 The performance evaluation was done for the training subsets, test subsets (one-forth of observations not used for the model training), and 10-repeats 10-fold cross-validation [@Viering2023]. 
-The learning curves for the best-performing GBM models are presented in __Supplementary Figure S13__, for the remaining algorithms comparable results were obtained (not shown). 
+The learning curves for the best-performing GBM models are presented in __Supplementary Figure S15__, for the remaining algorithms comparable results were obtained (not shown). 
 As inferred from substantial differences of accuracy and Cohen's $\kappa$ between the training, test, and cross-validation subsets even for the largest sizes of the training data, models of insufficient DLCO suffer from over-fitting, i.e.
@@
 rating of physical performance impairment (ECOG) and 
-exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S14__, __Supplementary Table S9__).
+exertional dyspnea (mMRC) (__Figure \@ref(fig:fig-varimp-dlco-red)__, __Supplementary Figure S16__, __Supplementary Table S10__).
 
-Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S15A__).
-In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated.
+Analysis of a correlation graph of the most influential explanatory factors for the predictions of DLCO and insufficient DLCO revealed associations between CT readouts of lung damage severity (CTSS, opacity, and high opacity, presence of any lesions, GGO and reticulations) and between readouts of acute COVID-19 severity (COVID-19 severity, ICU stay, hospital stay, anti-infective treatment during acute infection) (__Supplementary Figure S17A__).
+In particular, CTSS and opacity (correlation coefficient, Spearman's $\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'opacity_percent') %>% .$rho %>% signif(2)`), CTSS and high opacity ($\rho$ = `r corr_lft$test$ct %>% filter(variable1 == 'CTSS', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`), and opacity and high opacity (`r corr_lft$test$ct %>% filter(variable1 == 'opacity_percent', variable2 == 'high_opacity_percent') %>% .$rho %>% signif(2)`) were strongly inter-correlated (__Supplementary Figure S3B__, __Supplementary Table S11__).
 CTSS values were significantly higher in observations with than without 
 GGO (difference of medians: `r blast(comp_ct$analysis_tbl, GGO_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'GGO_finding') %>% .$eff_size`) and
-reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`) (__Supplementary Figure S3B__, __Supplementary Table S10__).
+reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$CTSS) %>% map_dbl(median) %>% diff` points, `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'CTSS', split_factor == 'reticulation_finding') %>% .$eff_size`).
 Analogically, significantly higher opacity was detected in observations with than without 
@@
 reticulation (difference of medians: `r blast(comp_ct$analysis_tbl, reticulation_finding) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`% of lung, `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$significance`, effect size: `r comp_ct$test %>% filter(variable == 'opacity_percent', split_factor == 'reticulation_finding') %>% .$eff_size`).
-Effect sizes of these differences were large (__Supplementary Figure S15B__, __Supplementary Tables S11__).
+Effect sizes of these differences were large (__Supplementary Figure S17B__, __Supplementary Tables S12__).
 Collectively, the high co-linearity of the most important explanatory variables points suggests their redundancy. 
@@
 FVC (difference of medians: `r blast(lft_uniboot$analysis_tbl, FVC_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FVC_reduced', variable == 'opacity_percent')$eff_size`) and
-insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S12__).
+insufficient FEV1 (difference of medians: `r blast(lft_uniboot$analysis_tbl, FEV1_reduced) %>% map(~.x$opacity_percent) %>% map_dbl(median) %>% diff %>% signif(2)`, `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$significance`, effect size: `r filter(lft_uniboot$test, split_factor == 'FEV1_reduced', variable == 'opacity_percent')$eff_size`); sizes of those effects were, however, small (__Supplementary Table S13__).
 Furthermore, CTSS, opacity and high opacity correlated negatively with moderate effect size with DLCO expressed as percentage of the patient's reference (p < 0.001, Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'DLCO_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
-Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S13__).
+Interestingly, the CT readouts of structural lung damage also correlated significantly with FVC (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FVC_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`) and FEV1 (Spearman's $\rho$: `r lft_uniboot$cor_test %>% filter(variable1 == 'FEV1_percent') %>% .$rho %>% range %>% signif(2) %>% paste(collapse = ' to ')`); effect size of those associations was small (__Supplementary Table S14__).
 

### 2025-03-14 paper/markdown/supplementary_material.Rmd
Tuning grids in Supplementary Material
Showing the complete tuning grids in Supplementary Tables. Supplementary Figures for tuning of the models of DLCO < 80% and DLCO

@@
 
-* for SVM, `sigma`: width of the radial kernel, `C`: cost/penalty applied to misclassified observations
+* for SVM, `sigma`: width of the radial kernel, `C`: cost/penalty applied to misclassified observations.
+For the SVM algorithm, the optimal value of `sigma` was selected by running `sigest()` function from _kernlab_ package [@Karatzoglou2004] on the modeling data set and kept constant in the tuning grid.
 
@@
 For tuning and training of the machine learning models, the `train()` wrapper from the _caret_ package was employed; the tuning/training setting such as participant-blocked indexes for cross-validation folds and Youden's J cost function was specified by via a `trainControl()` object provided to the `train()` function [@Kuhn2008].
+The complete tuning grids, i.e.
+all investigated combinations of the ghyper-parameters, with the corresponding values of Youden's J and MAE cost functions are listed in __Supplementary Table S\@ref(tab:tab-tune-grids)__.
 The optimal values of the tuning parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.
@@
 
-```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm hyper-parameters by cross-validation-mediated tuning.'}
+```{r tab-tune-grids, tab.cap = "Selection of machine learning algorithm hyper-parameters by cross-validation-mediated tuning: the complete tuning grids.
+The selection criteria were the maximum of Youden's J statistic for the classification models.
+and the minimum of mean absolute error for the regression models.
+The table is available as a supplementary Excel file."}
+
+suppl_tabs$tune_grids[1, 1] %>% 
+  mutate(`Model type` = '') %>%
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(18)
+
+```
+
+\newpage
+
+```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm hyper-parameters by cross-validation-mediated tuning: the optimal combinations of the hyper-parameters.'}
 
@@
            ref_symbols = 'a') %>% 
-  footnote(1, 3, 
-           value = as_paragraph('mtry: number of randomly choosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node; size: number of neurons in the hidden layer of the neuronal network; decay: decay of weights in the the neuronal network; sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node.'),
+  footnote(1, 1, 
+           value = as_paragraph("Tuning criteria in participant-blocked 10-repeats cross-validation: maximum of Youden's J for DLCO < 80%, FVC < 80%, FEV1 < 80%; minimum of mean absolute error for DLCO, FVC, FEV1."),
            part = 'header', 
            ref_symbols = 'b') %>% 
+  footnote(1, 3, 
+           value = as_paragraph('mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node; size: number of neurons in the hidden layer of the neuronal network; decay: decay of weights in the neuronal network; sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
   theme_vanilla
@@
 
+```{r fig-class-tune, fig.width = figur::convert(suppl_figs$class_tune, to = 'in')$w, fig.height = figur::convert(suppl_figs$class_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning classification models of reduced diffusion capacity for carbon monoxide.'}
+
+suppl_figs$class_tune$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-class-tune). Choice of hyper-parameters for machine learning classification models of reduced diffusion capacity for carbon monoxide.__
+
+_Classification machine learning models of abnormalities of lung function testing (diffusion capacity of carbon monoxide [DLCO] < 80%, forced vital capacity [FVC] < 80%, forced expiratory volume in one second [FEV1] < 80% of the patient's reference) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms.
+The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._
+_Choice of the algorithms' hyper-parameters (tuning) was accomplished by maximizing Youden's J statistic ($J = sensitivity + specificity - 1$) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._
+_Trajectories of J values for combinations of the hyper-parameters of the models of DLCO < 80% are presented in line plots.
+The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions.
+The full results of tuning are presented in Supplementary Tables S\@ref(tab:tab-tune-grids) and S\@ref(tab:tab-tuning)._
+
+_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._
+
+\newpage
+
+```{r fig-reg-tune, fig.width = figur::convert(suppl_figs$reg_tune, to = 'in')$w, fig.height = figur::convert(suppl_figs$reg_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning regression models of diffusion capacity for carbon monoxide.'}
+
+suppl_figs$reg_tune$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-reg-tune). Choice of hyper-parameters for machine learning regression models of diffusion capacity for carbon monoxide.__
+
+_Regression machine learning models of parameters of lung function testing (diffusion capacity of carbon monoxide [DLCO], forced vital capacity [FVC], forced expiratory volume in one second [FEV1]) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms.
+The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._
+_Choice of the algorithms' hyper-parameters (tuning) was accomplished by minimizing mean absolute errors (MAE) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._ 
+_Trajectories of MAE values for combinations of the hyper-parameters of the models of DLCO are presented in line plots.
+The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions.
+The full results of tuning are presented in Supplementary Tables S\@ref(tab:tab-tune-grids) and S\@ref(tab:tab-tuning)._
+
+_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._
+
+\newpage
+
 ```{r fig-fvc-fev-reduced-models, fig.width = figur::convert(suppl_figs$fvc_fev_reduced_models, to = 'in')$w, fig.height = figur::convert(suppl_figs$fvc_fev_reduced_models, to = 'in')$h, fig.cap = 'Evaluation of performance of machine learning classification models at prediction of insufficient forced vital capacity and forced expiratory volume in one second.'}

### 2025-03-14 LFT scripts/DLCO_findings_tuning.R

@@ -3,6 +3,11 @@
 # 2) NNet: neural network with a single hidden layer
 # 3) support vector machines with radial Kernel
 # 4) Stochastic Gradient Boosting, i.e. Caret's GBM method
+#
+# for SVM: the optimal value of sigma (width of the Gaussian kernel) is 
+# determined via a test run of caret's train, i.e. internally by calling 
+# `kernlab::sigest()`
+
 
   insert_head()
   

### 2025-03-14 LFT scripts/FEV1_findings_tuning.R

@@ -2,6 +2,12 @@
 # 1) Random Forests (Ranger), 
 # 2) NNet: neural network with a single hidden layer
 # 3) support vector machines with radial Kernel
+# 4) Gradient boosted machines
+#
+# for SVM: the optimal value of sigma (width of the Gaussian kernel) is 
+# determined via a test run of caret's train, i.e. internally by calling 
+# `kernlab::sigest()`
+
 
   insert_head()
   

### 2025-03-14 LFT scripts/FEV1_tuning.R

@@ -2,6 +2,11 @@
 # 1) Random Forests (Ranger), 
 # 2) NNet: neural network with a single hidden layer
 # 3) support vector machines with radial Kernel
+# 4) Gradient Boosted Machines
+#
+# for SVM: the optimal value of sigma (width of the Gaussian kernel) is 
+# determined via a test run of caret's train, i.e. internally by calling 
+# `kernlab::sigest()`
 
   insert_head()
   

### 2025-03-14 paper scripts/supplementary_figures.R

@@ -228,6 +228,47 @@
               caption = paste('Modeling strategy.'), 
               w = 180, 
               h = 2524/3404 * 180)
+  
+# Tuning of the models of DLCO < 80% and DLCO --------
+  
+  insert_msg('Tuning process plots')
+  
+  suppl_figs[c('class_tune', 'reg_tune')] <- 
+    list(dlco_red_tune$tuning_plots, 
+         dlco_tune$tuning_plots) %>%
+    map(~plot_grid(.x$nnet + 
+                     labs(title = paste('Tuning:', globals$algo_labs["nnet"])) + 
+                     facet_grid(.~ size) + 
+                     guides(x = guide_axis(angle = 45)) + 
+                     theme(legend.position = 'none', 
+                           axis.text.x = element_text(size = 7)) + 
+                     scale_color_manual(values = rep('steelblue4', 10)), 
+                   plot_grid(.x$ranger + 
+                               labs(title = paste('Tuning:', globals$algo_labs["ranger"])) + 
+                               theme(legend.position = 'bottom'), 
+                             .x$svmRadial + 
+                               labs(title = paste('Tuning:', globals$algo_labs["svmRadial"])), 
+                             ncol = 2, 
+                             rel_widths = c(2, 1), 
+                             align = 'hv', 
+                             axis = 'tblr'),
+                   .x$gbm + 
+                     labs(title = paste('Tuning:', globals$algo_labs["gbm"])) + 
+                     theme(legend.position = 'bottom'), 
+                   nrow = 3))
+  
+  suppl_figs[c('class_tune', 'reg_tune')]  <- 
+    suppl_figs[c('class_tune', 'reg_tune')] %>% 
+    list(label = c('tuning_classification_reduced_dlco', 
+                   'tuning_regression_dlco'),
+         ref_name = names(.),
+         caption = paste('Choice of hyper-parameters for machine learning', 
+                         c('classification', 'regression'), 'models of', 
+                         c('reduced diffusion capacity for carbon monoxide.', 
+                           'diffusion capacity for carbon monoxide.'))) %>% 
+    pmap(as_figure,
+         w = 190, 
+         h = 230)
 
 # evaluation of the models of findings in FVC and FEV ------
   

### 2025-03-14 paper scripts/supplementary_tables.R

@@ -136,9 +136,68 @@
                            'and counts within the complete observation set.'))) %>% 
     pmap(mdtable)
   
-# Tuning of the machine learning models ---------
+# Tuning of the machine learning models, tuning grids --------
   
-  insert_msg('tuning of the machine learning models')
+  insert_msg('Tuning of the machine learning models, tuning grids')
+
+  ## stashing all conditions, algorithms, and responses into 
+  ## a single table
+  
+  suppl_tabs$tune_grids$classification <- 
+    list(DLCO_reduced = dlco_red_tune, 
+         FVC_reduced = fvc_red_tune, 
+         FEV1_reduced = fev1_red_tune) %>% 
+    map(~.x$models) %>% 
+    map(format_tune_grid, 
+        cost_name = 'J')
+  
+  suppl_tabs$tune_grids$regression <- 
+    list(DLCO = dlco_tune, 
+         FVC = fvc_tune, 
+         FEV1 = fev1_tune) %>% 
+    map(~.x$models) %>% 
+    map(format_tune_grid, 
+        cost_name = 'MAE')
+
+  ## appending with the response and algorithm labels
+  
+  suppl_tabs$tune_grids <- 
+    suppl_tabs$tune_grids %>% 
+    map(compress, names_to = 'response') %>% 
+    list(x = ., 
+         y = c("Youden's J", "Mean absolute error"), 
+         z = c('J', 'MAE')) %>% 
+    pmap(function(x, y, z) x %>% 
+           mutate(x, 
+                  Response = exchange(response, covild$lft_lexicon),
+                  Algorithm = globals$algo_labs[algorithm], 
+                  `Hyper-parameters` = grid_condition, 
+                  `Cost function name` = y, 
+                  `Cost function value` = .data[[z]])) %>% 
+    map(select, 
+        Response, Algorithm, `Hyper-parameters`, 
+        `Cost function name`, `Cost function value`) %>% 
+    compress(names_to = 'Model type') %>% 
+    relocate(`Model type`)
+  
+  ## MD table object 
+  
+  suppl_tabs$tune_grids <- suppl_tabs$tune_grids %>% 
+    as_mdtable(label = 'tune_grids', 
+               ref_name = 'tune_grids', 
+               caption =  paste('Selection of machine learning algorithm', 
+                                'hyper-parameters by cross-validation-mediated', 
+                                'tuning: the complete tuning grids.', 
+                                'The selection criteria were the maximum of', 
[…]

### 2025-03-14 tools/tools.R

@@ -1162,7 +1162,7 @@
     
   }
   
-# Tuning plots ------
+# Tuning plots and tuning results tables ------
   
   plot_ranger_tuning <- function(model, 
                                  plot_title = NULL) {
@@ -1439,6 +1439,79 @@
            subtitle = '10-repeats 10-fold cross-validation', 
            x = x_lab,
            y = y_lab)
+    
+  }
+  
+  format_tune_grid_ <- function(tune_df, 
+                                cost_name = 'J', 
+                                hyper_parmeters = names(tune_df)[names(tune_df) != cost_name], 
+                                signif_digits = 3) {
+    
+    ## converts a data frame with tuning results to a two-column 
+    ## data frame. The first column, `grid_condition` contains named values 
+    ## of the hyper-parameters as a string (hyper_parameter = value). 
+    ## The second column contains values of the cost function, whose name is 
+    ## defined by `cost_name` argument. 
+    ## `signif_digits` specifies significant digits used for rounding of the 
+    ## hyper-parameter values and the cost function values. 
+    
+    tune_df <- tune_df %>% 
+      select(all_of(c(cost_name, hyper_parmeters))) %>% 
+      map_dfc(function(x) if(is.numeric(x)) signif(x, signif_digits) else x)
+    
+    tune_df[hyper_parmeters] <- 
+      map2_dfc(hyper_parmeters, 
+               tune_df[hyper_parmeters], 
+               paste, sep = ' = ')
+    
+    tune_df$grid_condition <- tune_df[hyper_parmeters] %>% 
+      as.list %>% 
+      map(as.list) %>% 
+      transpose %>% 
+      map(reduce, paste, sep = ', ') %>% 
+      reduce(c)
+    
+    tune_df[c('grid_condition', cost_name)]
+    
+  }
+  
+  format_tune_grid <- function(model_lst, 
+                               cost_name = 'J', 
+                               signif_digits = 3) {
+    
+    ## creates a data frame with values of the cost function 
+    ## (its name is provided by `cost_name` argument) for combinations 
+    ## of the hype-parameters parameters. 
+    ## `model_lst` is a list of caret models fitted with different algorithms 
+    ## for one response
+    
+    ## metadata ---------
+    
+    grid_tbl <- model_lst %>% 
+      map(~.x$result)
+    
+    stat_names <- grid_tbl %>% 
+      map(names) %>% 
[…]

### 2025-03-15 paper/markdown/rebuttal_letter2.Rmd
Rebuttal Letter for the second revision
And additional references in Supplementary Methods

---
title: "Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery"
subtitle: "Rebuttal Letter 2"
author: "Department of Radiology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: ct.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

_Dear Anna, Dear Gerlig, I've already adjusted the nummeration of Supplementary figures and Supplementary Tables in the second rebuttal letter._

# Reviewer 1

## Concern 1

### Reviewer, first round

The paper lacks details about hyper-parameters for machine learning models. 
For example, what was the C and class weight value for the SVM model?

### Response, first round

Thanks for this point. 
We agree that the information on hyper-parameters in the initial submission was too specific and hard to understand for many clinical readers. 
In the revised manuscript, we briefly describe the hyper-parameters in Supplementary Methods/Multi-parameter modeling of lung function. 
These descriptions were also added as footnotes of __Supplementary Table S7__.

### Reviewer, second round

Providing only good parameters is not sufficient. 
There is a need to demonstrate the effect of different parameters so that it can be concluded that your selected value is good.

### Response, second round

Thanks for this point. 
In the revised manuscript, we provide the complete tuning grids of hyper-parameters with the corresponding values of the cost functions (Youden's J for the classification models and mean absolute error [MAE] for the regression models) as __Supplementary Table S6__. 
Additionally, we show visualizations of the J and MAE as a function of combinations of the hyper-parameters for classification and regression models of DLCO < 80% and DLCO in __Supplementary Figures S9__ and __S10__. 
Plots of the tuning process for FVC < 80%, FEV1 < 80%, FVC, and FEV1 are presented in __Reviewer Figures \@ref(fig:fig-class-fvc-tune)__ - __\@ref(fig:fig-reg-fev1-tune)__ attached to the rebuttal letter.

## Concern 2

### Reviewer, first round

Have you considered the GridSearchCV for model hyper-parameters or set it manually?

### Response, first round

Thanks for this question. 
The hyper-parameters were found by searching grids of values with the maximum Youden’s J statistic for the classification models, and minimum mean absolute error (MAE) for the regression models in 10-repeats 10-fold patient-blocked cross-validation as the selection criteria. This tuning procedure was executed by calling `train()` function from caret package [@Kuhn2008], which, together with our open-source package [_caretExtra_](https://github.com/PiotrTymoszuk/caretExtra), belongs to our standard toolbox for explainable machine learning. This information is provided in Supplementary Methods/Multi-parameter modeling of lung function.

### Reviewer, second round

Have you specifically considered the GridSeachCV (https://scikit-learn.org/stable/modules/generated/sklearn.model_selection.GridSearchCV.html) to find the best parameters? If yes, then which range of values (parameters grid) have you provided to find the best values?

### Response, second round

Thanks for the suggestion - no, we have not considered _GridSearchCV_ for tuning of hyper-parameters in the current paper, but we may consider it for future projects. 
Please note, that our approach to tuning via repeated cross-validation, which uses a well established and highly cited R package _caret_, does not differ from other similar works in the field of respiratory medicine [@Sharifi2020; @Murdaca2021; @Nikolaou2021] (references included in __Supplementary Methods__). 
While, we appreciate your suggestion for _GridSearchCV_, we do not believe that the tuning process based on patient-blocked repeated cross-validation would yield substantially different or better machine learning models if we used _GridSearchCV_ instead of _caret_.

## Concern 3

### Reviewer, second round

Furthermore, the details of practical implication and also conclusion are not well written. 
Please thoroughly improve the paper in terms of practicality and write-up.

### Response, third round

The Conclusion section has been re-written to improve structure, readability, and impact.

\newpage

# Review Figures

```{r fig-class-fvc-tune, fig.width = figur::convert(rev2_figs$class_fvc_tune, to = 'in')$w, fig.height = figur::convert(rev2_figs$class_fvc_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning classification models of reduced forced vital capacity.'}

rev2_figs$class_fvc_tune$plot

```

__Review Figure \@ref(fig:fig-class-fvc-tune). Choice of hyper-parameters for machine learning classification models of reduced forced vital capacity.__

_Classification machine learning models of abnormalities of lung function testing (diffusion capacity of carbon monoxide [DLCO] < 80%, forced vital capacity [FVC] < 80%, forced expiratory volume in one second [FEV1] < 80% of the patient's reference) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms. The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._ 
_Choice of the algorithms' hyper-parameters (tuning) was accomplished by maximizing Youden's J statistic ($J = sensitivity + specificity - 1$) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._ 
_Trajectories of J values for combinations of the hyper-parameters of the models of FVC < 80% are presented in line plots. The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions._

_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._

\newpage

```{r fig-class-fev1-tune, fig.width = figur::convert(rev2_figs$class_fev1_tune, to = 'in')$w, fig.height = figur::convert(rev2_figs$class_fev1_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning classification models of reduced forced expiratory volume in one second.'}

rev2_figs$class_fev1_tune$plot

```

__Review Figure \@ref(fig:fig-class-fev1-tune). Choice of hyper-parameters for machine learning classification models of reduced forced expiratory volume in one second.__

_Classification machine learning models of abnormalities of lung function testing (diffusion capacity of carbon monoxide [DLCO] < 80%, forced vital capacity [FVC] < 80%, forced expiratory volume in one second [FEV1] < 80% of the patient's reference) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms. The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._ 
_Choice of the algorithms' hyper-parameters (tuning) was accomplished by maximizing Youden's J statistic ($J = sensitivity + specificity - 1$) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._ 
_Trajectories of J values for combinations of the hyper-parameters of the models of FEV1 < 80% are presented in line plots. The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions._

_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._

\newpage

```{r fig-reg-fvc-tune, fig.width = figur::convert(rev2_figs$reg_fvc_tune, to = 'in')$w, fig.height = figur::convert(rev2_figs$reg_fvc_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning regression models of forced vital capacity.'}

rev2_figs$reg_fvc_tune$plot

```

__Review Figure \@ref(fig:fig-reg-fvc-tune). Choice of hyper-parameters for machine learning regression models of forced vital capacity.__

_Regression machine learning models of parameters of lung function testing (diffusion capacity of carbon monoxide [DLCO], forced vital capacity [FVC], forced expiratory volume in one second [FEV1]) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms. The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._ 
_Choice of the algorithms' hyper-parameters (tuning) was accomplished by minimizing mean absolute errors (MAE) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._ 
_Trajectories of MAE values for combinations of the hyper-parameters of the models of FVC are presented in line plots. The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions._

_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._

\newpage

```{r fig-reg-fev1-tune, fig.width = figur::convert(rev2_figs$reg_fev1_tune, to = 'in')$w, fig.height = figur::convert(rev2_figs$reg_fev1_tune, to = 'in')$h, fig.cap = 'Choice of hyper-parameters for machine learning regression models of forced expiratory volume in one second.'}

rev2_figs$reg_fev1_tune$plot

```

__Figure \@ref(fig:fig-reg-fev1-tune). Choice of hyper-parameters for machine learning regression models of forced expiratory volume in one second.__

_Regression machine learning models of parameters of lung function testing (diffusion capacity of carbon monoxide [DLCO], forced vital capacity [FVC], forced expiratory volume in one second [FEV1]) were constructed with the neural network, random forest, support vector machines with radial kernel (SVM), and gradient boosted machines (GBM) algorithms. The explanatory variables included demographic information, information on course of acute COVID-19, longitudinally recorded persistent symptoms, and longitudinally recorded readouts of structural lung damage via computed tomography of the chest._ 
_Choice of the algorithms' hyper-parameters (tuning) was accomplished by minimizing mean absolute errors (MAE) for out-of-fold predictions in patient-blocked 10-repeats 10-fold cross-validation._ 
_Trajectories of MAE values for combinations of the hyper-parameters of the models of FEV1 are presented in line plots. The optimal combinations of the hyper-parameters used for training of the final models are displayed in the plot captions._

_size: number of neurons in the hidden layer of the neuronal model; decay: decay of weights in the neuronal network; mtry: number of randomly chosen explanatory variables for construction of decision trees; splitrule: rule used for splitting of explanatory variable at branches of decision trees; min.node.size: minimal size of the decision tree node;  sigma: sigma parameter that defines width of the radial kernel; C: cost/penalty applied to misclassified observations; n.tree: number of decision trees; interaction.depth/int: number of splits in the decision tree; shrinkage: regularization penalty applied to predictions by single decision trees; n.minobsinnode: minimal size of the decision tree node._

\newpage

# References

### 2025-03-15 paper scripts/render.R

@@ -23,11 +23,16 @@
                                         reference_docx = 'ms_template.docx'), 
          output_dir = './paper')
   
-# Rendering the Rebuttal Letter -------
+# Rendering the Rebuttal Letters -------
   
-  insert_msg('Rendering the rebuttal letter')
+  insert_msg('Rendering the rebuttal letters')
   
   render('./paper/markdown/rebuttal_letter.Rmd', 
+         output_format = word_document2(number_sections = FALSE, 
+                                        reference_docx = 'ms_template.docx'), 
+         output_dir = './paper')
+  
+  render('./paper/markdown/rebuttal_letter2.Rmd', 
          output_format = word_document2(number_sections = FALSE, 
                                         reference_docx = 'ms_template.docx'), 
          output_dir = './paper')

### 2025-03-15 paper scripts/reviewer_figures2.R

@@ -1,0 +1,76 @@
+# Figures for Reviewer 1 in the second rounf od revision
+
+  insert_head()
+  
+# container ------
+  
+  rev2_figs <- list()
+  
+# Tuning of classification and regression models of FVC and FEV -------
+  
+  insert_msg('Tuning process of models of FVC and FEV')
+  
+  rev2_figs[c('class_fvc_tune', 
+              'class_fev1_tune',
+              'reg_fvc_tune', 
+              'reg_fev1_tune')] <- 
+    list(fvc_red_tune, fev1_red_tune, fvc_tune, fev1_tune) %>% 
+    map(~.$tuning_plots) %>% 
+    map(~plot_grid(.x$nnet + 
+                     labs(title = paste('Tuning:', globals$algo_labs["nnet"])) + 
+                     facet_grid(.~ size) + 
+                     guides(x = guide_axis(angle = 45)) + 
+                     theme(legend.position = 'none', 
+                           axis.text.x = element_text(size = 7)) + 
+                     scale_color_manual(values = rep('steelblue4', 10)), 
+                   plot_grid(.x$ranger + 
+                               labs(title = paste('Tuning:', globals$algo_labs["ranger"])) + 
+                               theme(legend.position = 'bottom'), 
+                             .x$svmRadial + 
+                               labs(title = paste('Tuning:', globals$algo_labs["svmRadial"])), 
+                             ncol = 2, 
+                             rel_widths = c(2, 1), 
+                             align = 'hv', 
+                             axis = 'tblr'),
+                   .x$gbm + 
+                     labs(title = paste('Tuning:', globals$algo_labs["gbm"])) + 
+                     theme(legend.position = 'bottom'), 
+                   nrow = 3))
+  
+  rev2_figs[c('class_fvc_tune', 
+              'class_fev1_tune',
+              'reg_fvc_tune', 
+              'reg_fev1_tune')]  <- 
+    rev2_figs[c('class_fvc_tune', 
+                'class_fev1_tune',
+                'reg_fvc_tune', 
+                'reg_fev1_tune')] %>% 
+    list(label = c('tuning_classification_reduced_fvc', 
+                   'tuning_classification_reduced_fev1', 
+                   'tuning_regression_fvc',  
+                   'tuning_regression_fev1'),
+         ref_name = names(.),
+         caption = paste('Choice of hyper-parameters for machine learning', 
+                         c(rep('classification', 2), rep('regression', 2)), 
+                         'models of', 
[…]

### 2025-03-15 paper.R

@@ -39,6 +39,7 @@
     './paper scripts/figures.R', 
     './paper scripts/supplementary_figures.R', 
     './paper scripts/reviewer_figures.R', 
+    './paper scripts/reviewer_figures2.R', 
     './paper scripts/render.R') %>% 
     source_all(message = TRUE, crash = TRUE)
   
